---
title: "The fiscal health index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The fiscal health index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(fiscal)
```

The `get_*()` metrics answer "how does this organization look on *this* ratio?"
The **fiscal-health index** answers a harder question: "how does this
organization look *overall*?" It combines many standardized indicators into a
small number of empirical dimensions and, optionally, a single composite score.

Rather than hard-code a formula, `fiscal` builds a **reusable measurement model**
in explicit stages. Each stage is a separate function that returns an
inspectable object, so the pipeline is auditable and the fitted model can be
re-applied to later panel years without re-estimating anything.

```
fiscal_diagnostics()  ->  fiscal_triage()  ->  fiscal_dimensions()
        (screen indicators)     (decide keep/drop)   (discover structure)
                                                            |
fiscal_health_index() <- fiscal_weights() <- fiscal_health_score() <- fiscal_health_fit()
   (composite score)       (combine)          (apply model)            (freeze the model)
```

Two stages deliberately require human judgment --- `fiscal_triage()` recommends
but lets you override, and `fiscal_dimensions()` discovers but does not *name*
dimensions. The wrapper `fiscal_health()` chains the mechanical stages once you
have reviewed the structure.

> This subsystem uses the `psych` package for factor-analytic diagnostics; make
> sure it is installed.

## Prepare indicators

Indicators should be the **standardized** (`_z`) versions of the metrics, so they
share a common scale. Compute them with `compute_all()` and select a curated,
non-redundant set spanning the fiscal-health domains:

```{r}
scored <- compute_all(dat10k, metrics = "z", verbose = FALSE)

indicators <- scored[, c(
  "current_z", "cash_assets_z", "cash_burn_z",      # liquidity
  "debt_assets_z", "equity_z",                       # solvency
  "surplus_margin_z", "op_reserve_z",                # performance / reserves
  "prog_exp_z", "overhead_z",                        # expense structure
  "donations_rev_z", "earned_income_z",              # revenue structure
  "netassets_growth_z"                               # growth
)]

dim(indicators)
```

## Stage 1 — diagnose

`fiscal_diagnostics()` computes variable-level diagnostics that distinguish
indicators participating in a shared multivariate structure from those that are
redundant, standalone, or noisy (correlation summaries, KMO/MSA, communalities,
and so on).

```{r}
diagnostics <- fiscal_diagnostics(indicators)
diagnostics
```

## Stage 2 — triage

`fiscal_triage()` turns those diagnostics into an auditable recommendation:
which indicators to **retain** for dimension estimation, which to keep as
**standalone** measures, and which to drop as redundant or noise. Recommendations
can be overridden with the `overrides` argument.

```{r}
triage <- fiscal_triage(diagnostics)
triage
```

## Stage 3 — discover dimensions

`fiscal_dimensions()` uses PCA (or exploratory factor analysis) to propose
empirical dimensions from the retained indicators. It discovers and describes
groupings; it does not name them or orient the scores --- that is your call.

```{r}
dimensions <- fiscal_dimensions(
  indicators,
  triage   = triage,
  method   = "pca",
  nfactors = 3,
  seed     = 42
)
dimensions
```

## Stage 4 — fit a reusable model

`fiscal_health_fit()` freezes the reviewed dimensions into a scoring
specification. The fitted object stores indicator orientation, centers, scales,
missing-value treatment, and dimension coefficients --- everything needed to score
new data identically later.

```{r}
fit <- fiscal_health_fit(indicators, dimensions = dimensions)
fit
```

## Stage 5 — score

`fiscal_health_score()` applies the fitted model. No parameters are re-estimated;
it just transforms and combines indicators into dimension scores (prefixed
`fh_`). Because the model is frozen, the **same `fit`** can score a different
panel year on the same scale.

```{r}
scored_dims <- fiscal_health_score(indicators, fit)

head(grep("^fh_Dimension", names(scored_dims), value = TRUE))
```

## Stage 6 — weight

`fiscal_weights()` defines how dimension scores and standalone indicators combine
into the composite. The default is equal weighting; `min_coverage` controls how
much of an observation's weighted inputs must be present before it receives a
score.

```{r}
weights <- fiscal_weights(fit, min_coverage = 0.5)
weights
```

## Stage 7 — the composite index

`fiscal_health_index()` produces the single composite score (and a coverage
column recording how much information each score is based on).

```{r}
indexed <- fiscal_health_index(scored_dims, weights)

summary(indexed$fiscal_health)
```

## The shortcut: `fiscal_health()`

Once you have reviewed the triage and dimensions, `fiscal_health()` runs the
mechanical stages --- fit, score, weight, index --- in one call:

```{r}
result <- fiscal_health(indicators, dimensions = dimensions)
result
```

It deliberately does **not** automate diagnostics, triage, or dimension discovery,
because those stages require substantive review.

## Scoring new data with a fitted model

The payoff of separating *fit* from *score* is reuse. Fit the model once on a
reference sample (say, one panel year), then score every other year on the same
scale by passing the stored `fit`:

```{r, eval = FALSE}
fit <- fiscal_health_fit(indicators_2019, dimensions = reviewed_dims)

scores_2020 <- fiscal_health_score(indicators_2020, fit)
scores_2021 <- fiscal_health_score(indicators_2021, fit)
```

This keeps scores comparable across years --- exactly what you need when the index
feeds a longitudinal analysis of the panel you built in
`vignette("panel-workflow")`.
