---
title: "Building a fiscal health index: a measurement approach"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a fiscal health index: a measurement approach}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 5
)
has_psych <- requireNamespace("psych", quietly = TRUE) &&
             requireNamespace("GPArotation", quietly = TRUE)
knitr::opts_chunk$set(eval = has_psych)
```

Nobody observes *fiscal health* directly. What we observe are dozens of
accounting ratios, each a noisy, partial reflection of an organization's
financial condition. This vignette treats fiscal health the way a psychometrician
treats an unobserved trait: as a **latent construct** that we measure by
combining indicators. It follows the factor-analysis recipe in
[Measurement with factor analysis](https://lecy.github.io/econometrics/lectures2/measurement-factor-analysis.html),
applied to 2023 Form 990 filers in `dat10k`.

For the software pipeline that turns a reviewed model into reusable scores, see
`vignette("fiscal-health-workflow")`. This vignette is about the *measurement
decisions* that come before it.

## What the model assumes

Factor analysis assumes that each standardized indicator is a weighted
combination of a few shared latent factors plus a part unique to that indicator:

$$X = \Lambda F + \epsilon$$

- $X$: the observed indicators (standardized fiscal ratios)
- $F$: the latent factors, the dimensions of fiscal health we want to recover
- $\Lambda$: the **loadings**, how strongly each indicator reflects each factor
- $\epsilon$: the **uniqueness**, variance the factors do not explain,
  including measurement error and ratio-specific quirks

Three outputs matter most. **Loadings** tell us what each factor means.
**Communality** ($h^2$) is the share of an indicator's variance explained by the
factors. **Factor scores** are each organization's estimated position on each
dimension, the building blocks of an index.

> **A caveat specific to financial ratios.** The model is *reflective*: it assumes
> the latent condition causes the indicators to move together. Accounting ratios
> also move together *mechanically*, because they share numerators and
> denominators (equity is literally one minus the debt ratio). Part of the
> measurement work below is separating substantive covariation from arithmetic.

## 1. Setup

```{r setup}
library(fiscal)
library(psych)        # KMO(), cortest.bartlett(), fa.parallel(), fa(), alpha()
library(GPArotation)  # oblique rotation used by psych::fa()
```

## 2. Conceptualize the construct and choose candidate indicators

Measurement starts with theory, not data. We define **fiscal health** as an
organization's capacity to keep delivering its mission through ordinary shocks.
The nonprofit finance literature (Tuckman & Chang 1991; Greenlee & Trussel 2000;
Bowman 2011) suggests at least two facets:

- **Resilience (stocks):** cash and reserves relative to spending, and how much
  of the balance sheet is financed by debt.
- **Operating performance (flows):** whether revenues cover expenses year to
  year.

Content validity also tells us what fiscal health is *not*. The share of spending
on programs or administration, and how much revenue comes from fees, describe a
**business model** or an **efficiency** norm. We include three such indicators on
purpose: if the model is measuring fiscal condition, they should *not* load on
its factors (a check on discriminant validity).

`fiscal_metrics()` is the data dictionary. It records each metric's category and
the direction in which it is healthier.

```{r dictionary}
reg <- fiscal_metrics()

candidates <- c(
  # resilience
  "days_cash_ops", "op_reserve", "liquid_assets_months", "cash_assets", "debt_assets",
  # operating performance
  "surplus_margin", "profit_predepr", "return_assets",
  # business model / efficiency (discriminant checks)
  "self_suff", "prog_exp", "expenses_admin",
  # asset intensity
  "assets_rev"
)

dd <- reg[match(candidates, reg$metric), c("metric", "function_name", "category", "direction")]
dd$domain <- c(rep("resilience", 5), rep("performance", 3),
               rep("business model", 3), "asset intensity")
knitr::kable(dd, row.names = FALSE)
```

Many metrics in the registry were screened out before this point, and each
screen is itself a measurement decision:

- **Composition measures** (revenue mix, expense mix, asset mix) have direction
  `"context"`: more is not clearly better or worse.
- **Dollar levels** (`cash_on_hand`) measure size, not condition.
- **Structural missingness:** current, quick, and cash-liquidity ratios are
  undefined when current liabilities are zero, which is true of a large share of
  filers.
- **Algebraic identities:** `equity` is `1 - debt_assets`, `months_cash_ops` is a
  rescaled `days_cash_ops`, and `overhead` is `1 - prog_exp`. Identities make the
  correlation matrix singular and inflate a construct's apparent weight.
- **Signed denominators:** ratios over net assets (`debt_equity`,
  `return_netassets`, `netassets_growth`) flip sign when net assets are negative,
  so they are not monotone in health.

### The analysis sample

Most indicators require full Form 990 detail (Part IX functional expenses and
Part X balance sheet), so the sample is full-990 filers with positive revenue,
expenses, and assets.

```{r sample}
d <- dat10k[dat10k$RETURN_TYPE == "990" &
            dat10k$F9_08_REV_TOT_TOT  > 0 & !is.na(dat10k$F9_08_REV_TOT_TOT) &
            dat10k$F9_09_EXP_TOT_TOT  > 0 & !is.na(dat10k$F9_09_EXP_TOT_TOT) &
            dat10k$F9_10_ASSET_TOT_EOY > 0 & !is.na(dat10k$F9_10_ASSET_TOT_EOY), ]

scored <- compute_all(d, metrics = c("ratio", "z"), append_to_df = FALSE,
                      verbose = FALSE)
nrow(scored)
```

## 3. Standardize and orient the indicators

### Why the `_z` columns

Factor analysis works on the correlation matrix, the covariance matrix of
z-scores. Plain z-scores of fiscal ratios, though, are dominated by a handful of
organizations with near-zero denominators. Every `get_*()` function therefore
also returns a `_z` column: the ratio is winsorized, passed through a
distribution-aware transformation, and standardized (`vignette("normalization")`).
The transformation is monotone, so it preserves the rank ordering of organizations.

Compare three raw ratios with their `_z` versions:

```{r raw-vs-z, fig.height = 6}
show <- c("op_reserve", "debt_assets", "surplus_margin")
op <- par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (m in show) hist(scored[[m]], breaks = 60, main = m, xlab = "raw", col = "grey70", border = "white")
for (m in show) hist(scored[[paste0(m, "_z")]], breaks = 60, main = paste0(m, "_z"),
                     xlab = "normalized", col = "#1f6f78", border = "white")
par(op)
```

The raw histograms are nearly empty: a few extreme values stretch each axis so far
that almost every organization lands in a single bar. The normalized versions use
the whole scale. Some shapes are not bell curves, and should not be: about a
quarter of filers report no liabilities at all, so `debt_assets_z` has a single
spike for that tied group, placed at its middle rank, followed by the rest of the
distribution.

A useful check that the transformation has not distorted the relationships:
Pearson correlations of the `_z` columns should track the rank (Spearman)
correlations of the raw ratios.

```{r z-check}
raw <- scored[candidates]
raw[] <- lapply(raw, function(x) { x[!is.finite(x)] <- NA; x })
Z   <- scored[paste0(candidates, "_z")]
names(Z) <- candidates

r_spearman <- cor(raw, method = "spearman", use = "pairwise.complete.obs")
r_pearson  <- cor(Z, use = "pairwise.complete.obs")
low <- lower.tri(r_pearson)
round(cor(r_spearman[low], r_pearson[low]), 3)   # agreement across all pairs
```

### Orient every indicator so that higher is healthier

Loadings are easiest to read when every indicator points the same way. We flip
the sign of indicators whose registry direction is `"lower"`.

```{r orient}
flip <- dd$metric[dd$direction == "lower"]
flip
X <- Z
X[flip] <- lapply(X[flip], function(x) -x)
used <- complete.cases(X)   # keep track of the rows that enter the model
X <- X[used, ]
dim(X)
```

## 4. Is the data factorable?

Before fitting anything, look at the correlation structure. Ordering the
indicators by hierarchical clustering places blocks of correlated indicators
next to each other; those blocks are the factors we expect to find.

```{r corr-heatmap, fig.height = 6}
R <- cor(X)
ord <- hclust(as.dist(1 - abs(R)))$order
Ro <- R[ord, ord]

pal <- colorRampPalette(c("#bd5416", "#f0efec", "#245b8a"))(101)
op <- par(mar = c(9, 9, 2, 2))
image(1:ncol(Ro), 1:nrow(Ro), t(Ro)[, nrow(Ro):1], zlim = c(-1, 1), col = pal,
      axes = FALSE, xlab = "", ylab = "", main = "Correlations (clustered)")
axis(1, at = 1:ncol(Ro), labels = colnames(Ro), las = 2, cex.axis = 0.8)
axis(2, at = nrow(Ro):1, labels = rownames(Ro), las = 1, cex.axis = 0.8)
text(rep(1:ncol(Ro), each = nrow(Ro)), rep(nrow(Ro):1, ncol(Ro)),
     sprintf("%.2f", Ro), cex = 0.6)
par(op)
```

Two standard checks:

- **Kaiser-Meyer-Olkin (KMO)** measures how much of each correlation is shared
  with the other indicators. Above 0.70 is good; indicators below 0.50 share
  little with the set and are candidates for removal.
- **Bartlett's test** asks whether the correlation matrix differs from an
  identity matrix. A small p-value means there is shared variance to factor.

```{r factorability}
kmo  <- KMO(R)
bart <- cortest.bartlett(R, n = nrow(X))

round(kmo$MSA, 2)
sort(round(kmo$MSAi, 2))
bart$p.value
```

The overall KMO of `r round(kmo$MSA, 2)` is mediocre but workable, and Bartlett's
test rejects an identity matrix. The item-level values already flag
`r paste0("\x60", names(kmo$MSAi)[kmo$MSAi < 0.5], "\x60", collapse = " and ")`
as sharing little with the other indicators.

## 5. How many factors?

A **scree plot** shows the eigenvalues in descending order; look for the elbow.
**Parallel analysis** compares them with eigenvalues from random data of the
same size and keeps the factors that beat the random benchmark.

```{r parallel, fig.height = 4.5}
set.seed(2023)
pa <- fa.parallel(X, fm = "pa", fa = "fa", n.iter = 50,
                  main = "Scree plot and parallel analysis")
pa$nfact
round(pa$fa.values, 2)
```

With thousands of organizations, even trivial factors beat random noise, so
parallel analysis recommends `r pa$nfact` factors. The scree plot tells a
simpler story: two large factors, a much smaller third, then a flat tail.
The number of factors is a judgment that balances statistical evidence against
interpretability, so we compare two and three.

```{r n-factors}
fa2 <- fa(X, nfactors = 2, fm = "pa", rotate = "oblimin")
fa3 <- fa(X, nfactors = 3, fm = "pa", rotate = "oblimin")

print(fa3$loadings, cutoff = 0.30, sort = TRUE)
```

The third factor is defined by `cash_assets` and the reversed `assets_rev`:
balance sheets that are small and cash-heavy relative to revenue. That is
**asset intensity**, a feature of the business model, not a dimension of fiscal
health. We keep two factors.

## 6. Fit the factor model

`psych::fa()` fits the model:

- `nfactors = 2`, as chosen above
- `fm = "pa"`: principal axis factoring, which does not assume multivariate
  normality (fiscal ratios are far from normal)
- `rotate = "oblimin"`: an **oblique** rotation. There is no reason to believe
  that resilience and operating performance are uncorrelated (surpluses build
  reserves), so we let the factors correlate and estimate that correlation.

```{r fit-initial}
print(fa2, digits = 2, cut = 0.30, sort = TRUE)
```

## 7. Interpret the loadings and refine

Rules of thumb: |loading| ≥ 0.30 is potentially meaningful, and ≥ 0.40 is strong.
A **loadings heatmap** shows the whole structure at once, with each indicator's
communality alongside.

```{r loading-plot-fn}
plot_loadings <- function(f, main = "") {
  L  <- unclass(f$loadings)
  h2 <- f$communality
  primary <- apply(abs(L), 1, which.max)
  peak    <- apply(abs(L), 1, max)
  o <- order(ifelse(peak < 0.30, ncol(L) + 1, primary), -peak)
  L <- L[o, , drop = FALSE]; h2 <- h2[o]

  pal <- colorRampPalette(c("#bd5416", "#f0efec", "#245b8a"))(101)
  layout(matrix(1:2, 1), widths = c(3, 1.2))
  op <- par(mar = c(3, 11, 4.5, 0.5))
  Lc <- pmax(pmin(L, 1), -1)   # Heywood loadings can exceed 1; clamp the colors only
  image(1:ncol(L), 1:nrow(L), t(Lc[nrow(L):1, , drop = FALSE]), zlim = c(-1, 1),
        col = pal, axes = FALSE, xlab = "", ylab = "")
  title(main, line = 2.5)
  axis(3, at = 1:ncol(L), labels = colnames(L), tick = FALSE, line = -0.5)
  axis(2, at = nrow(L):1, labels = rownames(L), las = 1, tick = FALSE)
  txt <- ifelse(abs(L) >= 0.30, sprintf("%.2f", L), "")
  text(rep(1:ncol(L), each = nrow(L)), rep(nrow(L):1, ncol(L)), txt, cex = 0.8)
  par(mar = c(3, 0.5, 4.5, 1))
  barplot(rev(pmin(h2, 1)), horiz = TRUE, xlim = c(0, 1), names.arg = "",
          col = "grey55", border = NA, space = 0.2)
  title("Communality", line = 2.5)
  par(op); layout(1)
}
```

```{r loadings-initial, fig.height = 5}
plot_loadings(fa2, "Initial two-factor model")
```

Reading down the columns:

- **Resilience:** operating reserves, liquid unrestricted net assets, and days of
  cash load together, with debt (reversed) loading on the same factor.
- **Operating performance:** surplus margin, the pre-depreciation margin, and
  return on assets.

The rows at the bottom are just as informative. **Program-expense share,
administrative share, and self-sufficiency load on neither factor** and have
communalities near zero. The construct we are recovering is financial condition,
not spending efficiency or revenue model, which is the discriminant validity we
hoped to see. `cash_assets` also has almost no communality; it measures the
composition of the asset base, not its adequacy.

`assets_rev` is a different kind of problem. After orientation (its registry
direction is `"lower"`), it loads *negatively* on resilience: organizations with
more assets per dollar of revenue also hold larger reserves. The indicator's
meaning is ambiguous, so it cannot be scored in a single healthy direction.

**Refinement.** Following the rules above, drop the indicators with low KMO or
low communality, and the indicator whose direction is ambiguous:

```{r refine}
drop <- c("self_suff", "prog_exp", "expenses_admin", "cash_assets", "assets_rev")
items <- setdiff(candidates, drop)
X2 <- X[items]

round(KMO(cor(X2))$MSA, 2)

fit <- fa(X2, nfactors = 2, fm = "pa", rotate = "oblimin", scores = "tenBerge")
print(fit, digits = 2, cut = 0.30, sort = TRUE)
```

KMO rises to `r round(KMO(cor(X2))$MSA, 2)`, and the root mean square of the
residuals is `r round(fit$rms, 2)`: two factors reproduce the correlations among
these indicators closely. (Chi-square based statistics such as RMSEA penalize
any misfit heavily with samples this large and non-normal indicators; the goal
here is an interpretable measure, not a close-fitting confirmatory model.)

### Name the factors

```{r name}
L <- unclass(fit$loadings)
res_col  <- which.max(abs(L["op_reserve", ]))
perf_col <- which.max(abs(L["surplus_margin", ]))
factor_names <- setNames(c("resilience", "performance"),
                         colnames(L)[c(res_col, perf_col)])
factor_names
```

```{r loadings-final, fig.height = 4}
f_named <- fit
colnames(f_named$loadings) <- factor_names[colnames(fit$loadings)]
plot_loadings(f_named, "Refined model")
```

### About the Heywood case

`op_reserve` has a communality at or slightly above 1.00 (an *ultra-Heywood*
case), which is impossible for a true variance share. The reason is visible in
the correlation matrix: operating reserves and liquid unrestricted net assets are
correlated `r round(cor(X2$op_reserve, X2$liquid_assets_months), 2)`. The two
formulas differ only in whether mortgage debt is added back, so the resilience
factor is defined almost entirely by this near-duplicate pair. The same is true
of the surplus margin and the pre-depreciation margin
(`r round(cor(X2$surplus_margin, X2$profit_predepr), 2)`). The solution remains
usable, but the next section checks that it does not depend on the estimator, and
the refinement notes below return to redundancy.

### Factor correlation

```{r phi}
Phi <- fit$Phi
dimnames(Phi) <- list(factor_names[colnames(fit$loadings)], factor_names[colnames(fit$loadings)])
round(Phi, 2)
```

Resilience and performance are positively but modestly correlated
(`r round(Phi[1, 2], 2)`). They are related, distinct dimensions, which matters
for how an overall index should be built.

## 8. Does the solution depend on the estimator?

Refit with maximum likelihood and compare the loadings. Congruence above 0.95
means the factors are essentially the same.

```{r estimator}
fit_ml <- fa(X2, nfactors = 2, fm = "ml", rotate = "oblimin")
round(factor.congruence(fit, fit_ml), 2)
```

Each principal-axis factor has a maximum-likelihood twin with congruence near
1.00, so the two dimensions are not an artifact of the estimation method.

## 9. Extract the new features

Factor scores give every organization a value on each dimension. With an
oblique rotation, the ten Berge method preserves the factor correlations.

```{r scores}
scores <- as.data.frame(fit$scores)
names(scores) <- paste0("fh_", factor_names[names(scores)])
head(round(scores, 2))
round(cor(scores), 2)
```

## 10. Weighted scores versus a simple index

How much does factoring buy over the simplest approach, averaging the oriented
z-scores of the indicators that define each dimension? Cronbach's alpha tells us
whether those indicators are reliable as a unit-weighted scale.

```{r simple-index}
dims <- list(
  resilience  = c("days_cash_ops", "op_reserve", "liquid_assets_months", "debt_assets"),
  performance = c("surplus_margin", "profit_predepr", "return_assets")
)

compare <- do.call(rbind, lapply(names(dims), function(k) {
  simple <- rowMeans(X2[dims[[k]]])
  data.frame(
    dimension         = k,
    items             = length(dims[[k]]),
    cronbach_alpha    = round(psych::alpha(X2[dims[[k]]])$total$raw_alpha, 2),
    r_simple_vs_score = round(cor(simple, scores[[paste0("fh_", k)]]), 2)
  )
}))
compare
```

```{r simple-plot, fig.height = 4}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
for (k in names(dims)) {
  plot(rowMeans(X2[dims[[k]]]), scores[[paste0("fh_", k)]],
       pch = 19, cex = 0.4, col = grDevices::adjustcolor("grey30", 0.3), bty = "n",
       xlab = "Simple index (mean z)", ylab = "Factor score", main = k)
}
par(op)
```

Both dimensions are reliable scales, and the factor scores correlate highly with
the simple averages. Weighting matters most when some indicators are much noisier
than others. The scoring weights show where the two approaches part ways:

```{r score-weights}
W <- fit$weights
colnames(W) <- factor_names[colnames(W)]
round(W, 2)
```

The resilience score is essentially `op_reserve` alone, the Heywood indicator,
and it even gives small *negative* weights to the margin indicators (an oblique
model uses them to net out the correlation between the factors). The performance
score leans mostly on `surplus_margin`. A unit-weighted average spreads the weight
across all of a dimension's indicators, so it is less sensitive to any one ratio;
with near-duplicate indicators like these, that robustness is a good reason to
prefer the simple index until the redundancy is resolved.

## 11. From dimensions to an index

The two dimensions answer different questions, so the most informative output is
a **profile**: an organization can have deep reserves and a current-year deficit
(drawing down savings on purpose), or a healthy surplus and almost no cushion.
When a single number is needed, the simplest defensible composite weights the
dimensions equally.

```{r composite}
index <- data.frame(scores,
                    fiscal_health = rowMeans(scale(scores)))
summary(index$fiscal_health)
```

### Validity: does the index separate organizations known to differ?

A measure should rank organizations in ways that match conditions we can verify
directly in the raw filings (known-groups validity). The conditions below are
*not* inputs to the index: a cash balance that fell by a quarter or more, a
decline in total net assets, negative total net assets, and current liabilities
exceeding current assets.

```{r validity}
stopifnot(identical(scored$EIN2, d$EIN2))   # compute_all() preserves row order
d_used <- d[used, ]
s_used <- scored[used, ]

quintile <- cut(index$fiscal_health,
                quantile(index$fiscal_health, 0:5 / 5),
                include.lowest = TRUE, labels = paste0("Q", 1:5))
share <- function(condition) round(tapply(condition, quintile, mean, na.rm = TRUE), 2)

validity <- data.frame(
  quintile                 = levels(quintile),
  cash_fell_25pct          = share(s_used$cash_burn < 0.75),
  net_assets_declined      = share(s_used$netassets_growth < 0),
  negative_net_assets      = share(d_used$F9_10_NAFB_TOT_EOY < 0),
  current_ratio_below_1    = share(s_used$current < 1)
)
knitr::kable(validity, row.names = FALSE)
```

Every warning sign is most common in the bottom quintile. Declining and negative
net assets, and current liabilities exceeding current assets, fall steadily as the
index rises. Sharp cash declines drop by half between the first and second
quintiles and then level off: a falling cash balance is common even among healthy
organizations that move cash into investments or spend down planned reserves.

## Notes on refinement

This vignette follows the exploratory recipe step by step. Before treating the
index as a validated instrument:

- **Resolve redundancy.** Operating reserves and liquid unrestricted net assets,
  and the two margin measures, are near-duplicates. Keep one of each pair, or
  combine each pair into a single indicator, and refit.
- **Add indicators that broaden each dimension.** With redundant pairs removed,
  each factor rests on very few distinct measures. Candidates that survive the
  screens in section 2 are scarce, which is itself a finding about what the Form
  990 can measure.
- **Confirm the structure on new data.** Fit a confirmatory factor analysis on
  another tax year before relying on the scores.
- **Check measurement invariance.** Hospitals, universities, and small
  community organizations may not load the same way; compare the model across
  subsectors (`nteev2_subsector`) and size groups.
- **Freeze the model for reuse.** Once the structure is settled,
  `fiscal_dimensions()`, `fiscal_health_fit()`, and `fiscal_health_score()` store
  the orientation, centering, and weights so later years are scored on the same
  scale. See `vignette("fiscal-health-workflow")`.
