---
title: "The fiscal health workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The fiscal health workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(fiscal)
```

The `get_*()` metrics answer "how does this organization look on *this* ratio?"
The **fiscal-health index** answers a harder question: "how does this
organization look *overall*?" It combines many standardized indicators into a
small number of empirical dimensions and, optionally, a single composite score.

Rather than hard-code a formula, `fiscal` builds a **reusable measurement model**
in explicit stages. Each stage is a separate function that returns an
inspectable object, so the pipeline is auditable and the fitted model can be
re-applied to later panel years without re-estimating anything.

```
fiscal_diagnostics()  ->  fiscal_triage()  ->  fiscal_dimensions()
        (screen indicators)     (decide keep/drop)   (discover structure)
                                                            |
fiscal_health_index() <- fiscal_weights() <- fiscal_health_score() <- fiscal_health_fit()
   (composite score)       (combine)          (apply model)            (freeze the model)
```

Two stages deliberately require human judgment --- `fiscal_triage()` recommends
but lets you override, and `fiscal_dimensions()` discovers but does not *name*
dimensions. The wrapper `fiscal_health()` chains the mechanical stages once you
have reviewed the structure.

> This subsystem uses the `psych` package for factor-analytic diagnostics; make
> sure it is installed. For the measurement reasoning behind each stage (what
> counts as an indicator, how many dimensions to keep, how to validate the
> result), see `vignette("fiscal-health-index")`.

## Prepare indicators

Indicators should be the normalized (`_z`) versions of the metrics, so they share
a common scale. Most inputs exist only on the full Form 990, so start from full
990 filers with positive revenue, expenses, and assets, then compute the metrics
with `compute_all()`:

```{r}
full990 <- dat10k[dat10k$RETURN_TYPE == "990" &
                  dat10k$F9_08_REV_TOT_TOT   > 0 & !is.na(dat10k$F9_08_REV_TOT_TOT) &
                  dat10k$F9_09_EXP_TOT_TOT   > 0 & !is.na(dat10k$F9_09_EXP_TOT_TOT) &
                  dat10k$F9_10_ASSET_TOT_EOY > 0 & !is.na(dat10k$F9_10_ASSET_TOT_EOY), ]

scored <- compute_all(full990, metrics = "z", verbose = FALSE)
```

Select candidate indicators that are directional in `fiscal_metrics()` and
defined for nearly every filer. The list deliberately includes a few problems
for the pipeline to catch: `equity_z` is an algebraic mirror of `debt_assets_z`,
and the program- and administrative-expense shares describe spending patterns
rather than financial condition.

```{r}
indicators <- scored[, c(
  "days_cash_ops_z", "op_reserve_z", "liquid_assets_months_z",   # reserves
  "debt_assets_z", "equity_z",                                   # solvency
  "surplus_margin_z", "profit_predepr_z", "return_assets_z",     # performance
  "prog_exp_z", "expenses_admin_z", "self_suff_z"                # business model
)]

dim(indicators)
```

## Stage 1 — diagnose

`fiscal_diagnostics()` computes variable-level diagnostics that distinguish
indicators participating in a shared multivariate structure from those that are
redundant, standalone, or noisy (correlation summaries, KMO/MSA, communalities,
and so on).

```{r}
diagnostics <- fiscal_diagnostics(indicators)
diagnostics

diagnostics$redundancy_pairs
```

The diagnostics find the three near-duplicate pairs: `equity_z` mirrors
`debt_assets_z`, and the two reserve measures and the two margin measures differ
only in small adjustments. The expense and self-sufficiency shares share almost
no variance with the rest of the set.

## Stage 2 — triage

`fiscal_triage()` turns those diagnostics into an auditable recommendation:
which indicators to **retain** for dimension estimation, which to keep as
**standalone** measures, and which to drop as redundant or noise.

```{r}
triage <- fiscal_triage(diagnostics)
triage$decisions[, c("variable", "decision", "reason")]
```

The triage keeps one indicator from each redundant pair and drops the
administrative-expense share as noise. It proposes keeping `prog_exp_z` and
`self_suff_z` as **standalone** measures because they are unrelated to the other
indicators. A statistical rule cannot tell *why* an indicator stands alone. Here
it is because these shares describe a spending pattern and a revenue model rather
than financial condition, so we override the recommendation and exclude them:

```{r}
triage <- fiscal_triage(
  diagnostics,
  overrides = c(prog_exp_z = "noise", self_suff_z = "noise")
)
triage$selected_for_dimensions
triage$retained_standalone
```

## Stage 3 — discover dimensions

`fiscal_dimensions()` uses PCA (or exploratory factor analysis) to propose
empirical dimensions from the retained indicators. When `nfactors` is `NULL`,
parallel analysis chooses the number. It discovers and describes groupings; it
does not name them --- that is your call.

```{r}
dimensions <- fiscal_dimensions(
  indicators,
  triage = triage,
  method = "pca",
  seed   = 42
)
dimensions
dimensions$assignments[, c("variable", "primary_dimension", "primary_loading")]
```

Reserves and solvency indicators form one dimension; the margin and return
indicators form the other. Name them after their content:

```{r}
a <- dimensions$assignments
perf <- a$primary_dimension[a$variable == "surplus_margin_z"]
dimension_names <- setNames(
  ifelse(dimensions$multi_indicator_dimensions == perf, "performance", "resilience"),
  dimensions$multi_indicator_dimensions
)
dimension_names
```

## Stage 4 — fit a reusable model

`fiscal_health_fit()` freezes the reviewed dimensions into a scoring
specification. The fitted object stores indicator orientation (inferred from
`fiscal_metrics()` directions), centers, scales, missing-value treatment, and
dimension coefficients --- everything needed to score new data identically later.

```{r}
fit <- fiscal_health_fit(indicators, dimensions = dimensions,
                         dimension_names = dimension_names)
fit
```

## Stage 5 — score

`fiscal_health_score()` applies the fitted model. No parameters are re-estimated;
it just transforms and combines indicators into dimension scores (prefixed
`fh_`). Because the model is frozen, the **same `fit`** can score a different
panel year on the same scale.

```{r}
scored_dims <- fiscal_health_score(indicators, fit)

grep("^fh_", names(scored_dims), value = TRUE)
round(cor(scored_dims$fh_resilience, scored_dims$fh_performance, use = "complete.obs"), 2)
```

## Stage 6 — weight

`fiscal_weights()` defines how dimension scores and standalone indicators combine
into the composite. The default is equal weighting; `min_coverage` controls how
much of an observation's weighted inputs must be present before it receives a
score.

```{r}
weights <- fiscal_weights(fit, min_coverage = 0.5)
weights
```

## Stage 7 — the composite index

`fiscal_health_index()` produces the single composite score (and a coverage
column recording how much information each score is based on).

```{r}
indexed <- fiscal_health_index(scored_dims, weights)

summary(indexed$fiscal_health)
```

## The shortcut: `fiscal_health()`

Once you have reviewed the triage and dimensions, `fiscal_health()` runs the
mechanical stages --- fit, score, weight, index --- in one call:

```{r}
result <- fiscal_health(indicators, dimensions = dimensions,
                        fit_args = list(dimension_names = dimension_names))
result
```

It deliberately does **not** automate diagnostics, triage, or dimension discovery,
because those stages require substantive review.

## Scoring new data with a fitted model

The payoff of separating *fit* from *score* is reuse. Fit the model once on a
reference sample (say, one panel year), then score every other year on the same
scale by passing the stored `fit`:

```{r, eval = FALSE}
fit <- fiscal_health_fit(indicators_2023, dimensions = reviewed_dims,
                         dimension_names = dimension_names)

scores_2024 <- fiscal_health_score(indicators_2024, fit)
scores_2025 <- fiscal_health_score(indicators_2025, fit)
```

This keeps scores comparable across years --- exactly what you need when the index
feeds a longitudinal analysis of the panel you built in
`vignette("panel-workflow")`.
