---
title: "Getting started with fiscal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fiscal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

`fiscal` turns IRS 990 financial data into a standard set of nonprofit
fiscal-health metrics --- liquidity, leverage, margins, efficiency, and
reserves --- each computed the same way every time so results are comparable
across organizations and years.

Two ideas carry the whole package:

1. **Every metric is a `get_*()` function** that appends its result to your data
   frame. There are about fifty of them, catalogued in
   `vignette("metrics-catalog")`.
2. **Every metric returns four versions** of itself --- the raw ratio plus
   winsorized, standardized, and percentile-rank variants --- so outliers and
   scale never distort comparisons.

`compute_all()` runs the entire battery in one call.

## Where fiscal fits

`fiscal` is the **scoring** layer. Its companion package,
[**panel990**](https://github.com/Nonprofit-Open-Data-Collective/panel990),
is the **data** layer: it retrieves and normalizes the underlying 990 e-file
panels. The usual workflow is:

> **panel990** — assemble & clean a 990 panel &nbsp; → &nbsp; **fiscal** — score every organization-year

The panel-facing functions in `fiscal` (`get_panel()`, `deduplicate()`,
`panel_smooth()`, …) are thin, convenience wrappers over `panel990`; see
`vignette("panel-workflow")`.

## Install

`fiscal` depends on `panel990`; installing from GitHub pulls it in automatically
through the package's `Remotes:` field.

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("nonprofit-open-data-collective/fiscal")
```

## Quick start

The package bundles `dat10k`, a 10,000-row sample of 990 financials (both full
990 and 990-EZ filers), so everything below runs offline with no download.

```{r}
library(fiscal)

dim(dat10k)
table(dat10k$RETURN_TYPE)
```

Score the entire battery of fiscal-health metrics and append it to your data:

```{r}
scored <- compute_all(dat10k, verbose = FALSE)

# the metric columns that were appended
new_cols <- setdiff(names(scored), names(dat10k))
length(new_cols)
head(new_cols)
```

Or compute a single metric. Each `get_*()` function defaults to the correct 990
e-file field names but accepts your own column names:

```{r}
df <- get_debt_assets_ratio(dat10k)

df[1:5, c("debt_assets", "debt_assets_w", "debt_assets_z", "debt_assets_p")]
```

## The four versions of every metric

`get_debt_assets_ratio()` (like every metric) appends four columns:

| Column           | Meaning                                             |
|------------------|-----------------------------------------------------|
| `debt_assets`    | the raw financial ratio                             |
| `debt_assets_w`  | winsorized (extreme tails clipped)                  |
| `debt_assets_z`  | standardized as a z-score (from the winsorized values) |
| `debt_assets_p`  | percentile rank, 1–100                              |

The winsorized, standardized, and percentile versions exist so that outliers and
differences in scale don't dominate cross-organization comparisons or any
downstream index. What each transformation actually does --- and how the package
chooses one automatically --- is covered in `vignette("normalization")`.

```{r}
summary(df[, c("debt_assets", "debt_assets_w", "debt_assets_z", "debt_assets_p")])
```

Set `summarize = TRUE` on any metric to print these summaries and plot the four
density curves as you compute:

```{r, eval = FALSE}
df <- get_debt_assets_ratio(dat10k, summarize = TRUE)
```

## Controlling the output

`compute_all()` (and each `get_*()`) takes a `metrics` argument that selects
which of the four versions to return:

| Value     | Returns                    |
|-----------|----------------------------|
| `"ratio"` | raw financial ratio        |
| `"w"`     | winsorized version         |
| `"z"`     | z-score (standardized)     |
| `"p"`     | percentile rank (1–100)    |

```{r}
# only the raw ratios and their percentile ranks
scored_rp <- compute_all(dat10k, metrics = c("ratio", "p"), verbose = FALSE)
```

`append_to_df` controls the shape of the result:

| Value            | Returns                                              |
|------------------|------------------------------------------------------|
| `TRUE` (default) | your original data frame with metric columns appended |
| `FALSE`          | e-file identifier columns + metric columns only      |

```{r}
metrics_only <- compute_all(dat10k, append_to_df = FALSE, verbose = FALSE)
metrics_only[1:3, 1:6]
```

The `get_*()` functions are also pipe-enabled, so metrics can be chained:

```{r, eval = FALSE}
library(dplyr)

df <- dat10k |>
  get_debt_assets_ratio() |>
  get_current_ratio() |>
  get_program_expenses_ratio()
```

## 990 vs 990-EZ

`dat10k` mixes full 990 and 990-EZ filers. Because the two forms report different
fields, some metrics are computable for both and some only for full-990 filers ---
each metric's help entry (and the catalog) states the scope under **Calculated
For**. `fiscal` handles the difference automatically: blank financials are
interpreted correctly and, where a field has a 990-EZ equivalent, the two are
coalesced. This is the subject of `vignette("form-scope")`.

## Where to go next

- **`vignette("metrics-catalog")`** --- every ratio, with formula, inputs,
  definitional range, and benchmarks.
- **`vignette("normalization")`** --- what the `_w` / `_z` / `_p` columns are and
  how transformations are chosen.
- **`vignette("panel-workflow")`** --- build, clean, and score a multi-year panel.
- **`vignette("panel-smoothing")`** --- smoothing noisy longitudinal financials.
- **`vignette("fiscal-health-index")`** --- combine metrics into a composite
  fiscal-health index.
- **`vignette("form-scope")`** --- 990 vs 990-EZ field scope and sanitization.
