---
title: "990 vs 990-EZ: form scope & sanitization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{990 vs 990-EZ: form scope & sanitization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(fiscal)
```

Nonprofits file one of two forms depending on size: the full **Form 990** or the
shorter **Form 990-EZ**. They report *different fields* --- the 990-EZ has no
detailed Part VIII revenue, Part IX functional expenses, or Part X balance sheet.
A dataset of real filers therefore mixes two field vocabularies, and a metric that
blindly divides two 990-only columns would produce nonsense (or silent `NA`s) for
every EZ filer.

`fiscal` handles this with three coordinated pieces:

1. **`detect_ez_rows()`** --- which rows are EZ filers.
2. **field scopes** (`get_pz_fields()`, `get_pc_fields()`) --- which fields exist
   on which form.
3. **`sanitize_financials()`** --- form-aware imputation of blank financials,
   plus a two-column *coalescing* pattern in the metrics themselves.

The bundled `dat10k` mixes both forms:

```{r}
table(dat10k$RETURN_TYPE)
```

## Identifying EZ filers

`detect_ez_rows()` returns a logical vector marking the 990-EZ rows. It prefers the
`RETURN_TYPE` column and falls back to a structural test (rows with Part I revenue
but no Part VIII total revenue are almost certainly EZ filers):

```{r}
ez <- detect_ez_rows(dat10k)
table(ez)
```

## Field scopes: PZ vs PC

Fields are grouped by the forms they appear on:

- **PZ-scope** --- fields available on **both** 990 and 990-EZ (mostly Part I
  summary lines). Retrieved with `get_pz_fields()`.
- **PC-scope** --- fields available on the **full 990 only** (Parts VIII, IX, X
  detail). Retrieved with `get_pc_fields()`.

```{r}
length(get_pz_fields())   # available to both forms
length(get_pc_fields())   # full-990 only
```

This scope split is exactly why each metric's help entry (and the catalog) states
a **Calculated For** scope: a ratio built only from PZ fields works for *990 +
990-EZ* filers, while one that needs any PC field is *990 filers only*.

## Sanitizing blank financials

On a 990, a blank financial line almost always means "zero", not "unknown" --- an
organization with no grants payable simply leaves the line empty. Treating those
blanks as `NA` would drop otherwise-valid organizations from a ratio.
`sanitize_financials()` imputes zero for blank financial fields, but **respects
form scope** so it never invents data a form doesn't collect.

By default (no arguments) it delegates to `panel990::panel_normalize()`, so
`fiscal` and `panel990` share a single imputation engine and field-scope source of
truth:

```{r, eval = FALSE}
clean <- sanitize_financials(dat10k)   # form-aware, uses the shared panel990 engine
```

Every `get_*()` function calls this internally (controlled by its `sanitize`
argument, `TRUE` by default), so you rarely call it yourself.

### The scope-aware rule, made concrete

The imputation follows three rules. Passing explicit field sets uses the
transparent local imputer, which makes the behavior easy to see:

```{r}
df <- data.frame(
  EIN2                = c("990org", "EZorg", "nonfiler"),
  RETURN_TYPE         = c("990",    "990EZ", "990"),
  F9_01_REV_TOT_CY    = c(500000,   300000,  NA),   # PZ: both forms
  F9_08_REV_TOT_TOT   = c(NA,       NA,      NA),    # PC: full 990 only
  F9_08_REV_CONTR_TOT = c(NA,       NA,      NA),    # PC: full 990 only
  stringsAsFactors    = FALSE
)

sanitize_financials(
  df,
  pz_vars = "F9_01_REV_TOT_CY",
  pc_vars = c("F9_08_REV_TOT_TOT", "F9_08_REV_CONTR_TOT")
)
```

Reading the result:

- **`990org`** --- its PC fields are imputed to `0` (a full-990 filer that left
  Part VIII blank really did report zero).
- **`EZorg`** --- its PC fields stay `NA`: the 990-EZ never collects Part VIII, so
  zero would be a fabrication. PZ fields *would* be imputed for EZ filers.
- **`nonfiler`** --- every financial field is `NA`, so the row is left completely
  untouched (a record with no financial data is not silently turned into zeros).

## Coalescing 990 / 990-EZ fields

Some metrics have a genuine equivalent on both forms --- total revenue is Part VIII
line 12 on the 990 but Part I line 9 on the 990-EZ. Those metrics accept **two**
column names and coalesce them, preferring the full-990 value and falling back to
the EZ line:

```{r, eval = FALSE}
# default arguments show the coalescing pattern (990 first, EZ fallback second)
get_debt_assets_ratio(
  df,
  debt   = c("F9_10_LIAB_TOT_EOY",  "F9_01_NAFB_LIAB_TOT_EOY"),
  assets = c("F9_10_ASSET_TOT_EOY", "F9_01_NAFB_ASSET_TOT_EOY")
)
```

This is why metrics like the debt-to-asset ratio, equity ratio, and surplus margin
are **Calculated For** *990 + 990-EZ* filers, while purely PC-scope metrics (the
current ratio, overhead ratio, days of cash) are *990 filers only*. The catalog
(`vignette("metrics-catalog")`) lists the scope for every metric.

## Summary

| Piece | Role |
|-------|------|
| `detect_ez_rows()` | flag which rows are 990-EZ filers |
| `get_pz_fields()` / `get_pc_fields()` | which fields exist on both forms vs full-990 only |
| `sanitize_financials()` | impute blank financials to zero, respecting form scope and never touching all-NA rows |
| two-column arguments | coalesce a 990 field with its 990-EZ fallback |

Together they let a single call to `compute_all()` score a mixed panel of 990 and
990-EZ filers correctly --- see `vignette("fiscal")` to get started.
