---
title: "Normalization & transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Normalization & transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 3.2,
  dpi        = 96
)
library(fiscal)
```

Every metric in `fiscal` returns four columns --- the raw ratio plus winsorized
(`_w`), standardized (`_z`), and percentile-rank (`_p`) versions. This article
explains what those transformations do, the `range` codes that control them, and
the distribution-aware machinery behind `find_best_normalization()`.

Why bother? Nonprofit financial ratios are notoriously badly behaved: heavy right
tails (a few enormous organizations), structural zeros (a nonprofit with no debt),
hard bounds (a program-expense ratio can't exceed 1), and the occasional
data-entry outlier. Feeding raw ratios into a z-score or a composite index lets
those pathologies dominate. Normalization tames them so that comparisons and
indices reflect *position*, not scale artifacts.

## Three steps behind every metric column

For a raw ratio, the package:

1. **Winsorizes** --- clips extreme tails to a percentile bound (`_w`).
2. **Standardizes** --- centers and scales the winsorized values into a z-score (`_z`).
3. **Ranks** --- expresses each value as a percentile, 1–100 (`_p`).

`apply_transformations()` is the engine that does all three at once. It is what
each `get_*()` function calls internally:

```{r}
set.seed(1)
ratio <- c(rlnorm(500, 0, 0.6), 40, 55)   # a right-skewed ratio with two outliers

tf <- apply_transformations(ratio, winsorize = 0.98, range = "zp")
str(tf)
```

```{r}
summary(data.frame(raw = tf$raw, w = tf$winsorized, z = tf$z, p = tf$pctile))
```

Notice the two outliers (40, 55) are pulled back in the winsorized column, which
in turn keeps the z-score and percentile well behaved.

## The `range` codes

Winsorization and transformation both need to know a variable's **theoretical
support** --- a program-expense ratio lives in `[0, 1]`; a growth rate can be
negative; a dollar amount is non-negative and unbounded above. The `range`
argument encodes this:

| Code      | Support             | Typical metric                         |
|-----------|---------------------|----------------------------------------|
| `"np"`    | negative → positive | growth rates, margins                  |
| `"zp"`    | zero → positive     | dollar amounts, days of cash           |
| `"zo"`    | zero → one          | expense/revenue share ratios           |
| `"nz"`    | negative → zero     | (rare) strictly non-positive measures  |
| `"lo;hi"` | a custom pair       | e.g. `"0;10"` for a bounded index      |

The support determines *where* winsorization clips and *which* transformation is
appropriate. `winsorize_x()` exposes the winsorization step directly:

```{r}
w <- winsorize_x(ratio, range = "zp", winsorize = 0.98)

c(bottom = w$bottom, top = w$top)
sum(w$is_sentinel)     # observations pushed to a bound
```

## `find_best_normalization()`: distribution-aware transformation

Standardizing a badly skewed variable produces a badly skewed z-score. For
multivariate work --- correlations, factor analysis, the fiscal-health index ---
we want each indicator to be *approximately normal* first.
`find_best_normalization()` does this by choosing, per variable, one of four
transformations based on the shape of its **stable interior** (the non-sentinel,
non-missing values left after winsorization).

The procedure is always the same:

1. **Winsorize** and set aside sentinel (boundary) observations so pile-up at the
   bounds can't distort the fit.
2. **Diagnose** the stable interior --- its support, boundary mass, and zero
   inflation.
3. **Select** a transformation from those diagnostics (or accept an explicit
   `vtype`).
4. **Fit** the transformation's parameters and a robust center/scale
   (median / MAD).
5. **Report** before-and-after skewness and kurtosis so you can see whether the
   transformation actually helped.

Separating *fit* from *apply* means the same fitted model can score a later panel
year on an identical scale --- see `apply_normalization()` at the end.

### How the transformation is chosen

`guess_normalize_type()` implements the decision rule. Writing `p0` for the share
of values at zero, `p1` for the share at one, and using a `boundary_mass_cutoff`
of `0.10`:

```
is the variable effectively in [0, 1]?  (range "zo", or >98% of values in [0,1])
├── yes
│    ├── p0 > 0.10  ........................  hurdle       (zero-inflated)
│    ├── p1 > 0.10  ........................  rank_normal  (piled up at the top)
│    └── otherwise  ........................  logit        (well-spread proportion)
└── no
     ├── non-negative? (range "zp", or min >= 0)  ....  asinh   (heavy right tail)
     └── mixed / irregular support  ................  rank_normal
```

The four transformations, and the distribution pathology each one targets:

### asinh — heavy-tailed magnitudes

**Pathology.** Non-negative amounts with a long right tail: a handful of huge
organizations stretch the scale (dollar figures, days of cash). **Transform.**
The inverse hyperbolic sine `asinh(x / s)`, scaled by `s`, the median absolute
non-zero value of the stable interior. It behaves like the identity near zero and
like a log in the tails, is defined *at* zero, and accommodates negatives.

```{r}
set.seed(1)
x_asinh <- rlnorm(2000, meanlog = 10, sdlog = 1.3)   # e.g. total-asset dollars

fit <- find_best_normalization(x_asinh, range = "zp", verbose = FALSE)
c(type = fit$transform_type,
  raw_skew = round(fit$raw_skewness, 2), fit_skew = round(fit$fit_skewness, 2),
  raw_kurt = round(fit$raw_kurtosis, 2), fit_kurt = round(fit$fit_kurtosis, 2))

plot_normalize_x(x_asinh, range = "zp")
```

A raw skewness above 2 collapses toward zero, and the extreme kurtosis is tamed.

### logit — well-behaved proportions

**Pathology.** A ratio bounded in `[0, 1]` with mass spread through the middle and
little pile-up at the ends (many expense- and revenue-share ratios). **Transform.**
Rescale to `[0, 1]` relative to the theoretical support, then apply `qlogis()`
(the log-odds), which stretches the bounded scale onto the whole real line.

```{r}
set.seed(2)
x_logit <- rbeta(2000, 2, 8)   # a right-skewed share ratio

fit <- find_best_normalization(x_logit, range = "zo", verbose = FALSE)
c(type = fit$transform_type,
  raw_skew = round(fit$raw_skewness, 2), fit_skew = round(fit$fit_skewness, 2))

plot_normalize_x(x_logit, range = "zo")
```

### hurdle — structural zeros

**Pathology.** A `[0, 1]` variable with a spike of *exact* zeros --- organizations
that simply don't do the thing (no fundraising, no grants, no debt) --- mixed with
a continuous positive part. Treating the zeros as ordinary small numbers is wrong;
they are qualitatively different. **Transform.** Zeros are retained as `0`; the
positive part is transformed separately (rank or logit) and rescaled to a positive
interval. The trigger is `p0 > boundary_mass_cutoff`.

```{r}
set.seed(4)
x_hurdle <- c(rep(0, 900), rbeta(1100, 2, 5))   # ~45% structural zeros

fit <- find_best_normalization(x_hurdle, range = "zo", verbose = FALSE)
c(type = fit$transform_type, p0_reason = fit$guess_reason)

plot_normalize_x(x_hurdle, range = "zo")
```

The zero spike is preserved as its own group while the positive component is
spread out.

### rank_normal — irregular or mixed shapes

**Pathology.** Anything the parametric transforms can't handle gracefully:
bimodal distributions, mixed positive/negative support with no clean model, or a
`[0, 1]` variable piled up at the *top*. **Transform.** A rank-based inverse
normal transform maps the empirical ranks onto a Gaussian. It preserves *ordering*
but not distances, so it is robust to almost any shape --- at the cost of discarding
the original spacing.

```{r}
set.seed(3)
x_rank <- c(rnorm(1000, -3), rnorm(1000, 4))   # bimodal, mixed support

fit <- find_best_normalization(x_rank, range = "np", verbose = FALSE)
c(type = fit$transform_type,
  raw_kurt = round(fit$raw_kurtosis, 2), fit_kurt = round(fit$fit_kurtosis, 2))

plot_normalize_x(x_rank, range = "np")
```

The bimodal shape (strongly negative kurtosis) is pulled toward a single normal
mode.

## One-step and reusable normalization

`normalize_x()` runs fit-and-apply in a single call --- handy for a one-off
variable:

```{r}
z <- normalize_x(x_asinh, range = "zp", verbose = FALSE)
round(c(mean = mean(z), sd = sd(z)), 3)
```

When you need the *same* transformation applied to new data (a different panel
year, a holdout sample), fit once and re-apply, so both are scored on an identical
scale:

```{r}
fit  <- find_best_normalization(x_asinh, range = "zp", verbose = FALSE)
z_new <- apply_normalization(rlnorm(50, 10, 1.3), fit = fit)
```

Force a specific transformation with `vtype` if you want to override the
auto-detection:

```{r, eval = FALSE}
normalize_x(x, range = "np", vtype = "rank_normal")
```

## How this connects to the metrics

The `_w` / `_z` / `_p` columns on every metric come from the winsorize →
standardize → rank path shown at the top (`apply_transformations()`). The
distribution-aware transforms in this article --- asinh, logit, hurdle,
rank_normal --- are what you reach for when building the **fiscal-health index**
(`vignette("fiscal-health-index")`), where each indicator should be approximately
normal before it enters a correlation or factor model.
