---
title: "Normalization & transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Normalization & transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 3.2,
  dpi        = 96
)
library(fiscal)
```

Every metric in `fiscal` returns four columns --- the raw ratio plus winsorized
(`_w`), standardized (`_z`), and percentile-rank (`_p`) versions. This article
explains what those transformations do, the `range` codes that control them, and
the distribution-aware machinery behind `find_best_normalization()`.

Why bother? Nonprofit financial ratios are notoriously badly behaved: heavy right
tails (a few enormous organizations), structural zeros (a nonprofit with no debt),
hard bounds (a program-expense ratio can't exceed 1), and the occasional
data-entry outlier. Feeding raw ratios into a z-score or a composite index lets
those pathologies dominate. Normalization tames them so that comparisons and
indices reflect *position*, not scale artifacts.

## Three steps behind every metric column

For a raw ratio, the package:

1. **Winsorizes** --- clips extreme tails to a percentile bound (`_w`).
2. **Normalizes** --- applies a distribution-aware transformation (chosen by
   `find_best_normalization()`, below) and robustly centers and scales the
   result into a z-score (`_z`).
3. **Ranks** --- expresses each raw value as a percentile, 1–100 (`_p`).

`apply_transformations()` is the engine that does all three at once. It is what
each `get_*()` function calls internally:

```{r}
set.seed(1)
ratio <- c(rlnorm(500, 0, 0.6), 40, 55)   # a right-skewed ratio with two outliers

tf <- apply_transformations(ratio, winsorize = 0.98, range = "zp")
str(tf)
```

```{r}
summary(data.frame(raw = tf$raw, w = tf$winsorized, z = tf$z, p = tf$pctile))
```

Notice the two outliers (40, 55) are pulled back in the winsorized column, which
keeps them from dominating the z-score. The percentile rank depends only on
ordering, so extreme values cannot distort it either way.

## The `range` codes

Winsorization and transformation both need to know a variable's **theoretical
support** --- a program-expense ratio lives in `[0, 1]`; a growth rate can be
negative; a dollar amount is non-negative and unbounded above. The `range`
argument encodes this:

| Code      | Support             | Typical metric                         |
|-----------|---------------------|----------------------------------------|
| `"np"`    | negative → positive | growth rates, margins                  |
| `"zp"`    | zero → positive     | dollar amounts, days of cash           |
| `"zo"`    | zero → one          | expense/revenue share ratios           |
| `"nz"`    | negative → zero     | (rare) strictly non-positive measures  |
| `"lo;hi"` | a custom pair       | e.g. `"0;10"` for a bounded index      |

The support determines *where* winsorization clips and *which* transformation is
appropriate. `winsorize_x()` exposes the winsorization step directly:

```{r}
w <- winsorize_x(ratio, range = "zp", winsorize = 0.98)

c(bottom = w$bottom, top = w$top)
sum(w$is_sentinel)     # observations pushed to a bound
```

## `find_best_normalization()`: distribution-aware transformation

Standardizing a badly skewed variable produces a badly skewed z-score. For
multivariate work --- correlations, factor analysis, the fiscal-health index ---
we want each indicator to be *approximately normal* first.
`find_best_normalization()` does this by choosing, per variable, one of three
transformations (a fourth, `hurdle`, is available on request) based on the shape
of its **stable interior** (the non-sentinel, non-missing values left after
winsorization).

The procedure is always the same:

1. **Winsorize** and set aside sentinel (boundary) observations so pile-up at the
   bounds can't distort the fit.
2. **Diagnose** the stable interior --- its support, boundary mass, and zero
   inflation.
3. **Select** a transformation from those diagnostics (or accept an explicit
   `vtype`).
4. **Fit** the transformation's parameters and a robust center/scale
   (median / MAD, falling back to mean / SD when the MAD is zero because more
   than half of the transformed values are tied).
5. **Report** before-and-after skewness and kurtosis so you can see whether the
   transformation actually helped.

Separating *fit* from *apply* means the same fitted model can score a later panel
year on an identical scale --- see `apply_normalization()` at the end.

### How the transformation is chosen

`guess_normalize_type()` implements the decision rule. Writing `p0` for the share
of values at zero, `p1` for the share at one, and using a `boundary_mass_cutoff`
of `0.10`:

```
is the variable effectively in [0, 1]?  (range "zo", or >98% of values in [0,1])
├── yes
│    ├── p0 > 0.10  ........................  rank_normal  (zero-inflated)
│    ├── p1 > 0.10  ........................  rank_normal  (piled up at the top)
│    └── otherwise  ........................  logit        (well-spread proportion)
└── no
     ├── non-negative? (range "zp", or min >= 0)  ....  asinh   (heavy right tail)
     └── mixed / irregular support  ................  rank_normal
```

`hurdle` never comes out of this rule; it is used only when you ask for it with
`vtype = "hurdle"`.

The four transformations, and the distribution pathology each one targets:

### asinh — heavy-tailed magnitudes

**Pathology.** Non-negative amounts with a long right tail: a handful of huge
organizations stretch the scale (dollar figures, days of cash). **Transform.**
The inverse hyperbolic sine `asinh(x / s)`, scaled by `s`, the median absolute
non-zero value of the stable interior. It behaves like the identity near zero and
like a log in the tails, is defined *at* zero, and accommodates negatives.

```{r}
set.seed(1)
x_asinh <- rlnorm(2000, meanlog = 10, sdlog = 1.3)   # e.g. total-asset dollars

fit <- find_best_normalization(x_asinh, range = "zp", verbose = FALSE)
c(type = fit$transform_type,
  raw_skew = round(fit$raw_skewness, 2), fit_skew = round(fit$fit_skewness, 2),
  raw_kurt = round(fit$raw_kurtosis, 2), fit_kurt = round(fit$fit_kurtosis, 2))

plot_normalize_x(x_asinh, range = "zp")
```

A raw skewness above 2 drops sharply, and the extreme kurtosis is tamed.

### logit — well-behaved proportions

**Pathology.** A ratio bounded in `[0, 1]` with mass spread through the middle and
little pile-up at the ends (many expense- and revenue-share ratios). **Transform.**
Rescale to `[0, 1]` relative to the theoretical support, then apply `qlogis()`
(the log-odds), which stretches the bounded scale onto the whole real line.
Before the log-odds, values are clamped to the `(1 - winsorize) / 2` and
`1 - (1 - winsorize) / 2` quantiles of the strictly interior values (those not
exactly 0 or 1). An exact 0 or 1 is therefore scored at the edge of the interior
distribution instead of at an extreme log-odds that would dominate the center,
the scale, and any correlation.

```{r}
set.seed(2)
x_logit <- rbeta(2000, 2, 8)   # a right-skewed share ratio

fit <- find_best_normalization(x_logit, range = "zo", verbose = FALSE)
c(type = fit$transform_type,
  raw_skew = round(fit$raw_skewness, 2), fit_skew = round(fit$fit_skewness, 2))

plot_normalize_x(x_logit, range = "zo")
```

### hurdle — structural zeros (on request)

**Pathology.** A `[0, 1]` variable with a spike of *exact* zeros --- organizations
that simply don't do the thing (no fundraising, no grants, no debt) --- mixed with
a continuous positive part. You may want the zeros treated as a qualitatively
different group rather than as ordinary small numbers. **Transform.** Zeros are
retained as `0`; the positive part is transformed separately (rank or logit) and
rescaled to a positive interval (`hurdle_range`, default `[1, 100]`).

Hurdle is **not** selected automatically. After standardization the zero block
becomes a single spike at one end of the scale, apart from the positive values,
which distorts correlations with other indicators. Zero-inflated proportions
(`p0 > boundary_mass_cutoff`) are therefore routed to `rank_normal`, which scores
the zero block at its mid-rank. Request the hurdle transform explicitly with
`vtype = "hurdle"`:

```{r}
set.seed(4)
x_hurdle <- c(rep(0, 900), rbeta(1100, 2, 5))   # ~45% structural zeros

# auto-detection sends a zero-inflated proportion to rank_normal
fit <- find_best_normalization(x_hurdle, range = "zo", verbose = FALSE)
c(type = fit$transform_type, p0_reason = fit$guess_reason)

# the hurdle transform must be requested
fit_h <- find_best_normalization(x_hurdle, range = "zo", vtype = "hurdle",
                                 verbose = FALSE)
fit_h$transform_type

plot_normalize_x(x_hurdle, range = "zo", vtype = "hurdle")
```

The zero spike is preserved as its own group while the positive component is
spread out.

### rank_normal — irregular or mixed shapes

**Pathology.** Anything the parametric transforms can't handle gracefully:
bimodal distributions, mixed positive/negative support with no clean model, or a
`[0, 1]` variable piled up at zero or one. **Transform.** A rank-based inverse
normal transform maps the empirical mid-ranks onto a Gaussian. Tied values (a
block of exact zeros, say) share the score at the middle of their block, and
because the reference distribution keeps the winsorized tail observations, clipped
values are scored at their actual quantile rather than at the most extreme normal
score. It preserves *ordering* but not distances, so it is robust to almost any
shape --- at the cost of discarding the original spacing.

```{r}
set.seed(3)
x_rank <- c(rnorm(1000, -3), rnorm(1000, 4))   # bimodal, mixed support

fit <- find_best_normalization(x_rank, range = "np", verbose = FALSE)
c(type = fit$transform_type,
  raw_kurt = round(fit$raw_kurtosis, 2), fit_kurt = round(fit$fit_kurtosis, 2))

plot_normalize_x(x_rank, range = "np")
```

The bimodal shape (strongly negative kurtosis) is pulled toward a single normal
mode.

## One-step and reusable normalization

`normalize_x()` runs fit-and-apply in a single call --- handy for a one-off
variable:

```{r}
z <- normalize_x(x_asinh, range = "zp", verbose = FALSE)
round(c(mean = mean(z), sd = sd(z)), 3)
```

When you need the *same* transformation applied to new data (a different panel
year, a holdout sample), fit once and re-apply, so both are scored on an identical
scale:

```{r}
fit  <- find_best_normalization(x_asinh, range = "zp", verbose = FALSE)
z_new <- apply_normalization(rlnorm(50, 10, 1.3), fit = fit)
```

Force a specific transformation with `vtype` if you want to override the
auto-detection:

```{r, eval = FALSE}
normalize_x(x, range = "np", vtype = "rank_normal")
```

## How this connects to the metrics

The `_w` / `_z` / `_p` columns on every metric come from the winsorize →
normalize → rank path shown at the top (`apply_transformations()`). The `_z`
column uses exactly the machinery in this article: `apply_transformations()`
fits `find_best_normalization()` with the metric's `range` (auto-detecting the
transformation, so never `hurdle`) and scores the ratio with
`apply_normalization()`. The same transforms matter when building the
**fiscal-health index** (`vignette("fiscal-health-index")`), where each indicator
should be approximately normal before it enters a correlation or factor model.
