---
title: "Panel smoothing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Panel smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(fiscal)
```

`panel_smooth()` applies a rolling window average within each organization's time
series to reduce year-to-year noise in financial variables. It returns a data
frame with **identical dimensions and column names** --- only the selected
numeric columns are replaced with smoothed values --- so the smoothed panel drops
straight into `compute_all()` or any downstream step.

This article explains the window logic, the three weighting schemes, and how
missing values are handled. For where smoothing sits in the larger
retrieve → clean → score pipeline, see `vignette("panel-workflow")`.

## Demo data

A tiny three-organization, four-year panel:

```{r}
df <- data.frame(
  id   = rep(c("A", "B", "C"), each = 4),
  x    = 1:12,
  y    = c(-0.200, -0.864, -0.437, -0.975,
           -0.898,  0.223, -0.093,  1.588,
            0.527,  0.994, -0.139, -0.389),
  z    = c(0, 0, 1, 1, -1, 0, 1, 0, 0, 1, 1, 1),
  year = rep(2021:2024, times = 3)
)

knitr::kable(df)
```

## Conceptual overview

For a given window size `N` (which must be odd):

* **Beginning of the panel** --- use the first `N` observations.
* **Middle of the panel** --- use a symmetric window centered on the observation,
  including `(N - 1) / 2` lags and `(N - 1) / 2` leads.
* **End of the panel** --- use the last `N` observations.

This guarantees two things:

* every observation is replaced with a smoothed value, and
* the panel dimensions remain unchanged.

## Weighting schemes

Within each window, observations can be weighted three ways.

### Equal weights

All values in the window receive equal weight. For `N = 3`:

```
(1/3, 1/3, 1/3)
```

### Half weights

The focal observation receives 50% of the weight; the remaining 50% is
distributed evenly among its neighbors. For `N = 3` centered:

```
(0.25, 0.50, 0.25)
```

### Decay weights

Weights decline geometrically with distance from the focal observation, following
a half-life rule (`1, 1/2, 1/4, …`) before being normalized to sum to one.

* Centered (`N = 3`): `(0.25, 0.50, 0.25)`
* Edge (`N = 3`): `(0.5714, 0.2857, 0.1429)`

## Missing-value handling

* Missing values are dropped within each window.
* The remaining weights are re-normalized so they still sum to one.
* If *every* value in a window is missing, the result is `NA`.

For example, a window `{2, NA, 4}` with weights `{0.50, 0.25, 0.25}` collapses to
`{2, 4}` with re-normalized weights `{0.625, 0.375}`.

## Worked example: the math

Take organization `A`, variable `x = 1, 2, 3, 4`, with `window = 3`.

**Equal weights** --- each smoothed value is the simple mean of its window:

```
2021 uses {1,2,3} -> 2
2022 uses {1,2,3} -> 2
2023 uses {2,3,4} -> 3
2024 uses {2,3,4} -> 3
```

giving `c(2, 2, 3, 3)`.

**Half weights** --- edges shift the half-weight to the boundary value:

```
2021 (edge):     weights (0.50, 0.25, 0.25) on {1,2,3} -> 1.75
2022 (centered): weights (0.25, 0.50, 0.25) on {1,2,3} -> 2.00
2023 (centered): weights (0.25, 0.50, 0.25) on {2,3,4} -> 3.00
2024 (edge):     weights (0.25, 0.25, 0.50) on {2,3,4} -> 3.25
```

giving `c(1.75, 2.00, 3.00, 3.25)`.

**Decay weights** --- geometric decay from the focal point:

```
2021 (edge):     (0.5714, 0.2857, 0.1429) on {1,2,3} -> 1.5714
2022 (centered): (0.25,   0.50,   0.25)   on {1,2,3} -> 2.0000
2023 (centered): (0.25,   0.50,   0.25)   on {2,3,4} -> 3.0000
2024 (edge):     (0.1429, 0.2857, 0.5714) on {2,3,4} -> 3.4286
```

giving approximately `c(1.5714, 2.0000, 3.0000, 3.4286)`.

The examples below reproduce these values in the `x` column.

## Using `panel_smooth()`

The call signature names the panel structure explicitly. Here `vars` is a custom
character vector of columns to smooth; on a real 990 panel you would instead pass
`"PZ"`, `"PC"`, or `"ALL"` to smooth the standard financial fields (see below).

### Equal weights

```{r}
df_equal <- panel_smooth(
  df      = df,
  window  = 3,
  time    = "year",
  id      = "id",
  vars    = c("x", "y", "z"),
  weights = "equal"
)

knitr::kable(df_equal, digits = 4)
```

### Half weights

```{r}
df_half <- panel_smooth(
  df      = df,
  window  = 3,
  time    = "year",
  id      = "id",
  vars    = c("x", "y", "z"),
  weights = "half"
)

knitr::kable(df_half, digits = 4)
```

### Decay weights

```{r}
df_decay <- panel_smooth(
  df      = df,
  window  = 3,
  time    = "year",
  id      = "id",
  vars    = c("x", "y", "z"),
  weights = "decay"
)

knitr::kable(df_decay, digits = 4)
```

## Choosing what to smooth on a 990 panel

On an assembled 990 panel, `vars` usually names a field *scope* rather than
individual columns:

| Value            | Columns smoothed                                  |
|------------------|---------------------------------------------------|
| `"PZ"`           | PZ-scope fields (990 + 990-EZ), via `get_pz_fields()` |
| `"PC"`           | PC-scope fields (990 only), via `get_pc_fields()`     |
| `"ALL"`          | the union of PZ and PC fields                      |
| character vector | a custom list of column names (as above)           |

```{r, eval = FALSE}
# smooth the PZ-scope financial fields with a 3-year window
panel_smoothed <- panel_smooth(panel_clean, vars = "PZ", window = 3)

# then score the cleaned, smoothed panel
panel_ratios <- compute_all(panel_smoothed)
```

## Summary

`panel_smooth()`:

* preserves panel dimensions and column names,
* handles the beginning and end of each series without dropping observations,
* supports equal, half, and decay weighting, and
* re-normalizes weights around missing values.

It is particularly useful for stabilizing noisy longitudinal financial measures
while keeping each time point interpretable.
