---
title: "Building a 990 panel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a 990 panel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(fiscal)
```

A multi-year fiscal-health analysis needs a *clean* panel: one row per
organization per year, noise-tolerant financial fields, and a clear picture of
which organizations are actually observed across time. `fiscal` provides a small
set of functions --- thin wrappers over
[**panel990**](https://github.com/Nonprofit-Open-Data-Collective/panel990) ---
that take you from raw e-file downloads to a scored panel:

```
get_panel()          # 1. retrieve & stack multiple years
   |
inspect_duplicates() # 2. see duplicate filings
deduplicate()        #    reduce to one row per org-year
   |
panel_composition()  # 3. classify who is observed when
panel_summary()      #    describe the panel
panel_filter_types() #    keep the organizations you want
   |
panel_smooth()       # 4. smooth noisy financial series
compute_all_panel()  # 5. score every organization-year
```

The retrieval step needs a network connection, so it is shown but not run here.
Every later step runs live on a small synthetic panel.

## 1. Retrieve a panel

`get_panel()` downloads multiple years of IRS 990 e-file data and stacks them into
one long-format panel, stamping each row with `TAX_YEAR` and (optionally)
attaching Business Master File metadata.

```{r, eval = FALSE}
# a four-year panel with the default financial tables
panel <- get_panel(years = 2019:2022)

dim(panel)
table(panel$TAX_YEAR)

# skip the BMF join (faster; attach later with merge_bmf())
panel <- get_panel(years = 2019:2022, include_bmf = FALSE)
```

For the rest of this article we use a small hand-built panel that mimics the
messiness of real e-file data --- amended returns, group returns, and a gap:

```{r}
mk <- function(ein, yr, grp = "", amd = "", stamp, rev) {
  data.frame(
    EIN2 = ein, TAX_YEAR = yr,
    RETURN_GROUP_X = grp, RETURN_PARTIAL_X = "", RETURN_AMENDED_X = amd,
    RETURN_TIME_STAMP = stamp,
    F9_08_REV_TOT_TOT = rev, F9_09_EXP_TOT_TOT = round(rev * 0.9),
    stringsAsFactors = FALSE
  )
}

panel <- do.call(rbind, list(
  mk("A", 2019, stamp = "2020-05-01", rev = 500000),
  mk("A", 2020, stamp = "2021-05-01", rev = 520000),
  mk("A", 2021, stamp = "2022-05-01", rev = 560000),
  mk("B", 2019, stamp = "2020-05-01", rev = 300000),
  mk("B", 2019, amd = "X", stamp = "2020-09-01", rev = 310000),  # amended duplicate
  mk("B", 2021, stamp = "2022-05-01", rev = 340000),             # 2020 missing
  mk("C", 2020, grp = "X", stamp = "2021-05-01", rev = 800000),  # group return
  mk("C", 2020, stamp = "2021-05-20", rev = 810000),
  mk("C", 2021, stamp = "2022-05-01", rev = 790000)
))

knitr::kable(panel[, c("EIN2", "TAX_YEAR", "RETURN_GROUP_X",
                       "RETURN_AMENDED_X", "F9_08_REV_TOT_TOT")])
```

## 2. Remove duplicate filings

E-file data can contain several filings for the same organization-year: amended
returns, partial-year returns, and group returns. `inspect_duplicates()`
*diagnoses* the problem without changing the data:

```{r}
dups <- inspect_duplicates(panel)

dups$summary_types
```

`deduplicate()` then reduces the panel to at most one record per `EIN2` ×
`TAX_YEAR` using ordered rules --- drop group returns, drop partial-year returns,
then keep the most recent filing by `RETURN_TIME_STAMP`. If a rule would drop
*every* record for an organization-year, those rows are rescued and passed to the
next rule.

```{r}
panel_dd <- deduplicate(panel, verbose = FALSE)

nrow(panel)      # before
nrow(panel_dd)   # after
```

## 3. Describe the panel

Who is actually observed, and when? `panel_composition()` classifies each
organization's presence across the panel window. Ask for the per-ID
classification table with `return_classification = TRUE`:

```{r}
cls <- panel_composition(panel_dd, return_classification = TRUE, print_table = FALSE)

cls[, c("EIN2", "panel_type", "panel_year_count", "panel_gap_count")]
```

Each organization is labelled `persistent`, `entrant`, `exit`, `transient`, or
`empty`, with a `panel_spell` of `seamless` or `segmented` (a segmented spell has
a gap year). To attach these columns to the data instead, use
`append_classification = TRUE`; `panel_summary()` then reports the overall shape
of the panel:

```{r}
panel_cls <- panel_composition(panel_dd, append_classification = TRUE, print_table = FALSE)

panel_summary(panel_cls, print_table = FALSE)
```

Keep only the organizations you want to analyze with `panel_filter_types()` ---
for example, organizations observed throughout the window:

```{r}
persistent <- panel_filter_types(panel_dd, panel_types = cls, keep = "persistent")

unique(persistent$EIN2)
```

`panel_impute()` can fill short internal gaps for selected panel types before
smoothing or scoring:

```{r}
panel_imp <- panel_impute(
  panel_dd,
  panel_types = cls,
  types       = "persistent",
  vars        = "F9_08_REV_TOT_TOT"
)
```

## 4. Smooth noisy series

`panel_smooth()` replaces selected numeric fields with a rolling within-organization
average, leaving the panel's shape unchanged. On a real panel you would pass a
field *scope* (`"PZ"`, `"PC"`, or `"ALL"`); here we name the columns directly:

```{r}
panel_sm <- panel_smooth(
  panel_dd,
  vars   = c("F9_08_REV_TOT_TOT", "F9_09_EXP_TOT_TOT"),
  window = 3
)
```

`panel_smooth()` has its own article covering window logic and the equal / half /
decay weighting schemes: see `vignette("panel-smoothing")`.

## 5. Score every organization-year

`compute_all_panel()` is the panel-aware sibling of `compute_all()`: it computes
the full battery of fiscal-health metrics **within each `TAX_YEAR`**, so that the
winsorization, standardization, and percentile ranks are all year-relative.

```{r}
panel_scored <- compute_all_panel(panel_sm, metrics = c("ratio", "z"), verbose = FALSE)

new_cols <- setdiff(names(panel_scored), names(panel_sm))
length(new_cols)
head(new_cols)
```

The result is a clean, smoothed, scored panel --- one row per organization-year,
ready for longitudinal analysis or a composite index
(`vignette("fiscal-health-index")`).

## Recap

| Step | Function(s) | Purpose |
|------|-------------|---------|
| Retrieve | `get_panel()` | download & stack years |
| Deduplicate | `inspect_duplicates()`, `deduplicate()` | one row per org-year |
| Describe | `panel_composition()`, `panel_summary()`, `panel_filter_types()`, `panel_impute()` | who is observed, when; subset & fill |
| Smooth | `panel_smooth()` | reduce year-to-year noise |
| Score | `compute_all_panel()` | year-relative fiscal-health metrics |
