---
title: "Construction-induced correlation among fiscal ratios"
output: rmarkdown::html_vignette
params:
  efile_path: ""
  reps: 30
  seed: 2023
  save_results: ""
vignette: >
  %\VignetteIndexEntry{Construction-induced correlation among fiscal ratios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  dpi        = 96
)
options(width = 110)
```

```{r packages}
library(fiscal)
library(psych)
library(splines)
set.seed(params$seed)
```

## The question

The fiscal health index factors a correlation matrix of 14 ratios. Those ratios
are built from about 20 Form 990 lines, and many of them share a line: the
same denominator (total expenses, total assets), the same numerator (cash, the
surplus), or a line that sits on top of one ratio and underneath another. Two
ratios that share a line correlate even when nothing links the organizations'
finances. Shares of one whole correlate negatively by definition.

This article asks three questions:

1. **How much of each pairwise correlation is construction?** We build a
   *construction null*: the same 990 lines, shuffled across organizations so
   that every substantive relationship is destroyed while the arithmetic of the
   ratios, the accounting identities, and the pipeline (rules, imputation,
   winsorizing, asinh) are kept.
2. **Does the construction null manufacture factors?** If PCA on arithmetic
   alone returns components that look like the health dimensions, those
   dimensions are at least partly definitional.
3. **Is there one pre-processing fix that removes the artifact consistently
   across all 91 pairs**, without destroying variables or the matrix?
4. **Does redesigning the variable set by link type work better** than
   transforming the ratios?

By default the article runs on the bundled `dat10k` sample (2021). Set the
`efile_path` parameter to an e-file extract (for example the deduplicated 2023
file) to run it on a full tax year.

## 1. Components and ratios

The ratio rules and imputation are the measurement model from
`vignette("variance-preserving-normalization")`. They are split into two
steps here, **lines** and **ratios**, so the null can shuffle lines and then
rebuild ratios exactly as the real data are built.

```{r helpers}
fix64 <- function(x) {
  if (inherits(x, "integer64")) return(as.numeric(x))
  if (is.double(x)) {
    nz <- x[!is.na(x) & x != 0]
    if (length(nz) && all(abs(nz) < 1e-300)) {
      class(x) <- "integer64"
      return(as.numeric(x))
    }
  }
  x
}
pct_rank    <- function(x) { r <- rank(x, na.last = "keep"); (r - 0.5) / sum(!is.na(x)) }
rank_normal <- function(x) qnorm(pct_rank(x))
fill_median <- function(x) { x[is.na(x)] <- median(x, na.rm = TRUE); x }
winsor <- function(x, p = 0.01) {
  q <- quantile(x, c(p, 1 - p), na.rm = TRUE)
  pmin(pmax(x, q[[1]]), q[[2]])
}
winsor_normalize <- function(x, p = 0.01) {
  w  <- winsor(x, p)
  sk <- psych::skew(w)
  if (sk > 1) {
    s <- median(abs(w[w != 0])); w <- asinh(w / s)
  } else if (sk < -1) {
    r <- max(w) - w; s <- median(r[r != 0]); w <- -asinh(r / s)
  }
  as.numeric(scale(w))
}
cond_impute <- function(target, predictors, miss) {
  y   <- rank_normal(target)
  X   <- as.data.frame(lapply(predictors, rank_normal))
  fit <- stats::lm(y ~ ., data = cbind(y = y, X)[!miss, ])
  z   <- stats::predict(fit, newdata = X[miss, , drop = FALSE]) +
         stats::rnorm(sum(miss), 0, summary(fit)$sigma)
  out <- target
  out[miss] <- stats::quantile(target[!miss], stats::pnorm(z), names = FALSE, na.rm = TRUE)
  out
}
```

```{r lines-ratios}
# Step 1: the 990 lines each ratio is built from.
efile_lines <- function(d) {
  d <- as.data.frame(d)
  d <- d[d$RETURN_TYPE == "990", ]
  fin <- function(f) {
    x <- if (f %in% names(d)) suppressWarnings(as.numeric(fix64(d[[f]]))) else rep(0, nrow(d))
    x[is.na(x)] <- 0
    x
  }
  k <- data.frame(
    cash    = fin("F9_10_ASSET_CASH_EOY") + fin("F9_10_ASSET_SAVING_EOY"),
    recv    = fin("F9_10_ASSET_PLEDGE_NET_EOY") + fin("F9_10_ASSET_ACC_NET_EOY"),
    othca   = fin("F9_10_ASSET_INV_SALE_EOY") + fin("F9_10_ASSET_EXP_PREPAID_EOY"),
    ap      = fin("F9_10_LIAB_ACC_PAYABLE_EOY"),
    cl      = fin("F9_10_LIAB_ACC_PAYABLE_EOY") + fin("F9_10_LIAB_GRANT_PAYABLE_EOY"),
    A = fin("F9_10_ASSET_TOT_EOY"), L = fin("F9_10_LIAB_TOT_EOY"),
    NA_eoy = fin("F9_10_NAFB_TOT_EOY"), NA_boy = fin("F9_10_NAFB_TOT_BOY"),
    UNA = fin("F9_10_NAFB_UNRESTRICT_EOY"), RES = fin("F9_10_NAFB_RESTRICT_EOY"),
    land = fin("F9_10_ASSET_LAND_BLDG_NET_EOY"), mtg = fin("F9_10_LIAB_MTG_NOTE_EOY"),
    R = fin("F9_08_REV_TOT_TOT"), E = fin("F9_09_EXP_TOT_TOT"), dep = fin("F9_09_EXP_DEPREC_TOT"),
    progrev = fin("F9_08_REV_PROG_TOT_TOT"),
    P = fin("F9_09_EXP_TOT_PROG"), M = fin("F9_09_EXP_TOT_MGMT"), Fu = fin("F9_09_EXP_TOT_FUNDR"))
  k[k$E > 0, ]
}

# Step 2: rules, conditional imputation, winsorize + normalize.
candidates <- c("current", "months_cash", "liquid_months", "cash_assets", "debt_assets",
                "payables_una", "surplus_margin", "return_assets", "na_growth", "self_suff",
                "admin_exp", "fundr_exp", "land_assets", "assets_rev")

lines_to_frame <- function(k) {
  k$ca   <- k$cash + k$recv + k$othca
  k$una  <- ifelse(k$UNA == 0 & k$RES == 0, k$NA_eoy, k$UNA)
  k$fsum <- k$P + k$M + k$Fu
  k$opE  <- ifelse(k$E - k$dep > 0, k$E - k$dep, k$E)
  rules <- with(k, data.frame(
    current        = ifelse(cl > 0, ca / cl, NA),
    months_cash    = (cash + recv) / (opE / 12),
    liquid_months  = (una - land + mtg) / (E / 12),
    cash_assets    = ifelse(A > 0, pmin(cash / A, 1), NA),
    debt_assets    = ifelse(A > 0, L / A, ifelse(L > 0, NA, 0)),
    payables_una   = ifelse(ap > 0, ap / (ap + pmax(una, 0)), 0),
    surplus_margin = (R - E) / pmax(R, E),
    return_assets  = ifelse(A > 0, (R - E) / A, NA),
    na_growth      = ifelse(NA_eoy == 0 & NA_boy == 0, 0,
                            (NA_eoy - NA_boy) / ((abs(NA_eoy) + abs(NA_boy)) / 2)),
    self_suff      = progrev / E,
    admin_exp      = ifelse(fsum > 0, M / fsum, NA),
    fundr_exp      = ifelse(fsum > 0, Fu / fsum, NA),
    land_assets    = ifelse(A > 0, land / A, 0),
    assets_rev     = A / ifelse(R > 0, R, E)))
  if (any(k$A <= 0 & k$L > 0))
    rules$debt_assets[k$A <= 0 & k$L > 0] <- max(rules$debt_assets, na.rm = TRUE)
  frame <- rules
  preds <- as.data.frame(lapply(
    rules[, c("liquid_months", "surplus_margin", "debt_assets", "months_cash")], fill_median))
  frame$current <- cond_impute(rules$current, preds, k$cl == 0)
  base_preds <- c("liquid_months", "surplus_margin", "months_cash", "self_suff")
  for (v in names(frame)) if (anyNA(frame[[v]])) {
    frame[[v]] <- cond_impute(frame[[v]], frame[, setdiff(base_preds, v), drop = FALSE],
                              is.na(frame[[v]]))
  }
  frame
}
normalize_frame <- function(frame) as.data.frame(lapply(frame[candidates], winsor_normalize))
lines_to_ratios <- function(k) normalize_frame(lines_to_frame(k))
```

```{r data}
if (nzchar(params$efile_path)) {
  efile <- readRDS(params$efile_path)
  source_label <- basename(params$efile_path)
} else {
  data(dat10k, package = "fiscal")
  efile <- dat10k
  source_label <- "dat10k (2021 sample)"
}
k <- efile_lines(efile)
rm(efile); invisible(gc())
frame_obs <- lines_to_frame(k)
X <- normalize_frame(frame_obs)
R_obs <- cor(X)
data.frame(source = source_label, organizations = format(nrow(X), big.mark = ","),
           ratios = ncol(X))
```

## 2. The pairwise audit

Each of the 91 pairs is classified from the formulas alone, before looking at
any data:

* **closure**: both numerators are disjoint parts of the shared denominator;
  forced negative.
* **shared_den**: same denominator (or two denominators that are nearly the
  same flow, E and R); forced positive for positive numerators.
* **reciprocal**: a line on top of one ratio sits underneath the other; forced
  negative.
* **shared_num**: the same line in both numerators; forced in the direction
  of the signs.
* **identity**: linked through an accounting identity (NA = A - L,
  change in NA ~ R - E) rather than a shared line.
* **none**: no shared line.

`role` records whether the shared term is part of what *both* ratios are meant
to measure (**construct**), of neither (**artifact**), or of one (**mixed**).
`fix` is the repair that fits the pair taken alone.

```{r audit}
audit <- read.table(sep = "|", header = TRUE, strip.white = TRUE, text = "
v1|v2|type|shared|sign|role|fix
current|months_cash|shared_num|cash, recv|+|construct|keep
current|liquid_months|identity|ca in una; cl in L|+|construct|keep
current|cash_assets|shared_num|cash|+|artifact|benchmark
current|debt_assets|reciprocal|ca in A; cl in L|-|mixed|benchmark
current|payables_una|reciprocal|ap|-|artifact|benchmark
current|surplus_margin|none||0||none
current|return_assets|reciprocal|ca in A (signed numerator)|mixed|artifact|benchmark
current|na_growth|none||0||none
current|self_suff|none||0||none
current|admin_exp|none||0||none
current|fundr_exp|none||0||none
current|land_assets|identity|ca and land in A|-|artifact|benchmark
current|assets_rev|shared_num|ca in A|+|mixed|benchmark
months_cash|liquid_months|shared_den|E; cash, recv in una|+|construct|keep
months_cash|cash_assets|shared_num|cash|+|artifact|benchmark
months_cash|debt_assets|reciprocal|cash, recv in A|-|artifact|benchmark
months_cash|payables_una|identity|cash, recv in una|-|artifact|benchmark
months_cash|surplus_margin|shared_den|E (signed numerator)|+|artifact|resid_E
months_cash|return_assets|shared_den|E (+); cash in A (-)|mixed|artifact|resid_E
months_cash|na_growth|identity|dNA ~ R - E|+|artifact|resid_E
months_cash|self_suff|shared_den|E|+|artifact|resid_E
months_cash|admin_exp|shared_den|E|+|artifact|resid_E
months_cash|fundr_exp|shared_den|E|+|artifact|resid_E
months_cash|land_assets|identity|cash and land in A|-|artifact|benchmark
months_cash|assets_rev|shared_num|cash in A; E ~ R|+|mixed|benchmark
liquid_months|cash_assets|identity|cash in una|+|artifact|benchmark
liquid_months|debt_assets|identity|L lowers una|-|mixed|benchmark
liquid_months|payables_una|reciprocal|una|-|artifact|benchmark
liquid_months|surplus_margin|shared_den|E (signed numerator)|+|artifact|resid_E
liquid_months|return_assets|shared_den|E (+); una in A (-)|mixed|artifact|resid_E
liquid_months|na_growth|identity|una in NA|mixed|mixed|benchmark
liquid_months|self_suff|shared_den|E|+|artifact|resid_E
liquid_months|admin_exp|shared_den|E|+|artifact|resid_E
liquid_months|fundr_exp|shared_den|E|+|artifact|resid_E
liquid_months|land_assets|shared_num|land (opposite signs)|-|construct|keep
liquid_months|assets_rev|identity|una in A; E ~ R|+|mixed|benchmark
cash_assets|debt_assets|shared_den|A|+|artifact|resid_A
cash_assets|payables_una|none||0||none
cash_assets|surplus_margin|none||0||none
cash_assets|return_assets|shared_den|A (signed numerator)|mixed|artifact|resid_A
cash_assets|na_growth|none||0||none
cash_assets|self_suff|none||0||none
cash_assets|admin_exp|none||0||none
cash_assets|fundr_exp|none||0||none
cash_assets|land_assets|closure|parts of A|-|artifact|log_ratio
cash_assets|assets_rev|reciprocal|A|-|artifact|benchmark
debt_assets|payables_una|shared_num|ap in L; una = A - L - RES|+|construct|keep
debt_assets|surplus_margin|none||0||none
debt_assets|return_assets|shared_den|A (signed numerator)|mixed|artifact|resid_A
debt_assets|na_growth|identity|NA = A - L|mixed|artifact|benchmark
debt_assets|self_suff|none||0||none
debt_assets|admin_exp|none||0||none
debt_assets|fundr_exp|none||0||none
debt_assets|land_assets|shared_den|A|+|mixed|resid_A
debt_assets|assets_rev|reciprocal|A|-|artifact|benchmark
payables_una|surplus_margin|none||0||none
payables_una|return_assets|none||0||none
payables_una|na_growth|identity|una in NA|mixed|artifact|benchmark
payables_una|self_suff|none||0||none
payables_una|admin_exp|none||0||none
payables_una|fundr_exp|none||0||none
payables_una|land_assets|none||0||none
payables_una|assets_rev|identity|una in A|-|artifact|benchmark
surplus_margin|return_assets|shared_num|R - E|+|construct|combine
surplus_margin|na_growth|identity|dNA ~ R - E|+|construct|combine
surplus_margin|self_suff|shared_num|progrev in R; E|+|artifact|resid_E
surplus_margin|admin_exp|shared_den|E|+|artifact|resid_E
surplus_margin|fundr_exp|shared_den|E|+|artifact|resid_E
surplus_margin|land_assets|none||0||none
surplus_margin|assets_rev|reciprocal|R|-|artifact|benchmark
return_assets|na_growth|identity|dNA ~ R - E|+|construct|combine
return_assets|self_suff|shared_num|progrev in R|+|artifact|benchmark
return_assets|admin_exp|shared_den|E|+|artifact|resid_E
return_assets|fundr_exp|shared_den|E|+|artifact|resid_E
return_assets|land_assets|shared_den|A (signed numerator)|mixed|artifact|resid_A
return_assets|assets_rev|reciprocal|A; R|-|artifact|benchmark
na_growth|self_suff|identity|progrev in R; dNA ~ R - E|+|artifact|benchmark
na_growth|admin_exp|none||0||none
na_growth|fundr_exp|none||0||none
na_growth|land_assets|none||0||none
na_growth|assets_rev|identity|NA in A|mixed|artifact|benchmark
self_suff|admin_exp|shared_den|E|+|artifact|resid_E
self_suff|fundr_exp|shared_den|E|+|artifact|resid_E
self_suff|land_assets|none||0||none
self_suff|assets_rev|reciprocal|progrev in R; E ~ R|mixed|artifact|benchmark
admin_exp|fundr_exp|closure|parts of E|-|artifact|log_ratio
admin_exp|land_assets|none||0||none
admin_exp|assets_rev|shared_den|E ~ R|+|artifact|resid_E
fundr_exp|land_assets|none||0||none
fundr_exp|assets_rev|shared_den|E ~ R|+|artifact|resid_E
land_assets|assets_rev|reciprocal|A|-|artifact|benchmark
", colClasses = "character", na.strings = NULL)
stopifnot(nrow(audit) == choose(length(candidates), 2))
audit$r_obs <- R_obs[cbind(audit$v1, audit$v2)]
table(audit$type)
table(audit$fix)
```

## 3. The construction null

The null keeps each organization's size and every accounting identity, and
destroys everything else. Each 990 line is divided by a **size reference**,
shuffled across organizations independently of every other line, multiplied
back, and the totals are rebuilt from their parts (A = sum of asset lines,
L = sum of liability lines, NA = A - L, NA_boy from the surplus). The ratios are
then rebuilt with the same rules, imputation, and transformations as the real
data.

Coupling through a denominator is an artifact only to the extent that the
denominator varies independently of size, so the null depends on the size
reference. We run two:

* **flow anchor**: size = total expenses. Expense-denominator coupling is
  absent by definition; balance-sheet coupling is present.
* **stock anchor**: size = total assets. Expenses become an independent draw
  relative to assets, so expense-denominator coupling is present.

```{r null}
# `draw(x, line)` reassigns one size-scaled line across organizations. The
# default is an independent shuffle; section 10 swaps in a draw that plants
# known factors.
null_lines <- function(k, anchor = c("flow", "stock", "joint"),
                       draw = function(x, line) x[sample.int(length(x))]) {
  anchor <- match.arg(anchor)
  # "joint" (section 11): balance-sheet lines relative to assets, flow lines
  # relative to the organization's own expenses, which are never shuffled.
  gp     <- k$cl - k$ap
  otherA <- pmax(k$A - k$cash - k$recv - k$othca - k$land, 0)
  otherL <- pmax(k$L - k$cl - k$mtg, 0)
  res_sh <- ifelse(k$UNA >= 0 & k$RES > 0, k$RES / (k$UNA + k$RES), 0)
  na_oth <- (k$NA_eoy - k$NA_boy) - (k$R - k$E)
  size <- if (anchor == "flow") k$E else ifelse(k$A > 0, k$A, k$E)
  E    <- if (anchor == "stock") draw(k$E / size, "E") * size else k$E
  bs <- function(x, line) draw(x / size, line) * size   # balance-sheet lines
  fl <- function(x, line) draw(x / k$E, line) * E       # flow lines, relative to expenses
  out <- k
  out$cash <- bs(k$cash, "cash"); out$recv <- bs(k$recv, "recv"); out$othca <- bs(k$othca, "othca")
  out$land <- bs(k$land, "land"); out$ap <- bs(k$ap, "ap"); out$mtg <- bs(k$mtg, "mtg")
  out$cl <- out$ap + bs(gp, "gp")
  out$A  <- out$cash + out$recv + out$othca + out$land + bs(otherA, "otherA")
  out$L  <- out$cl + out$mtg + bs(otherL, "otherL")
  out$NA_eoy <- out$A - out$L
  out$RES <- ifelse(out$NA_eoy > 0, draw(res_sh, "res_sh") * out$NA_eoy, 0)
  out$UNA <- out$NA_eoy - out$RES
  out$E <- E
  out$R <- fl(k$R, "R"); out$dep <- fl(k$dep, "dep"); out$progrev <- fl(k$progrev, "progrev")
  out$P <- fl(k$P, "P"); out$M <- fl(k$M, "M"); out$Fu <- fl(k$Fu, "Fu")
  out$NA_boy <- out$NA_eoy - (out$R - out$E) - fl(na_oth, "na_oth")
  out
}
```

## 4. Candidate fixes

All fixes are applied to each variable, not to each correlation, except F1,
which is included to show why pairwise correction fails.

* **F0 as built**: the winsorized, normalized frame.
* **F1 pairwise null subtraction**: R_obs - (R_null - I), one adjustment per
  pair.
* **F2 common size residualization**: every ratio is regressed on natural
  splines of log expenses and log assets; the residuals are re-standardized.
* **F3 denominator-specific residualization**: each ratio is regressed only on
  the scale variable in its own denominator (E, A, or nothing for `current`,
  `payables_una`, and `na_growth`).

```{r fixes}
den_scale <- c(current = "none", months_cash = "E", liquid_months = "E", cash_assets = "A",
               debt_assets = "A", payables_una = "none", surplus_margin = "E",
               return_assets = "A", na_growth = "none", self_suff = "E", admin_exp = "E",
               fundr_exp = "E", land_assets = "A", assets_rev = "E")

residualize <- function(X, k, mode = c("common", "denominator")) {
  mode <- match.arg(mode)
  BE <- ns(log(k$E), df = 4)
  BA <- ns(log1p(pmax(k$A, 0)), df = 4)
  basis <- list(E = cbind(1, BE), A = cbind(1, BA), both = cbind(1, BE, BA), none = NULL)
  out <- lapply(names(X), function(v) {
    b <- if (mode == "common") "both" else den_scale[[v]]
    if (b == "none") return(X[[v]])
    as.numeric(scale(lm.fit(basis[[b]], X[[v]])$residuals))
  })
  names(out) <- names(X)
  as.data.frame(out)
}
size_basis <- cbind(1, ns(log(k$E), df = 4), ns(log1p(pmax(k$A, 0)), df = 4))
size_r2 <- vapply(names(X), function(v) 1 - var(lm.fit(size_basis, X[[v]])$residuals) / var(X[[v]]), 0)
round(sort(size_r2, decreasing = TRUE), 3)
```

`size_r2` is the share of each ratio's variance explained by the size splines.
A value near 1 means residualizing on size erases the variable.

```{r run-null}
reps <- params$reps
run_null <- function(anchor) {
  lapply(seq_len(reps), function(i) {
    kn <- null_lines(k, anchor)
    Xn <- lines_to_ratios(kn)
    list(F0 = cor(Xn),
         F2 = cor(residualize(Xn, kn, "common")),
         F3 = cor(residualize(Xn, kn, "denominator")))
  })
}
null_runs <- list(flow = run_null("flow"), stock = run_null("stock"))

mean_mat <- function(runs, fix) Reduce(`+`, lapply(runs, `[[`, fix)) / length(runs)
eig_q95  <- function(runs, fix) apply(sapply(runs, function(r) eigen(r[[fix]], TRUE, only.values = TRUE)$values),
                                      1, quantile, 0.95)
R_null <- lapply(null_runs, function(runs) lapply(c(F0 = "F0", F2 = "F2", F3 = "F3"),
                                                   function(f) mean_mat(runs, f)))
for (a in names(R_null)) audit[[paste0("r_null_", a)]] <- R_null[[a]]$F0[cbind(audit$v1, audit$v2)]
```

## 5. How much of each correlation is construction?

```{r null-vs-observed, fig.height = 5}
type_col <- c(closure = "#8A3A2C", shared_den = "#2C5A4C", reciprocal = "#C07A2C",
              shared_num = "#4A6FA5", identity = "#7A5C99", none = "#9AA59E")
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (a in c("flow", "stock")) {
  x <- audit[[paste0("r_null_", a)]]
  plot(x, audit$r_obs, pch = 19, col = type_col[audit$type], xlim = c(-1, 1), ylim = c(-1, 1),
       xlab = paste("null r,", a, "anchor"), ylab = "observed r", main = paste(a, "anchor"))
  abline(0, 1, lty = 2); abline(h = 0, v = 0, col = "grey80")
}
legend("bottomright", legend = names(type_col), col = type_col, pch = 19, cex = 0.7, bty = "n")
par(op)
```

```{r null-by-type}
audit$r_null_max <- ifelse(abs(audit$r_null_flow) >= abs(audit$r_null_stock),
                           audit$r_null_flow, audit$r_null_stock)
by_type <- do.call(rbind, lapply(split(audit, audit$type), function(s) data.frame(
  type = s$type[1], pairs = nrow(s),
  mean_abs_obs = mean(abs(s$r_obs)),
  mean_abs_null_flow = mean(abs(s$r_null_flow)),
  mean_abs_null_stock = mean(abs(s$r_null_stock)),
  share_of_obs = sum(abs(s$r_null_max)) / sum(abs(s$r_obs)))))
knitr::kable(by_type[order(-by_type$mean_abs_null_flow), ], digits = 3, row.names = FALSE)

# Does the formula-based sign prediction match the null?
signed <- audit[audit$sign %in% c("+", "-") & abs(audit$r_null_max) >= 0.05, ]
sign_hit <- mean(sign(signed$r_null_max) == ifelse(signed$sign == "+", 1, -1))
none_ok  <- mean(abs(audit$r_null_max[audit$type == "none"]) < 0.05)
c(signed_pairs_with_null_r_ge_05 = nrow(signed), sign_agreement = round(sign_hit, 3),
  none_pairs_below_05 = round(none_ok, 3))
```

```{r largest-null}
top <- audit[order(-abs(audit$r_null_max)), c("v1", "v2", "type", "role", "sign",
                                               "r_obs", "r_null_flow", "r_null_stock")]
knitr::kable(head(top, 20), digits = 2, row.names = FALSE,
             caption = "The 20 pairs with the largest construction-null correlation")
```

## 6. Does arithmetic alone produce components?

```{r eigen, fig.height = 4}
eig_obs <- eigen(R_obs, TRUE, only.values = TRUE)$values
eig_rand <- apply(replicate(20, eigen(cor(matrix(rnorm(nrow(X) * ncol(X)), ncol = ncol(X))),
                                      TRUE, only.values = TRUE)$values), 1, quantile, 0.95)
eig_flow  <- eig_q95(null_runs$flow, "F0")
eig_stock <- eig_q95(null_runs$stock, "F0")
matplot(seq_along(eig_obs), cbind(eig_obs, eig_rand, eig_flow, eig_stock), type = "b",
        pch = 19, lty = 1, col = c("black", "grey60", "#2C5A4C", "#C07A2C"),
        xlab = "component", ylab = "eigenvalue")
legend("topright", c("observed", "random normal (95th pct)", "null, flow anchor (95th pct)",
                     "null, stock anchor (95th pct)"),
       col = c("black", "grey60", "#2C5A4C", "#C07A2C"), pch = 19, bty = "n", cex = 0.8)
# Components retained: leading eigenvalues that exceed the reference, stopping at the first that does not.
n_retain <- function(obs, ref) { x <- c(obs > ref, FALSE); which.min(x) - 1 }
retain <- c(random_normal = n_retain(eig_obs, eig_rand), null_flow = n_retain(eig_obs, eig_flow),
            null_stock = n_retain(eig_obs, eig_stock))
retain
knitr::kable(data.frame(component = seq_along(eig_obs), observed = eig_obs, random_normal = eig_rand,
                        null_flow = eig_flow, null_stock = eig_stock), digits = 2)
```

```{r loadings}
anchors <- c(performance = "return_assets", reserves = "months_cash",
             leverage = "debt_assets", asset_structure = "land_assets",
             expense = "fundr_exp")
pca5 <- function(R) {
  L <- unclass(psych::principal(R, nfactors = 5, rotate = "oblimin", n.obs = nrow(X))$loadings)
  perms <- function(n) if (n == 1) list(1L) else
    do.call(c, lapply(seq_len(n), function(i) lapply(perms(n - 1), function(p) c(i, setdiff(seq_len(n), i)[p]))))
  P <- perms(5)
  A <- abs(L[anchors, , drop = FALSE])
  best <- P[[which.max(vapply(P, function(p) sum(A[cbind(1:5, p)]), 0))]]
  L <- L[, best]
  L <- sweep(L, 2, sign(diag(L[anchors, ])), `*`)   # anchor loads positively
  colnames(L) <- names(anchors)
  L
}
show_loadings <- function(L, caption) {
  L[abs(L) < 0.30] <- NA
  knitr::kable(L, digits = 2, caption = caption)
}
congruence <- function(A, B) diag(psych::factor.congruence(A, B, digits = 3))

L_obs <- pca5(R_obs)
show_loadings(L_obs, "F0 as built: five-component oblimin loadings")
L_null_flow  <- pca5(R_null$flow$F0)
L_null_stock <- pca5(R_null$stock$F0)
show_loadings(L_null_flow, "Construction null, flow anchor")
show_loadings(L_null_stock, "Construction null, stock anchor")
round(rbind(null_flow = congruence(L_null_flow, L_obs),
            null_stock = congruence(L_null_stock, L_obs)), 2)
```

Congruence here compares each null component with the observed component of
the same label. Values of 0.85 or more mean the arithmetic alone reproduces
that component's loading pattern.

## 7. Do the fixes remove the artifact consistently?

```{r fix-null}
off <- function(R) R[upper.tri(R)]
fix_null <- do.call(rbind, lapply(names(R_null), function(a) do.call(rbind, lapply(
  c("F0", "F2", "F3"), function(f) {
    r <- off(R_null[[a]][[f]])
    data.frame(anchor = a, fix = f, mean_abs_null_r = mean(abs(r)),
               max_abs_null_r = max(abs(r)), pairs_abs_ge_10 = sum(abs(r) >= 0.10))
  }))))
knitr::kable(fix_null, digits = 3, row.names = FALSE)

# Remaining null correlation by pair type, after each fix
pair_idx <- cbind(match(audit$v1, candidates), match(audit$v2, candidates))
type_fix <- do.call(rbind, lapply(names(R_null), function(a) {
  m <- sapply(c("F0", "F2", "F3"), function(f) tapply(abs(R_null[[a]][[f]][pair_idx]), audit$type, mean))
  data.frame(anchor = a, type = rownames(m), m, row.names = NULL)
}))
knitr::kable(type_fix, digits = 3, row.names = FALSE,
             caption = "Mean |null r| by pair type and fix")
```

```{r fix-observed}
X_F2 <- residualize(X, k, "common")
X_F3 <- residualize(X, k, "denominator")
R_F2 <- cor(X_F2); R_F3 <- cor(X_F3)
R_F1 <- R_obs - (R_null$flow$F0 - diag(ncol(X)))
R_F1s <- R_obs - (R_null$stock$F0 - diag(ncol(X)))

min_eig <- sapply(list(F0 = R_obs, F1_flow = R_F1, F1_stock = R_F1s, F2 = R_F2, F3 = R_F3),
                  function(R) min(eigen(R, TRUE, only.values = TRUE)$values))
round(min_eig, 3)

ev <- function(R) eigen(R, TRUE, only.values = TRUE)$values
retain_fix <- rbind(
  F0 = c(n_retain(eig_obs, eig_flow), n_retain(eig_obs, eig_stock)),
  F2 = c(n_retain(ev(R_F2), eig_q95(null_runs$flow, "F2")),
         n_retain(ev(R_F2), eig_q95(null_runs$stock, "F2"))),
  F3 = c(n_retain(ev(R_F3), eig_q95(null_runs$flow, "F3")),
         n_retain(ev(R_F3), eig_q95(null_runs$stock, "F3"))))
colnames(retain_fix) <- c("vs null flow", "vs null stock")
retain_fix

R_F1pd <- as.matrix(Matrix::nearPD(R_F1, corr = TRUE)$mat)
dimnames(R_F1pd) <- dimnames(R_obs)
L_F1 <- pca5(R_F1pd); L_F2 <- pca5(R_F2); L_F3 <- pca5(R_F3)
show_loadings(L_F1, "F1 pairwise null subtraction (flow anchor, nearest PD)")
show_loadings(L_F2, "F2 common size residualization")
show_loadings(L_F3, "F3 denominator-specific residualization")
round(rbind(F1 = congruence(L_F1, L_obs), F2 = congruence(L_F2, L_obs),
            F3 = congruence(L_F3, L_obs)), 2)
```

## 8. Findings

```{r findings-values, include = FALSE}
cg_stock <- congruence(L_null_stock, L_obs)
cg_flow  <- congruence(L_null_flow, L_obs)
share    <- setNames(by_type$share_of_obs, by_type$type)
fn <- function(x, d = 2) formatC(x, format = "f", digits = d)
fx <- function(a, f, col = "mean_abs_null_r") fix_null[fix_null$anchor == a & fix_null$fix == f, col]
```

**Most of the correlation structure is construction.** For closure,
reciprocal, identity, and shared-numerator pairs, the construction null
reproduces `r fn(100 * min(share[c("closure", "reciprocal", "identity", "shared_num")]), 0)`%
or more of the observed absolute correlation. Only shared-denominator pairs
(`r fn(100 * share[["shared_den"]], 0)`%) and unlinked pairs
(`r fn(100 * share[["none"]], 0)`%) carry mostly non-arithmetic
correlation. The formula-based sign prediction matches the null for
`r fn(100 * sign_hit, 0)`% of the `r nrow(signed)` pairs where the null
correlation reaches 0.05.

**Arithmetic alone reproduces the dimensions.** With assets as the size
reference, PCA on the null matrix recovers the observed performance, reserves,
leverage, and asset-structure components with congruence
`r fn(cg_stock[["performance"]])`, `r fn(cg_stock[["reserves"]])`,
`r fn(cg_stock[["leverage"]])`, and `r fn(cg_stock[["asset_structure"]])`.
The null's leading eigenvalues (`r paste(fn(eig_stock[1:3]), collapse = ", ")`)
sit next to the observed ones (`r paste(fn(eig_obs[1:3]), collapse = ", ")`).
Against a construction null, parallel analysis retains
`r retain[["null_stock"]]` component, not `r retain[["random_normal"]]`.
The components are shared-ingredient components: `surplus_margin`,
`return_assets`, and `na_growth` are one number (the surplus) on three scales,
and `months_cash`, `liquid_months`, and `assets_rev` all inherit the ratio of
assets to expenses.

**No single fix is consistent across the set.**

* *Pairwise null subtraction (F1)* produces an invalid correlation matrix
  (smallest eigenvalue `r fn(min_eig[["F1_flow"]])`), and after projection
  to the nearest valid matrix the components no longer resemble any
  interpretable structure.
* *Common size residualization (F2)* removes most shared-denominator artifact
  (mean |null r| for those pairs falls from
  `r fn(type_fix$F0[type_fix$anchor == "flow" & type_fix$type == "shared_den"], 3)` to
  `r fn(type_fix$F2[type_fix$anchor == "flow" & type_fix$type == "shared_den"], 3)`),
  but leaves closure and shared-numerator artifact untouched and erases
  `r fn(100 * size_r2[["assets_rev"]], 0)`% of `assets_rev`. Reserves and
  asset structure are rebuilt around what is left (congruence
  `r fn(congruence(L_F2, L_obs)[["reserves"]])` and
  `r fn(congruence(L_F2, L_obs)[["asset_structure"]])`).
* *Denominator-specific residualization (F3)* keeps the matrix valid and the
  structure intact, but lowers the average null correlation only from
  `r fn(fx("flow", "F0"), 3)` to `r fn(fx("flow", "F3"), 3)`.

The artifact lives in different places for different link types, so the repair
has to be chosen by link type. The one step that applies to all 91 pairs is the
construction null itself, used as a benchmark: for retention decisions, for
reading loadings, and for reporting how much of each correlation exceeds
what the arithmetic implies.

## 9. Repairing the variable set by link type

Section 7 transformed the 14 ratios. This section instead **redesigns the set**,
choosing one repair per link type, and runs the redesigned set through the same
construction null.

| Link type | Ratios affected | Repair | New variable |
|---|---|---|---|
| Closure | `cash_assets`, `land_assets` | one bounded contrast, 0 when both parts are 0 | `cash_land` = (cash - land) / (cash + land) |
| Closure | `admin_exp`, `fundr_exp` | one bounded contrast, imputed when both parts are 0 | `admin_fundr` = (M - Fu) / (M + Fu) |
| Shared numerator, construct | `surplus_margin`, `return_assets`, `na_growth` | enter the surplus once | `surplus` = mean of the three normalized ratios |
| Stock over flow | `months_cash`, `liquid_months`, `assets_rev` | split into a share of assets and one scale | `liquid_assets_share` = (cash + recv) / A; `una_liquid_share` = (una - land + mtg) / A; `assets_exp` = A / E |
| Reciprocal, identity | `current`, `debt_assets`, `payables_una`, `self_suff` | kept as built | unchanged |

Because months of cash = 12 x `liquid_assets_share` x `assets_exp` (up to the
depreciation adjustment), the redesign keeps the information in the original
ratios. The difference is that the assets-to-expenses scale enters once, as its
own variable, instead of being hidden in three.

```{r redesign-def}
redesign_vars <- c("current", "liquid_assets_share", "una_liquid_share", "assets_exp",
                   "cash_land", "debt_assets", "payables_una", "surplus", "self_suff",
                   "admin_fundr")

redesign <- function(k, frame) {
  una <- ifelse(k$UNA == 0 & k$RES == 0, k$NA_eoy, k$UNA)
  new <- with(k, data.frame(
    liquid_assets_share = ifelse(A > 0, pmin((cash + recv) / A, 1), NA),
    una_liquid_share    = ifelse(A > 0, (una - land + mtg) / A, NA),
    assets_exp          = A / E,
    cash_land           = ifelse(cash + land > 0, pmin(pmax((cash - land) / (cash + land), -1), 1), 0),
    admin_fundr         = ifelse(M + Fu > 0, pmin(pmax((M - Fu) / (M + Fu), -1), 1), NA)))
  base_preds <- c("liquid_months", "surplus_margin", "months_cash", "self_suff")
  for (v in names(new)) if (anyNA(new[[v]])) {
    new[[v]] <- cond_impute(new[[v]], frame[, base_preds], is.na(new[[v]]))
  }
  Z <- as.data.frame(lapply(cbind(frame[c("current", "debt_assets", "payables_una", "self_suff",
                                          "surplus_margin", "return_assets", "na_growth")], new),
                            winsor_normalize))
  Z$surplus <- as.numeric(scale(rowMeans(Z[c("surplus_margin", "return_assets", "na_growth")])))
  Z[redesign_vars]
}

X_S2 <- redesign(k, frame_obs)
R_S2 <- cor(X_S2)
```

```{r redesign-null}
set.seed(params$seed + 1)
run_null_S2 <- function(anchor) {
  lapply(seq_len(reps), function(i) {
    kn <- null_lines(k, anchor)
    cor(redesign(kn, lines_to_frame(kn)))
  })
}
null_S2 <- list(flow = run_null_S2("flow"), stock = run_null_S2("stock"))
R_null_S2 <- lapply(null_S2, function(runs) Reduce(`+`, runs) / length(runs))
eig_q95_S2 <- lapply(null_S2, function(runs)
  apply(sapply(runs, function(r) eigen(r, TRUE, only.values = TRUE)$values), 1, quantile, 0.95))
eig_rand_S2 <- apply(replicate(20, eigen(cor(matrix(rnorm(nrow(X_S2) * ncol(X_S2)), ncol = ncol(X_S2))),
                                         TRUE, only.values = TRUE)$values), 1, quantile, 0.95)
```

### 9.1 How much construction is left?

```{r redesign-summary}
set_summary <- function(label, R_o, R_f, R_s, e_rand, e_f, e_s) {
  up <- upper.tri(R_o)
  data.frame(set = label, variables = ncol(R_o), pairs = sum(up),
             mean_abs_obs = mean(abs(R_o[up])),
             mean_abs_null_flow = mean(abs(R_f[up])), mean_abs_null_stock = mean(abs(R_s[up])),
             share_pairs_null_ge_10 = mean(pmax(abs(R_f[up]), abs(R_s[up])) >= 0.10),
             retain_random = n_retain(ev(R_o), e_rand),
             retain_null_flow = n_retain(ev(R_o), e_f), retain_null_stock = n_retain(ev(R_o), e_s))
}
s2_summary <- rbind(
  set_summary("F0: 14 ratios as built", R_obs, R_null$flow$F0, R_null$stock$F0, eig_rand, eig_flow, eig_stock),
  set_summary("S2: redesigned set", R_S2, R_null_S2$flow, R_null_S2$stock,
              eig_rand_S2, eig_q95_S2$flow, eig_q95_S2$stock))
knitr::kable(s2_summary, digits = 3, row.names = FALSE)

knitr::kable(data.frame(component = seq_len(ncol(R_S2)), observed = ev(R_S2), random_normal = eig_rand_S2,
                        null_flow = eig_q95_S2$flow, null_stock = eig_q95_S2$stock),
             digits = 2, caption = "Redesigned set: eigenvalues")
```

```{r redesign-pairs}
ij <- which(upper.tri(R_S2), arr.ind = TRUE)
s2_pairs <- data.frame(v1 = redesign_vars[ij[, 1]], v2 = redesign_vars[ij[, 2]],
                       r_obs = R_S2[ij], r_null_flow = R_null_S2$flow[ij],
                       r_null_stock = R_null_S2$stock[ij])
# Beyond null: the observed correlation minus the null that lies closer to it.
s2_pairs$beyond_null <- with(s2_pairs, ifelse(abs(r_obs - r_null_flow) < abs(r_obs - r_null_stock),
                                              r_obs - r_null_flow, r_obs - r_null_stock))
s2_pairs <- s2_pairs[order(-abs(s2_pairs$r_obs)), ]
knitr::kable(head(s2_pairs, 20), digits = 2, row.names = FALSE,
             caption = "Redesigned set: the 20 largest observed correlations and their nulls")
```

The stock-over-flow decomposition is the clearest test. Before the redesign,
`months_cash` and `liquid_months` had a null correlation of
`r fn(R_null$flow$F0["months_cash", "liquid_months"])` (flow) and
`r fn(R_null$stock$F0["months_cash", "liquid_months"])` (stock). Their
asset-share counterparts have
`r fn(R_null_S2$flow["liquid_assets_share", "una_liquid_share"])` and
`r fn(R_null_S2$stock["liquid_assets_share", "una_liquid_share"])`.

### 9.2 What structure does the redesigned set carry?

```{r redesign-loadings}
perms <- function(n) if (n == 1) list(1L) else
  do.call(c, lapply(seq_len(n), function(i) lapply(perms(n - 1), function(p) c(i, setdiff(seq_len(n), i)[p]))))
pca_k <- function(R, nf) {
  L <- unclass(psych::principal(R, nfactors = nf, rotate = if (nf > 1) "oblimin" else "none",
                                n.obs = nrow(X))$loadings)
  colnames(L) <- paste0("C", seq_len(nf))
  L
}
# Reorder and re-sign the columns of L to best match ref; return matched loadings and congruence.
match_to <- function(L, ref) {
  C <- psych::factor.congruence(L, ref, digits = 3)
  P <- perms(ncol(ref))
  p <- P[[which.max(vapply(P, function(q) sum(abs(C[cbind(q, seq_len(ncol(ref)))])), 0))]]
  s <- sign(C[cbind(p, seq_len(ncol(ref)))])
  L <- sweep(L[, p, drop = FALSE], 2, s, `*`)
  colnames(L) <- colnames(ref)
  list(L = L, congruence = setNames(abs(C[cbind(p, seq_len(ncol(ref)))]), colnames(ref)))
}

k_S2 <- max(2, n_retain(ev(R_S2), eig_rand_S2))
L_S2 <- pca_k(R_S2, k_S2)
L_S2_flow  <- match_to(pca_k(R_null_S2$flow, k_S2), L_S2)
L_S2_stock <- match_to(pca_k(R_null_S2$stock, k_S2), L_S2)
show_loadings(L_S2, paste0("Redesigned set, observed: ", k_S2, " components"))
show_loadings(L_S2_stock$L, "Redesigned set, construction null (stock anchor)")
round(rbind(null_flow = L_S2_flow$congruence, null_stock = L_S2_stock$congruence), 2)
```

### 9.3 Findings for the redesign

```{r redesign-values, include = FALSE}
f0s <- s2_summary[1, ]; s2s <- s2_summary[2, ]
```

* **Construction correlation.** Mean |null r| is
  `r fn(s2s$mean_abs_null_flow, 3)` (flow) and `r fn(s2s$mean_abs_null_stock, 3)`
  (stock) in the redesigned set, against `r fn(f0s$mean_abs_null_flow, 3)` and
  `r fn(f0s$mean_abs_null_stock, 3)` for the as-built ratios. The share of pairs
  with a null correlation of 0.10 or more is
  `r fn(100 * s2s$share_pairs_null_ge_10, 0)`%, against
  `r fn(100 * f0s$share_pairs_null_ge_10, 0)`%. The observed mean |r| is
  `r fn(s2s$mean_abs_obs, 3)`, against `r fn(f0s$mean_abs_obs, 3)`.
* **Where the construction correlation now sits.** The largest null
  correlations are
  `r s2_top <- s2_pairs[order(-pmax(abs(s2_pairs$r_null_flow), abs(s2_pairs$r_null_stock))), ][1:4, ]; paste(sprintf("%s x %s (%s)", s2_top$v1, s2_top$v2, fn(ifelse(abs(s2_top$r_null_flow) > abs(s2_top$r_null_stock), s2_top$r_null_flow, s2_top$r_null_stock))), collapse = "; ")`.
  Putting every balance-sheet quantity over total assets turns stock-over-flow
  coupling into composition: the asset shares, `debt_assets`, and
  `una_liquid_share` are parts of the same balance sheet.
* **Retention against the null.** Parallel analysis keeps
  `r s2s$retain_random` components against random data, and
  `r s2s$retain_null_flow` (flow) and `r s2s$retain_null_stock` (stock) against
  the construction null, compared with `r f0s$retain_null_flow` and
  `r f0s$retain_null_stock` for the as-built ratios.
* **How much of the remaining structure is arithmetic.** Congruence of the
  null components with the observed `r k_S2`-component solution is
  `r paste(fn(L_S2_stock$congruence), collapse = ", ")` (stock anchor) and
  `r paste(fn(L_S2_flow$congruence), collapse = ", ")` (flow anchor).
* **What the data add beyond arithmetic.** The largest gaps between an observed
  correlation and the nearer null are
  `r s2_gap <- s2_pairs[order(-abs(s2_pairs$beyond_null)), ][1:5, ]; paste(sprintf("%s x %s (%s observed, %s beyond null)", s2_gap$v1, s2_gap$v2, fn(s2_gap$r_obs), fn(s2_gap$beyond_null)), collapse = "; ")`.
  These are the relationships a factor model of this set would be entitled to
  interpret.

## 10. A factor model with construction held fixed

Sections 7 and 9 tried to remove construction correlation from the data. This
section keeps the data as they are and puts the arithmetic into the model. The
construction null gives the correlation matrix the ratios would have if only
arithmetic linked them. Call its off-diagonal part **C**. The factor model is

    Sigma = Lambda Lambda' + Psi + C

with C **fixed**, and Lambda and Psi estimated by maximum likelihood. Unlike
the pairwise subtraction in section 7, the model is fit to the observed
matrix, which is a valid correlation matrix. The factors only have to explain
covariance the arithmetic does not already account for.

Two questions:

1. **Does it work when the truth is known?** Plant known factors in the 990
   lines, rebuild the ratios, and check whether the model recovers them where
   ordinary factor analysis does not.
2. **What does it find in the real data?** Fit it to the 14 as-built ratios,
   to the redesigned set from section 9, and to a composition set that treats
   the balance sheet as parts of one total.

### 10.1 The model

```{r offset-model}
# ML fit of Sigma = L L' + diag(psi) + C, with C's off-diagonal fixed.
fit_offset <- function(S, C = NULL, nf, n.obs) {
  p <- ncol(S)
  if (is.null(C)) C <- matrix(0, p, p)
  diag(C) <- 0
  if (nf > 0) {
    e  <- eigen(S - C, symmetric = TRUE)
    L0 <- e$vectors[, seq_len(nf), drop = FALSE] %*%
          diag(sqrt(pmax(e$values[seq_len(nf)] - 1, 0.05)), nf)
    L0 <- L0 * pmin(1, 0.9 / sqrt(pmax(rowSums(L0^2), 1e-8)))
  } else {
    L0 <- matrix(0, p, 0)
  }
  psi0 <- pmax(1 - rowSums(L0^2), 0.05)
  bump <- 0
  while (inherits(try(chol(tcrossprod(L0) + diag(psi0 + bump) + C), silent = TRUE), "try-error"))
    bump <- bump + 0.1
  unpack <- function(th) list(L = matrix(th[seq_len(p * nf)], p, nf), psi = exp(th[p * nf + seq_len(p)]))
  sigma  <- function(u) tcrossprod(u$L) + diag(u$psi) + C
  fn <- function(th) {
    ch <- try(chol(sigma(unpack(th))), silent = TRUE)
    if (inherits(ch, "try-error")) return(1e10)
    2 * sum(log(diag(ch))) + sum(S * chol2inv(ch))
  }
  gr <- function(th) {
    u  <- unpack(th)
    ch <- try(chol(sigma(u)), silent = TRUE)
    if (inherits(ch, "try-error")) return(rep(0, length(th)))
    Si <- chol2inv(ch)
    G  <- Si - Si %*% S %*% Si
    c(2 * G %*% u$L, diag(G) * u$psi)
  }
  opt <- optim(c(L0, log(psi0 + bump)), fn, gr, method = "BFGS", control = list(maxit = 5000))
  u   <- unpack(opt$par)
  Sig <- sigma(u)
  Fml <- opt$value - as.numeric(determinant(S)$modulus) - p
  df  <- p * (p + 1) / 2 - (p * nf + p - nf * (nf - 1) / 2)
  chi <- (n.obs - 1) * Fml
  low <- lower.tri(S, diag = TRUE)
  L <- u$L
  rownames(L) <- colnames(S)
  list(L = L, psi = u$psi, converged = opt$convergence == 0, df = df,
       rmsea = sqrt(max(chi - df, 0) / (max(df, 1) * (n.obs - 1))),
       srmr = sqrt(mean((S - Sig)[low]^2)))
}

# Oblimin pattern (for display) and structure (for comparison with truth).
rotate_fit <- function(L) {
  if (ncol(L) < 2) {
    colnames(L) <- paste0("F", seq_len(ncol(L)))
    return(list(pattern = L, structure = L))
  }
  r <- GPArotation::oblimin(L)
  P <- unclass(r$loadings)
  colnames(P) <- paste0("F", seq_len(ncol(P)))
  S <- P %*% r$Phi
  dimnames(S) <- dimnames(P)
  list(pattern = P, structure = S)
}

null_mean_offdiag <- function(null_list) {
  C <- Reduce(`+`, null_list) / length(null_list)
  diag(C) <- 0
  C
}

# Parallel analysis for the fixed-construction model: eigenvalues of S - C
# against those of each null draw minus C.
offset_retain <- function(S, null_list) {
  C <- null_mean_offdiag(null_list)
  q <- apply(sapply(null_list, function(R) ev(R - C)), 1, quantile, 0.95)
  n_retain(ev(S - C), q)
}
rand_eig <- function(p, n) apply(replicate(20, ev(cor(matrix(rnorm(n * p), ncol = p)))),
                                 1, quantile, 0.95)
```

### 10.2 Test against planted factors

The planted data use the same lines and the same flow-anchored shuffle as the
construction null. The difference is that some lines are reassigned through
a Gaussian copula driven by three latent factors, so each line keeps its exact
distribution but co-moves with the others that share its factor:

* **liquidity**: cash, receivables, other current assets up; payables and
  grants payable down;
* **capital**: land and buildings, mortgages, other liabilities, depreciation
  up;
* **performance**: revenue relative to expenses, other net-asset changes,
  program revenue up.

The "true" loading of each ratio on each factor is its correlation with the
planted factor. A second scenario plants nothing, so any factor found there is
false.

```{r planted}
line_loadings <- rbind(
  cash = c(0.7, 0, 0), recv = c(0.5, 0, 0), othca = c(0.3, 0, 0), ap = c(-0.4, 0, 0),
  gp = c(-0.3, 0, 0), land = c(0, 0.7, 0), mtg = c(0, 0.6, 0), otherL = c(0, 0.3, 0),
  dep = c(0, 0.5, 0), R = c(0, 0, 0.7), na_oth = c(0, 0, 0.4), progrev = c(0, 0, 0.3))
colnames(line_loadings) <- c("liquidity", "capital", "performance")

planted_lines <- function(k, lam) {
  n  <- nrow(k)
  Fm <- matrix(rnorm(n * ncol(lam)), n, ncol(lam), dimnames = list(NULL, colnames(lam)))
  draw <- function(x, line) {
    l <- if (line %in% rownames(lam)) lam[line, ] else rep(0, ncol(lam))
    if (all(l == 0)) return(x[sample.int(n)])
    s <- drop(Fm %*% l) + sqrt(1 - sum(l^2)) * rnorm(n)
    sort(x)[rank(s, ties.method = "first")]
  }
  list(k = null_lines(k, "flow", draw), F = Fm)
}

reps_p <- min(reps, 10)
run_scenario <- function(lam) {
  pl <- planted_lines(k, lam)
  Xp <- lines_to_ratios(pl$k)
  list(S = cor(Xp), T = cor(Xp, pl$F),
       nulls = lapply(c(flow = "flow", stock = "stock"), function(a)
         lapply(seq_len(reps_p), function(i) cor(lines_to_ratios(null_lines(pl$k, a))))))
}
set.seed(params$seed + 2)
scen <- list(none = run_scenario(line_loadings * 0), planted = run_scenario(line_loadings))
eig_rand_14 <- rand_eig(length(candidates), nrow(X))

select_table <- function(S, nulls, e_rand) {
  null_q <- function(a) apply(sapply(nulls[[a]], ev), 1, quantile, 0.95)
  data.frame(random_pa = n_retain(ev(S), e_rand),
             construction_pa_flow = n_retain(ev(S), null_q("flow")),
             construction_pa_stock = n_retain(ev(S), null_q("stock")),
             fixed_construction_flow = offset_retain(S, nulls$flow),
             fixed_construction_stock = offset_retain(S, nulls$stock))
}
planted_sel <- cbind(scenario = c("nothing planted", "3 factors planted"), true_factors = c(0, 3),
                     rbind(select_table(scen$none$S, scen$none$nulls, eig_rand_14),
                           select_table(scen$planted$S, scen$planted$nulls, eig_rand_14)))
knitr::kable(planted_sel, row.names = FALSE, caption = "Number of factors retained")
```

```{r planted-recovery}
sp <- scen$planted
C_pf <- null_mean_offdiag(sp$nulls$flow)
C_ps <- null_mean_offdiag(sp$nulls$stock)
recover <- function(fit) {
  rf <- rotate_fit(fit$L)
  m  <- match_to(rf$structure, sp$T)
  list(congruence = m$congruence, structure = m$L)
}
planted_fits <- list(
  `ordinary ML factor analysis`       = fit_offset(sp$S, NULL, 3, nrow(X)),
  `fixed construction, flow null`     = fit_offset(sp$S, C_pf, 3, nrow(X)),
  `fixed construction, stock null`    = fit_offset(sp$S, C_ps, 3, nrow(X)))
planted_rec <- lapply(planted_fits, recover)

# Control: the same three fits on data with nothing planted, scored against
# the planted structure. A method that "recovers" the planted structure here
# is reading arithmetic, not the planted factors.
sn <- scen$none
control_fits <- list(
  `ordinary ML factor analysis`    = fit_offset(sn$S, NULL, 3, nrow(X)),
  `fixed construction, flow null`  = fit_offset(sn$S, null_mean_offdiag(sn$nulls$flow), 3, nrow(X)),
  `fixed construction, stock null` = fit_offset(sn$S, null_mean_offdiag(sn$nulls$stock), 3, nrow(X)))
control_rec <- lapply(control_fits, recover)

recovery_table <- data.frame(
  method = names(planted_rec),
  planted = sapply(planted_rec, function(r) mean(r$congruence)),
  nothing_planted = sapply(control_rec, function(r) mean(r$congruence)),
  t(sapply(planted_rec, `[[`, "congruence")), row.names = NULL, check.names = FALSE)
recovery_table$detection_margin <- recovery_table$planted - recovery_table$nothing_planted
knitr::kable(recovery_table, digits = 2, row.names = FALSE,
             caption = "Mean congruence with the planted structure (3 factors), with and without planting")

show_loadings(sp$T, "Planted structure: correlation of each ratio with each factor")
show_loadings(planted_rec$`ordinary ML factor analysis`$structure, "Ordinary ML factor analysis")
show_loadings(planted_rec$`fixed construction, flow null`$structure, "Fixed construction, flow null")

planted_fit_k <- do.call(rbind, lapply(names(scen), function(s) {
  sc <- scen[[s]]
  Cf <- null_mean_offdiag(sc$nulls$flow); Cs <- null_mean_offdiag(sc$nulls$stock)
  do.call(rbind, lapply(0:4, function(nf) {
    f <- list(ordinary = fit_offset(sc$S, NULL, nf, nrow(X)),
              fixed_flow = fit_offset(sc$S, Cf, nf, nrow(X)),
              fixed_stock = fit_offset(sc$S, Cs, nf, nrow(X)))
    data.frame(scenario = s, factors = nf,
               t(unlist(lapply(f, function(x) c(rmsea = x$rmsea, srmr = x$srmr)))))
  }))
}))
knitr::kable(planted_fit_k, digits = 3, row.names = FALSE,
             caption = "Fit by number of factors, both scenarios")
```

### 10.3 The balance sheet as parts of one total

The composition set replaces every balance-sheet ratio with a share of total
assets from one of the two partitions of the balance sheet. On the asset side
the parts are liquid assets (cash, receivables, other current assets), land and
buildings, and other assets. On the funding side they are liabilities,
unrestricted net assets, and restricted net assets. One part per side is left
out so the shares are not perfectly collinear; when restricted net assets are
not reported (as in `dat10k`), the funding side is liabilities alone. Flow
measures enter as ratios to expenses. The closure among the shares is left in
the data on purpose: the fixed C in the model accounts for it.

```{r composition}
composition <- function(k, frame, with_una) {
  una <- ifelse(k$UNA == 0 & k$RES == 0, k$NA_eoy, k$UNA)
  new <- with(k, data.frame(
    liquid_share = ifelse(A > 0, (cash + recv + othca) / A, NA),
    fixed_share  = ifelse(A > 0, land / A, NA),
    liab_share   = ifelse(A > 0, L / A, NA),
    una_share    = ifelse(A > 0, una / A, NA),
    assets_exp   = A / E))
  if (!with_una) new$una_share <- NULL
  base_preds <- c("liquid_months", "surplus_margin", "months_cash", "self_suff")
  for (v in names(new)) if (anyNA(new[[v]])) {
    new[[v]] <- cond_impute(new[[v]], frame[, base_preds], is.na(new[[v]]))
  }
  as.data.frame(lapply(cbind(new, frame[c("surplus_margin", "self_suff", "admin_exp", "fundr_exp")]),
                       winsor_normalize))
}
with_una <- any(k$RES != 0)
X_S3 <- composition(k, frame_obs, with_una)
R_S3 <- cor(X_S3)

set.seed(params$seed + 3)
null_S3 <- lapply(c(flow = "flow", stock = "stock"), function(a) lapply(seq_len(reps), function(i) {
  kn <- null_lines(k, a)
  cor(composition(kn, lines_to_frame(kn), with_una))
}))
names(X_S3)
```

### 10.4 The real data

```{r real-offset}
real_sets <- list(
  `as built (14 ratios)` = list(S = R_obs, nulls = lapply(null_runs, function(runs) lapply(runs, `[[`, "F0"))),
  `redesigned (section 9)` = list(S = R_S2, nulls = null_S2),
  `composition` = list(S = R_S3, nulls = null_S3))

real_sel <- do.call(rbind, lapply(names(real_sets), function(nm) {
  s <- real_sets[[nm]]
  cbind(set = nm, variables = ncol(s$S), select_table(s$S, s$nulls, rand_eig(ncol(s$S), nrow(X))))
}))
knitr::kable(real_sel, row.names = FALSE, caption = "Number of factors retained, real data")

real_fit_k <- do.call(rbind, lapply(names(real_sets), function(nm) {
  s <- real_sets[[nm]]
  Cf <- null_mean_offdiag(s$nulls$flow); Cs <- null_mean_offdiag(s$nulls$stock)
  do.call(rbind, lapply(0:4, function(nf) {
    f <- list(ordinary = fit_offset(s$S, NULL, nf, nrow(X)),
              fixed_flow = fit_offset(s$S, Cf, nf, nrow(X)),
              fixed_stock = fit_offset(s$S, Cs, nf, nrow(X)))
    data.frame(set = nm, factors = nf, t(unlist(lapply(f, function(x) c(rmsea = x$rmsea, srmr = x$srmr)))))
  }))
}))
knitr::kable(real_fit_k, digits = 3, row.names = FALSE, caption = "Fit by number of factors, real data")
```

```{r real-loadings, results = "asis"}
real_loadings <- lapply(names(real_sets), function(nm) {
  s  <- real_sets[[nm]]
  out <- list()
  for (a in c("flow", "stock")) {
    nf <- max(1, min(4, real_sel$fixed_construction_flow[real_sel$set == nm] * (a == "flow") +
                        real_sel$fixed_construction_stock[real_sel$set == nm] * (a == "stock")))
    fit <- fit_offset(s$S, null_mean_offdiag(s$nulls[[a]]), nf, nrow(X))
    out[[a]] <- rotate_fit(fit$L)$pattern
    cat(show_loadings(out[[a]], paste0(nm, ": fixed construction, ", a, " null, ", nf, " factor(s)")),
        sep = "\n")
    cat("\n\n")
  }
  out
})
names(real_loadings) <- names(real_sets)
```

### 10.5 Findings

```{r offset-values, include = FALSE}
ps <- planted_sel
rt <- recovery_table
pr <- setNames(rt$planted, rt$method)
pn <- setNames(rt$nothing_planted, rt$method)
rs <- real_sel
```

* **False factors.** With nothing planted, ordinary parallel analysis keeps
  `r ps$random_pa[1]` factor(s). Parallel analysis against the construction
  null keeps `r ps$construction_pa_flow[1]` (flow) and
  `r ps$construction_pa_stock[1]` (stock). The fixed-construction model keeps
  `r ps$fixed_construction_flow[1]` (flow) and `r ps$fixed_construction_stock[1]`
  (stock).
* **True factors.** With three factors planted, the same methods keep
  `r ps$random_pa[2]`, `r ps$construction_pa_flow[2]`,
  `r ps$construction_pa_stock[2]`, `r ps$fixed_construction_flow[2]`,
  and `r ps$fixed_construction_stock[2]`.
* **Recovery.** Mean congruence with the planted structure is
  `r fn(pr[["ordinary ML factor analysis"]])` for ordinary factor analysis,
  `r fn(pr[["fixed construction, flow null"]])` with the matching (flow)
  construction null held fixed, and `r fn(pr[["fixed construction, stock null"]])`
  with the mismatched (stock) null. On data with nothing planted, the same
  fits score `r fn(pn[["ordinary ML factor analysis"]])`,
  `r fn(pn[["fixed construction, flow null"]])`, and
  `r fn(pn[["fixed construction, stock null"]])` against the planted structure.
  Only the gap between the two is evidence that a method detects the planted
  factors rather than the arithmetic.
* **Real data.** The fixed-construction model keeps
  `r paste(sprintf("%s (flow) and %s (stock) factors for the %s set", rs$fixed_construction_flow, rs$fixed_construction_stock, rs$set), collapse = "; ")`.
* **Single-variable factors.** Of the factors shown in 10.4,
  `r singles <- unlist(lapply(real_loadings, function(s) lapply(s, function(P) apply(abs(P), 2, max) >= 0.95))); sum(singles)`
  of `r length(singles)` have a loading of 0.95 or more on one variable. A
  factor like that is not a shared dimension. It absorbs the misfit between
  that variable's observed correlations and the ones the construction null
  implies, which is what happens when the null does not describe how the real
  lines are built.

## 11. A conditional construction null

Both nulls so far break the real relationship between expenses and assets: the
flow null keeps expenses and scrambles assets relative to them, and the stock
null does the reverse. Section 10 showed the fixed-construction model is only
as good as its null. This section builds a **conditional** null: organizations
are grouped into cells of similar expenses and similar assets per dollar of
expenses, each organization keeps its own expenses, and every size-scaled line
is shuffled only within its cell.

The conditional null is stricter than the other two. It keeps the arithmetic
**and** every relationship that runs through size and asset intensity. A
correlation that exceeds it cannot be explained by construction or by how big
and how asset-heavy organizations are.

```{r joint-null-def}
cell_bins <- if (nrow(k) > 1e5) 20 else 10

# Cells: bins of expenses, then bins of assets per dollar of expenses within each.
size_cells <- function(k, bins) {
  eb <- cut(rank(k$E, ties.method = "first"), bins, labels = FALSE)
  ae <- ifelse(k$A > 0, k$A / k$E, -1)
  ab <- ave(ae, eb, FUN = function(v) cut(rank(v, ties.method = "first"), bins, labels = FALSE))
  (eb - 1) * bins + ab
}

# Reassign values within cells in the order of a score: random for a shuffle,
# factor-driven for planted structure.
within_cells <- function(cell, score = function(line, n) runif(n)) {
  force(cell)
  function(x, line) {
    s   <- score(line, length(x))
    out <- numeric(length(x))
    out[order(cell, s)] <- x[order(cell, x)]
    out
  }
}

cells_obs <- size_cells(k, cell_bins)
c(cells = length(unique(cells_obs)), median_orgs_per_cell = median(table(cells_obs)))
```

### 11.1 How much of the observed structure does it absorb?

```{r joint-null-run}
set.seed(params$seed + 4)
joint_runs <- lapply(seq_len(reps), function(i) {
  kn <- null_lines(k, "joint", within_cells(cells_obs))
  fr <- lines_to_frame(kn)
  list(F0 = cor(normalize_frame(fr)), S2 = cor(redesign(kn, fr)),
       S3 = cor(composition(kn, fr, with_una)))
})
R_joint <- lapply(c(F0 = "F0", S2 = "S2", S3 = "S3"), function(s)
  Reduce(`+`, lapply(joint_runs, `[[`, s)) / reps)

audit$r_null_joint <- R_joint$F0[cbind(audit$v1, audit$v2)]
joint_by_type <- do.call(rbind, lapply(split(audit, audit$type), function(s) data.frame(
  type = s$type[1], pairs = nrow(s), mean_abs_obs = mean(abs(s$r_obs)),
  mean_abs_null_flow = mean(abs(s$r_null_flow)), mean_abs_null_stock = mean(abs(s$r_null_stock)),
  mean_abs_null_joint = mean(abs(s$r_null_joint)))))
knitr::kable(joint_by_type, digits = 3, row.names = FALSE,
             caption = "Mean |r| by link type: observed and three nulls (14 as-built ratios)")

eig_joint <- apply(sapply(joint_runs, function(r) ev(r$F0)), 1, quantile, 0.95)
L_null_joint <- pca5(R_joint$F0)
show_loadings(L_null_joint, "Conditional null: five-component oblimin loadings")
joint_summary <- data.frame(
  null = c("flow", "stock", "joint"),
  mean_abs_null_r = c(mean(abs(off(R_null$flow$F0))), mean(abs(off(R_null$stock$F0))),
                      mean(abs(off(R_joint$F0)))),
  retained = c(retain[["null_flow"]], retain[["null_stock"]], n_retain(eig_obs, eig_joint)),
  rbind(congruence(L_null_flow, L_obs), congruence(L_null_stock, L_obs),
        congruence(L_null_joint, L_obs)))
knitr::kable(joint_summary, digits = 2, row.names = FALSE,
             caption = "Mean |null r|, components retained, and null congruence with the observed components")

gap_joint <- audit[order(-abs(audit$r_obs - audit$r_null_joint)),
                   c("v1", "v2", "type", "r_obs", "r_null_flow", "r_null_stock", "r_null_joint")]
knitr::kable(head(gap_joint, 10), digits = 2, row.names = FALSE,
             caption = "Largest gaps between observed r and the conditional null")
```

### 11.2 Planted-factor test, rerun

The planted data are now generated inside the same cells: each line is
reassigned within its cell in the order of the planted factor score. The
conditional null is therefore the correct one, and the global flow null is the
wrong one.

```{r joint-planted}
planted_within <- function(cell, lam, Fm) within_cells(cell, function(line, n) {
  l <- if (line %in% rownames(lam)) lam[line, ] else rep(0, ncol(lam))
  if (all(l == 0)) runif(n) else drop(Fm %*% l) + sqrt(1 - sum(l^2)) * rnorm(n)
})

run_scenario_joint <- function(lam) {
  n  <- nrow(k)
  Fm <- matrix(rnorm(n * ncol(lam)), n, ncol(lam), dimnames = list(NULL, colnames(lam)))
  kp <- null_lines(k, "joint", planted_within(cells_obs, lam, Fm))
  Xp <- lines_to_ratios(kp)
  cp <- size_cells(kp, cell_bins)   # cells rebuilt from the planted data, as they would be in practice
  list(S = cor(Xp), T = cor(Xp, Fm),
       nulls = list(
         joint = lapply(seq_len(reps_p), function(i)
           cor(lines_to_ratios(null_lines(kp, "joint", within_cells(cp))))),
         flow  = lapply(seq_len(reps_p), function(i) cor(lines_to_ratios(null_lines(kp, "flow"))))))
}
set.seed(params$seed + 5)
scen_j <- list(none = run_scenario_joint(line_loadings * 0),
               planted = run_scenario_joint(line_loadings))

select_by <- function(S, nulls, e_rand) {
  out <- data.frame(random_pa = n_retain(ev(S), e_rand))
  for (a in names(nulls)) {
    q <- apply(sapply(nulls[[a]], ev), 1, quantile, 0.95)
    out[[paste0("construction_pa_", a)]] <- n_retain(ev(S), q)
    out[[paste0("fixed_construction_", a)]] <- offset_retain(S, nulls[[a]])
  }
  out
}
joint_planted_sel <- cbind(scenario = c("nothing planted", "3 factors planted"), true_factors = c(0, 3),
                           rbind(select_by(scen_j$none$S, scen_j$none$nulls, eig_rand_14),
                                 select_by(scen_j$planted$S, scen_j$planted$nulls, eig_rand_14)))
knitr::kable(joint_planted_sel, row.names = FALSE, caption = "Number of factors retained (conditional planting)")

recover_to <- function(fit, Tm) {
  m <- match_to(rotate_fit(fit$L)$structure, Tm)
  list(congruence = m$congruence, structure = m$L)
}
Tj <- scen_j$planted$T
joint_fits <- function(sc) list(
  `ordinary ML factor analysis`   = fit_offset(sc$S, NULL, 3, nrow(X)),
  `fixed construction, joint null` = fit_offset(sc$S, null_mean_offdiag(sc$nulls$joint), 3, nrow(X)),
  `fixed construction, flow null`  = fit_offset(sc$S, null_mean_offdiag(sc$nulls$flow), 3, nrow(X)))
jp_rec <- lapply(joint_fits(scen_j$planted), recover_to, Tm = Tj)
jn_rec <- lapply(joint_fits(scen_j$none), recover_to, Tm = Tj)
joint_recovery <- data.frame(
  method = names(jp_rec),
  planted = sapply(jp_rec, function(r) mean(r$congruence)),
  nothing_planted = sapply(jn_rec, function(r) mean(r$congruence)),
  t(sapply(jp_rec, `[[`, "congruence")), row.names = NULL, check.names = FALSE)
joint_recovery$detection_margin <- joint_recovery$planted - joint_recovery$nothing_planted
knitr::kable(joint_recovery, digits = 2, row.names = FALSE,
             caption = "Mean congruence with the planted structure (3 factors), conditional planting")

joint_planted_fit_k <- do.call(rbind, lapply(names(scen_j), function(s) {
  sc <- scen_j[[s]]
  Cj <- null_mean_offdiag(sc$nulls$joint); Cf <- null_mean_offdiag(sc$nulls$flow)
  do.call(rbind, lapply(0:4, function(nf) {
    f <- list(ordinary = fit_offset(sc$S, NULL, nf, nrow(X)),
              fixed_joint = fit_offset(sc$S, Cj, nf, nrow(X)),
              fixed_flow = fit_offset(sc$S, Cf, nf, nrow(X)))
    data.frame(scenario = s, factors = nf,
               t(unlist(lapply(f, function(x) c(rmsea = x$rmsea, srmr = x$srmr)))))
  }))
}))
knitr::kable(joint_planted_fit_k, digits = 3, row.names = FALSE,
             caption = "Fit by number of factors, conditional planting")
```

### 11.3 Real data, rerun

```{r joint-real, results = "asis"}
joint_sets <- list(
  `as built (14 ratios)`   = list(S = R_obs, nulls = list(joint = lapply(joint_runs, `[[`, "F0"))),
  `redesigned (section 9)` = list(S = R_S2,  nulls = list(joint = lapply(joint_runs, `[[`, "S2"))),
  `composition`            = list(S = R_S3,  nulls = list(joint = lapply(joint_runs, `[[`, "S3"))))

joint_real_sel <- do.call(rbind, lapply(names(joint_sets), function(nm) {
  s <- joint_sets[[nm]]
  cbind(set = nm, variables = ncol(s$S), select_by(s$S, s$nulls, rand_eig(ncol(s$S), nrow(X))))
}))
cat(knitr::kable(joint_real_sel, row.names = FALSE, caption = "Number of factors retained, conditional null"),
    sep = "\n")
cat("\n\n")

joint_real_fit_k <- do.call(rbind, lapply(names(joint_sets), function(nm) {
  s  <- joint_sets[[nm]]
  Cj <- null_mean_offdiag(s$nulls$joint)
  do.call(rbind, lapply(0:4, function(nf) {
    f <- list(ordinary = fit_offset(s$S, NULL, nf, nrow(X)), fixed_joint = fit_offset(s$S, Cj, nf, nrow(X)))
    data.frame(set = nm, factors = nf, t(unlist(lapply(f, function(x) c(rmsea = x$rmsea, srmr = x$srmr)))))
  }))
}))
cat(knitr::kable(joint_real_fit_k, digits = 3, row.names = FALSE,
                 caption = "Fit by number of factors, conditional null"), sep = "\n")
cat("\n\n")

joint_real_loadings <- lapply(names(joint_sets), function(nm) {
  s  <- joint_sets[[nm]]
  nf <- max(1, min(4, joint_real_sel$fixed_construction_joint[joint_real_sel$set == nm]))
  P  <- rotate_fit(fit_offset(s$S, null_mean_offdiag(s$nulls$joint), nf, nrow(X))$L)$pattern
  cat(show_loadings(P, paste0(nm, ": fixed construction, conditional null, ", nf, " factor(s)")), sep = "\n")
  cat("\n\n")
  P
})
names(joint_real_loadings) <- names(joint_sets)

gap_pairs <- function(S, C) {
  ij <- which(upper.tri(S), arr.ind = TRUE)
  d  <- data.frame(v1 = colnames(S)[ij[, 1]], v2 = colnames(S)[ij[, 2]], r_obs = S[ij], r_null_joint = C[ij])
  d$beyond <- d$r_obs - d$r_null_joint
  d[order(-abs(d$beyond)), ]
}
joint_gaps <- lapply(names(joint_sets), function(nm) {
  s <- joint_sets[[nm]]
  g <- gap_pairs(s$S, Reduce(`+`, s$nulls$joint) / length(s$nulls$joint))
  cat(knitr::kable(head(g, 6), digits = 2, row.names = FALSE,
                   caption = paste0(nm, ": largest correlations beyond the conditional null")), sep = "\n")
  cat("\n\n")
  g
})
names(joint_gaps) <- names(joint_sets)
```

### 11.4 Findings

```{r joint-values, include = FALSE}
jr <- joint_recovery
jps <- joint_planted_sel
jrs <- joint_real_sel
jsingle <- sapply(joint_real_loadings, function(P) sum(apply(abs(P), 2, max) >= 0.95))
jn_fac  <- sapply(joint_real_loadings, ncol)
jfit4   <- joint_real_fit_k[joint_real_fit_k$factors == 4, ]
jfit0   <- joint_real_fit_k[joint_real_fit_k$factors == 0, ]
pf0     <- joint_planted_fit_k[joint_planted_fit_k$factors == 0, ]
```

* **Absorption.** The conditional null reproduces a mean |r| of
  `r fn(joint_summary$mean_abs_null_r[3], 3)` across the 91 pairs, against
  `r fn(mean(abs(audit$r_obs)), 3)` observed. Its components match the
  observed ones with congruence
  `r paste(fn(unlist(joint_summary[3, names(anchors)])), collapse = ", ")`
  (`r paste(names(anchors), collapse = ", ")`).
* **Planted test.** With the conditional null held fixed, zero factors fit the
  nothing-planted data with SRMR `r fn(pf0$fixed_joint.srmr[pf0$scenario == "none"], 3)`.
  Mean congruence with the planted structure is
  `r fn(jr$planted[2])` when factors are planted and
  `r fn(jr$nothing_planted[2])` when they are not (margin
  `r fn(jr$detection_margin[2])`). Ordinary factor analysis scores
  `r fn(jr$planted[1])` and `r fn(jr$nothing_planted[1])`
  (margin `r fn(jr$detection_margin[1])`); the wrong, global flow null gives a
  margin of `r fn(jr$detection_margin[3])`.
* **Real data fit.** With zero factors, the conditional null leaves SRMR of
  `r paste(sprintf("%s (%s)", fn(jfit0$fixed_joint.srmr, 3), jfit0$set), collapse = ", ")`.
  With four factors:
  `r paste(sprintf("%s (%s)", fn(jfit4$fixed_joint.srmr, 3), jfit4$set), collapse = ", ")`.
* **Single-variable factors.** `r sum(jsingle)` of `r sum(jn_fac)` factors
  load 0.95 or more on one variable.

```{r save, include = FALSE}
if (nzchar(params$save_results)) {
  saveRDS(list(source = source_label, n = nrow(X), reps = reps, audit = audit,
               by_type = by_type, sign_hit = sign_hit, n_signed = nrow(signed), none_ok = none_ok,
               size_r2 = size_r2, eig = list(obs = eig_obs, rand = eig_rand, flow = eig_flow,
                                             stock = eig_stock), retain = retain,
               L = list(obs = L_obs, null_flow = L_null_flow, null_stock = L_null_stock,
                        F1 = L_F1, F2 = L_F2, F3 = L_F3),
               fix_null = fix_null, type_fix = type_fix, min_eig = min_eig,
               retain_fix = retain_fix, R_obs = R_obs, R_null = R_null,
               R_F2 = R_F2, R_F3 = R_F3,
               S2 = list(summary = s2_summary, pairs = s2_pairs, R = R_S2, R_null = R_null_S2,
                         eig = list(obs = ev(R_S2), rand = eig_rand_S2, flow = eig_q95_S2$flow,
                                    stock = eig_q95_S2$stock),
                         k = k_S2, L = L_S2, L_null_flow = L_S2_flow, L_null_stock = L_S2_stock),
               joint = list(cells = length(unique(cells_obs)), audit = audit, by_type = joint_by_type,
                            summary = joint_summary, gaps_F0 = gap_joint, R = R_joint,
                            L_null = L_null_joint, eig = eig_joint,
                            planted_sel = joint_planted_sel, recovery = joint_recovery,
                            planted_fit_k = joint_planted_fit_k, real_sel = joint_real_sel,
                            real_fit_k = joint_real_fit_k, real_loadings = joint_real_loadings,
                            real_gaps = joint_gaps, planted_T = Tj,
                            planted_structure = lapply(jp_rec, `[[`, "structure")),
               offset = list(planted_sel = planted_sel, planted_rec = planted_rec,
                             control_rec = control_rec, recovery_table = recovery_table,
                             planted_fits = planted_fits, planted_T = scen$planted$T,
                             planted_fit_k = planted_fit_k, real_sel = real_sel,
                             real_fit_k = real_fit_k, real_loadings = real_loadings,
                             S3_vars = names(X_S3), R_S3 = R_S3,
                             R_null_S3 = lapply(null_S3, null_mean_offdiag))),
          params$save_results)
}
```
