---
title: "Variance-preserving normalization of fiscal ratios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variance-preserving normalization of fiscal ratios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 2.4,
  dpi        = 96
)
options(width = 110)
```

```{css, echo = FALSE}
/* Callout and key-idea styling from the fiscal-health walkthroughs
   (dev/health-index/walkthrough.css). Scoped to blockquotes and .concepts so
   the pkgdown / html_vignette theme is otherwise untouched. */
blockquote {
  background: #fbf0e7;
  border: 0;
  border-left: 3px solid #bd5416;
  border-radius: 3px;
  padding: 1rem 1.25rem;
  margin: 1.5rem 0;
  font-size: 1rem;
  color: #1f2429;
}
blockquote p:last-child { margin-bottom: 0; }
.concepts {
  background: #153243;
  color: #fff;
  border-radius: 10px;
  padding: 1.5rem 1.75rem 1.25rem;
  margin: 1.5rem 0 2rem;
}
.concepts p, .concepts li { color: rgba(255, 255, 255, .9); }
.concepts strong { color: #dc965a; }
.concepts li::marker { color: #dc965a; }
.concepts code { background: rgba(255, 255, 255, .12); color: #fff; }
```

```{r packages}
library(fiscal)
library(psych)
library(corrplot)
set.seed(2023)
```

## Why the measurement model comes first

A fiscal-health index is only as good as the correlation matrix it is built
from. Two decisions made *before* any factor analysis shape that matrix:

1. **Which organizations have a defined value for each ratio.** Many ratios are
   undefined for large, non-random groups: organizations with no current
   liabilities, no unrestricted net assets, or no functional expense detail.
   How those cases are handled (dropped, set to a fixed value, or imputed)
   decides who is in the correlation matrix and where they sit.
2. **How extreme values are tamed.** Nonprofit ratios have tails running into
   the thousands. Winsorizing, log-type transformations, and rank-based scores
   all change Pearson correlations differently.

Different choices lead to different dimensions. A study of AA/NH/PI nonprofits
(Urban Institute) combined 13 ratios into four dimensions: growth potential,
leverage, solvency, and organizational slack. That study winsorized at the
5th/95th percentiles, logged skewed ratios, and z-scored before PCA.
`vignette("fiscal-health-index")`-style workflows that drop more indicators and
use rank-normal scores find fewer dimensions.

This article works through the measurement model for each ratio:

* the raw ratio and the cases where it is undefined;
* a rule or imputation for each kind of undefined case, with evidence for
  whether those organizations are *better than*, *worse than*, or *typical of*
  the rest;
* whether winsorization should come before or after normalization;
* how four versions of the data (raw with rules applied, winsorized,
  winsorized and normalized, and rank-normal) compare as inputs to a
  correlation-based model.

The goal is transformations that **maximize coverage** (fewest undefined cases)
and **preserve meaningful covariance** (correlations that reflect how the
ratios actually move together, not artifacts of tails, ties, or arbitrary
fills).

<div class="concepts">

**Key ideas**

- **Undefined is not missing at random.** When a ratio can't be computed, ask
  whether those organizations are better off, worse off, or typical, and let
  related ratios (*anchors*) answer.
- **Rule when the direction is known; impute when it isn't.** Negative net
  assets mean *worst* leverage, which calls for a bounded reformulation. No
  current liabilities is an empirical question, which calls for conditional
  imputation.
- **Winsorize before estimating anything.** With percentile bounds and fixed
  monotone transforms the order doesn't matter. With z-score clipping or
  data-estimated parameters it does.
- **Variance-preserving normalization:** winsorize, compress remaining skew
  with asinh, then z-score. It keeps spacing within the body of each
  distribution while taming the tails.
- **Fidelity:** how closely Pearson correlations after transformation track the
  rank correlations of the raw ratios.

</div>

## Data

The bundled `dat10k` sample holds 10,000 e-filed 2021 returns. We keep full
Form 990 filers with positive total expenses: the organizations for which the
Part IX and Part X ratios exist.

One repair is needed first. In the current `dat10k`, several financial columns
hold 64-bit integer (`integer64`) values stored as plain doubles without their
class attribute. Read as numbers, they come out as ~1e-318. `fix64()` detects
that pattern and restores the true values. It does nothing to columns that are
already correct.

```{r data}
fix64 <- function(x) {
  if (inherits(x, "integer64")) return(as.numeric(x))
  if (is.double(x)) {
    nz <- x[!is.na(x) & x != 0]
    if (length(nz) && all(abs(nz) < 1e-300)) {
      class(x) <- "integer64"
      return(as.numeric(x))
    }
  }
  x
}
requireNamespace("bit64", quietly = TRUE)

d <- as.data.frame(dat10k)
d <- d[d$RETURN_TYPE == "990", ]

# Financial fields: repaired, numeric, and blank = 0 (full-990 filers
# leave lines blank when the amount is zero).
fin <- function(f) {
  x <- if (f %in% names(d)) suppressWarnings(as.numeric(fix64(d[[f]]))) else rep(0, nrow(d))
  x[is.na(x)] <- 0
  x
}

k <- data.frame(
  cash    = fin("F9_10_ASSET_CASH_EOY") + fin("F9_10_ASSET_SAVING_EOY"),
  recv    = fin("F9_10_ASSET_PLEDGE_NET_EOY") + fin("F9_10_ASSET_ACC_NET_EOY"),
  othca   = fin("F9_10_ASSET_INV_SALE_EOY") + fin("F9_10_ASSET_EXP_PREPAID_EOY"),
  ap      = fin("F9_10_LIAB_ACC_PAYABLE_EOY"),
  cl      = fin("F9_10_LIAB_ACC_PAYABLE_EOY") + fin("F9_10_LIAB_GRANT_PAYABLE_EOY"),
  A       = fin("F9_10_ASSET_TOT_EOY"),
  L       = fin("F9_10_LIAB_TOT_EOY"),
  NA_eoy  = fin("F9_10_NAFB_TOT_EOY"),
  NA_boy  = fin("F9_10_NAFB_TOT_BOY"),
  UNA     = fin("F9_10_NAFB_UNRESTRICT_EOY"),
  RES     = fin("F9_10_NAFB_RESTRICT_EOY"),
  land    = fin("F9_10_ASSET_LAND_BLDG_NET_EOY"),
  mtg     = fin("F9_10_LIAB_MTG_NOTE_EOY"),
  R       = fin("F9_08_REV_TOT_TOT"),
  E       = fin("F9_09_EXP_TOT_TOT"),
  dep     = fin("F9_09_EXP_DEPREC_TOT"),
  progrev = fin("F9_08_REV_PROG_TOT_TOT"),
  P       = fin("F9_09_EXP_TOT_PROG"),
  M       = fin("F9_09_EXP_TOT_MGMT"),
  Fu      = fin("F9_09_EXP_TOT_FUNDR"),
  c3      = d$F9_00_EXEMPT_STAT_501C3_X %in% "X"
)
n_990 <- nrow(k)
k <- k[k$E > 0, ]

# Derived components
k$ca   <- k$cash + k$recv + k$othca                    # current assets
k$una  <- ifelse(k$UNA == 0 & k$RES == 0, k$NA_eoy, k$UNA)  # unrestricted NA (non-SFAS 117 filers)
k$fsum <- k$P + k$M + k$Fu                             # functional expense columns
k$opE  <- ifelse(k$E - k$dep > 0, k$E - k$dep, k$E)    # cash operating expenses

c(full_990 = n_990, positive_expenses = nrow(k))
```

## 1. The ratios and their undefined cases

The table defines each ratio in its **textbook** form (numerator ÷ denominator)
and lists what can go wrong. The ratios cover the domains in the AA/NH/PI study
(liquidity, leverage and solvency, performance, expense slack, asset structure)
so the dimension question can be revisited on a like-for-like basis.

| Ratio | Textbook formula | Undefined or misleading when | Proposed rule |
|---|---|---|---|
| `current` | current assets ÷ (payables + grants payable) | current liabilities = 0 (~40%) | Impute (Section 2) |
| `months_cash` | (cash + savings + receivables) ÷ monthly cash expenses | never (expenses > 0) | none |
| `liquid_months` | (unrestricted NA − land/buildings + mortgages) ÷ monthly expenses | unrestricted NA blank for non-SFAS 117 filers | Use total NA when both NA classes are blank |
| `cash_assets` | (cash + savings) ÷ total assets | assets = 0 | Impute |
| `debt_assets` | liabilities ÷ assets | assets = 0 | 0 if no liabilities; sample maximum if liabilities > 0 |
| `debt_una` | liabilities ÷ unrestricted NA | unrestricted NA ≤ 0 (sign flips) | Bounded share: L ÷ (L + max(UNA, 0)) |
| `payables_una` | payables ÷ unrestricted NA | unrestricted NA ≤ 0 | Bounded share: AP ÷ (AP + max(UNA, 0)) |
| `surplus_margin` | (revenue − expenses) ÷ revenue | revenue ≤ 0; tiny revenue gives huge negatives | Symmetric margin: (R − E) ÷ max(R, E) |
| `return_assets` | (revenue − expenses) ÷ assets | assets = 0 | Impute |
| `na_growth` | (NA_eoy − NA_boy) ÷ NA_boy | NA_boy ≤ 0 (sign flips) | Symmetric growth: Δ ÷ mean(abs(NA_eoy), abs(NA_boy)) |
| `self_suff` | program revenue ÷ expenses | never | none |
| `prog_exp`, `admin_exp`, `fundr_exp` | functional column ÷ total expenses | functional columns not reported | Share of the functional sum; impute when unreported |
| `land_assets` | net land/buildings ÷ assets | assets = 0 | 0 (no property) |
| `assets_rev` | assets ÷ revenue | revenue ≤ 0 | Use expenses as the scale when revenue ≤ 0 |

```{r ratios}
dv <- function(n, dd) { x <- ifelse(dd == 0, NA_real_, n / dd); x[!is.finite(x)] <- NA; x }

naive <- with(k, data.frame(
  current = dv(ca, cl), months_cash = dv(cash + recv, opE / 12),
  liquid_months = dv(una - land + mtg, E / 12), cash_assets = dv(cash, A),
  debt_assets = dv(L, A), debt_una = dv(L, una), payables_una = dv(ap, una),
  surplus_margin = dv(R - E, R), return_assets = dv(R - E, A),
  na_growth = dv(NA_eoy - NA_boy, NA_boy), self_suff = dv(progrev, E),
  prog_exp = dv(P, E), admin_exp = dv(M, E), fundr_exp = dv(Fu, E),
  land_assets = dv(land, A), assets_rev = dv(A, R)))

rules <- with(k, data.frame(
  current        = ifelse(cl > 0, ca / cl, NA),
  months_cash    = (cash + recv) / (opE / 12),
  liquid_months  = (una - land + mtg) / (E / 12),
  cash_assets    = ifelse(A > 0, pmin(cash / A, 1), NA),
  debt_assets    = ifelse(A > 0, L / A, ifelse(L > 0, NA, 0)),
  debt_una       = ifelse(L > 0, L / (L + pmax(una, 0)), 0),
  payables_una   = ifelse(ap > 0, ap / (ap + pmax(una, 0)), 0),
  surplus_margin = (R - E) / pmax(R, E),
  return_assets  = ifelse(A > 0, (R - E) / A, NA),
  na_growth      = ifelse(NA_eoy == 0 & NA_boy == 0, 0,
                          (NA_eoy - NA_boy) / ((abs(NA_eoy) + abs(NA_boy)) / 2)),
  self_suff      = progrev / E,
  prog_exp       = ifelse(fsum > 0, P / fsum, NA),
  admin_exp      = ifelse(fsum > 0, M / fsum, NA),
  fundr_exp      = ifelse(fsum > 0, Fu / fsum, NA),
  land_assets    = ifelse(A > 0, land / A, 0),
  assets_rev     = A / ifelse(R > 0, R, E)))
rules$debt_assets[k$A <= 0 & k$L > 0] <- max(rules$debt_assets, na.rm = TRUE)

coverage <- data.frame(
  textbook = colMeans(!is.na(naive)),
  rules    = colMeans(!is.na(rules)))
round(coverage, 3)
c(complete_rows_textbook = mean(complete.cases(naive)),
  complete_rows_rules    = mean(complete.cases(rules)))
```

Rules alone don't raise coverage much, because the largest gaps are cases where
no *rule* can recover a value: the current ratio with no current liabilities,
and unreported functional expenses. Those need imputation (Section 2).

The textbook column is also misleading on its own. It counts a debt-to-equity
ratio computed on *negative* net assets as defined, even though that value
ranks an insolvent organization as the least leveraged.

## 2. Undefined cases: better, worse, or typical?

When a ratio is undefined, the key question is whether those organizations are
doing **better than average**, **worse than average**, or are **typical** on
the concept the ratio measures. The data can answer it. Take the organizations
where the ratio is undefined and see where they rank on **anchor ratios**:
related measures that are defined for them and do not share the missing
component. If they sit near the 50th percentile on the anchors, they are
typical. Well above or below 50 means they are better or worse.

Percentile ranks are computed over all organizations:

```{r helpers}
pct_rank <- function(x) { r <- rank(x, na.last = "keep"); (r - 0.5) / sum(!is.na(x)) }
rank_normal <- function(x) qnorm(pct_rank(x))
```

### 2.1 The current ratio when there are no current liabilities

```{r current-groups}
grp <- with(k, ifelse(cl > 0, "defined (CL > 0)",
               ifelse(ca > 0, "CL = 0, CA > 0", "CL = 0, CA = 0")))
round(prop.table(table(grp)), 3)

anchors <- data.frame(
  liquid_months  = pct_rank(rules$liquid_months),
  surplus_margin = pct_rank(rules$surplus_margin),
  low_debt       = 1 - pct_rank(rules$debt_assets),
  cash_assets    = pct_rank(rules$cash_assets),
  size_expenses  = pct_rank(k$E))
anchor_profile <- aggregate(anchors, list(group = grp),
                            function(v) round(median(v, na.rm = TRUE), 2))
anchor_profile
```

```{r current-distribution}
cur_def <- rules$current[grp == "defined (CL > 0)"]
round(quantile(cur_def, c(.10, .25, .50, .75, .90, .99)), 2)
```

What the anchors say:

* **CL = 0, CA > 0** (about `r round(100 * mean(grp == "CL = 0, CA > 0"))`% of
  filers). These organizations rank above the median on liquid reserves and
  cash share, and far above it on low debt. They are also *smaller* than
  typical. Their liquidity is better than average, but not extreme.
* **CL = 0, CA = 0** (about `r round(100 * mean(grp == "CL = 0, CA = 0"), 1)`%).
  They have almost no liquid reserves (near the bottom on liquid months and
  cash share). But they owe nothing, and their operating results are only
  somewhat below average. "No buffer, no obligations" is weak liquidity, not
  the worst liquidity.

**Why a cap of 2 or 3 is the wrong ceiling.** The 2–3 rule of thumb comes from
business balance sheets. On the 990, "current liabilities" here are only
payables and grants payable. Deferred revenue, accrued liabilities, and notes
aren't separately identified. The observed nonprofit current ratio is
therefore much larger: the median defined value is
`r round(median(cur_def), 1)`. A cap of 3 would place organizations with no
current liabilities at roughly the
`r round(100 * mean(cur_def <= 3))`th percentile, *below* typical. The anchors
say they belong *above* typical. Any cap should be set on the observed
distribution, not borrowed from corporate finance.

We compare five treatments:

* **drop:** leave undefined (NA).
* **median:** the median of defined values.
* **cap:** 3 when CA > 0 and 0 when CA = 0 (the rule-of-thumb proposal).
* **anchor:** a *percentile-matched* value. Each group gets the quantile of the
  defined distribution that matches the group's mean anchor percentile.
* **conditional:** stochastic *row-plus-column* imputation. Regress the
  rank-normal current ratio on rank-normal anchor ratios among defined cases.
  Predict each undefined organization from *its own* anchors (the row). Add a
  random residual so the imputed values keep the observed variance. Then map
  the normal score back through the defined distribution's quantile function
  (the column).

```{r current-options}
cond_impute <- function(target, predictors, miss) {
  y  <- rank_normal(target)
  X  <- as.data.frame(lapply(predictors, rank_normal))
  fit <- lm(y ~ ., data = cbind(y = y, X)[!miss, ])
  z  <- predict(fit, newdata = X[miss, , drop = FALSE]) +
        rnorm(sum(miss), 0, summary(fit)$sigma)
  out <- target
  out[miss] <- quantile(target[!miss], pnorm(z), names = FALSE, na.rm = TRUE)
  out
}
fill_median <- function(x) { x[is.na(x)] <- median(x, na.rm = TRUE); x }

miss  <- grp != "defined (CL > 0)"
preds <- as.data.frame(lapply(
  rules[, c("liquid_months", "surplus_margin", "debt_assets", "months_cash")], fill_median))
q_anchor <- tapply(rowMeans(anchors[, 1:3], na.rm = TRUE), grp, mean)

current_options <- list(
  drop        = rules$current,
  median      = ifelse(miss, median(cur_def), rules$current),
  cap         = ifelse(grp == "CL = 0, CA > 0", 3,
                       ifelse(grp == "CL = 0, CA = 0", 0, rules$current)),
  anchor      = ifelse(miss, quantile(cur_def, q_anchor[grp], names = FALSE), rules$current),
  conditional = cond_impute(rules$current, preds, miss))

compare_current <- t(sapply(current_options, function(x) c(
  coverage          = mean(!is.na(x)),
  r_months_cash     = cor(rank_normal(x), rank_normal(rules$months_cash), use = "pair"),
  r_liquid_months   = cor(rank_normal(x), rank_normal(rules$liquid_months), use = "pair"),
  r_debt_assets     = cor(rank_normal(x), rank_normal(rules$debt_assets), use = "pair"),
  pctile_CL0_CApos  = mean(pct_rank(x)[grp == "CL = 0, CA > 0"]),
  pctile_CL0_CA0    = mean(pct_rank(x)[grp == "CL = 0, CA = 0"]))))
round(compare_current, 3)
```

How to read the table. The `r_*` columns show how the current ratio correlates
with related liquidity and leverage measures under each treatment. The
`drop` row is the benchmark on defined cases only. The `pctile_*` columns show
where each treatment places the undefined groups; compare them with the anchor
evidence above.

* **median** fills 100% of rows but pulls every undefined case to the middle.
  That weakens the current ratio's correlations with related measures by
  roughly a third. It treats these organizations as typical, and the anchors
  say they aren't.
* **cap** places organizations with CA > 0 *below* typical and puts the
  CA = 0 group at the bottom of the distribution. It all but erases the ratio's
  relationship with leverage.
* **anchor** places groups where the anchors say they belong. But it gives
  every member the same value, and those ties dilute correlations.
* **conditional** keeps the correlations at or somewhat above the
  defined-case benchmark. It places both groups consistently with the anchor
  evidence, without being told to. It is the variance-preserving choice. The
  small increase for ratios that were also used as predictors (liquid months,
  debt to assets) is the mild inflation built into conditional imputation. The
  random residual keeps it small.

> **Recommendation: current ratio.** Use conditional (row-plus-column)
> imputation for organizations with no current liabilities. If a transparent
> fixed rule is required, use the *percentile-matched* anchor rule. Don't use a
> fixed cap from corporate benchmarks. On this data, "no current liabilities
> and some current assets" means *better than typical*, "no current assets and
> no current liabilities" means *below typical but not the floor*, and neither
> means *typical*.
>
> *Caveat:* conditional imputation assumes that, given its anchor ratios, an
> undefined organization resembles defined organizations with the same anchors.
> The residual noise keeps imputed values from inflating correlations with the
> anchors. Re-check the anchor profile whenever the anchors change.

### 2.2 Negative denominators: leverage on unrestricted net assets

About `r round(100 * mean(k$una < 0), 1)`% of filers have negative unrestricted
net assets, and `r round(100 * mean(k$una == 0), 1)`% have none. For them,
liabilities ÷ unrestricted NA is negative or infinite. A negative value would
rank an insolvent organization as the *least* leveraged.

Unlike the current ratio, the direction here isn't in doubt. An organization
whose liabilities exceed its unrestricted equity is at the *worst* end of
leverage. Imputation would wrongly pull it toward typical, so the fix should
be a rule. We compare four rules:

* **drop:** NA when UNA ≤ 0.
* **floor $1:** bottom-code UNA at $1.
* **floor 1% of assets:** bottom-code UNA at 1% of total assets.
* **bounded share:** debt ÷ (debt + max(UNA, 0)). It lies in [0, 1] and equals
  1 when there is no unrestricted equity.

```{r leverage}
insolvent <- k$una <= 0 & k$L > 0
leverage_options <- with(k, list(
  drop        = ifelse(una > 0, L / una, NA),
  floor_1     = L / pmax(una, 1),
  floor_1pct  = L / pmax(una, 0.01 * pmax(A, 1)),
  share       = rules$debt_una))

round(t(sapply(leverage_options, function(x) c(
  coverage               = mean(!is.na(x)),
  spearman_debt_assets   = cor(x, rules$debt_assets, method = "spearman", use = "pair"),
  spearman_liquid_months = cor(x, rules$liquid_months, method = "spearman", use = "pair"),
  spearman_size_insolvent = if (all(is.na(x[insolvent]))) NA else
    cor(x[insolvent], k$L[insolvent], method = "spearman")))), 3)
```

* **floor $1** turns the ratio into a *dollar amount* for insolvent
  organizations. Among them, the ratio ranks organizations exactly by
  liabilities (Spearman ρ with dollars of debt = 1), which makes it a size
  measure.
* **floor 1% of assets** fixes the scale problem, but the 1% is arbitrary.
* **bounded share** needs no arbitrary constant, is defined for everyone, reads
  as a proportion ("share of capital that is borrowed"), and ranks insolvent
  organizations at the worst end.

> **Recommendation: leverage ratios with net-asset denominators.** Re-express
> them as the bounded debt share of capital. It is a monotone transformation of
> the textbook ratio wherever the textbook ratio is valid: D ÷ (D + E) =
> (D/E) ÷ (1 + D/E). So no information is lost, and the rule decides only how
> the invalid cases are placed.
>
> **Implication for dimensions:** once the negative denominators are handled,
> leverage on *unrestricted* equity ranks organizations almost identically to
> debt-to-assets (ρ ≈ `r round(cor(rules$debt_una, rules$debt_assets, method = "spearman"), 2)`).
> A separate "leverage" dimension distinct from "solvency" therefore can't come
> from this ratio. If one appears, look for sign-flip artifacts or ratios that
> share a denominator.

### 2.3 Growth rates with a non-positive base

Net-asset growth divides by beginning net assets. That base is zero or negative
for `r round(100 * mean(k$NA_boy <= 0), 1)`% of filers. The **symmetric
(arc) growth rate**, change ÷ mean(abs(start), abs(end)), is defined whenever
either value is nonzero. It is bounded in [−2, 2] and orders ordinary cases
the same way the textbook rate does.

```{r growth}
growth_options <- with(k, list(
  textbook      = naive$na_growth,
  drop_nonpos   = ifelse(NA_boy > 0, (NA_eoy - NA_boy) / NA_boy, NA),
  symmetric     = rules$na_growth))
round(t(sapply(growth_options, function(x) c(
  coverage = mean(!is.na(x)), sd = sd(x, na.rm = TRUE),
  p01 = unname(quantile(x, .01, na.rm = TRUE)), p99 = unname(quantile(x, .99, na.rm = TRUE)),
  spearman_surplus = cor(x, rules$surplus_margin, method = "spearman", use = "pair")))), 3)
```

The symmetric rate covers everyone. It keeps the same association with
operating surplus as the textbook rate computed on positive bases only. And its
standard deviation is on a sensible scale instead of being dominated by
near-zero bases. The same logic motivates the **symmetric surplus margin**,
(R − E) ÷ max(R, E). It is defined for every organization with expenses, and
it keeps the same ordering as the textbook margin where revenue is positive:

```{r surplus-check}
ok <- k$R > 0
c(spearman_symmetric_vs_textbook = cor(rules$surplus_margin[ok], naive$surplus_margin[ok],
                                       method = "spearman"),
  textbook_p01 = unname(quantile(naive$surplus_margin, .01, na.rm = TRUE)),
  symmetric_p01 = unname(quantile(rules$surplus_margin, .01)))
```

### 2.4 Functional expenses that were never reported

Part IX columns B–D (program, management, fundraising) are required of
501(c)(3) and 501(c)(4) organizations but optional for other filers. A zero in
every functional column means **not reported**, not "spent nothing on
programs".

```{r functional}
unreported <- k$fsum == 0
round(tapply(unreported, ifelse(k$c3, "501(c)(3)", "other 501(c)"), mean), 3)
```

The textbook program-expense ratio gives these organizations a value of 0: the
worst possible program ratio and the best possible overhead ratio. The rule
sets them to missing, and they are imputed conditionally below. The shares are
also computed over the *functional sum*, so they add to exactly 1. For the same
reason, **only two of the three shares can enter a correlation-based model**:
program share = 1 − admin share − fundraising share. Entering all three builds
a closure factor into the matrix, a "slack" component that exists by
arithmetic.

### 2.5 The analysis frame

Every remaining undefined value is filled by conditional imputation from ratios
that are defined for everyone. The current ratio uses the imputation from
Section 2.1.

```{r analysis-frame}
frame <- rules
frame$current <- current_options$conditional
imputed_share <- colMeans(is.na(rules))
base_preds <- c("liquid_months", "surplus_margin", "months_cash", "self_suff")
for (v in names(frame)) if (anyNA(frame[[v]])) {
  frame[[v]] <- cond_impute(frame[[v]], frame[, setdiff(base_preds, v), drop = FALSE],
                            is.na(frame[[v]]))
}
stopifnot(!anyNA(frame))
round(imputed_share[imputed_share > 0], 3)

# Closure: program share is implied by admin and fundraising shares.
model_vars <- setdiff(names(frame), "prog_exp")
```

## 3. Transformations

Every correlation below is computed on the analysis frame from Section 2.5.
Rules and imputations are applied first, so the versions differ only in how
tails and shape are handled.

| Version | Steps | What it keeps | What it gives up |
|---|---|---|---|
| **Raw** | rules applied, nothing else | original units | tails dominate Pearson r |
| **Winsorized** | clip at the 1st/99th percentiles | units within bounds | skewed bodies remain |
| **Winsorized + normalized** | winsorize → asinh(x / s) if abs(skew) > 1 → z-score | interval spacing within the body; reusable parameters | some residual skew |
| **Package `_z`** | `apply_transformations()` (auto-selected asinh / logit / hurdle / rank-normal, median/MAD) | the package's current behavior | see Section 5 |
| **Rank-normal** | rank → normal quantile | ordering only | all spacing information |

The **winsorized + normalized** version is the variance-preserving candidate.
Winsorizing uses percentile bounds. The asinh transform (which behaves like a
log in the tails but is defined at zero and for negatives) is applied only to
ratios that remain skewed after winsorization. Left-skewed ratios, such as a
program share piled up near 1, are reflected first so the long tail is the one
compressed. The asinh scale `s` is the *median* absolute nonzero value, a
robust statistic that outliers can't move. The final z-score is linear and so
changes no correlations.

```{r transforms}
winsor <- function(x, p = 0.01) {
  q <- quantile(x, c(p, 1 - p), na.rm = TRUE)
  pmin(pmax(x, q[[1]]), q[[2]])
}
winsor_normalize <- function(x, p = 0.01) {
  w <- winsor(x, p)
  sk <- psych::skew(w)
  if (sk > 1) {                       # long right tail
    s <- median(abs(w[w != 0]))
    w <- asinh(w / s)
  } else if (sk < -1) {               # long left tail: reflect, compress, flip back
    r <- max(w) - w
    s <- median(r[r != 0])
    w <- -asinh(r / s)
  }
  as.numeric(scale(w))
}
ranges <- c(current = "zp", months_cash = "zp", liquid_months = "np", cash_assets = "zo",
            debt_assets = "zp", debt_una = "zo", payables_una = "zo", surplus_margin = "np",
            return_assets = "np", na_growth = "np", self_suff = "zp", prog_exp = "zo",
            admin_exp = "zo", fundr_exp = "zo", land_assets = "zo", assets_rev = "zp")
package_z <- function(x, v) suppressWarnings(apply_transformations(x, 0.98, range = ranges[[v]])$z)

versions <- list(
  raw                    = frame,
  winsorized             = as.data.frame(lapply(frame, winsor)),
  winsorized_normalized  = as.data.frame(lapply(frame, winsor_normalize)),
  package_z              = as.data.frame(setNames(lapply(names(frame), function(v)
                              package_z(frame[[v]], v)), names(frame))),
  rank_normal            = as.data.frame(lapply(frame, rank_normal)))

asinh_applied <- sapply(frame, function(x) abs(psych::skew(winsor(x))) > 1)
```

### 3.1 Each ratio: raw distribution and transformed versions

Each row of histograms shows one ratio: raw (rules applied), winsorized,
winsorized + normalized, and rank-normal. The raw panel shows the whole range,
including extreme values. The table under each group reports textbook and
rule-based coverage, the share imputed, raw quantiles, and skewness at each
step.

```{r ratio-card-fn}
ratio_card <- function(v) {
  op <- par(mfrow = c(1, 4), mar = c(2.5, 1, 2, 1), cex.main = 0.85)
  on.exit(par(op))
  panels <- list(raw = frame[[v]], winsorized = versions$winsorized[[v]],
                 `wins + norm` = versions$winsorized_normalized[[v]],
                 `rank-normal` = versions$rank_normal[[v]])
  for (nm in names(panels)) {
    hist(panels[[nm]], breaks = 50, col = "#6d7681", border = "white",
         main = paste0(v, ": ", nm), yaxt = "n", xlab = "", ylab = "")
  }
}
card_table <- function(vars) {
  q <- t(sapply(vars, function(v) quantile(frame[[v]], c(.01, .25, .5, .75, .99))))
  colnames(q) <- c("p01", "p25", "p50", "p75", "p99")
  data.frame(
    textbook_cov = round(coverage[vars, "textbook"], 3),
    rules_cov    = round(coverage[vars, "rules"], 3),
    imputed      = round(imputed_share[vars], 3),
    signif(q, 3),
    skew_raw     = round(sapply(vars, function(v) psych::skew(frame[[v]])), 1),
    skew_wins    = round(sapply(vars, function(v) psych::skew(versions$winsorized[[v]])), 2),
    asinh        = asinh_applied[vars],
    skew_norm    = round(sapply(vars, function(v) psych::skew(versions$winsorized_normalized[[v]])), 2))
}
```

#### Liquidity

```{r cards-liquidity, fig.height = 2.2}
for (v in c("current", "months_cash", "liquid_months", "cash_assets")) ratio_card(v)
card_table(c("current", "months_cash", "liquid_months", "cash_assets"))
```

`current` and `months_cash` are non-negative with extreme right tails. After
winsorizing they are still heavily skewed, so they get the asinh step.
`liquid_months` can be negative (property and debt exceed unrestricted equity).
asinh handles both signs, so negative reserves stay meaningfully below zero
instead of being truncated. `cash_assets` is a bounded share and needs little
more than clipping.

#### Leverage

```{r cards-leverage, fig.height = 2.2}
for (v in c("debt_assets", "debt_una", "payables_una")) ratio_card(v)
card_table(c("debt_assets", "debt_una", "payables_una"))
```

The bounded shares pile up at 0 (no debt) and, for `debt_una`, at 1 (no
unrestricted equity). Those spikes are real, qualitatively distinct states.
Rank-normal breaks the ties by averaging and spreads the spike across the
middle of the normal scale. The winsorized + normalized version keeps the
spike as a spike.

#### Performance

```{r cards-performance, fig.height = 2.2}
for (v in c("surplus_margin", "return_assets", "na_growth")) ratio_card(v)
card_table(c("surplus_margin", "return_assets", "na_growth"))
```

Because the margin and growth rate are symmetric, they are bounded, so
winsorizing is nearly a no-op for them. The definition does the work that
winsorizing does for the textbook versions.

#### Expense structure

```{r cards-expense, fig.height = 2.2}
for (v in c("prog_exp", "admin_exp", "fundr_exp")) ratio_card(v)
card_table(c("prog_exp", "admin_exp", "fundr_exp"))
```

`fundr_exp` is zero-inflated: most organizations report no fundraising
expense. Section 5 shows what that does to the package's hurdle
transformation.

#### Asset structure and revenue model

```{r cards-structure, fig.height = 2.2}
for (v in c("land_assets", "assets_rev", "self_suff")) ratio_card(v)
card_table(c("land_assets", "assets_rev", "self_suff"))
```

## 4. Does it matter whether winsorization comes before normalization?

Sometimes. It depends on whether the normalization *estimates anything from
the data*. We use the current ratio and months of cash to show each case.

```{r order-setup}
x <- frame$current
y <- frame$months_cash
zscore <- function(v) (v - mean(v)) / sd(v)
s_fixed <- median(abs(x[x != 0]))
```

**(a) A fixed monotone transform with percentile winsorizing: order does not
matter.** Clipping at the 1st and 99th percentiles and then applying asinh
gives the same values as applying asinh and then clipping at the 1st and 99th
percentiles, because a monotone function maps percentiles to percentiles.

```{r order-a}
max(abs(asinh(winsor(x) / s_fixed) - winsor(asinh(x / s_fixed))))
```

**(b) z-scoring first, then clipping at a fixed z threshold: order matters a
lot.** Outliers inflate the standard deviation, so almost nothing reaches
|z| > 3. The body of the distribution gets squeezed into a narrow band, and
correlations with other ratios are attenuated.

```{r order-b}
z_then_clip <- function(v) pmin(pmax(zscore(v), -3), 3)
c(share_beyond_3sd           = mean(abs(zscore(x)) > 3),
  sd_after_z_then_clip       = sd(z_then_clip(x)),
  r_winsorize_then_z         = cor(zscore(winsor(x)), zscore(winsor(y))),
  r_z_then_clip              = cor(z_then_clip(x), z_then_clip(y)))
```

**(c) Estimated transform parameters: order matters in proportion to how
sensitive the estimator is.** If the asinh scale is estimated with a moment
statistic (the standard deviation) instead of the median, estimating it before
winsorizing lets the outliers pick the scale:

```{r order-c}
s_before <- sd(x)
s_after  <- sd(winsor(x))
c(scale_before_winsorizing = s_before, scale_after_winsorizing = s_after,
  r_between_resulting_scores = cor(asinh(winsor(x) / s_before), asinh(winsor(x) / s_after)))
```

> **Rule of thumb.** Winsorize first whenever anything downstream is estimated
> from the data: a standard deviation, a transformation parameter (a log
> offset, Box–Cox λ, a logit rescaling range), or a centering constant. Use
> percentile bounds, not z-score bounds. With fixed monotone transforms and
> percentile winsorizing, the order is irrelevant. `find_best_normalization()`
> already winsorizes before fitting.

## 5. Correlation matrices

All four versions (plus the package `_z` columns) use the same variables in the
same order. The order and the outlined blocks come from hierarchical
clustering of the rank-normal matrix, so a block that "dissolves" in another
panel shows how much that transformation changes the structure. `prog_exp` is
excluded because of the closure described in Section 2.4.

```{r corr-matrices}
cmats <- lapply(versions, function(df) cor(df[, model_vars], use = "pairwise.complete.obs"))

hc      <- hclust(as.dist(1 - abs(cmats$rank_normal)), method = "complete")
ord     <- hc$order
n_block <- 5
blocks  <- cutree(hc, k = n_block)[ord]
rect_index <- c(which(!duplicated(blocks)), length(blocks))

plot_corr <- function(C, title) {
  C <- C[ord, ord]
  res <- corrplot(C, order = "original", tl.col = "black", tl.cex = 0.75,
                  na.label = "NA", title = title, mar = c(0, 0, 1.5, 0))
  corrRect(res, index = rect_index, col = "#bd5416", lwd = 2)
}
```

The blocks, listed in plot order:

```{r blocks}
split(names(blocks), factor(blocks, levels = unique(blocks)))
```

```{r corr-raw, fig.width = 6.5, fig.height = 6.5}
plot_corr(cmats$raw, "Raw ratios (rules applied)")
```

```{r corr-winsorized, fig.width = 6.5, fig.height = 6.5}
plot_corr(cmats$winsorized, "Winsorized (1st/99th percentile)")
```

```{r corr-normalized, fig.width = 6.5, fig.height = 6.5}
plot_corr(cmats$winsorized_normalized, "Winsorized + normalized (asinh when skewed, z-score)")
```

```{r corr-package, fig.width = 6.5, fig.height = 6.5}
plot_corr(cmats$package_z, "Package _z (apply_transformations)")
```

```{r corr-ranknormal, fig.width = 6.5, fig.height = 6.5}
plot_corr(cmats$rank_normal, "Rank-normal")
```

### 5.1 Scorecard

Each version is scored on:

* **shape:** mean absolute skewness and excess kurtosis across ratios;
* **fidelity:** the mean absolute difference between the version's Pearson
  correlations and the Spearman (rank) correlations of the raw ratios. Spearman
  is the benchmark because it reflects the monotone association in the data,
  whatever the tails look like;
* **linear fidelity:** the mean correlation of each transformed ratio with its
  winsorized original. It measures how much interval spacing survives;
* **signal:** mean |r| off the diagonal, and the number of pairs with |r| ≥ 0.3;
* **internal consistency:** mean |r| within the hierarchical-clustering blocks
  shown above versus between them, and overall KMO;
* **structure:** the number of components retained by parallel analysis.

```{r scorecard}
S <- cor(frame[, model_vars], method = "spearman")
off <- upper.tri(S)
same_block <- outer(cutree(hc, k = n_block), cutree(hc, k = n_block), "==")[model_vars, model_vars]

score_version <- function(df, name) {
  df <- df[, model_vars]
  C  <- cor(df, use = "pairwise.complete.obs")
  all_na <- colSums(!is.na(df)) == 0
  Cz <- C; Cz[is.na(Cz)] <- 0; diag(Cz) <- 1
  pa <- suppressMessages(suppressWarnings(
    psych::fa.parallel(Cz, n.obs = nrow(df), fa = "pc", plot = FALSE, n.iter = 20)))
  c(all_na_columns        = sum(all_na),
    mean_abs_skew         = mean(abs(sapply(df[!all_na], psych::skew, na.rm = TRUE))),
    mean_excess_kurtosis  = mean(sapply(df[!all_na], psych::kurtosi, na.rm = TRUE)),
    distortion_vs_rank    = mean(abs(C - S)[off], na.rm = TRUE),
    linear_fidelity       = mean(sapply(model_vars[!all_na], function(v)
                              cor(df[[v]], versions$winsorized[[v]], use = "pair"))),
    mean_abs_r            = mean(abs(C)[off], na.rm = TRUE),
    pairs_abs_r_ge_0.3    = sum(abs(C)[off] >= 0.3, na.rm = TRUE),
    within_block_abs_r    = mean(abs(C)[off & same_block], na.rm = TRUE),
    between_block_abs_r   = mean(abs(C)[off & !same_block], na.rm = TRUE),
    KMO                   = psych::KMO(Cz)$MSA,
    parallel_components   = pa$ncomp)
}
invisible(capture.output(
  scorecard <- t(mapply(score_version, versions, names(versions)))))

# Sensitivity: winsorizing at the 5th/95th percentiles, as in the AA/NH/PI study
invisible(capture.output(sens <- rbind(
  winsorized_5_95            = score_version(as.data.frame(lapply(frame, winsor, p = 0.05)), "w5"),
  winsorized_normalized_5_95 = score_version(as.data.frame(lapply(frame, winsor_normalize, p = 0.05)), "wn5"))))
round(rbind(scorecard, sens), 3)
```

Ratios for which the package `_z` column is entirely missing:

```{r package-na}
names(versions$package_z)[colSums(!is.na(versions$package_z)) == 0]
```

How to read the scorecard:

* **Raw → winsorized** is the largest single improvement. Clipping 2% of values
  cuts distortion by more than a third and more than doubles the typical
  correlation.
  Tails were hiding most of the covariance.
* **Winsorized → winsorized + normalized** cuts distortion roughly in half
  again. It brings skewness and kurtosis near normal, and it raises both
  within-block correlation and the number of meaningful pairs. Linear fidelity
  stays high: the transformed values are still close to linear in the
  winsorized originals.
* **Rank-normal** has the least distortion and the strongest blocks, but the
  lowest linear fidelity. Its gains come from discarding spacing.
* **Package `_z`** is competitive on the correlation metrics for the ratios it
  scores. But it returns an all-`NA` column for a zero-inflated share, which
  silently removes that ratio from any model, and its shape metrics exclude
  that column. Its median/MAD scaling also leaves variances unequal across
  ratios.
* **The 5th/95th percentile rows** show the winsorizing trade-off. Clipping 10%
  of values instead of 2% moves every version closer to the rank benchmark and
  lowers linear fidelity. The AA/NH/PI study's 5/95 choice is defensible for
  correlation structure, but it treats the top and bottom 5% of organizations
  as identical to the 5th and 95th percentile values.

## 6. Recommendations

**Undefined cases: decide direction first, then choose rule or imputation.**

| Situation | Example | Direction known? | Treatment |
|---|---|---|---|
| Denominator ≤ 0 and the ratio's meaning implies the worst (or best) state | leverage on negative unrestricted NA; growth from a negative base | yes | **Rule:** bounded or symmetric reformulation that is monotone with the textbook ratio where it is valid |
| Denominator = 0 because the component doesn't exist | current ratio with no current liabilities | **ask the data:** anchor profile | **Conditional (row-plus-column) stochastic imputation**, or percentile-matched rule if a fixed value is required |
| Component not reported | functional expenses of non-501(c)(3) filers | no: missing, not zero | Conditional imputation; never zero |
| Both numerator and denominator are zero | no current assets and no current liabilities | ask the data | Imputation placed them below typical, not at zero |

* **Fixed benchmark caps** (for example, current ratio = 3) should not be
  borrowed from corporate finance. On the 990 they sit well below the median
  of the observed distribution.
* **Bottom-coding a denominator at $1** turns a ratio into a dollar amount for
  the affected cases.
* **Median imputation** is only right when the anchor profile says the group is
  typical. For every undefined group examined here it isn't, and median fills
  weaken the correlations the index depends on.

**Transformations.**

* **Raw ratios** (even with rules applied) should not enter a correlation-based
  model. A few extreme values drive the Pearson correlations, and block
  structure is hard to see.
* **Winsorizing alone** is not enough. Skewness and kurtosis stay high, and
  correlations are still visibly distorted relative to the rank benchmark.
* **Winsorized + normalized** (percentile winsorizing, then asinh with a robust
  median scale for ratios that stay skewed, then a z-score) gets most of the
  way to rank-normal fidelity and internal consistency. It keeps interval
  spacing within the body of each distribution, and its parameters (bounds,
  scale, mean, standard deviation) can be stored and reapplied to another
  year. **This is the recommended default for index construction.**
* **Rank-normal** gives the highest fidelity and the cleanest blocks, and it is
  a good robustness check. But it throws away spacing: a 2-month and a
  20-month reserve are distinguished only by rank. It also turns real spikes
  (no debt; no unrestricted equity) into arbitrary spreads. Scoring a new year
  needs the reference-year distribution.
* **Package `_z`** performs comparably on correlations for the ratios it
  scores. But its hurdle and robust-MAD path fails on zero-inflated shares: a
  ratio whose median is zero gets a MAD of zero and comes back entirely `NA`
  (see the output above). That needs a fix before the `_z` columns are used
  for index construction.
* **Winsorizing level** is a separate choice from the transformation. 1/99
  keeps more real variation; 5/95 gives cleaner correlations. Report both when
  dimensions are sensitive to it.

**Order of operations.** Winsorize (at percentiles) before estimating any
normalization parameter. Never clip on z-scores computed from unwinsorized
data.

**Implications for identifying dimensions.**

* **Blocks.** With the full set of ratios measured this way, parallel analysis
  retains more components than a two-dimension solution. In this sample the
  clustering separates:
  * an **operating performance** block (surplus margin, return on assets,
    net-asset growth);
  * a **reserves** block (months of cash, liquid months, asset intensity);
  * a **balance-sheet structure and leverage** block (the debt ratios, together
    with the current ratio, cash share, and land share).

  Expense-structure ratios do *not* form a single block. Admin share stands
  apart, and fundraising share groups with self-sufficiency.
* **Ratios that share inputs.** Some apparent dimensions come from definitions,
  not from organizational behavior. Program, admin, and fundraising shares sum
  to 1 by construction. Leverage ratios on unrestricted equity rank
  organizations almost identically to debt-to-assets once negative
  denominators are handled. The next step is to re-run dimension discovery on
  the winsorized + normalized frame with one indicator per such family, then
  check which blocks survive that restriction.
