---

title: "From Financial Ratios to Distinct Nonprofit Financial Dimensions"
output:
html_document:
toc: true
toc_depth: 3
word_document: default
----------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(dplyr)
library(psych)
library(scales)
library(purrr)
library(tidyr)
library(stringr)
library(tibble)
```

# Overview

This document describes a reproducible workflow for moving from a large set of nonprofit financial ratios to a smaller set of distinct financial dimensions. The goal is not merely to reduce the number of variables, but to identify:

1. clusters of indicators that appear to measure a common underlying dimension,
2. singleton indicators that do not cluster with others but may still represent meaningful standalone constructs, and
3. variables that appear to behave mostly as noise and should be excluded from the final measurement system.

The underlying logic is pragmatic. Classical exploratory factor analysis (EFA) is attractive because it aims to recover latent constructs. In practice, however, financial-ratio systems often contain a mixture of:

* highly overlapping indicators,
* small clusters of related indicators,
* engineered ratios that behave as standalone measures, and
* weak or noisy indicators.

Because of this, a pure EFA solution may fit poorly even when the variables are substantively useful. Principal components analysis (PCA) is therefore used here as a descriptive clustering tool to identify stable dimensions. Those dimensions may then be refined into composite indices or one-factor measurement models for coherent subsets of variables.

# Conceptual strategy

The workflow separates variables into three classes:

## 1. Clustered indicators

These are variables that correlate with one another and appear to define a common dimension. These are candidates for:

* PCA-based interpretation,
* composite index construction, and
* optional one-factor models within each subset.

## 2. Standalone but meaningful indicators

These are variables that do not cluster strongly with others, but may still represent a distinct and substantively meaningful aspect of nonprofit financial condition. These should generally be retained in the final framework as individual indicators rather than forced into an index.

## 3. Weak or noisy indicators

These are variables with very low shared variance, weak correlation with the rest of the system, or unstable behavior across samples. These are candidates for removal from the final measurement framework.

# Working assumptions

This workflow assumes the starting object is a data frame `dd` containing numeric financial ratios.

# Step 1: Transform variables to reduce skew and align scales

The first step is to winsorize extreme tails, rescale variables, optionally log-transform strongly skewed variables, and standardize the result.

```{r transform-functions}
rescale_ratio <- function(x) {
  x[is.na(x)] <- median(x, na.rm = TRUE)

  x05 <- quantile(x, 0.05, na.rm = TRUE)
  x95 <- quantile(x, 0.95, na.rm = TRUE)

  x[x < x05] <- x05
  x[x > x95] <- x95

  x <- scales::rescale(x, to = c(1, 100))

  if (abs(mean(x, na.rm = TRUE) - median(x, na.rm = TRUE)) > 5) {
    x <- log(x)
  }

  x <- as.numeric(scale(x))
  x
}
```

```{r build-analysis-frame}
dr <- sapply(dd, rescale_ratio, simplify = TRUE) |> as.data.frame()
```

# Notes on transformation

This transformation is intentionally pragmatic rather than purely distribution-theoretic. It reduces the influence of outliers, puts variables on comparable scales, and makes the analysis more stable across financial ratios that may have very different raw units.

A future refinement would be to track, for each variable, whether the log transform was applied and whether the transformed variable remains substantively interpretable.

# Step 2: Remove redundant indicators

Highly redundant variables do not add substantive information and can destabilize factor-based methods.

```{r redundancy-check}
cor_mat <- cor(dr, use = "pairwise.complete.obs")

high_corr_idx <- which(abs(cor_mat) > 0.90 & abs(cor_mat) < 1, arr.ind = TRUE)
high_corr_idx
```

```{r redundancy-table}
high_corr_pairs <- as.data.frame(high_corr_idx) |>
  filter(row < col) |>
  mutate(
    var1 = rownames(cor_mat)[row],
    var2 = colnames(cor_mat)[col],
    r = cor_mat[cbind(row, col)]
  ) |>
  select(var1, var2, r)

high_corr_pairs
```

## Decision rule

For highly redundant pairs, choose one of three actions:

1. drop one variable based on weaker interpretability,
2. retain the conceptually preferable variable, or
3. combine them into a composite if they are substantively interchangeable.

The choice should be driven by theory and interpretability, not correlation alone.

# Step 3: Evaluate shared variance using KMO

The Kaiser-Meyer-Olkin statistic helps distinguish variables that participate in a shared multivariate structure from those that behave idiosyncratically.

```{r kmo-check}
kmo <- psych::KMO(cor_mat)
kmo$MSA
kmo$MSAi
```

## Rules of thumb

* below 0.50: problematic
* 0.50 to 0.60: weak
* 0.60 to 0.70: marginal but potentially usable
* above 0.70: good

Variables with low individual KMO are not necessarily meaningless. They may instead be standalone indicators. For that reason, low KMO should trigger classification, not automatic deletion.

# Step 4: Inspect weak overall connectedness

Some variables may simply have weak relationships with the rest of the ratio system.

```{r mean-abs-correlation}
mean_abs_cor <- apply(abs(cor_mat), 1, mean)
mean_abs_cor |> sort()
```

Variables with very low average absolute correlation may be:

* noise,
* overly specialized ratios, or
* meaningful but unique indicators.

This step is especially useful for flagging variables for closer substantive review.

# Step 5: Assess whether a global factor model is plausible

Before fitting any serious factor model, inspect the determinant of the correlation matrix.

```{r determinant-check}
det(cor_mat)
```

A determinant very close to zero suggests redundancy or near-singularity. That does not invalidate PCA, but it may make EFA unstable.

# Step 6: Use PCA to identify empirical dimensions

Because the ratio system may contain local clusters rather than a single coherent latent system, PCA is used here as a descriptive tool to identify stable dimensions.

```{r pca-screen}
pca_full <- psych::principal(dr, nfactors = min(8, ncol(dr)), rotate = "varimax")
print(pca_full$loadings, cutoff = 0.30)
```

## How PCA is used here

PCA is not being treated as a strict latent-variable model. Instead, it is used to answer a practical question:

Which indicators tend to move together strongly enough to justify interpretation as a common financial dimension?

Stable clusters revealed by PCA may be labeled and used as provisional dimensions if they:

* have at least two substantively coherent indicators,
* show reasonably strong loadings,
* are interpretable as a nonprofit financial construct, and
* are stable across alternative samples or time periods.

# Step 7: Classify variables into dimensions, standalone indicators, and noise

A useful next step is to classify variables based on their loadings, KMO, and overall connectedness.

```{r classify-variables}
loading_mat <- unclass(pca_full$loadings)
max_loading <- apply(abs(loading_mat), 1, max, na.rm = TRUE)
n_over_30 <- apply(abs(loading_mat) >= 0.30, 1, sum, na.rm = TRUE)
primary_component <- colnames(loading_mat)[apply(abs(loading_mat), 1, which.max)]

variable_diagnostics <- tibble(
  variable = rownames(loading_mat),
  kmo = kmo$MSAi[rownames(loading_mat)],
  mean_abs_cor = mean_abs_cor[rownames(loading_mat)],
  max_loading = max_loading[rownames(loading_mat)],
  n_loadings_over_30 = n_over_30,
  primary_component = primary_component
) |>
  mutate(
    preliminary_class = case_when(
      kmo < 0.50 & max_loading < 0.40 ~ "likely_noise",
      kmo < 0.50 & max_loading >= 0.40 ~ "standalone_or_local_dimension",
      n_loadings_over_30 >= 1 & max_loading >= 0.50 ~ "clustered_dimension_member",
      TRUE ~ "review"
    )
  )

variable_diagnostics |> arrange(preliminary_class, desc(max_loading))
```

## Interpretation logic

### Clustered dimension member

A variable belongs to a coherent dimension if it has:

* a strong primary loading,
* a clear substantive interpretation, and
* at least one partner variable that loads on the same component.

### Standalone or local dimension

A variable may still be worth keeping even if it has low KMO, if it appears to measure a distinct and important domain of fiscal condition.

### Likely noise

A variable is a stronger candidate for exclusion if it has:

* low KMO,
* very low average correlation,
* weak PCA loadings, and
* no clear theoretical importance.

# Step 8: Build indices for coherent subsets

For stable multi-item dimensions, there are two defensible approaches.

## Option A: Standardized mean index

This is transparent and robust.

```{r index-builder-function}
build_index <- function(data, vars) {
  rowMeans(scale(data[, vars]), na.rm = TRUE)
}
```

Example:

```{r example-index, eval=FALSE}
# df_out$performance_index <- build_index(dr, c("podpm_w", "sm_w", "roa_w"))
```

## Option B: One-factor model within each coherent subset

This is appropriate when a small set of variables appears to measure a single construct.

```{r one-factor-function}
fit_one_factor <- function(data, vars, fm = "minres") {
  psych::fa(data[, vars], nfactors = 1, fm = fm, rotate = "none", scores = "regression")
}
```

Example:

```{r one-factor-example, eval=FALSE}
# perf_fa <- fit_one_factor(dr, c("podpm_w", "sm_w", "roa_w"))
# head(perf_fa$scores)
```

## Recommendation

A reasonable measurement-development strategy is:

1. use PCA to identify coherent subsets,
2. label those subsets substantively,
3. fit one-factor models within subsets that appear unidimensional,
4. compare one-factor scores to simple standardized averages, and
5. retain the simpler index unless the factor score yields a meaningful improvement.

This is methodologically defensible because the one-factor model is no longer being applied to the entire heterogeneous ratio system, only to subsets that appear plausibly unidimensional.

# Step 9: Create a final measurement frame

The final output should contain:

* one column for each multi-item index or one-factor score,
* one column for each retained standalone indicator, and
* metadata describing how each measure was constructed.

A simple scaffold is shown below.

```{r final-frame-scaffold}
final_measures <- tibble(id = seq_len(nrow(dr)))

# Example placeholders:
# final_measures$performance_index <- build_index(dr, c("podpm_w", "sm_w", "roa_w"))
# final_measures$capital_structure_index <- build_index(dr, c("doci_w", "luna_w", "dar_w"))
# final_measures$growth_index <- build_index(dr, c("dgdr_w", "eidr_w"))
# final_measures$specialized_indicator <- dr$some_variable

final_measures
```

# Step 10: Create a metadata table for transparency

```{r metadata-scaffold}
measure_metadata <- tibble(
  final_measure = character(),
  measure_type = character(),   # index, factor_score, standalone_indicator
  source_variables = character(),
  rationale = character()
)

measure_metadata
```

# Recommended decision framework

A final variable or index should be retained if it satisfies one of the following:

## A. Cluster-based dimension

* at least two indicators move together,
* the resulting construct is interpretable,
* the dimension is stable across samples or time periods.

## B. Standalone meaningful indicator

* the indicator has clear theoretical importance,
* it captures a concept not represented elsewhere,
* even if it does not correlate strongly with the rest of the system.

## C. Exclude as likely noise

* weak connectedness,
* poor interpretability,
* unstable behavior,
* no distinct conceptual contribution.

# Provisional methodological conclusion

PCA can be used to identify and label stable empirical dimensions in a financial-ratio system, even if a full latent factor model is not appropriate. Those PCA-derived groupings can then be refined into simple indices or one-factor measurement models when the grouped variables appear substantively coherent and empirically unidimensional.

Variables that do not cluster should not automatically be discarded. Instead, they should be evaluated as either:

* standalone dimensions worth retaining, or
* weak/noisy ratios best excluded from the final framework.

This produces a hybrid measurement model: a final data frame containing both multi-item dimensions and unique standalone indicators.

# Variable-level diagnostics table

The next step is to create a unified variable-diagnostics table that combines the main statistics used to classify indicators as retained variables, redundant variables, standalone but meaningful indicators, or likely noise.

```{r diagnostic-functions}
compute_variable_diagnostics <- function(data,
                                         corr_cutoff = 0.90,
                                         kmo_low = 0.50,
                                         communality_low = 0.20,
                                         mean_abs_cor_low = 0.10,
                                         smc_high = 0.90,
                                         vif_high = 10,
                                         near_zero_sd = 1e-6,
                                         nfactors_comm = NULL,
                                         rotate = "oblimin",
                                         fm = "minres") {

  x <- data |> dplyr::select(where(is.numeric)) |> as.data.frame()

  if (ncol(x) < 2) stop("Need at least two numeric variables.")

  # remove zero-variance variables before correlation
  sds <- sapply(x, sd, na.rm = TRUE)
  zero_var <- names(sds)[is.na(sds) | sds <= near_zero_sd]

  x_use <- x[, sds > near_zero_sd & !is.na(sds), drop = FALSE]
  cor_mat <- cor(x_use, use = "pairwise.complete.obs")

  # choose a small communality model automatically if not supplied
  if (is.null(nfactors_comm)) {
    nfactors_comm <- max(1, min(3, floor(ncol(x_use) / 3)))
  }

  # basic diagnostics
  kmo_obj <- psych::KMO(cor_mat)
  msa_i <- kmo_obj$MSAi
  smc <- psych::smc(cor_mat)
  vif <- 1 / (1 - smc)
  mean_abs_cor <- apply(abs(cor_mat), 1, mean, na.rm = TRUE)
  max_abs_cor <- apply(abs(cor_mat), 1, function(v) max(v[v < 0.999999], na.rm = TRUE))
  det_full <- det(cor_mat)

  # delta determinant and log-determinant change
  delta_det <- sapply(seq_len(ncol(cor_mat)), function(i) det(cor_mat[-i, -i, drop = FALSE]) - det_full)
  names(delta_det) <- colnames(cor_mat)

  delta_log_det <- sapply(seq_len(ncol(cor_mat)), function(i) {
    d_sub <- det(cor_mat[-i, -i, drop = FALSE])
    log(abs(d_sub) + 1e-12) - log(abs(det_full) + 1e-12)
  })
  names(delta_log_det) <- colnames(cor_mat)

  # anti-image diagonal from inverse correlation matrix
  inv_cor <- tryCatch(solve(cor_mat), error = function(e) NULL)
  anti_image_diag <- if (!is.null(inv_cor)) 1 / diag(inv_cor) else rep(NA_real_, ncol(cor_mat))
  names(anti_image_diag) <- colnames(cor_mat)

  # one quick communality screen
  fa_try <- tryCatch(
    psych::fa(x_use, nfactors = nfactors_comm, fm = fm, rotate = rotate),
    error = function(e) NULL,
    warning = function(w) invokeRestart("muffleWarning")
  )

  communality <- if (!is.null(fa_try)) fa_try$communality else rep(NA_real_, ncol(x_use))
  names(communality) <- colnames(x_use)

  # pairwise redundancy information
  cor_long <- as.data.frame(as.table(cor_mat), stringsAsFactors = FALSE) |>
    dplyr::rename(var1 = Var1, var2 = Var2, r = Freq) |>
    dplyr::filter(var1 != var2) |>
    dplyr::mutate(abs_r = abs(r))

  redundancy_pairs <- cor_long |>
    dplyr::filter(abs_r >= corr_cutoff) |>
    dplyr::rowwise() |>
    dplyr::mutate(pair_id = paste(sort(c(var1, var2)), collapse = "~~")) |>
    dplyr::ungroup() |>
    dplyr::distinct(pair_id, .keep_all = TRUE)

  redundant_with <- lapply(colnames(cor_mat), function(v) {
    redundancy_pairs |>
      dplyr::filter(var1 == v | var2 == v) |>
      dplyr::mutate(other = ifelse(var1 == v, var2, var1)) |>
      dplyr::pull(other)
  })
  names(redundant_with) <- colnames(cor_mat)

  n_redundant_links <- sapply(redundant_with, length)

  diagnostics <- tibble::tibble(
    variable = colnames(cor_mat),
    sd = sds[colnames(cor_mat)],
    kmo_msa = as.numeric(msa_i[colnames(cor_mat)]),
    communality = as.numeric(communality[colnames(cor_mat)]),
    smc = as.numeric(smc[colnames(cor_mat)]),
    vif = as.numeric(vif[colnames(cor_mat)]),
    mean_abs_cor = as.numeric(mean_abs_cor[colnames(cor_mat)]),
    max_abs_cor = as.numeric(max_abs_cor[colnames(cor_mat)]),
    anti_image_diag = as.numeric(anti_image_diag[colnames(cor_mat)]),
    delta_det = as.numeric(delta_det[colnames(cor_mat)]),
    delta_log_det = as.numeric(delta_log_det[colnames(cor_mat)]),
    n_redundant_links = as.integer(n_redundant_links[colnames(cor_mat)]),
    redundant_with = vapply(redundant_with[colnames(cor_mat)], function(z) paste(z, collapse = "; "), character(1))
  ) |>
    dplyr::mutate(
      flag_zero_variance = sd <= near_zero_sd,
      flag_low_kmo = kmo_msa < kmo_low,
      flag_low_communality = !is.na(communality) & communality < communality_low,
      flag_high_redundancy = max_abs_cor >= corr_cutoff,
      flag_weak_correlation = mean_abs_cor < mean_abs_cor_low,
      flag_high_smc = smc >= smc_high,
      flag_high_vif = vif >= vif_high,
      prelim_class = dplyr::case_when(
        flag_zero_variance ~ "noise",
        flag_high_redundancy ~ "redundant_candidate",
        flag_low_kmo & flag_low_communality & flag_weak_correlation ~ "noise",
        flag_low_kmo & flag_low_communality ~ "standalone_or_noise",
        flag_low_communality ~ "standalone_candidate",
        TRUE ~ "retain"
      )
    )

  # append zero-variance variables if any were removed before the correlation step
  if (length(zero_var) > 0) {
    zero_tbl <- tibble::tibble(
      variable = zero_var,
      sd = sds[zero_var],
      kmo_msa = NA_real_,
      communality = NA_real_,
      smc = NA_real_,
      vif = NA_real_,
      mean_abs_cor = NA_real_,
      max_abs_cor = NA_real_,
      anti_image_diag = NA_real_,
      delta_det = NA_real_,
      delta_log_det = NA_real_,
      n_redundant_links = NA_integer_,
      redundant_with = NA_character_,
      flag_zero_variance = TRUE,
      flag_low_kmo = NA,
      flag_low_communality = NA,
      flag_high_redundancy = NA,
      flag_weak_correlation = NA,
      flag_high_smc = NA,
      flag_high_vif = NA,
      prelim_class = "noise"
    )
    diagnostics <- dplyr::bind_rows(diagnostics, zero_tbl)
  }

  attr(diagnostics, "cor_mat") <- cor_mat
  attr(diagnostics, "redundancy_pairs") <- redundancy_pairs
  attr(diagnostics, "kmo_object") <- kmo_obj
  attr(diagnostics, "fa_screen") <- fa_try
  diagnostics
}
```

## Definitions of diagnostic statistics

* `sd`: standard deviation of the transformed variable. Near-zero values indicate almost no variation.
* `kmo_msa`: variable-level Kaiser-Meyer-Olkin measure of sampling adequacy. Higher values indicate that the variable shares variance with the rest of the system.
* `communality`: proportion of variance explained by the quick common-factor screen. Low values suggest the variable does not belong well in a common-factor model.
* `smc`: squared multiple correlation, the R-squared from regressing one variable on all others. High values indicate redundancy and contribution to near-singularity.
* `vif`: variance inflation factor derived from the correlation matrix. High values indicate multicollinearity.
* `mean_abs_cor`: average absolute correlation with all other variables. Low values indicate weak overall connectedness.
* `max_abs_cor`: largest absolute pairwise correlation with any other variable. High values indicate that the variable may be nearly interchangeable with another.
* `anti_image_diag`: diagonal of the anti-image matrix. Smaller values generally indicate weaker common variance structure.
* `delta_det`: change in the determinant of the correlation matrix when the variable is removed.
* `delta_log_det`: log-scale version of the determinant change, easier to compare when determinants are very small.
* `n_redundant_links`: number of other variables exceeding the redundancy threshold with this variable.
* `redundant_with`: names of variables exceeding the redundancy threshold with this variable.

## Rules of thumb

* Near-zero variance: `sd <= 1e-6`
* Low KMO: `kmo_msa < 0.50`
* Weak KMO: `0.50 <= kmo_msa < 0.60`
* Low communality: `communality < 0.20`
* Weak overall connectedness: `mean_abs_cor < 0.10`
* High redundancy: `max_abs_cor >= 0.90`
* High SMC: `smc >= 0.90`
* High VIF: `vif >= 10`

These are screening heuristics rather than absolute decision rules. Final decisions should combine diagnostics with theory and interpretability.

# Variable triage report

The next function turns the diagnostics table into a structured decision report.

```{r triage-function}
triage_variables <- function(diagnostics,
                             keep_from_redundant = c("higher_kmo", "higher_mean_abs_cor", "first")) {

  keep_from_redundant <- match.arg(keep_from_redundant)

  d <- diagnostics |> dplyr::arrange(variable)

  choose_keeper <- function(sub) {
    if (nrow(sub) == 1) return(sub$variable)
    if (keep_from_redundant == "higher_kmo") {
      sub |> dplyr::arrange(dplyr::desc(kmo_msa), dplyr::desc(mean_abs_cor), variable) |> dplyr::slice(1) |> dplyr::pull(variable)
    } else if (keep_from_redundant == "higher_mean_abs_cor") {
      sub |> dplyr::arrange(dplyr::desc(mean_abs_cor), dplyr::desc(kmo_msa), variable) |> dplyr::slice(1) |> dplyr::pull(variable)
    } else {
      sort(sub$variable)[1]
    }
  }

  # Build redundancy groups from adjacency list
  vars_red <- d$variable[isTRUE(any(d$flag_high_redundancy, na.rm = TRUE))]
  adjacency <- lapply(d$variable, function(v) {
    txt <- d$redundant_with[d$variable == v]
    if (length(txt) == 0 || is.na(txt) || txt == "") character(0) else strsplit(txt, ";\s*")[[1]]
  })
  names(adjacency) <- d$variable

  visited <- setNames(rep(FALSE, length(adjacency)), names(adjacency))
  redundancy_groups <- list()

  for (v in names(adjacency)) {
    neigh <- adjacency[[v]]
    if (visited[[v]] || length(neigh) == 0) next
    queue <- c(v)
    comp <- character(0)
    while (length(queue) > 0) {
      cur <- queue[1]
      queue <- queue[-1]
      if (visited[[cur]]) next
      visited[[cur]] <- TRUE
      comp <- unique(c(comp, cur))
      nxt <- adjacency[[cur]]
      queue <- unique(c(queue, nxt[!visited[nxt]]))
    }
    redundancy_groups[[length(redundancy_groups) + 1]] <- sort(comp)
  }

  redundancy_tbl <- purrr::map_dfr(redundancy_groups, function(g) {
    sub <- d |> dplyr::filter(variable %in% g)
    keeper <- choose_keeper(sub)
    tibble::tibble(
      group = paste(g, collapse = " | "),
      kept = keeper,
      dropped = paste(setdiff(g, keeper), collapse = "; ")
    )
  })

  dropped_redundant <- if (nrow(redundancy_tbl) > 0) unique(unlist(strsplit(redundancy_tbl$dropped[redundancy_tbl$dropped != ""], ";\s*"))) else character(0)
  kept_redundant <- if (nrow(redundancy_tbl) > 0) redundancy_tbl$kept else character(0)

  build_reasons <- function(row) {
    out <- character(0)
    if (isTRUE(row$flag_zero_variance)) {
      out <- c(out, sprintf("Near-zero variance (sd = %.4f).", row$sd))
    }
    if (isTRUE(row$flag_high_redundancy)) {
      out <- c(out, sprintf("Highly redundant with %s (max |r| = %.3f).", row$redundant_with, row$max_abs_cor))
    }
    if (isTRUE(row$flag_low_kmo)) {
      out <- c(out, sprintf("Low KMO / MSA (%.3f), suggesting weak shared variance with the rest of the system.", row$kmo_msa))
    }
    if (isTRUE(row$flag_low_communality)) {
      out <- c(out, sprintf("Low communality (%.3f), suggesting poor fit in a common-factor model.", row$communality))
    }
    if (isTRUE(row$flag_weak_correlation)) {
      out <- c(out, sprintf("Weak overall correlation structure (mean absolute correlation = %.3f).", row$mean_abs_cor))
    }
    if (isTRUE(row$flag_high_smc)) {
      out <- c(out, sprintf("Highly predictable from the remaining variables (SMC = %.3f).", row$smc))
    }
    if (isTRUE(row$flag_high_vif)) {
      out <- c(out, sprintf("High multicollinearity (VIF = %.2f).", row$vif))
    }
    out
  }

  d$reasons <- purrr::pmap(d, function(...) build_reasons(as.list(list(...))))

  keep_vars <- d |> dplyr::filter(!variable %in% dropped_redundant) |>
    dplyr::filter(!(flag_zero_variance %in% TRUE)) |>
    dplyr::filter(!(flag_low_kmo %in% TRUE & flag_low_communality %in% TRUE & flag_weak_correlation %in% TRUE)) |>
    dplyr::pull(variable)

  low_comm_standalone <- d |>
    dplyr::filter(!variable %in% dropped_redundant) |>
    dplyr::filter(flag_low_communality %in% TRUE) |>
    dplyr::filter(!(flag_low_kmo %in% TRUE & flag_weak_correlation %in% TRUE)) |>
    dplyr::select(variable, kmo_msa, communality, mean_abs_cor, reasons)

  suspected_noise <- d |>
    dplyr::filter(
      flag_zero_variance %in% TRUE |
      (flag_low_kmo %in% TRUE & flag_low_communality %in% TRUE & flag_weak_correlation %in% TRUE)
    ) |>
    dplyr::select(variable, sd, kmo_msa, communality, mean_abs_cor, reasons)

  dropped_report <- d |>
    dplyr::filter(variable %in% c(dropped_redundant, suspected_noise$variable, low_comm_standalone$variable)) |>
    dplyr::mutate(reason_bullets = purrr::map_chr(reasons, function(x) paste0("- ", x, collapse = "
"))) |>
    dplyr::select(variable, reason_bullets)

  list(
    keep_for_subsequent_steps = keep_vars,
    highly_redundant_groups = redundancy_tbl,
    low_communality_but_possible_standalone = low_comm_standalone,
    suspected_noise = suspected_noise,
    dropped_report = dropped_report,
    full_diagnostics = d
  )
}
```

# Example workflow

```{r example-workflow, eval=FALSE}
diag_tbl <- compute_variable_diagnostics(dr)

# inspect the unified diagnostics table
mdiag <- diag_tbl |>
  dplyr::arrange(prelim_class, dplyr::desc(max_abs_cor), kmo_msa)
mdiag

triage <- triage_variables(diag_tbl)

triage$keep_for_subsequent_steps
triage$highly_redundant_groups
triage$low_communality_but_possible_standalone
triage$suspected_noise
triage$dropped_report
```

# Notes on interpretation

The triage function is intentionally conservative. It distinguishes between three different reasons a variable may be excluded from the multivariate measurement model:

1. **Redundant**: the variable is essentially interchangeable with another.
2. **Standalone but low-communality**: the variable may still be conceptually useful, but it does not belong well in a common dimension.
3. **Likely noise**: the variable has little variance, little shared structure, and little sign of being a stable dimension.

The final choice should always combine empirical diagnostics with substantive judgment about the meaning of the ratio.

# Next steps

The next version of this document should add:

1. a function that automatically turns retained clusters into candidate indices,
2. stability checks across subsamples or years,
3. optional alpha or omega estimates for multi-item indices,
4. a decision log for variables dropped, retained, or treated as standalone, and
5. final export code for the refined measurement dataset.
