---
title: "Should Reserves and Performance Carry Separate Weight?"
subtitle: "Accumulation, overlap, and incremental validity of the health dimensions"
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
    css: walkthrough.css
params:
  panel_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2019-2024-imputed.rds")
  results_file: ""
  seed: 2023
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 130, knitr.kable.NA = "")
```

# The question

H5 showed that about 69 cents of every surplus dollar ends the year in cash or
investments. Reserves are therefore largely the running total of past
performance. `fiscal_weights()` weights dimensions equally by default, so an
index that scores reserves and performance separately counts the same history
twice: once as the stock, once as this year's increment to it.

Three things decide whether that is a problem:

1. **Accumulation.** How much of today's reserves is explained by past reserves
   plus the surpluses since?
2. **Overlap.** How correlated are the dimension scores in a given year?
3. **Incremental validity.** Does performance predict later trouble *beyond*
   reserves, and does reserves predict beyond performance? Weights should
   follow the answer, not the count of dimensions.

```{r packages}
suppressMessages({
  library(data.table)
  library(bit64)
  library(knitr)
  library(splines)
})
source("build-ratio-frame.R")   # the measurement model: rules, imputation, winsorize + normalize
set.seed(params$seed)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

# 1. Dimension scores by year

Scores use the retained indicator sets from `dimension-discovery-2023.Rmd`,
each member winsorized and normalized, then averaged. All three are oriented so
that **higher is healthier**, which means leverage enters with its sign
flipped.

```{r scores}
d <- as.data.table(readRDS(params$panel_file))
d <- d[RETURN_TYPE == "990" & substr(TAX_PERIOD_END_DATE, 6, 7) == "12" & TAX_YEAR %in% 2019:2024]
keep <- d[TAX_YEAR %in% 2019:2023, .N, by = EIN2][N == 5, EIN2]
in2024 <- unique(d[TAX_YEAR == 2024, EIN2])
d <- d[EIN2 %chin% keep]

MEMBERS <- list(reserves = c("months_cash", "liquid_months", "assets_rev", "current"),
                performance = c("surplus_margin", "return_assets", "na_growth"),
                leverage = c("debt_assets", "payables_una"))

year_scores <- function(y) {
  built <- build_ratio_frame(d[TAX_YEAR == y], seed = params$seed)
  z <- as.data.frame(lapply(built$frame[unlist(MEMBERS)], winsor_normalize))
  out <- data.table(EIN2 = built$ein, year = y)
  for (nm in names(MEMBERS)) {
    s <- rowMeans(z[, MEMBERS[[nm]], drop = FALSE])
    out[[nm]] <- as.numeric(scale(if (nm == "leverage") -s else s))
  }
  k <- built$components
  out[, `:=`(surplus_per_expense = (k$R - k$E) / k$E, E = k$E, A = k$A,
             cash_months = 12 * (k$cash + k$recv) / pmax(k$E, 1))]
  out[]
}
S <- rbindlist(lapply(2019:2023, year_scores))
kable(S[, .(organizations = .N, mean_reserves = mean(reserves), mean_performance = mean(performance),
            mean_leverage = mean(leverage)), by = year][order(year)], digits = 3,
      format.args = list(big.mark = ","))
```

# 2. Overlap between the dimensions

```{r overlap}
wide <- dcast(S, EIN2 ~ year, value.var = c("reserves", "performance", "leverage",
                                            "surplus_per_expense", "cash_months", "E"))
setnames(wide, gsub("_(\\d{4})$", "\\1", names(wide)))
wide <- wide[complete.cases(wide)]
kable(round(cor(S[year == 2023, .(reserves, performance, leverage)], method = "spearman"), 3),
      caption = "Dimension scores, 2023")
kable(data.frame(
  comparison = c("reserves 2023 vs. performance 2023",
                 "reserves 2023 vs. mean performance 2019-2022",
                 "reserves 2023 vs. reserves 2019",
                 "performance 2023 vs. performance 2019"),
  spearman = c(cor(wide$reserves2023, wide$performance2023, method = "spearman"),
               cor(wide$reserves2023, rowMeans(wide[, .(performance2019, performance2020,
                                                        performance2021, performance2022)]),
                   method = "spearman"),
               cor(wide$reserves2023, wide$reserves2019, method = "spearman"),
               cor(wide$performance2023, wide$performance2019, method = "spearman"))),
  digits = 3)
```

A single year's performance is nearly uncorrelated with the stock it feeds, and
performance barely persists while reserves persist strongly. That is the shape
of a flow feeding a slow-moving stock.

# 3. Accumulation: is reserves the running total?

```{r accumulation}
ols <- function(y, W) {
  XtXi <- solve(crossprod(W)); b <- drop(XtXi %*% crossprod(W, y))
  e <- drop(y - W %*% b)
  V <- XtXi %*% crossprod(W * e) %*% XtXi * nrow(W) / (nrow(W) - ncol(W))
  list(coef = data.table(term = colnames(W), estimate = b, se = sqrt(diag(V))),
       fitted = drop(W %*% b), r2 = 1 - var(e) / var(y))
}
wz <- function(x) { q <- quantile(x, c(0.01, 0.99), na.rm = TRUE)
                    as.numeric(scale(pmin(pmax(x, q[[1]]), q[[2]]))) }
wide[, cum_surplus := wz(surplus_per_expense2020 + surplus_per_expense2021 +
                         surplus_per_expense2022 + surplus_per_expense2023)]
acc <- rbind(
  data.table(model = "reserves 2019 only",
             r2 = ols(wide$reserves2023, cbind(1, r19 = wide$reserves2019))$r2),
  data.table(model = "cumulative surplus only",
             r2 = ols(wide$reserves2023, cbind(1, cum = wide$cum_surplus))$r2),
  data.table(model = "both",
             r2 = ols(wide$reserves2023, cbind(1, r19 = wide$reserves2019, cum = wide$cum_surplus))$r2))
kable(acc, digits = 4, caption = "Explaining the 2023 reserves score")
both <- ols(wide$reserves2023, cbind(1, reserves_2019 = wide$reserves2019,
                                     cumulative_surplus = wide$cum_surplus))
kable(both$coef[term != ""], digits = 4, caption = "Coefficients (standardized)")

# In natural units: months of cash gained per dollar of cumulative surplus.
# Months of cash and surplus ratios are heavy-tailed, so both are winsorized at
# 1/99 before this regression; untrimmed, a handful of filers dominate it.
wins <- function(x) { q <- quantile(x, c(0.01, 0.99), na.rm = TRUE); pmin(pmax(x, q[[1]]), q[[2]]) }
cum_raw <- wins(wide$surplus_per_expense2020 + wide$surplus_per_expense2021 +
                wide$surplus_per_expense2022 + wide$surplus_per_expense2023)
mc <- ols(wins(wide$cash_months2023), cbind(1, cash_months_2019 = wins(wide$cash_months2019),
                                            cumulative_surplus_per_expense = cum_raw))
kable(mc$coef, digits = 3,
      caption = "Months of cash in 2023 (winsorized), per dollar of cumulative surplus per dollar of expenses")
```

# 4. Incremental validity

Scores are measured in 2021 and used to predict 2022-2023. The question for
weighting is not whether each dimension predicts on its own, but how much each
adds to the others.

```{r validity}
flows <- d[TAX_YEAR %in% 2021:2023, .(EIN2, year = TAX_YEAR,
                                      R = as.numeric(F9_08_REV_TOT_TOT),
                                      E = as.numeric(F9_09_EXP_TOT_TOT),
                                      na = as.numeric(F9_10_NAFB_TOT_EOY),
                                      cash = as.numeric(F9_10_ASSET_CASH_EOY) +
                                             as.numeric(F9_10_ASSET_SAVING_EOY))]
for (cl in c("R", "E", "na", "cash")) flows[[cl]][is.na(flows[[cl]])] <- 0
o <- dcast(flows, EIN2 ~ year, value.var = c("R", "E", "na", "cash"))
setnames(o, gsub("_(\\d{4})$", "\\1", names(o)))
o <- o[complete.cases(o)]
o[, `:=`(both_years_deficit = as.numeric(R2022 < E2022 & R2023 < E2023),
         any_deficit = as.numeric(R2022 < E2022 | R2023 < E2023),
         na_fell = as.numeric(na2023 < na2021),
         cash_fell_month = as.numeric(12 * (cash2023 - cash2021) / pmax(E2021, 1) < -1),
         no2024 = as.numeric(!(EIN2 %chin% in2024)))]
m <- merge(o, S[year == 2021, .(EIN2, reserves, performance, leverage, E, A)], by = "EIN2")

auc <- function(score, y) {
  r <- rank(score); n1 <- as.numeric(sum(y == 1)); n0 <- as.numeric(sum(y == 0))
  (sum(r[y == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
ctrl <- cbind(intercept = 1, ns(log(pmax(m$E, 1)), df = 4))
OUT <- c(both_years_deficit = "both_years_deficit", any_deficit = "any_deficit",
         na_fell = "na_fell", cash_fell_month = "cash_fell_month", no2024 = "no2024")
race <- rbindlist(lapply(names(OUT), function(nm) {
  y <- m[[OUT[[nm]]]]
  fit <- function(cols) {
    W <- if (length(cols)) cbind(ctrl, as.matrix(m[, ..cols])) else ctrl
    f <- ols(y, W); list(auc = auc(f$fitted, y), coef = f$coef)
  }
  base <- fit(character()); res <- fit("reserves"); per <- fit("performance")
  lev <- fit("leverage"); all3 <- fit(c("reserves", "performance", "leverage"))
  cf <- all3$coef
  data.table(outcome = nm, rate = mean(y),
             auc_size_only = base$auc, auc_reserves = res$auc, auc_performance = per$auc,
             auc_leverage = lev$auc, auc_all_three = all3$auc,
             gain_performance_over_reserves = all3$auc - fit(c("reserves", "leverage"))$auc,
             gain_reserves_over_performance = all3$auc - fit(c("performance", "leverage"))$auc,
             beta_reserves = cf[term == "reserves", estimate],
             beta_performance = cf[term == "performance", estimate],
             beta_leverage = cf[term == "leverage", estimate])
}))
kable(race[, .(outcome, rate, auc_size_only, auc_reserves, auc_performance, auc_leverage,
               auc_all_three)], digits = 4, caption = "AUC by predictor set (2021 scores, 2022-2023 outcomes)")
kable(race[, .(outcome, gain_performance_over_reserves, gain_reserves_over_performance,
               beta_reserves, beta_performance, beta_leverage)], digits = 4,
      caption = "Incremental AUC and coefficients in the joint model (per SD, higher score = healthier)")
```

**Level or change?** If performance is simply the annual increment to reserves,
the change in the reserves score should carry the same information. This
compares performance against the two-year change in reserves.

```{r level-change}
mm <- merge(m, wide[, .(EIN2, reserves2019, reserves2021)], by = "EIN2")
mm[, d_reserves := wz(reserves2021 - reserves2019)]
lc <- rbindlist(lapply(names(OUT), function(nm) {
  y <- mm[[OUT[[nm]]]]
  c0 <- cbind(intercept = 1, ns(log(pmax(mm$E, 1)), df = 4))
  f_r  <- ols(y, cbind(c0, reserves = mm$reserves))
  f_rp <- ols(y, cbind(c0, reserves = mm$reserves, performance = mm$performance))
  f_rd <- ols(y, cbind(c0, reserves = mm$reserves, change_in_reserves = mm$d_reserves))
  f_all <- ols(y, cbind(c0, reserves = mm$reserves, performance = mm$performance,
                        change_in_reserves = mm$d_reserves))
  data.table(outcome = nm, auc_reserves = auc(f_r$fitted, y),
             auc_plus_performance = auc(f_rp$fitted, y),
             auc_plus_change = auc(f_rd$fitted, y),
             auc_plus_both = auc(f_all$fitted, y),
             beta_performance = f_all$coef[term == "performance", estimate],
             beta_change = f_all$coef[term == "change_in_reserves", estimate])
}))
kable(lc, digits = 4, caption = "Performance vs. the change in reserves")
```

# 5. Implied weights

If weights should track how much each dimension contributes to predicting
trouble, one defensible rule is to set them proportional to each dimension's
incremental AUC, averaged across outcomes, rather than to give each an equal
share.

```{r weights}
gains <- rbindlist(lapply(names(OUT), function(nm) {
  y <- m[[OUT[[nm]]]]
  fit_auc <- function(cols) {
    W <- if (length(cols)) cbind(ctrl, as.matrix(m[, ..cols])) else ctrl
    auc(ols(y, W)$fitted, y)
  }
  all3 <- fit_auc(c("reserves", "performance", "leverage"))
  data.table(outcome = nm,
             reserves = all3 - fit_auc(c("performance", "leverage")),
             performance = all3 - fit_auc(c("reserves", "leverage")),
             leverage = all3 - fit_auc(c("reserves", "performance")))
}))
avg <- colMeans(gains[, .(reserves, performance, leverage)])
kable(rbind(gains, data.table(outcome = "average", t(avg))), digits = 4,
      caption = "Incremental AUC by dimension")
kable(data.frame(dimension = names(avg),
                 equal_weight = rep(1 / 3, 3),
                 incremental_weight = pmax(avg, 0) / sum(pmax(avg, 0))),
      digits = 3, caption = "Equal weights vs. weights proportional to incremental AUC")
```

```{r save, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(scores = S[, .(n = .N, reserves = mean(reserves), performance = mean(performance),
                              leverage = mean(leverage)), by = year][order(year)],
               overlap_2023 = cor(S[year == 2023, .(reserves, performance, leverage)], method = "spearman"),
               persistence = c(reserves = cor(wide$reserves2023, wide$reserves2019, method = "spearman"),
                               performance = cor(wide$performance2023, wide$performance2019, method = "spearman"),
                               reserves_vs_performance = cor(wide$reserves2023, wide$performance2023,
                                                             method = "spearman")),
               accumulation = acc, accumulation_coef = both$coef, months = mc$coef,
               race = race, level_change = lc, gains = gains, implied = avg, n = nrow(m)),
          params$results_file)
}
```

```{r session}
sessionInfo()
```
