---
title: "H1-H3 on the Panel"
subtitle: "Year-by-year replication, and matching inside organizations at one, two, and four years"
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
    css: walkthrough.css
params:
  panel_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2019-2024-imputed.rds")
  results_file: ""
  seed: 2023
  perms: 200
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 130, knitr.kable.NA = "")
```

# Purpose

`matching-hypotheses-2023.Rmd` tested three financing hypotheses on one year of
cross-sectional data:

* **H1** long-term debt tracks fixed assets (supported, elasticity 0.51);
* **H2** short-term liabilities track non-cash current assets (supported, 0.20);
* **H3** organizations with more cash carry less long-term debt (inconclusive:
  they are less likely to borrow at all, but the amount is unrelated).

A cross-section compares a hospital to a food bank. The panel asks the question
the theory is actually about: **when this organization's property grows, does
its own debt grow?** H4 has already shown how far a within-organization answer
can diverge from a cross-sectional one, and the matching-gap validation showed
the cross-sectional benchmark mostly sorts stable types.

Three additions here:

1. **Year-by-year replication**, 2019-2023, on identical definitions.
2. **First differences inside organizations** at one, two, and four years.
   Property is lumpy and debt amortizes, so a single year catches only
   acquisition timing; a longer difference should be steadier if matching is
   real.
3. **A dynamic H3**: does cash on hand predict less *new* borrowing? That is
   the question the cross-section cannot answer, because borrowing itself puts
   cash on the balance sheet.

```{r packages}
suppressMessages({
  library(data.table)
  library(bit64)
  library(knitr)
  library(splines)
})
source("matching-gap.R")   # gap_lines(): the same line definitions as H1-H3
set.seed(params$seed)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

# 1. Sample

Organizations filing a full 990 with a December year-end in every year
2019-2023, so the same organizations appear in every estimate.

```{r data}
d <- as.data.table(readRDS(params$panel_file))
d <- d[RETURN_TYPE == "990" & substr(TAX_PERIOD_END_DATE, 6, 7) == "12" & TAX_YEAR %in% 2019:2023]
keep <- d[, .N, by = EIN2][N == 5, EIN2]
d <- d[EIN2 %chin% keep]

year_lines <- function(y) {
  m <- as.data.table(gap_lines(d[TAX_YEAR == y], "EOY"))
  setnames(m, "ein", "EIN2")
  m <- m[E > 0 & A > 0]
  m[, year := y][]
}
L <- rbindlist(lapply(2019:2023, year_lines))
kable(L[, .(organizations = .N, any_long_term_debt = mean(lt > 0), any_short_term = mean(st > 0),
            any_fixed_assets = mean(fixed > 0)), by = year][order(year)], digits = 3,
      format.args = list(big.mark = ","))
```

# 2. Year-by-year replication

The specification is the one from the pre-registered tests: log(1 + x) plus an
indicator for each of fixed assets, cash, and non-cash current assets, with
4-df splines of log expenses and log(1 + other assets) as size controls.

```{r cross-sections}
robust <- function(y, W) {
  XtXi <- solve(crossprod(W)); b <- drop(XtXi %*% crossprod(W, y))
  e <- drop(y - W %*% b)
  V <- XtXi %*% crossprod(W * e) %*% XtXi * nrow(W) / (nrow(W) - ncol(W))
  data.table(term = colnames(W), estimate = b, se = sqrt(diag(V)))
}
design <- function(m) {
  other <- pmax(m$A - m$cash - m$ncca - m$fixed, 0)
  sE <- ns(log(m$E), df = 4); colnames(sE) <- paste0("size_E", 1:4)
  sO <- ns(log1p(other), df = 4); colnames(sO) <- paste0("size_other", 1:4)
  cbind(log_fixed = log1p(m$fixed), any_fixed = as.numeric(m$fixed > 0),
        log_cash = log1p(m$cash), any_cash = as.numeric(m$cash > 0),
        log_ncca = log1p(m$ncca), any_ncca = as.numeric(m$ncca > 0),
        intercept = 1, sE, sO, no_other = as.numeric(other == 0))
}
per_year <- rbindlist(lapply(2019:2023, function(y) {
  m <- L[year == y]; W <- design(m)
  f1 <- robust(as.numeric(m$lt > 0), W)
  f2 <- robust(log(m$lt[m$lt > 0]), W[m$lt > 0, , drop = FALSE])
  f3 <- robust(log(m$st[m$st > 0]), W[m$st > 0, , drop = FALSE])
  data.table(year = y,
             H1_any_lt_fixed = f1[term == "log_fixed", estimate],
             H1_amount_lt_fixed = f2[term == "log_fixed", estimate],
             H2_amount_st_ncca = f3[term == "log_ncca", estimate],
             cross_check_lt_ncca = f2[term == "log_ncca", estimate],
             cross_check_st_fixed = f3[term == "log_fixed", estimate],
             H3_any_lt_cash = f1[term == "log_cash", estimate],
             H3_amount_lt_cash = f2[term == "log_cash", estimate])
}))
kable(per_year, digits = 4, caption = "Carrying coefficients by year (cross-sections)")
```

# 3. Matching inside organizations

Changes are scaled by expenses in the first year of the window, so a
coefficient reads as **cents of debt per dollar of new assets**. Rows outside
the 1st/99th percentile of any variable in the model are dropped, which keeps
the changes on their original scale rather than clipping them.

```{r differences}
wide <- dcast(L, EIN2 ~ year, value.var = c("fixed", "cash", "ncca", "lt", "st", "E", "A"))
setnames(wide, gsub("_(\\d{4})$", "\\1", names(wide)))
wide <- wide[complete.cases(wide)]

window_frame <- function(y0, y1) {
  g <- function(stem, y) wide[[paste0(stem, y)]]
  E0 <- g("E", y0)
  z <- data.table(
    EIN2 = wide$EIN2,
    d_lt = (g("lt", y1) - g("lt", y0)) / E0,
    d_st = (g("st", y1) - g("st", y0)) / E0,
    d_fixed = (g("fixed", y1) - g("fixed", y0)) / E0,
    d_cash = (g("cash", y1) - g("cash", y0)) / E0,
    d_ncca = (g("ncca", y1) - g("ncca", y0)) / E0,
    d_logE = log(g("E", y1)) - log(E0),
    cash_months0 = 12 * g("cash", y0) / E0,
    assets_exp0 = g("A", y0) / E0,
    lt0 = g("lt", y0) / E0, E0 = E0, A0 = g("A", y0))
  keep <- rep(TRUE, nrow(z))
  for (v in c("d_lt", "d_st", "d_fixed", "d_cash", "d_ncca", "d_logE")) {
    q <- quantile(z[[v]], c(0.01, 0.99), na.rm = TRUE)
    keep <- keep & z[[v]] >= q[[1]] & z[[v]] <= q[[2]]
  }
  z[keep]
}

fit_window <- function(z, label) {
  X <- cbind(intercept = 1, d_fixed = z$d_fixed, d_cash = z$d_cash, d_ncca = z$d_ncca,
             d_logE = z$d_logE, ns(log(z$E0), df = 4), ns(log1p(z$A0), df = 4))
  colnames(X)[6:13] <- c(paste0("size_E", 1:4), paste0("size_A", 1:4))
  lt <- robust(z$d_lt, X); st <- robust(z$d_st, X)
  rbind(
    data.table(window = label, n = nrow(z), outcome = "new long-term debt", predictor = "new fixed assets",
               estimate = lt[term == "d_fixed", estimate], se = lt[term == "d_fixed", se]),
    data.table(window = label, n = nrow(z), outcome = "new long-term debt", predictor = "new current assets",
               estimate = lt[term == "d_ncca", estimate], se = lt[term == "d_ncca", se]),
    data.table(window = label, n = nrow(z), outcome = "new short-term liabilities", predictor = "new current assets",
               estimate = st[term == "d_ncca", estimate], se = st[term == "d_ncca", se]),
    data.table(window = label, n = nrow(z), outcome = "new short-term liabilities", predictor = "new fixed assets",
               estimate = st[term == "d_fixed", estimate], se = st[term == "d_fixed", se]))
}

w1 <- window_frame(2022, 2023); w2 <- window_frame(2021, 2023); w4 <- window_frame(2019, 2023)
diffs <- rbind(fit_window(w1, "1 year (2022-2023)"), fit_window(w2, "2 years (2021-2023)"),
               fit_window(w4, "4 years (2019-2023)"))
kable(dcast(diffs, outcome + predictor ~ window, value.var = "estimate"), digits = 4,
      caption = "Cents of new debt per dollar of new assets")
kable(diffs[predictor == "new fixed assets" & outcome == "new long-term debt"], digits = 4,
      caption = "H1 within organizations, with standard errors")
```

**The cross-prediction.** Matching implies that new property draws long-term
debt more than it draws payables, and that new receivables and inventory do the
reverse.

```{r cross-prediction}
kable(dcast(diffs, window ~ outcome + predictor, value.var = "estimate"), digits = 4)
```

# 4. A dynamic H3

Does an organization that starts a window with more cash borrow less over it?
The cross-section could not separate this from the fact that borrowing itself
puts cash on the balance sheet; here cash is measured before the borrowing.

Starting cash months are heavy-tailed, so they enter winsorized at 1/99 and
standardized: the coefficient is per standard deviation of starting cash.

```{r dynamic-h3}
wz <- function(x) { q <- quantile(x, c(0.01, 0.99), na.rm = TRUE)
                    as.numeric(scale(pmin(pmax(x, q[[1]]), q[[2]]))) }
h3 <- rbindlist(lapply(list(`1 year (2022-2023)` = w1, `2 years (2021-2023)` = w2,
                            `4 years (2019-2023)` = w4), function(z) {
  X <- cbind(intercept = 1, cash_start = wz(z$cash_months0), d_fixed = z$d_fixed, d_ncca = z$d_ncca,
             d_logE = z$d_logE, lt0 = wz(z$lt0), ns(log(z$E0), df = 4),
             ns(log(pmax(z$assets_exp0, 0.01)), df = 4))
  colnames(X)[7:14] <- c(paste0("size_E", 1:4), paste0("asset_intensity", 1:4))
  f <- robust(z$d_lt, X)
  cbind(f[term == "cash_start", .(estimate, se)],
        median_cash_months = median(z$cash_months0),
        sd_cash_months = sd(pmin(pmax(z$cash_months0, quantile(z$cash_months0, 0.01)),
                                 quantile(z$cash_months0, 0.99))))
}), idcol = "window")
kable(h3, digits = 5,
      caption = "New long-term debt (as a share of starting expenses) per SD of starting cash")
```

# 5. Placebo: shuffled asset growth

The within-organization estimates have no construction null of their own, so
the headline coefficient is compared with refits in which new fixed assets are
shuffled among organizations of similar starting size and asset intensity.

```{r placebo}
size_cells <- function(E, A, bins = 20) {
  eb <- cut(rank(E, ties.method = "first"), bins, labels = FALSE)
  ab <- ave(A / E, eb, FUN = function(z) cut(rank(z, ties.method = "first"), bins, labels = FALSE))
  (eb - 1) * bins + ab
}
permute_within <- function(cells) {
  n <- length(cells); p <- integer(n)
  p[order(cells, seq_len(n))] <- order(cells, runif(n))
  p
}
placebo <- function(z, label, B = params$perms) {
  O <- cbind(intercept = 1, d_cash = z$d_cash, d_ncca = z$d_ncca, d_logE = z$d_logE,
             ns(log(z$E0), df = 4), ns(log1p(z$A0), df = 4))
  q <- qr(O); ry <- qr.resid(q, z$d_lt)
  cells <- size_cells(z$E0, z$A0)
  coef_for <- function(v) { rx <- qr.resid(q, v); sum(rx * ry) / sum(rx * rx) }
  obs <- coef_for(z$d_fixed)
  perms <- vapply(seq_len(B), function(i) coef_for(z$d_fixed[permute_within(cells)]), 0)
  data.table(window = label, observed = obs, perm_lo = quantile(perms, 0.005),
             perm_hi = quantile(perms, 0.995), clears = obs > quantile(perms, 0.995))
}
plac <- rbind(placebo(w1, "1 year (2022-2023)"), placebo(w2, "2 years (2021-2023)"),
              placebo(w4, "4 years (2019-2023)"))
kable(plac, digits = 4, row.names = FALSE,
      caption = "H1 within organizations against shuffled asset growth")
```

# 6. Findings

```{r findings, include = FALSE}
h1w <- diffs[outcome == "new long-term debt" & predictor == "new fixed assets"]
```

* **Year by year.** The cross-sectional coefficients appear in section 2;
  stability across 2019-2023 is the test of whether the pandemic changed the
  relationship.
* **Within organizations.** A dollar of new property comes with
  `r fmt(h1w$estimate[1])` of new long-term debt over one year,
  `r fmt(h1w$estimate[2])` over two, and `r fmt(h1w$estimate[3])` over four.
* **Dynamic H3.** The coefficient on starting cash months appears in section 4;
  a negative value means cash-rich organizations borrow less afterwards.

```{r save, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(per_year = per_year, diffs = diffs, h3 = h3, placebo = plac,
               n = list(w1 = nrow(w1), w2 = nrow(w2), w4 = nrow(w4), orgs = length(keep)),
               coverage = L[, .(organizations = .N, any_lt = mean(lt > 0), any_st = mean(st > 0),
                                any_fixed = mean(fixed > 0)), by = year][order(year)]),
          params$results_file)
}
```

```{r session}
sessionInfo()
```
