---
title: "H4: Does Earned Revenue Replace Fundraising?"
subtitle: "Pre-specified cross-section, plus the COVID earned-revenue shock"
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
    css: walkthrough.css
params:
  panel_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2019-2024-imputed.rds")
  results_file: ""
  seed: 2023
  perms: 200
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 120, knitr.kable.NA = "")
```

# Purpose

H4 from the ratio-correlation memo: **organizations that earn their revenue
spend less on raising it.** The correlation that motivated it is
`self_suff` x `fundr_exp` = -0.21, which is -0.24 beyond the conditional
construction null, one of the few relationships in the 14-ratio set that
survives that benchmark.

Two tests:

* **A. Pre-specified cross-section.** Fundraising expense on program service
  revenue, holding contributions and size. Written in the memo before
  estimation and repeated below unchanged.
* **B. The COVID shock (exploratory).** Program service revenue fell from 16.5%
  of revenue in 2019 to 10.6% in 2020 and recovered by 2023. If earned revenue
  and fundraising are substitutes, organizations that lost the most earned
  revenue should have raised fundraising effort afterwards. This design was not
  pre-specified; it is labeled exploratory throughout.

```{r packages}
suppressMessages({
  library(data.table)
  library(bit64)
  library(knitr)
  library(splines)
})
set.seed(params$seed)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

# 1. Pre-specification (test A)

| Outcome | Coefficient that carries the test | Support if | Reject if |
|---|---|---|---|
| Any fundraising expense; log fundraising expense | log program service revenue, holding log contributions and size | < 0 in both parts, with the contributions coefficient > 0 | >= 0 |

* **Sample.** Pooled tax years 2022 and 2023, December year-ends,
  organizations that filed a full 990 every year 2019-2023.
* **Models.** Linear probability model for any fundraising expense; OLS on log
  fundraising expense among filers reporting a positive amount. Predictors
  enter as log(1 + x) plus an indicator for x > 0. Size controls: 4-df natural
  splines of log expenses and log(1 + total assets).
* **Evidence rule.** Same as H1-H3: the estimate must have the predicted sign
  and lie beyond the 0.5th or 99.5th percentile of `r params$perms` refits in
  which the program-revenue block is shuffled within cells of expenses x assets
  per dollar of expenses.

# 2. Data

```{r data}
d <- as.data.table(readRDS(params$panel_file))
num <- function(x) { x <- as.numeric(x); x[is.na(x)] <- 0; x }
forms <- d[, .(n990 = sum(RETURN_TYPE == "990"), nez = sum(RETURN_TYPE == "990EZ")), by = EIN2]
x <- d[EIN2 %chin% forms[n990 >= 5 & nez == 0, EIN2] & RETURN_TYPE == "990" &
       substr(TAX_PERIOD_END_DATE, 6, 7) == "12"]
rm(d); invisible(gc())

v <- data.table(
  EIN2 = x$EIN2, year = x$TAX_YEAR,
  subsector = ifelse(is.na(x$nteev2_subsector) | x$nteev2_subsector == "", "unknown",
                     as.character(x$nteev2_subsector)),
  E = num(x$F9_09_EXP_TOT_TOT), A = num(x$F9_10_ASSET_TOT_EOY), R = num(x$F9_08_REV_TOT_TOT),
  fundr = pmax(num(x$F9_09_EXP_TOT_FUNDR), 0),
  prog  = pmax(num(x$F9_08_REV_PROG_TOT_TOT), 0),
  contr = pmax(num(x$F9_08_REV_CONTR_TOT), 0))
rm(x); invisible(gc())
v <- v[E > 0 & A > 0]
kable(v[, .(org_years = .N, any_fundraising = mean(fundr > 0), any_program_revenue = mean(prog > 0),
            median_program_share = median(prog / pmax(R, 1))), by = year][order(year)], digits = 3,
      format.args = list(big.mark = ","))
```

# 3. Test A: the pre-specified cross-section

```{r functions}
size_cells <- function(E, A, bins = 20) {
  eb <- cut(rank(E, ties.method = "first"), bins, labels = FALSE)
  ab <- ave(A / E, eb, FUN = function(z) cut(rank(z, ties.method = "first"), bins, labels = FALSE))
  (eb - 1) * bins + ab
}
permute_within <- function(cells) {
  n <- length(cells); p <- integer(n)
  p[order(cells, seq_len(n))] <- order(cells, runif(n))
  p
}
robust_fit <- function(y, W) {
  XtXi <- solve(crossprod(W))
  b <- drop(XtXi %*% crossprod(W, y))
  e <- drop(y - W %*% b)
  V <- XtXi %*% crossprod(W * e) %*% XtXi * nrow(W) / (nrow(W) - ncol(W))
  data.frame(term = colnames(W), estimate = b, se = sqrt(diag(V)), row.names = NULL)
}
# Coefficient on the first column of a block, everything else held fixed (FWL).
perm_block <- function(block, others, models, cells, B) {
  prep <- lapply(models, function(md) {
    q <- qr(others[md$rows, , drop = FALSE]); list(q = q, ry = qr.resid(q, md$y))
  })
  coef_for <- function(Bf) sapply(names(models), function(nm) {
    R <- qr.resid(prep[[nm]]$q, Bf[models[[nm]]$rows, , drop = FALSE])
    solve(crossprod(R), crossprod(R, prep[[nm]]$ry))[1]
  })
  list(observed = coef_for(block),
       perms = do.call(rbind, lapply(seq_len(B), function(i)
         coef_for(block[permute_within(cells), , drop = FALSE]))))
}
```

```{r test-a}
a <- v[year %in% 2022:2023]
blocks <- list(
  program = cbind(log_program = log1p(a$prog), any_program = as.numeric(a$prog > 0)),
  contributions = cbind(log_contributions = log1p(a$contr), any_contributions = as.numeric(a$contr > 0)))
sE <- ns(log(a$E), df = 4); colnames(sE) <- paste0("size_E", 1:4)
sA <- ns(log1p(a$A), df = 4); colnames(sA) <- paste0("size_A", 1:4)
Z <- cbind(intercept = 1, sE, sA, year2023 = as.numeric(a$year == 2023))
W <- cbind(do.call(cbind, blocks), Z)

models <- list(any_fundraising = list(rows = rep(TRUE, nrow(a)), y = as.numeric(a$fundr > 0)),
               log_fundraising = list(rows = a$fundr > 0, y = log(a$fundr[a$fundr > 0])))
coefs <- rbindlist(lapply(names(models), function(nm) {
  f <- robust_fit(models[[nm]]$y, W[models[[nm]]$rows, , drop = FALSE])
  data.table(model = nm, n = sum(models[[nm]]$rows),
             f[f$term %in% c("log_program", "log_contributions"), ])
}))

cells <- size_cells(a$E, a$A)
pp <- perm_block(blocks$program, cbind(blocks$contributions, Z), models, cells, params$perms)
pc <- perm_block(blocks$contributions, cbind(blocks$program, Z), models, cells, params$perms)
qs <- function(m, nm) quantile(m[, nm], c(0.005, 0.995))
coefs[, perm_lo := c(qs(pp$perms, "any_fundraising")[1], qs(pc$perms, "any_fundraising")[1],
                     qs(pp$perms, "log_fundraising")[1], qs(pc$perms, "log_fundraising")[1])]
coefs[, perm_hi := c(qs(pp$perms, "any_fundraising")[2], qs(pc$perms, "any_fundraising")[2],
                     qs(pp$perms, "log_fundraising")[2], qs(pc$perms, "log_fundraising")[2])]
coefs[, meets_rule := ifelse(term == "log_program", estimate < 0 & estimate < perm_lo,
                             estimate > 0 & estimate > perm_hi)]
kable(coefs, digits = 4, caption = "Test A: program revenue and contributions in both parts")

verdict_a <- if (all(coefs[term == "log_program", meets_rule])) "supported" else
             if (any(coefs[term == "log_program", estimate] >= 0)) "rejected" else "inconclusive"
```

The pre-specified verdict for H4 is **`r verdict_a`**.

**Within organizations.** The cross-section compares different organizations.
The panel also asks whether one organization's fundraising moves when its own
earned revenue moves, using 2019, 2022, and 2023 and removing each
organization's mean.

```{r fixed-effects}
fe <- v[year %in% c(2019, 2022, 2023)]
dm <- function(z, g) z - ave(z, g, FUN = mean)
g <- fe$EIN2
Wfe <- cbind(log_program = dm(log1p(fe$prog), g), log_contributions = dm(log1p(fe$contr), g),
             log_expenses = dm(log(fe$E), g), log_assets = dm(log1p(fe$A), g),
             y2022 = dm(as.numeric(fe$year == 2022), g), y2023 = dm(as.numeric(fe$year == 2023), g))
fe_fit <- robust_fit(dm(log1p(fe$fundr), g), Wfe)
kable(fe_fit[fe_fit$term %in% c("log_program", "log_contributions"), ], digits = 4,
      caption = "Within-organization (2019, 2022, 2023)")
```

# 4. Test B: the COVID earned-revenue shock (exploratory)

If the two are substitutes, an organization that lost earned revenue in the
pandemic should have leaned harder on fundraising afterwards. The shock is the
change in program revenue from 2019 to 2021; the response is the change in
fundraising expense from 2019 to 2023. Both are log changes. Subsector fixed
effects absorb the fact that the shock hit arts, education, and health
differently, and the 2019 levels of fundraising and contributions allow for
mean reversion.

```{r shock}
w <- dcast(v[year %in% c(2019, 2021, 2023)], EIN2 + subsector ~ year,
           value.var = c("fundr", "prog", "contr", "E", "A"))
setnames(w, gsub("_(\\d{4})$", "\\1", names(w)))
# Organizations that moved their fiscal year end drop out of the December
# filter in some years, leaving gaps in the reshape; those rows cannot be used.
need <- setdiff(names(w), c("EIN2", "subsector"))
complete <- complete.cases(w[, ..need])
cat(format(sum(!complete), big.mark = ","), "of", format(nrow(w), big.mark = ","),
    "organizations dropped for a missing December filing in 2019, 2021, or 2023\n")
w <- w[complete][prog2019 > 0 & E2019 > 0 & A2019 > 0]
w[, `:=`(shock = log1p(prog2021) - log1p(prog2019),
         d_fundraising = log1p(fundr2023) - log1p(fundr2019),
         d_contributions = log1p(contr2023) - log1p(contr2019))]
kable(w[, .(organizations = .N, median_shock = median(shock),
            pct_lost_earned_revenue = mean(shock < 0),
            median_change_fundraising = median(d_fundraising))], digits = 3,
      format.args = list(big.mark = ","))

sub <- model.matrix(~ subsector, data = as.data.frame(w))[, -1, drop = FALSE]
Zs <- cbind(intercept = 1, ns(log(w$E2019), df = 4), ns(log1p(w$A2019), df = 4),
            log_fundraising_2019 = log1p(w$fundr2019), any_fundraising_2019 = as.numeric(w$fundr2019 > 0),
            log_contributions_2019 = log1p(w$contr2019), sub)
colnames(Zs) <- make.unique(colnames(Zs))
shock_fit <- function(y) {
  f <- robust_fit(y, cbind(shock = w$shock, Zs))
  f[f$term == "shock", ]
}
shock_res <- rbind(
  cbind(outcome = "change in fundraising expense, 2019-2023", shock_fit(w$d_fundraising)),
  cbind(outcome = "change in contributions, 2019-2023", shock_fit(w$d_contributions)))
shock_res$per_sd <- shock_res$estimate * sd(w$shock)
kable(shock_res, digits = 4,
      caption = "Exploratory: response to the 2019-2021 earned-revenue shock")

# Extensive margin: did organizations with no fundraising in 2019 start by 2023?
started <- w[fundr2019 == 0]
start_fit <- robust_fit(as.numeric(started$fundr2023 > 0),
                        cbind(shock = started$shock,
                              intercept = 1, ns(log(started$E2019), df = 4),
                              model.matrix(~ subsector, data = as.data.frame(started))[, -1, drop = FALSE]))
kable(start_fit[start_fit$term == "shock", ], digits = 4,
      caption = paste0("Started fundraising by 2023, among the ",
                       format(nrow(started), big.mark = ","), " with none in 2019"))
```

# 5. Findings

```{r findings, include = FALSE}
cp <- coefs[term == "log_program"]
cc <- coefs[term == "log_contributions"]
sh <- shock_res[1, ]
```

* **Cross-section.** The program-revenue coefficient is
  `r fmt(cp$estimate[1], 4)` for having any fundraising expense and
  `r fmt(cp$estimate[2], 4)` for the amount, against shuffle bounds of
  `r fmt(cp$perm_lo[1], 4)` and `r fmt(cp$perm_lo[2], 4)`. The contributions
  coefficients are `r fmt(cc$estimate[1], 4)` and `r fmt(cc$estimate[2], 4)`.
  The pre-specified verdict is **`r verdict_a`**.
* **Within organizations.** A 1-unit change in log program revenue goes with a
  `r fmt(fe_fit$estimate[fe_fit$term == "log_program"], 4)` change in log
  fundraising expense.
* **The shock.** A one-standard-deviation loss of earned revenue between 2019
  and 2021 is followed by a `r fmt(-sh$per_sd, 3)` change in log fundraising
  expense by 2023 (coefficient `r fmt(sh$estimate, 4)`, SE
  `r fmt(sh$se, 4)`); a negative coefficient means substitution.

```{r save, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(coefs = coefs, verdict_a = verdict_a, fe = fe_fit, shock = shock_res,
               start = start_fit, n_shock = nrow(w), n_started = nrow(started),
               years = v[, .N, by = year][order(year)]), params$results_file)
}
```

```{r session}
sessionInfo()
```
