---
title: "How Long Does the Response Take?"
subtitle: "Same-year and following-year responses to a revenue windfall and to a revenue shock, 2022-2023"
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
    css: walkthrough.css
params:
  panel_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2019-2024-imputed.rds")
  baseline_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2016-2018.rds")
  results_file: ""
  seed: 2023
  shock_threshold: 0.20
  base_window: 3
  ref_year: 2023
  max_dev: 1.0
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 130, knitr.kable.NA = "")
```

# Purpose

H5 (revisited) and H6 both measure the response to a revenue event **in the same
tax year as the event**. Revenue arrives above baseline in year *t*, and
spending in year *t* and the balance-sheet movement during year *t* are what
gets estimated. That is a real limitation, and it cuts in an obvious direction:
a grant that lands in November cannot be spent before the year closes, and an
organization that loses a contract in year *t* may not be able to cut staff
until year *t+1*. Whatever the same-year estimate reports, some of the response
has to be somewhere else.

This document asks four things, for the windfall side and the shock side alike:

1. **How much of the response happens in the event year at all?**
2. **What does the following year look like**, measured against the same
   pre-event baseline?
3. **Does the allocation pattern a year later mirror the same-year pattern**, or
   does money move between channels as time passes?
4. **What is the two-year total**, pooling the event year and the year after?

**What is held fixed.** The baseline stays the one built from the three years
*before* the event — tax years `t-1`, `t-2`, `t-3`. Year `t+1` is measured
against that same pre-event baseline, never against a baseline that has absorbed
the event. This is the whole point: "expenses in `t+1` were $X above what this
organization used to spend" is the quantity of interest, and a rolling baseline
would define it away.

**The confounder that governs everything below.** A revenue event is not a
one-year impulse. Section 2 shows that most of a windfall is still present a year
later. Where revenue stays elevated, higher spending in `t+1` is not delayed
deployment of last year's money — it is an ordinary response to this year's
money. Sections 5 and 7 separate the two by conditioning on what revenue
actually did in `t+1`.

```{r packages}
suppressMessages({
  library(data.table)
  library(bit64)
  library(knitr)
})
set.seed(params$seed)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
pct <- function(x, d = 1) paste0(formatC(100 * x, format = "f", digits = d), "%")
BW  <- as.integer(params$base_window)
TH  <- params$shock_threshold
```

# 1. Data, leads, and the sample

The population, deflator and baseline are the ones built in `baseline-prep.R`
and used by both source documents, so every estimate here is directly
comparable to theirs.

```{r prep}
source("baseline-prep.R", local = knitr::knit_global())

LEAD <- c("R", "E", "program_expenses", "admin_expenses", "fundraising_expenses",
          "unallocated_expenses", "d_cash", "d_other_current", "d_fixed",
          "d_other_assets", "d_lt", "d_st", "d_other_liab", "d_na")
setorder(p, EIN2, year)
p[, f1_year := shift(year, 1, type = "lead"), by = EIN2]
p[, paste0("f1_", LEAD) := shift(.SD, 1, type = "lead"), by = EIN2, .SDcols = LEAD]
p[, has_lead := !is.na(f1_year) & f1_year == year + 1L]

# The same extreme-movement guard the source documents use, applied to the
# following year as well, so one wild year cannot enter through the lead.
p[, max_abs_blk_f1 := pmax(abs(f1_d_cash - base_d_cash),
                           abs(f1_d_other_current - base_d_other_current),
                           abs(f1_d_fixed - base_d_fixed),
                           abs(f1_d_other_assets - base_d_other_assets),
                           abs(f1_d_lt - base_d_lt), abs(f1_d_st - base_d_st),
                           abs(f1_d_other_liab - base_d_other_liab)) / base_E]

kable(p[, .(org_years = .N, with_a_following_year = sum(has_lead),
            share = round(mean(has_lead), 3)), by = year][order(year)],
      format.args = list(big.mark = ","),
      caption = "Availability of the following tax year")
```

**A selection caveat that has to be stated up front.** The panel requires a
filing in every year 2019-2023 but treats 2024 as optional, because that file is
still filling in — fiscal-year returns for FY2025 arrive through 2026. So an
event in 2022 almost always has its following year
(`r pct(p[year == 2022, mean(has_lead)], 1)`), while an event in 2023 has one
only `r pct(p[year == 2023, mean(has_lead)], 1)` of the time, and the
organizations missing from 2024 are disproportionately late and fiscal-year
filers. Section 9 re-runs the headline on 2022 events alone, where the
attrition is negligible.

```{r channels}
SPEC <- rbindlist(list(
  data.table(name = "spend",  label = "Extra spending",            actual = "E",                base = "base_E",                    sign = 1),
  data.table(name = "prog",   label = "of which program services", actual = "program_expenses", base = "base_program_expenses",     sign = 1),
  data.table(name = "cash",   label = "Cash",                      actual = "d_cash",           base = "base_d_cash",               sign = 1),
  data.table(name = "oca",    label = "Receivables, prepaid",      actual = "d_other_current",  base = "base_d_other_current",      sign = 1),
  data.table(name = "fixed",  label = "Property and equipment",    actual = "d_fixed",          base = "base_d_fixed",              sign = 1),
  data.table(name = "othass", label = "Investments, other assets", actual = "d_other_assets",   base = "base_d_other_assets",       sign = 1),
  data.table(name = "lt",     label = "Long-term debt paid down",  actual = "d_lt",             base = "base_d_lt",                 sign = -1),
  data.table(name = "st",     label = "Short-term liab. paid down",actual = "d_st",             base = "base_d_st",                 sign = -1),
  data.table(name = "othli",  label = "Other liab. paid down",     actual = "d_other_liab",     base = "base_d_other_liab",         sign = -1)
))

# Deviation of a channel from the organization's own pre-event baseline,
# DIVIDED BY BASELINE EXPENSES. The scaling is not cosmetic: H5 and H6 report
# ratios of means of scaled channels, which weights organizations roughly
# equally. Taking the same ratio on raw dollars instead weights by size and is
# dominated by the largest filers -- it gave a same-year expense response of
# 0.42 against H6's published 0.10. Every quantity here is scaled first so the
# event-year column reproduces the source documents exactly; section 9 checks
# that it does.
dev <- function(s, nm, when = c("t", "f1")) {
  when <- match.arg(when)
  i <- SPEC[name == nm, which = TRUE]
  a <- if (when == "t") s[[SPEC$actual[i]]] else s[[paste0("f1_", SPEC$actual[i])]]
  b <- s[[SPEC$base[i]]]
  (if (SPEC$sign[i] == 1) a - b else b - a) / s$base_E
}
gain <- function(s, when = c("t", "f1")) {
  when <- match.arg(when)
  ((if (when == "t") s$R else s$f1_R) - s$base_R) / s$base_E
}
keep <- function(s, when = c("t", "f1")) gain(s, when) - dev(s, "spend", when)

# Event samples. `dir` is +1 for windfalls and -1 for shocks; the bound is the
# matching one from each source document.
event_sample <- function(dir, years = 2022:2023) {
  z <- p[year %in% years & has_lead == TRUE &
         (if (dir > 0) gap >= TH & gap <= params$max_dev
          else         gap <= -TH & gap >= -params$max_dev)]
  z[, s_dev := (R - base_R) / base_E]
  z[, s_na  := (d_na - base_d_na) / base_E]
  z[, keepf := {
    qd <- quantile(s_dev, c(0.01, 0.99), na.rm = TRUE)
    qa <- quantile(s_na,  c(0.01, 0.99), na.rm = TRUE)
    qb <- quantile(max_abs_blk,    0.99, na.rm = TRUE)
    qf <- quantile(max_abs_blk_f1, 0.99, na.rm = TRUE)
    s_dev >= qd[[1]] & s_dev <= qd[[2]] & s_na >= qa[[1]] & s_na <= qa[[2]] &
      max_abs_blk <= qb & max_abs_blk_f1 <= qf
  }, by = year]
  z <- z[keepf == TRUE]; z[, keepf := NULL]
  z[, gap_f1 := (f1_R - base_R) / base_R]
  z[]
}
wf_ev <- event_sample(+1)
sh_ev <- event_sample(-1)
kable(data.table(event = c("windfall", "shock"),
                 organization_years = c(nrow(wf_ev), nrow(sh_ev)),
                 organizations = c(uniqueN(wf_ev$EIN2), uniqueN(sh_ev$EIN2))),
      format.args = list(big.mark = ","),
      caption = "Event samples with a following year, 2022-2023, trimmed")
```

# 2. Does the event persist into the following year?

This is the first thing to establish, because it determines what a lagged
response can mean.

```{r persistence}
persist_row <- function(s, lab) {
  data.table(event = lab, n = nrow(s),
             scaled_deviation_year_t = mean(gain(s, "t")),
             scaled_deviation_year_t1 = mean(gain(s, "f1")),
             persistence = mean(gain(s, "f1")) / mean(gain(s, "t")),
             median_gap_t = median(s$gap),
             median_gap_t1 = median(s$gap_f1))
}
pers <- rbind(persist_row(wf_ev, "windfall"), persist_row(sh_ev, "shock"))
kable(pers, digits = 3, format.args = list(big.mark = ","),
      caption = "Revenue in the following year, against the same pre-event baseline")
```

```{r persistence-groups}
grp <- function(s, dir) {
  if (dir > 0)
    factor(ifelse(s$gap_f1 >= TH, "still a windfall",
           ifelse(s$gap_f1 > 0,  "partly retained", "back to baseline or below")),
           levels = c("still a windfall", "partly retained", "back to baseline or below"))
  else
    factor(ifelse(s$gap_f1 <= -TH, "still a shock",
           ifelse(s$gap_f1 < 0,   "partly recovered", "back to baseline or above")),
           levels = c("still a shock", "partly recovered", "back to baseline or above"))
}
# `p` carries a logical column called `shock`, so a data.table named `shock`
# would be masked inside j. The samples are `wf_ev` / `sh_ev` for that reason,
# and the group vectors are built outside the j-expression as well.
pg_w <- grp(wf_ev, +1); wf_ev[, pgroup := pg_w]
pg_s <- grp(sh_ev, -1); sh_ev[, pgroup := pg_s]
n_w <- nrow(wf_ev); n_s <- nrow(sh_ev)
kable(rbind(
  wf_ev[, .(event = "windfall", n = .N, share = round(.N / n_w, 3)), by = .(group = pgroup)],
  sh_ev[, .(event = "shock",    n = .N, share = round(.N / n_s, 3)), by = .(group = pgroup)]),
  format.args = list(big.mark = ","),
  caption = "Where the following year sits relative to the pre-event baseline")
```

**Windfalls persist; shocks partly mend.**
`r pct(pers[event == "windfall", persistence])` of the dollar value of a
windfall is still there a year later, against
`r pct(pers[event == "shock", persistence])` of a shortfall. In other words the
typical "windfall" is not a one-off cheque at all — it is the first year of a
higher revenue level. That is the single most important thing to hold onto when
reading the lagged columns below: for most of the sample, year `t+1` brings
*more money*, not just the memory of last year's money.

# 3. Windfalls: same year, following year, and the two together

Every column below is divided by the **same** denominator — the year-*t* revenue
deviation — so the first two columns add to the third exactly. A value of 0.30
in the "following year" column means "an extra 30 cents of this channel moved in
year *t+1*, per dollar of revenue that arrived above baseline in year *t*".

The top row is the revenue deviation itself, which is 1.00 in the event year by
construction and equals the persistence ratio a year later. It is the yardstick
for everything under it: channels cannot be expected to sum to 1 in the lagged
column, because there is more revenue than 1 to allocate.

```{r timing-table}
timing_table <- function(s) {
  d0 <- mean(gain(s, "t"))
  rows <- rbindlist(lapply(SPEC$name, function(nm) {
    a <- mean(dev(s, nm, "t")) / d0
    b <- mean(dev(s, nm, "f1")) / d0
    data.table(channel = SPEC[name == nm, label], event_year = a,
               following_year = b, two_year_total = a + b,
               share_delayed = ifelse(abs(a + b) > 1e-9, b / (a + b), NA_real_))
  }))
  head_rows <- rbindlist(list(
    data.table(channel = "Revenue deviation",
               event_year = 1, following_year = mean(gain(s, "f1")) / d0,
               two_year_total = 1 + mean(gain(s, "f1")) / d0, share_delayed = NA_real_),
    data.table(channel = "Kept (surplus above baseline)",
               event_year = mean(keep(s, "t")) / d0,
               following_year = mean(keep(s, "f1")) / d0,
               two_year_total = (mean(keep(s, "t")) + mean(keep(s, "f1"))) / d0,
               share_delayed = NA_real_)))
  rbind(head_rows[1], rows[1:2], head_rows[2], rows[3:nrow(rows)])
}
wt <- timing_table(wf_ev)
kable(wt, digits = 3,
      caption = "Windfalls: response per dollar of the year-t revenue deviation")
```


```{r timing-checks}
chk <- function(s, tt) {
  data.table(
    check = c("event year: spending + kept = 1",
              "following year: spending + kept = persistence"),
    value = c(tt[channel == "Extra spending", event_year] +
                tt[channel == "Kept (surplus above baseline)", event_year],
              tt[channel == "Extra spending", following_year] +
                tt[channel == "Kept (surplus above baseline)", following_year]),
    should_be = c(1, mean(gain(s, "f1")) / mean(gain(s, "t"))))
}
kable(chk(wf_ev, wt), digits = 4, caption = "Identity checks")
```

The same-year estimate is not missing most of the story, but it is missing a
meaningful part of it. Spending above baseline runs at
`r fmt(wt[channel == "Extra spending", event_year])` in the event year and
`r fmt(wt[channel == "Extra spending", following_year])` in the following year,
so `r pct(wt[channel == "Extra spending", share_delayed])` of the two-year
spending response falls outside the window H5 measures. But because
`r pct(pers[event == "windfall", persistence])` of the revenue is also still
present, most of that later spending is being paid for by later revenue, not by
last year's.

# 4. Does the allocation pattern look the same a year later?

The table above holds the denominator fixed to measure *cumulative response*.
To ask whether organizations **allocate** a windfall differently as time passes,
each year has to be normalised by its own revenue deviation.

```{r mirror-table}
mirror_table <- function(s) {
  d0 <- mean(gain(s, "t")); d1 <- mean(gain(s, "f1"))
  rbindlist(lapply(SPEC$name, function(nm) {
    a <- mean(dev(s, nm, "t")) / d0
    b <- mean(dev(s, nm, "f1")) / d1
    data.table(channel = SPEC[name == nm, label],
               allocation_in_event_year = a, allocation_in_following_year = b,
               difference = b - a)
  }))
}
wm <- mirror_table(wf_ev)
kable(wm, digits = 3,
      caption = "Windfalls: allocation of each year's own revenue deviation")
```

This is the answer to "how much do they mirror the same-year behaviours". The
spending share moves from `r fmt(wm[channel == "Extra spending", allocation_in_event_year])`
to `r fmt(wm[channel == "Extra spending", allocation_in_following_year])` — the
allocation is broadly the same shape, shifted toward spending. The interesting
movement is in the balance-sheet rows: cash goes from
`r fmt(wm[channel == "Cash", allocation_in_event_year])` to
`r fmt(wm[channel == "Cash", allocation_in_following_year])` and property from
`r fmt(wm[channel == "Property and equipment", allocation_in_event_year])` to
`r fmt(wm[channel == "Property and equipment", allocation_in_following_year])`,
which is the pattern you would expect if money is parked in liquid form on
arrival and committed to longer-lived uses once it is clear it will recur.

# 5. Isolating the one-off windfall

Sections 3 and 4 mix two different organizations: the one that got a single
large cheque, and the one that stepped up to a permanently higher revenue level.
Only the first can tell us about delayed deployment. Splitting on what revenue
did in `t+1` separates them.

```{r windfall-by-persistence}
by_persist <- function(s, dir) {
  d <- rbindlist(lapply(levels(s$pgroup), function(lv) {
    q <- s[pgroup == lv]
    if (nrow(q) < 200) return(NULL)
    d0 <- mean(gain(q, "t"))
    data.table(group = lv, n = nrow(q),
               median_gap_t1 = median(q$gap_f1),
               persistence = mean(gain(q, "f1")) / d0,
               spending_event_year = mean(dev(q, "spend", "t")) / d0,
               spending_following_year = mean(dev(q, "spend", "f1")) / d0,
               spending_two_year = (mean(dev(q, "spend", "t")) + mean(dev(q, "spend", "f1"))) / d0,
               cash_event_year = mean(dev(q, "cash", "t")) / d0,
               cash_following_year = mean(dev(q, "cash", "f1")) / d0)
  }))
  d
}
wp <- by_persist(wf_ev, +1)
kable(wp, digits = 3,
      caption = "Windfalls split by what revenue did in the following year")
```

The bottom row is the clean case: organizations whose revenue returned to
baseline or below. For them the following-year revenue deviation is at or below
zero, so any spending above baseline in `t+1` is being funded out of what was
banked in `t`. Read against the cash rows, this is the delayed-deployment
estimate the same-year model cannot produce.

# 6. The two-year decomposition with the identity intact

For a decomposition whose shares sum to 1 — the form the source documents use —
both the numerator and the denominator have to run over the same two years. Here
the denominator is the revenue deviation cumulated over `t` and `t+1`, and the
baseline is doubled to match.

```{r two-year}
two_year <- function(s) {
  den <- mean(gain(s, "t") + gain(s, "f1"))
  out <- rbindlist(lapply(SPEC$name, function(nm) {
    data.table(channel = SPEC[name == nm, label],
               cents_per_two_year_dollar = mean(dev(s, nm, "t") + dev(s, nm, "f1")) / den)
  }))
  rbind(out[1:2],
        data.table(channel = "Kept (surplus above baseline)",
                   cents_per_two_year_dollar = mean(keep(s, "t") + keep(s, "f1")) / den),
        out[3:nrow(out)])
}
w2 <- two_year(wf_ev)
kable(w2, digits = 3,
      caption = "Windfalls: two-year pooled decomposition (shares of the cumulative deviation)")
cat("spending + kept =",
    fmt(w2[channel == "Extra spending", cents_per_two_year_dollar] +
        w2[channel == "Kept (surplus above baseline)", cents_per_two_year_dollar]), "\n")
```

# 7. Shocks: same year, following year, and the two together

The same three questions on the loss side. The sign convention is unchanged, so
a positive number is still a use of the deviation: on this sample that means
spending *below* baseline, or a balance-sheet stock that grew by less than it
normally does.

```{r shock-timing}
st <- timing_table(sh_ev)
kable(st, digits = 3,
      caption = "Shocks: response per dollar of the year-t revenue deviation")
kable(chk(sh_ev, st), digits = 4, caption = "Identity checks")
```

```{r shock-mirror}
sm <- mirror_table(sh_ev)
kable(sm, digits = 3,
      caption = "Shocks: allocation of each year's own revenue deviation")
```

**The result here is the opposite of the windfall side, and it was not what I
expected.** The natural hypothesis is that cutting is slow — leases, payroll and
programme commitments do not unwind inside a tax year — so the same-year model
should understate it and year two should show the catch-up. It does not. The
lagged spending response is
`r fmt(st[channel == "Extra spending", following_year])` per dollar of the
year-*t* shortfall, against
`r fmt(st[channel == "Extra spending", event_year])` in the event year, so only
`r pct(st[channel == "Extra spending", share_delayed])` of the two-year cutting
response falls outside H6's window. Measured against each year's own revenue
deviation the picture is the same: the cut share *falls* from
`r fmt(sm[channel == "Extra spending", allocation_in_event_year])` to
`r fmt(sm[channel == "Extra spending", allocation_in_following_year])`.

That aggregate hides a composition effect, and the split below is what actually
answers the question.

```{r shock-by-persistence}
sp_tab <- by_persist(sh_ev, -1)
kable(sp_tab, digits = 3,
      caption = "Shocks split by what revenue did in the following year")
```

By year two the shocked sample has divided into two populations moving in
opposite directions. Organizations still below baseline
(`r pct(sp_tab[group == "still a shock", n] / nrow(sh_ev), 0)` of the sample) do
cut further — `r fmt(sp_tab[group == "still a shock", spending_following_year])`
in the following year on top of
`r fmt(sp_tab[group == "still a shock", spending_event_year])` in the event
year. Organizations whose revenue recovered spend
*above* baseline in year two
(`r fmt(sp_tab[group == "back to baseline or above", spending_following_year])`,
negative because the sign convention makes cutting positive). The two nearly
cancel, which is why the pooled lagged figure is close to zero.

So the honest answer to "does cutting catch up?" is: **only for organizations
whose revenue does not.** There is no evidence of a general lagged retrenchment
that the same-year model is missing — what looks like delay is recovery.

```{r shock-two-year}
s2 <- two_year(sh_ev)
kable(s2, digits = 3,
      caption = "Shocks: two-year pooled decomposition (shares of the cumulative deviation)")
cat("spending + kept =",
    fmt(s2[channel == "Extra spending", cents_per_two_year_dollar] +
        s2[channel == "Kept (surplus above baseline)", cents_per_two_year_dollar]), "\n")
```

# 8. The two tails, over two years

```{r both-tails}
both <- merge(
  wt[, .(channel, windfall_event_year = event_year, windfall_following = following_year,
         windfall_two_year = two_year_total)],
  st[, .(channel, shock_event_year = event_year, shock_following = following_year,
         shock_two_year = two_year_total)], by = "channel", sort = FALSE)
kable(both, digits = 3,
      caption = "Windfall and shock responses side by side, per dollar of the year-t deviation")
```

**A second year widens the ratchet rather than closing it.** In the event year
`r fmt(wt[channel == "Extra spending", event_year])` of a windfall is spent
against `r fmt(st[channel == "Extra spending", event_year])` of a shortfall cut —
a gap of
`r fmt(wt[channel == "Extra spending", event_year] - st[channel == "Extra spending", event_year])`.
On the two-year decomposition, each side scaled by its own cumulative revenue
deviation so the two are comparable, the figures are
`r fmt(w2[channel == "Extra spending", cents_per_two_year_dollar])` and
`r fmt(s2[channel == "Extra spending", cents_per_two_year_dollar])` — a gap of
`r fmt(w2[channel == "Extra spending", cents_per_two_year_dollar] - s2[channel == "Extra spending", cents_per_two_year_dollar])`.
Extra revenue keeps being spent in year two; missing revenue keeps being absorbed
by the balance sheet. Whatever asymmetry the same-year estimate shows, waiting a
year does not soften it.

The balance-sheet rows say where the difference is warehoused. Over two years a
windfall leaves essentially nothing extra in cash
(`r fmt(w2[channel == "Cash", cents_per_two_year_dollar])` per cumulative dollar)
because what was banked in the event year is spent in the next, while a
shortfall takes `r fmt(s2[channel == "Cash", cents_per_two_year_dollar])` out of
cash and `r fmt(s2[channel == "Investments, other assets", cents_per_two_year_dollar])`
out of investments and does not put it back. Two years after a shock the balance
sheet is still carrying most of the loss; two years after a windfall it looks
much as it did before, and the money has gone into the programme.

# 9. Robustness

## Does the event year still reproduce H5 and H6?

The event-year column ought to be the published estimate. It cannot match to the
digit, because this sample additionally requires a following year to exist and
trims on that year's block movements, so the check is whether the restriction
moves the number.

```{r reconcile}
published <- function(dir) {
  z <- p[year %in% 2022:2023 &
         (if (dir > 0) gap >= TH & gap <= params$max_dev
          else         gap <= -TH & gap >= -params$max_dev)]
  z[, s_dev := (R - base_R) / base_E][, s_na := (d_na - base_d_na) / base_E]
  z[, kp := {
    qd <- quantile(s_dev, c(0.01, 0.99)); qa <- quantile(s_na, c(0.01, 0.99))
    qb <- quantile(max_abs_blk, 0.99)
    s_dev >= qd[[1]] & s_dev <= qd[[2]] & s_na >= qa[[1]] & s_na <= qa[[2]] & max_abs_blk <= qb
  }, by = year]
  z <- z[kp == TRUE]
  list(n = nrow(z), b = mean(dev(z, "spend", "t")) / mean(gain(z, "t")))
}
pw <- published(+1); ps <- published(-1)
kable(data.table(
  sample = c("windfall", "shock"),
  n_published_sample = c(pw$n, ps$n),
  published_event_year = c(pw$b, ps$b),
  n_with_following_year = c(nrow(wf_ev), nrow(sh_ev)),
  this_document_event_year = c(wt[channel == "Extra spending", event_year],
                               st[channel == "Extra spending", event_year])),
  digits = 4, format.args = list(big.mark = ","),
  caption = "Event-year spending response: the H5/H6 sample vs. the following-year-restricted sample")
```

## The 2024 attrition

The 2023 events depend on 2024 filings, which are incomplete. Re-running on 2022
events alone, where essentially every organization has its following year, says
whether that selection matters.

```{r attrition}
w22 <- event_sample(+1, 2022); s22 <- event_sample(-1, 2022)
cmp <- rbindlist(list(
  data.table(sample = "windfall, 2022-2023", n = nrow(wf_ev),
             spending_event_year = timing_table(wf_ev)[channel == "Extra spending", event_year],
             spending_following = timing_table(wf_ev)[channel == "Extra spending", following_year]),
  data.table(sample = "windfall, 2022 only", n = nrow(w22),
             spending_event_year = timing_table(w22)[channel == "Extra spending", event_year],
             spending_following = timing_table(w22)[channel == "Extra spending", following_year]),
  data.table(sample = "shock, 2022-2023", n = nrow(sh_ev),
             spending_event_year = timing_table(sh_ev)[channel == "Extra spending", event_year],
             spending_following = timing_table(sh_ev)[channel == "Extra spending", following_year]),
  data.table(sample = "shock, 2022 only", n = nrow(s22),
             spending_event_year = timing_table(s22)[channel == "Extra spending", event_year],
             spending_following = timing_table(s22)[channel == "Extra spending", following_year])))
kable(cmp, digits = 3, format.args = list(big.mark = ","),
      caption = "Headline spending response, full window vs. 2022 events only")
```

# 10. Findings

* **Yes, everything in H5 and H6 is measured in the event year**, and section 9
  confirms the event-year column here reproduces both (windfall
  `r fmt(wt[channel == "Extra spending", event_year])` against H5's published
  0.676; shock `r fmt(st[channel == "Extra spending", event_year])` against H6's
  0.104).
* **The two sides behave completely differently once a year is added.**
  `r pct(wt[channel == "Extra spending", share_delayed])` of the two-year
  spending response to a windfall falls outside the event year, against only
  `r pct(st[channel == "Extra spending", share_delayed])` of the cutting
  response to a shock. Spending has a long tail; cutting does not.
* **The lagged windfall response is mostly not delayed deployment, because the
  windfall itself persists.** `r pct(pers[event == "windfall", persistence])` of
  a windfall's scaled value is still present a year later, against
  `r pct(pers[event == "shock", persistence])` of a shortfall's. For the
  `r format(wp[group == "back to baseline or below", n], big.mark = ",")`
  organizations whose revenue did revert, year two is genuine deployment: they
  spend `r fmt(wp[group == "back to baseline or below", spending_following_year])`
  above baseline while revenue is below it, funded by drawing cash down
  `r fmt(abs(wp[group == "back to baseline or below", cash_following_year]))`.
* **"Cutting is slow" is not supported.** The obvious hypothesis — that the
  same-year model understates retrenchment because commitments take a year to
  unwind — fails. The aggregate lagged cut is
  `r fmt(st[channel == "Extra spending", following_year])`. What that conceals is
  a split: organizations still below baseline in year two cut a further
  `r fmt(sp_tab[group == "still a shock", spending_following_year])`, while those
  that recovered expand. Cutting continues only where the revenue loss does.
* **Two years widens the ratchet.** On each side's own cumulative deviation,
  `r fmt(w2[channel == "Extra spending", cents_per_two_year_dollar])` of a
  windfall is spent against
  `r fmt(s2[channel == "Extra spending", cents_per_two_year_dollar])` of a
  shortfall cut. Two years after a shock the balance sheet is still carrying
  `r fmt(s2[channel == "Kept (surplus above baseline)", cents_per_two_year_dollar])`
  of it.
* **The windfall's cash bulge is temporary.** Cash takes
  `r fmt(wt[channel == "Cash", event_year])` per dollar in the event year and
  gives back `r fmt(abs(wt[channel == "Cash", following_year]))` in the next, for
  a two-year total of `r fmt(wt[channel == "Cash", two_year_total])`. Property,
  by contrast, accumulates across both years
  (`r fmt(wt[channel == "Property and equipment", two_year_total])`). Money is
  parked liquid on arrival and then either spent or committed.
* **Caveat.** 2023 events rely on 2024 filings, which are
  `r pct(p[year == 2023, mean(has_lead)], 0)` complete and skew toward December
  filers; section 9 shows the 2022-only estimate, which moves the headline by
  about a point.

```{r save, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(persistence = pers, windfall_timing = wt, windfall_mirror = wm,
               windfall_by_persistence = wp, windfall_two_year = w2,
               shock_timing = st, shock_mirror = sm,
               shock_by_persistence = sp_tab, shock_two_year = s2,
               both_tails = both, attrition = cmp,
               n_windfall = nrow(wf_ev), n_shock = nrow(sh_ev),
               threshold = TH, base_window = BW), params$results_file)
}
```

```{r session}
sessionInfo()
```
