---
title: "H5 (revisited): Where Does a Windfall Go?"
subtitle: "Allocation of revenue that arrives above a three-year inflation-adjusted baseline, 2019-2024"
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
    css: walkthrough.css
params:
  panel_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2019-2024-imputed.rds")
  baseline_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2016-2018.rds")
  results_file: ""
  seed: 2023
  shock_threshold: 0.20
  base_window: 3
  ref_year: 2023
  max_gain: 1.0
  comp_share: 0.3333333
  comp_material: 0.10
  conc_high: 0.90
  conc_low: 0.60
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 130, knitr.kable.NA = "")
```

# Purpose

The original H5 (`h5-surplus-allocation.Rmd`) asked where a **surplus** goes: it
regressed each balance-sheet block's movement on the level of the surplus and
read the coefficients as cents per surplus dollar. That document stands. This
one asks a different and, for the index, more useful question:

**When an organization has an unusually good year, what does it do with the
money?**

The difference matters because a surplus is already the residual of a decision.
By the time a dollar shows up as surplus, the organization has decided not to
spend it. Starting from the revenue windfall instead puts the spend-or-keep
decision *inside* the estimate, which is where the interesting behaviour is.
Every dollar of revenue arriving above what an organization normally raises is
either spent in the same year or it is not; if it is not, it lands on the
balance sheet as cash, receivables, property, investments, or debt paid down.

This is the mirror of H6. The two documents share one data preparation
(`baseline-prep.R`), one population, one deflator and one baseline, so the
upside and downside estimates are directly comparable — section 11 puts them in
one table.

## What is inherited from the H6 rework

1. **Windfalls are measured against a trailing three-year average, not year
   *t-1*.** Against *t-1*, an organization recovering from one bad year looks
   identical to one receiving a windfall. Section 2 shows that
   almost a third of the year-over-year "windfalls" are exactly that.
2. **Everything is in constant `r params$ref_year` dollars** (CPI-U), applied
   before any baseline is formed.
3. **Every channel is reported twice**: cents per dollar of windfall, and the
   same dollars as a share of what that channel started with. Putting 25 cents
   into cash and 25 cents into property is the same number twice, but it may
   double the cash and add 2% to the buildings.

**A windfall** is an organization-year in which total real revenue comes in at
least `r sprintf("%.0f%%", params$shock_threshold * 100)` **above** the
organization's own inflation-adjusted average of the `r params$base_window`
prior tax years — the exact mirror of H6's shock.

**Scaling.** Everything is divided by **baseline expenses**, the trailing
three-year real average. Current-year expenses are an outcome here (spending
the windfall is one of the things being measured), so they cannot be the
denominator.

**Sign convention.** Throughout this document a positive number is a **use of
the windfall**: spending above the normal level, an asset that grew by more
than it normally does, or debt that grew by less than it normally does. On the
revenue side a positive number is a **source**: a line that came in above its
own baseline.

```{r packages}
suppressMessages({
  library(data.table)
  library(bit64)
  library(knitr)
})
set.seed(params$seed)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
pct <- function(x, d = 1) paste0(formatC(100 * x, format = "f", digits = d), "%")
BW  <- as.integer(params$base_window)
```

# 1. Data and the baseline

Blocks and revenue components are defined exactly as in H6; the tables in that
document's section 1 apply here unchanged.

```{r prep}
source("baseline-prep.R", local = knitr::knit_global())
kable(filings_by_year, format.args = list(big.mark = ","),
      caption = "Full-990 December filings available by tax year")
kable(p[, .(org_years = .N, organizations = uniqueN(EIN2)), by = year][order(year)],
      format.args = list(big.mark = ","),
      caption = "Organization-years with a complete three-year baseline")
```

# 2. What a windfall looks like

```{r windfall-distribution}
g <- p$gap
qs <- c(0.01, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 0.999)
dist <- data.table(statistic = c("median", "robust SD (IQR/1.349)", paste0("p", qs * 100)),
                   gap = round(c(median(g), iqr_sd(g), quantile(g, qs)), 4))
kable(dist, caption = "Revenue relative to a three-year inflation-adjusted baseline (0 = exactly on baseline)")

z20 <- (params$shock_threshold - median(g)) / iqr_sd(g)
by_size <- p[, .(organization_years = .N, median = median(gap), robust_sd = iqr_sd(gap),
                 p90 = quantile(gap, 0.90),
                 sd_of_a_20pct_windfall = (params$shock_threshold - median(gap)) / iqr_sd(gap),
                 share_windfall = mean(windfall)),
             by = size_band][order(size_band)]
kable(by_size, digits = 3, format.args = list(big.mark = ","),
      caption = "The 20% rule on the upside, by baseline revenue band")
```

A 20% windfall sits `r fmt(z20, 2)` robust standard deviations above the median
and occurs in `r pct(mean(p$windfall), 1)` of organization-years — marginally
*more* common than a 20% shortfall (`r pct(mean(p$shock), 1)`). As on the
downside, the same fixed percentage is a much weaker filter for small
organizations than for large ones.

**The upside tail is far fatter than the downside tail, and this is the one
place where the two sides cannot be treated symmetrically.** A shortfall is
bounded: revenue cannot fall more than 100% below baseline without going
negative, and only `r pct(mean(p$gap < -1), 2)` of organization-years do. There
is no corresponding ceiling.

```{r upper-tail}
b <- c(0.5, 1, 2, 5, 10, 50)
tail_tab <- data.table(
  bound = b,
  organization_years_above = vapply(b, function(v) sum(p$gap > v), 0L),
  share_of_all = vapply(b, function(v) mean(p$gap > v), 0),
  share_of_windfalls = vapply(b, function(v) sum(p$gap > v) / sum(p$windfall), 0))
kable(tail_tab, digits = 5,
      caption = "How many organization-years come in further than a given multiple above baseline")
cat("maximum gap observed:", fmt(max(p$gap), 1), "\n")
```

**Where to put the ceiling.** The mirror of H6's `-1.0` bound in robust standard
deviations is about `r fmt(median(g) + 3.5 * iqr_sd(g), 2)`, so the main
estimate caps the windfall at `r fmt(params$max_gain, 2)` — revenue at most
double its baseline. This is a much bigger exclusion than on the loss side:
it removes `r pct(sum(p$gap > params$max_gain) / sum(p$windfall), 1)` of
windfalls, against `r pct(sum(p$gap < -1) / sum(p$shock), 1)` of shocks at the
matching bound. That asymmetry is not a nuisance to be tuned away — it says
that organizations which more than double their revenue relative to baseline
are not having a good year, they are changing scale, and averaging them in with
ordinary windfalls would let a few hundred capital campaigns set the allocation
for everyone. Section 12 reports the estimate at other ceilings.

## Was the year-over-year rule picking up recoveries?

H6 found that a quarter of the organization-years the *t-1* rule called a shock
were really the reversion leg of a windfall. The same error runs the other way
here: an organization that had one bad year and returns to normal shows a large
year-over-year *gain* while nothing good has happened.

```{r recovery}
rec <- p[windfall_t1 == TRUE,
  .(n = .N,
    median_prior_year_vs_baseline = median(prior_windfall),
    share_prior_year_20pct_below_baseline = mean(prior_windfall < -0.20),
    median_current_vs_baseline = median(gap)),
  by = .(also_a_baseline_windfall = windfall)][order(-also_a_baseline_windfall)]
kable(rec, digits = 3,
      caption = "Organization-years the prior-year rule calls a windfall, split by whether the baseline rule agrees")

ov <- p[, .(t1_rule = sum(windfall_t1), baseline_rule = sum(windfall),
            both = sum(windfall & windfall_t1),
            t1_only = sum(windfall_t1 & !windfall),
            baseline_only = sum(windfall & !windfall_t1))]
kable(ov, format.args = list(big.mark = ","),
      caption = "Overlap of the two windfall definitions, all years")
```

`r pct(ov$t1_only / ov$t1_rule)` of the year-over-year "windfalls" are rejected
by the baseline rule — a **larger** contamination than the
the 25% H6 found on the loss side — and the rejected group is unmistakable:
their prior year sat a median
`r pct(rec[also_a_baseline_windfall == FALSE, median_prior_year_vs_baseline])`
below their own baseline, and their current year is
`r pct(rec[also_a_baseline_windfall == FALSE, median_current_vs_baseline])`
from it. They recovered; they did not gain.

# 3. Channels, sample and trimming

```{r channels}
# sign = +1 : actual minus baseline. sign = -1 : baseline minus actual.
# Either way a POSITIVE number is a use of the windfall (or, on the revenue
# rows, a source of it).
CH <- rbindlist(list(
  data.table(name="gov",       label="Government grants",         actual="gov",       base="base_gov",       capacity="base_gov",       sign=1, group="revenue"),
  data.table(name="dues",      label="Membership dues",           actual="dues",      base="base_dues",      capacity="base_dues",      sign=1, group="revenue"),
  data.table(name="events",    label="Fundraising events",        actual="events",    base="base_events",    capacity="base_events",    sign=1, group="revenue"),
  data.table(name="contr_oth", label="Other contributions",       actual="contr_oth", base="base_contr_oth", capacity="base_contr_oth", sign=1, group="revenue"),
  data.table(name="program",   label="Program revenue",           actual="program",   base="base_program",   capacity="base_program",   sign=1, group="revenue"),
  data.table(name="invest",    label="Investments",               actual="invest",    base="base_invest",    capacity="base_invest",    sign=1, group="revenue"),
  data.table(name="other",     label="Other revenue",             actual="other",     base="base_other",     capacity="base_other",     sign=1, group="revenue"),
  data.table(name="spend",     label="Extra spending",            actual="E",         base="base_E",         capacity="base_E",         sign=1, group="absorption"),
  data.table(name="sp_prog",   label="Program services",          actual="program_expenses",     base="base_program_expenses",     capacity="base_program_expenses",     sign=1, group="expense"),
  data.table(name="sp_admin",  label="Management and general",    actual="admin_expenses",       base="base_admin_expenses",       capacity="base_admin_expenses",       sign=1, group="expense"),
  data.table(name="sp_fund",   label="Fundraising",               actual="fundraising_expenses", base="base_fundraising_expenses", capacity="base_fundraising_expenses", sign=1, group="expense"),
  data.table(name="sp_unall",  label="Unallocated",               actual="unallocated_expenses", base="base_unallocated_expenses", capacity="base_unallocated_expenses", sign=1, group="expense"),
  data.table(name="bs_cash",   label="Cash",                      actual="d_cash",          base="base_d_cash",          capacity="boy_cash",          sign=1,  group="block"),
  data.table(name="bs_oca",    label="Receivables, prepaid",      actual="d_other_current", base="base_d_other_current", capacity="boy_other_current", sign=1,  group="block"),
  data.table(name="bs_fixed",  label="Property and equipment",    actual="d_fixed",         base="base_d_fixed",         capacity="boy_fixed",         sign=1,  group="block"),
  data.table(name="bs_othass", label="Investments, other assets", actual="d_other_assets",  base="base_d_other_assets",  capacity="boy_other_assets",  sign=1,  group="block"),
  data.table(name="bs_lt",     label="Long-term debt paid down",  actual="d_lt",            base="base_d_lt",            capacity="boy_lt",            sign=-1, group="block"),
  data.table(name="bs_st",     label="Short-term liabilities paid down", actual="d_st",     base="base_d_st",            capacity="boy_st",            sign=-1, group="block"),
  data.table(name="bs_othli",  label="Other liabilities paid down",     actual="d_other_liab", base="base_d_other_liab", capacity="boy_other_liab",   sign=-1, group="block")
))
CAP_PHRASE <- c(spend = "its normal budget", bs_cash = "the cash it held",
                bs_oca = "the receivables it held", bs_fixed = "the property it held",
                bs_othass = "the investments it held",
                bs_lt = "the long-term debt it owed", bs_st = "the short-term liabilities it owed",
                bs_othli = "the other liabilities it owed")

build_channels <- function(z) {
  z <- copy(z)
  z[, dol_gain := R - base_R][, s_gain := dol_gain / base_E]
  for (i in seq_len(nrow(CH))) {
    r <- CH[i]
    d <- if (r$sign == 1) z[[r$actual]] - z[[r$base]] else z[[r$base]] - z[[r$actual]]
    z[[paste0("dol_", r$name)]] <- d
    z[[paste0("s_",   r$name)]] <- d / z$base_E
    z[[paste0("cap_", r$name)]] <- z[[r$capacity]]
  }
  z[, dol_keep := dol_gain - dol_spend][, s_keep := dol_keep / base_E]   # the residual: kept
  z[, dol_na := d_na - base_d_na][, s_na := dol_na / base_E]
  z[]
}
p <- build_channels(p)

kable(data.table(
  check = c("revenue components sum to the windfall",
            "expense lines sum to the extra spending",
            "balance-sheet blocks sum to the extra net assets"),
  max_abs_error_dollars = c(
    p[, max(abs(dol_gov + dol_dues + dol_events + dol_contr_oth + dol_program + dol_invest + dol_other - dol_gain))],
    p[, max(abs(dol_sp_prog + dol_sp_admin + dol_sp_fund + dol_sp_unall - dol_spend))],
    p[, max(abs(dol_bs_cash + dol_bs_oca + dol_bs_fixed + dol_bs_othass + dol_bs_lt + dol_bs_st + dol_bs_othli - dol_na))])),
  digits = 6, caption = "Adding-up checks, before any trimming (constant dollars)")
```

Outliers are handled by **dropping organization-years, not clipping values**,
for the reason H5 and H6 both give: clipping each component separately breaks
the adding-up the design rests on. Three rules apply, within year — the 1st/99th
percentile of the scaled windfall, the 1st/99th of the net balance-sheet
movement, and the top 1% of the largest single block movement (blocks offset
each other, so an extreme swing between two of them passes the first two
filters while carrying enormous leverage).

```{r sample}
trim_sample <- function(z, trim = TRUE, bound = params$max_gain) {
  if (!is.null(bound)) z <- z[gap <= bound]
  if (!trim) return(z[])
  z[, keep := {
    qg <- quantile(s_gain, c(0.01, 0.99), na.rm = TRUE)
    qa <- quantile(s_na,   c(0.01, 0.99), na.rm = TRUE)
    qb <- quantile(max_abs_blk, 0.99, na.rm = TRUE)
    s_gain >= qg[[1]] & s_gain <= qg[[2]] & s_na >= qa[[1]] & s_na <= qa[[2]] &
      max_abs_blk <= qb
  }, by = year]
  z <- z[keep == TRUE]; z[, keep := NULL]; z[]
}
windfalls <- function(years) trim_sample(p[year %in% years & windfall == TRUE])
main_w <- windfalls(2022:2023)
cat("windfall organization-years, 2022-2023:", format(nrow(main_w), big.mark = ","),
    " organizations:", format(uniqueN(main_w$EIN2), big.mark = ","), "\n")
```

```{r estimands}
share_of_avg <- function(s, nm) mean(s[[paste0("s_", nm)]]) / mean(s$s_gain)
pct_capacity <- function(s, nm) {
  cap <- sum(s[[paste0("cap_", nm)]])
  if (!is.finite(cap) || cap <= 0) return(NA_real_)
  sum(s[[paste0("dol_", nm)]]) / cap
}
marginal <- function(s, nm) {
  f <- ols(s[[paste0("s_", nm)]], cbind(intercept = 1, gain = s$s_gain))
  c(b = f$b[["gain"]], se = f$se[["gain"]])
}
channel_table <- function(s, names_vec) {
  rbindlist(lapply(names_vec, function(nm) {
    m <- marginal(s, nm)
    data.table(channel = CH[name == nm, label],
               cents_per_dollar = share_of_avg(s, nm),
               pct_of_capacity  = pct_capacity(s, nm),
               marginal = m[["b"]], se = m[["se"]])
  }))
}
```

# 4. Where the windfall came from

```{r sources}
src <- channel_table(main_w, CH[group == "revenue", name])
kable(src, digits = 3,
      caption = "Which revenue lines came in above baseline, 2022-2023 windfalls")
cat("sum of shares:", fmt(sum(src$cents_per_dollar)), "\n")
```

# 5. First-line allocation: spent or kept

The identity `revenue - expenses = surplus` means a windfall must show up either
as spending above the organization's own baseline, or as a surplus above its
baseline. The two sum to exactly 1.

* **Extra spending** = actual expenses minus baseline expenses. Positive means
  the organization spent **more** than its own three-year norm — the windfall
  was used this year.
* **Kept** = the remainder, which lands on the balance sheet.

```{r allocation}
sp <- share_of_avg(main_w, "spend")
kp <- mean(main_w$s_keep) / mean(main_w$s_gain)
mS <- marginal(main_w, "spend")
alloc <- data.table(
  channel = c("Extra spending (above its own baseline)",
              "Kept (carried onto the balance sheet)"),
  cents_per_dollar = c(sp, kp),
  pct_of_capacity  = c(pct_capacity(main_w, "spend"), NA_real_),
  marginal = c(mS[["b"]], 1 - mS[["b"]]),
  se = c(mS[["se"]], mS[["se"]]))
kable(alloc, digits = 3,
      caption = "Of each dollar of windfall, how much is spent in the same year vs. kept")
cat("identity check, shares sum to:", fmt(sp + kp), "\n")
```

## What the extra spending buys

```{r spend-split}
spl <- channel_table(main_w, CH[group == "expense", name])
kable(spl, digits = 3,
      caption = "Where the extra spending goes (positive = that line came in above its baseline)")
cat("sums to the extra-spending share:", fmt(sum(spl$cents_per_dollar)), " vs. ", fmt(sp), "\n")
```

# 6. Where the balance sheet puts it

Each block enters as its movement during the windfall year measured against its
own baseline movement. An organization that normally adds $40k to cash a year
and this year added $140k has parked $100k of the windfall there.

```{r blocks}
bs <- channel_table(main_w, CH[group == "block", name])
kable(bs, digits = 3,
      caption = "Where the kept portion lands (positive = a use of the windfall)")
ident <- data.table(
  blocks_net = sum(bs$cents_per_dollar),
  net_assets_direct = mean(main_w$s_na) / mean(main_w$s_gain),
  kept = kp, n = nrow(main_w))
kable(ident, digits = 3,
      caption = "Identity check: blocks vs. the direct net-assets movement vs. the kept share")
```

The first two columns must agree; they are the same quantity computed two ways.
The third need not — net assets also move with unrealized investment gains and
donated capital, which do not pass through the surplus.

**The two columns weight organizations differently, and on this table they can
disagree in sign.** `cents_per_dollar` is a ratio of means of *scaled* channels,
so each organization enters in proportion to its own baseline expenses — a
small organization and a large one count roughly alike. `pct_of_capacity` is a
ratio of raw dollar totals, so it is dominated by the largest organizations in
the sample. Cash shows the effect plainly: the cents-per-dollar figure is
positive (`r fmt(bs[channel == "Cash", cents_per_dollar])`) while the aggregate
capacity figure is slightly negative
(`r fmt(bs[channel == "Cash", pct_of_capacity])`). Both are correct. Read
together they say that the typical organization puts a little of a windfall into
cash, while the large organizations that hold most of the sector's cash were
running their balances down over 2022-2023 faster than their own baseline
growth — the tail end of the COVID cash build-up. Where the two columns
disagree, the cents-per-dollar figure is the behavioural one and the capacity
figure is the sector aggregate.

The same caveat H6 records applies to the investments row: Part X is fair value,
so that block mixes deliberate investment of the windfall with market
revaluation that would have happened anyway. Read cash, receivables, property
and debt as behaviour; read investments as behaviour plus the market.

# 7. Is the windfall concentrated in one revenue stream?

This is the question the rework was asked for. A windfall can arrive as one
large grant or as a generally good year across every line, and there is no
reason to expect those to be treated the same way.

**Measure.** For each windfall organization-year, take the seven component
deviations from baseline, keep the positive ones, and compute the largest one's
share of their total. A value near 1 means essentially all of the good news came
from a single stream.

```{r concentration}
COMPS <- CH[group == "revenue"]
dev_mat <- vapply(COMPS$name, function(nm) p[[paste0("dol_", nm)]], numeric(nrow(p)))
pos_mat <- pmax(dev_mat, 0)
pos_tot <- rowSums(pos_mat)
# max.col rather than apply(): the same answer on 670k rows in a fraction of the time
top_idx <- max.col(pos_mat, ties.method = "first")
top_val <- pos_mat[cbind(seq_len(nrow(pos_mat)), top_idx)]
p[, top_share := ifelse(pos_tot > 0, top_val / pos_tot, NA_real_)]
p[, top_component := COMPS$label[top_idx]]
p[, concentration := factor(
  ifelse(is.na(top_share), NA_character_,
  ifelse(top_share >= params$conc_high, "single stream",
  ifelse(top_share >= params$conc_low,  "mostly one stream", "spread across lines"))),
  levels = c("single stream", "mostly one stream", "spread across lines"))]

main_w <- trim_sample(p[year %in% 2022:2023 & windfall == TRUE])   # refresh with the new columns
kable(data.table(statistic = c("p10", "p25", "median", "p75", "p90"),
                 top_component_share = round(quantile(main_w$top_share, c(.1,.25,.5,.75,.9),
                                                      na.rm = TRUE), 3)),
      caption = "Share of the windfall coming from its single largest contributing line")

conc <- rbindlist(lapply(levels(main_w$concentration), function(lv) {
  s <- main_w[concentration == lv]
  if (nrow(s) < 200) return(NULL)
  data.table(concentration = lv, n = nrow(s),
             median_top_share = median(s$top_share),
             windfall_vs_baseline_revenue = mean(s$dol_gain) / mean(s$base_R),
             spent = share_of_avg(s, "spend"),
             kept = mean(s$s_keep) / mean(s$s_gain),
             program_spending = share_of_avg(s, "sp_prog"),
             cash = share_of_avg(s, "bs_cash"),
             cash_pct_cap = pct_capacity(s, "bs_cash"),
             investments = share_of_avg(s, "bs_othass"),
             property = share_of_avg(s, "bs_fixed"),
             debt_paid_down = share_of_avg(s, "bs_lt") + share_of_avg(s, "bs_st"))
}))
kable(conc, digits = 3,
      caption = "Allocation by how concentrated the windfall is (cents per dollar of windfall)")
```

**Windfalls are overwhelmingly concentrated.** The median windfall takes
`r pct(median(main_w$top_share, na.rm = TRUE), 0)` of its good news from a single
revenue line, and
`r pct(mean(main_w$concentration == "single stream", na.rm = TRUE), 0)` of
windfall organization-years clear the `r pct(params$conc_high, 0)` bar. "A
generally good year across the board" is the exception, not the rule — which is
itself a reason to prefer the component view in section 8 over the pooled
estimate.

**But concentration barely moves the spend-or-keep decision.** The spending
share is
`r fmt(conc[concentration == "single stream", spent])`,
`r fmt(conc[concentration == "mostly one stream", spent])` and
`r fmt(conc[concentration == "spread across lines", spent])` across the three
groups — flat, and not even monotonic. Whatever governs how much of a windfall
gets spent, it is not how concentrated the windfall is.

Where concentration does show up is in **which** balance-sheet block takes the
kept portion. Single-stream windfalls put
`r fmt(conc[concentration == "single stream", cash])` per dollar into cash
against `r fmt(conc[concentration == "spread across lines", cash])` for spread
ones, while spread windfalls put more into property
(`r fmt(conc[concentration == "spread across lines", property])` against
`r fmt(conc[concentration == "single stream", property])`). That ordering is
consistent with a one-off grant being parked until it is clear what it is for,
while a broad-based good year is treated as durable enough to commit to
buildings — but the differences are small next to the component differences in
section 8, and the concentration measure is partly a proxy for which line the
money arrived in, so this should not be over-read.

```{r concentration-mix}
kable(dcast(main_w[!is.na(concentration), .N, by = .(top_component, concentration)],
            top_component ~ concentration, value.var = "N", fill = 0L),
      format.args = list(big.mark = ","),
      caption = "Which line is the largest contributor, by concentration")
```

# 8. Windfalls concentrated in a single revenue stream

Section 7 asks whether concentration matters. This section asks which stream.

**Definition.** A component windfall is an organization-year in which one
component comes in at least `r pct(params$comp_share, 0)` **above** its own
three-year baseline, *and* that rise is worth at least
`r pct(params$comp_material, 0)` of the organization's baseline total revenue.

**Denominator.** As in H6's mirror section, each component is scaled by **its
own rise**, not by the total windfall. When one line jumps, total revenue does
not necessarily rise by as much — other lines can fall at the same time, and
crowding-out of that kind is one of the things worth measuring. The three
shares below sum to exactly 1:

* **Offset by other revenue falling** — the share of the rise cancelled by other
  lines coming in below their own baselines.
* **Spent** — the share matched by spending above baseline.
* **Kept** — the remainder, onto the balance sheet.

```{r component-flags}
comp_flag <- function(z, nm) {
  base_c <- z[[paste0("base_", nm)]]
  rise   <- z[[paste0("dol_", nm)]]           # actual minus baseline, in dollars
  ok <- is.finite(base_c) & is.finite(rise) & base_c > 0
  ok & (rise / base_c) >= params$comp_share & (rise / z$base_R) >= params$comp_material
}
pool  <- p[year %in% 2022:2023]
flags <- vapply(COMPS$name, function(nm) comp_flag(pool, nm), logical(nrow(pool)))
pool[, nflags := rowSums(flags)]
kable(pool[, .(organization_years = .N), by = .(component_windfalls = nflags)][order(component_windfalls)],
      format.args = list(big.mark = ","),
      caption = "How many revenue components are in windfall in the same organization-year")
```

```{r component-allocation}
comp_alloc <- rbindlist(lapply(seq_len(nrow(COMPS)), function(i) {
  nm <- COMPS$name[i]
  s  <- trim_sample(pool[flags[, nm]])
  if (nrow(s) < 200) return(data.table(component = COMPS$label[i], n = nrow(s)))
  rise <- mean(s[[paste0("dol_", nm)]])
  data.table(
    component = COMPS$label[i], n = nrow(s), orgs = uniqueN(s$EIN2),
    median_line_rise = median(s[[paste0("dol_", nm)]] / s[[paste0("base_", nm)]]),
    offset_by_other_revenue = 1 - mean(s$dol_gain) / rise,
    spent = mean(s$dol_spend) / rise,
    kept  = mean(s$dol_keep)  / rise,
    program_spending = mean(s$dol_sp_prog) / rise,
    cash  = mean(s$dol_bs_cash) / rise,
    cash_pct_cap = pct_capacity(s, "bs_cash"),
    investments = mean(s$dol_bs_othass) / rise,
    property = mean(s$dol_bs_fixed) / rise,
    debt_paid_down = (mean(s$dol_bs_lt) + mean(s$dol_bs_st)) / rise)
}), fill = TRUE)
kable(comp_alloc, digits = 3,
      caption = "How a windfall in each revenue component is allocated (per dollar of that line's rise)")
cat("offset + spent + kept = ",
    paste(fmt(comp_alloc$offset_by_other_revenue + comp_alloc$spent + comp_alloc$kept),
          collapse = ", "), "\n")
```

The spread across components is much wider than anything in the aggregate, and
it maps onto H6's loss-side result line for line.

* **Program revenue is spent, almost entirely.**
  `r fmt(comp_alloc[component == "Program revenue", spent])` of a dollar of extra
  program revenue is matched by spending above baseline and only
  `r fmt(comp_alloc[component == "Program revenue", kept])` is kept. H6 found the
  mirror: `r "0.85"` of a *lost* program dollar was matched by cutting. Program
  revenue and program cost move together in both directions, which is what you
  would expect if they are two sides of one activity rather than a financing
  decision.
* **Contributed income is where discretion lives.** Other contributions keep
  `r fmt(comp_alloc[component == "Other contributions", kept])` per dollar and
  government grants `r fmt(comp_alloc[component == "Government grants", kept])`,
  against near zero for program revenue.
* **Half of some windfalls never reach the bottom line.** Other revenue is
  `r pct(comp_alloc[component == "Other revenue", offset_by_other_revenue], 0)`
  offset by other lines falling, fundraising events
  `r pct(comp_alloc[component == "Fundraising events", offset_by_other_revenue], 0)`,
  membership dues
  `r pct(comp_alloc[component == "Membership dues", offset_by_other_revenue], 0)`.
  These lines tend to move when something else is moving the other way, so the
  "windfall" is substantially a substitution.
* **The investments row is mechanical and should not be read as behaviour.** A
  windfall in the investments line is mostly realized gains on asset sales, and
  selling an appreciated asset is exactly what makes the investments *block* on
  the balance sheet fall. That is why its balance-sheet coefficient is large and
  negative (`r fmt(comp_alloc[component == "Investments", investments])` per
  dollar): the money did not go into investments, it came out of them and
  reappeared as revenue. The same accounting runs through the
  `r fmt(comp_alloc[component == "Investments", debt_paid_down])` debt figure.

# 9. Do reserves change what happens to a windfall?

H6 found that reserves buffer a shock. The upside question is whether an
organization that already holds a lot of cash banks a windfall or spends it.

```{r reserves}
main_w[, reserve_group := factor(c("low", "middle", "high")[
  pmin(3L, 1L + floor(3 * (frank(months_cash_boy, ties.method = "first") - 1) / .N))],
  levels = c("low", "middle", "high"))]
res <- rbindlist(lapply(levels(main_w$reserve_group), function(gp) {
  s <- main_w[reserve_group == gp]
  data.table(reserves = gp, n = nrow(s),
             median_months_cash = median(s$months_cash_boy),
             spent = share_of_avg(s, "spend"),
             program_spending = share_of_avg(s, "sp_prog"),
             kept = mean(s$s_keep) / mean(s$s_gain),
             cash = share_of_avg(s, "bs_cash"),
             cash_pct_cap = pct_capacity(s, "bs_cash"),
             property = share_of_avg(s, "bs_fixed"),
             debt_paid_down = share_of_avg(s, "bs_lt") + share_of_avg(s, "bs_st"))
}))
kable(res, digits = 3,
      caption = "Allocation by beginning-of-year reserves (cents per dollar of windfall)")
```

# 10. Continuity with the original H5

The original H5 regressed each block's movement on the level of the surplus,
scaled by the year's expenses, over all organization-years. Re-running that
specification on this sample — with baseline expenses as the denominator, so the
scaling matches the rest of this document — shows how the two framings relate.

```{r h5-continuity}
h5 <- trim_sample(p[year %in% 2022:2023], bound = NULL)
h5[, s_surplus := (R - E) / base_E]
W5 <- cbind(intercept = 1, surplus = h5$s_surplus)
BLK <- CH[group == "block"]
h5_tab <- rbindlist(lapply(seq_len(nrow(BLK)), function(i) {
  # the original H5 measured the block MOVEMENT, not its deviation from baseline
  y <- h5[[BLK$actual[i]]] / h5$base_E
  f <- ols(y, W5)
  data.table(block = BLK$label[i], cents_per_surplus_dollar = f$b[["surplus"]], se = f$se[["surplus"]])
}))
kable(h5_tab, digits = 3,
      caption = "H5-style regression: block movement on the surplus level, 2022-2023")
cat("asset blocks minus liability blocks:",
    fmt(sum(h5_tab$cents_per_surplus_dollar[1:4]) - sum(h5_tab$cents_per_surplus_dollar[5:7])),
    " | direct net-assets regression:", fmt(ols(h5$d_na / h5$base_E, W5)$b[["surplus"]]), "\n")
```

The two framings answer different questions and should not be expected to
agree. The H5 coefficient says where a dollar of *surplus* sits at year end; the
windfall estimate says what happens to a dollar of *revenue* that arrived above
normal, most of which never becomes surplus at all.

# 11. The two tails side by side

Because the sign convention is symmetric — every channel is a deviation from the
organization's own baseline, divided by the deviation in revenue — the same code
run on the shortfall sample reproduces H6 exactly. That is a check on both
documents as much as a comparison.

```{r two-tails}
# The -1 floor is applied before the percentile trim, exactly as H6 applies it,
# so this reproduces that document's sample rather than a near neighbour of it.
short_s <- trim_sample(p[year %in% 2022:2023 & shock == TRUE & gap >= -1], bound = NULL)
two <- data.table(
  channel = c("Expenses", "Surplus / kept", "Cash", "Investments, other assets",
              "Property and equipment", "Debt"),
  per_dollar_of_windfall = c(
    sp, kp, share_of_avg(main_w, "bs_cash"), share_of_avg(main_w, "bs_othass"),
    share_of_avg(main_w, "bs_fixed"),
    share_of_avg(main_w, "bs_lt") + share_of_avg(main_w, "bs_st")),
  per_dollar_of_shortfall = c(
    share_of_avg(short_s, "spend"), mean(short_s$s_keep) / mean(short_s$s_gain),
    share_of_avg(short_s, "bs_cash"), share_of_avg(short_s, "bs_othass"),
    share_of_avg(short_s, "bs_fixed"),
    share_of_avg(short_s, "bs_lt") + share_of_avg(short_s, "bs_st")))
kable(two, digits = 3,
      caption = "A dollar above baseline vs. a dollar below it, 2022-2023 (both as a share of the deviation)")
```

The expense row is the asymmetry that matters for the index: the share of a
windfall that is spent in the same year is several times the share of a
shortfall that is matched by cutting. Good news moves spending; bad news moves
the balance sheet.

# 12. Robustness

```{r contrasts}
wins <- list(`2019 (pre-COVID)` = 2019, `2020 (COVID)` = 2020, `2021 (relief year)` = 2021,
             `2022` = 2022, `2023` = 2023, `2024` = 2024)
by_year <- rbindlist(lapply(names(wins), function(nm) {
  s <- windfalls(wins[[nm]])
  if (nrow(s) < 200) return(NULL)
  data.table(window = nm, n = nrow(s), spent = share_of_avg(s, "spend"),
             kept = mean(s$s_keep) / mean(s$s_gain),
             cash = share_of_avg(s, "bs_cash"),
             investments = share_of_avg(s, "bs_othass"),
             property = share_of_avg(s, "bs_fixed"),
             debt_paid_down = share_of_avg(s, "bs_lt") + share_of_avg(s, "bs_st"))
}))
kable(by_year, digits = 3, caption = "Allocation by windfall year")
```

```{r sensitivity}
sens <- rbindlist(lapply(list(
  list(nm = "ceiling at +1.0 (as reported)", b = 1.0, t = params$shock_threshold),
  list(nm = "ceiling at +0.5",               b = 0.5, t = params$shock_threshold),
  list(nm = "ceiling at +2.0",               b = 2.0, t = params$shock_threshold),
  list(nm = "no ceiling",                    b = NULL, t = params$shock_threshold),
  list(nm = "threshold +10%, ceiling +1.0",  b = 1.0, t = 0.10),
  list(nm = "threshold +50%, ceiling +1.0",  b = 1.0, t = 0.50)), function(a) {
  s <- trim_sample(p[year %in% 2022:2023 & gap >= a$t], bound = a$b)
  data.table(definition = a$nm, n = nrow(s), spent = share_of_avg(s, "spend"),
             kept = mean(s$s_keep) / mean(s$s_gain),
             cash = share_of_avg(s, "bs_cash"),
             property = share_of_avg(s, "bs_fixed"))
}))
kable(sens, digits = 3, caption = "Sensitivity to the ceiling and the threshold")

alt <- rbindlist(lapply(list(
  list(nm = "3-year baseline, real (as reported)", v = "gap"),
  list(nm = "2-year baseline, real",               v = "gap2"),
  list(nm = "3-year baseline, nominal",            v = "gap_nom"),
  list(nm = "prior year only, real",               v = "gap_t1")), function(a) {
  gv  <- p[[a$v]]
  sel <- p$year %in% 2022:2023 & !is.na(gv) & gv >= params$shock_threshold
  s <- trim_sample(p[sel])
  data.table(definition = a$nm, n = nrow(s), spent = share_of_avg(s, "spend"),
             kept = mean(s$s_keep) / mean(s$s_gain), cash = share_of_avg(s, "bs_cash"))
}))
kable(alt, digits = 3, caption = "Sensitivity to how the baseline is built")

untrimmed <- trim_sample(p[year %in% 2022:2023 & windfall == TRUE], trim = FALSE)
kable(rbindlist(lapply(list(list(nm = "trimmed", s = main_w), list(nm = "untrimmed", s = untrimmed)),
  function(a) data.table(sample = a$nm, n = nrow(a$s), spent = share_of_avg(a$s, "spend"),
                         kept = mean(a$s$s_keep) / mean(a$s$s_gain),
                         cash = share_of_avg(a$s, "bs_cash")))),
  digits = 3, caption = "Trimmed vs. untrimmed")
```

# 13. The flow figure

```{r flow-figure, fig.width = 8.8, fig.height = 6.4, fig.retina = 2}
pal <- c(ink = "#1f2429", ink_mid = "#4d565f", ink_soft = "#6d7681",
         accent = "#bd5416", accent_ink = "#8f3f10", navy = "#153243",
         pumpkin = "#dc965a", surface = "#f3f4f6")

flow_figure <- function(sources, uses, use_capacity_label, title, subtitle, dropped = NULL) {
  op <- par(mar = c(0.4, 0.4, 3.0, 0.4), bg = "white", family = "sans")
  on.exit(par(op))
  gap <- 0.045; centre <- 0.5
  lab_s <- sprintf("%s  %.2f", names(sources), sources)
  lab_u <- sprintf("%s  %.2f", names(uses), uses)
  sources_n <- sources / sum(sources); uses_n <- uses / sum(uses)
  half <- function(v) (sum(v) + gap * (length(v) - 1)) / 2
  top_max <- centre + max(half(sources_n), half(uses_n), 0.5)
  bot_min <- centre - max(half(sources_n), half(uses_n), 0.5)
  head_y <- top_max + 0.085; arrow_y <- bot_min - 0.105
  legend_y <- arrow_y - 0.075
  foot_y <- legend_y - (if (length(dropped)) 0.185 else 0.115)
  plot.new(); plot.window(xlim = c(0, 1), ylim = c(foot_y - 0.02, head_y + 0.06))
  stack <- function(v) {
    y <- centre + half(v)
    out <- lapply(v, function(h) { r <- c(top = y, bot = y - h); y <<- y - h - gap; r })
    names(out) <- names(v); out
  }
  ribbon <- function(x0, x1, t0, b0, t1, b1, col, alpha) {
    t <- seq(0, 1, length.out = 80); s <- t * t * (3 - 2 * t); xs <- x0 + (x1 - x0) * t
    polygon(c(xs, rev(xs)), c(t0 + (t1 - t0) * s, rev(b0 + (b1 - b0) * s)),
            col = adjustcolor(col, alpha), border = NA)
  }
  chevron <- function(x, ytop, ybot, col) {
    ymid <- (ytop + ybot) / 2; h <- min(0.020, (ytop - ybot) / 2.6); w <- 0.012
    if (h <= 0.003) return(invisible())
    lines(c(x - w, x, x - w), c(ymid + h, ymid, ymid - h), col = col, lwd = 1.6)
  }
  xs <- c(0.250, 0.300); xt <- c(0.487, 0.537); xu <- c(0.700, 0.750)
  srcb <- stack(sources_n); useb <- stack(uses_n); top <- centre + 0.5
  place <- function(boxes, minsep) {
    y <- vapply(boxes, mean, 0); n <- length(y)
    if (n > 1) for (i in (n - 1):1) if (y[i] - y[i + 1] < minsep) y[i] <- y[i + 1] + minsep
    y
  }
  ys_lab <- place(srcb, 0.052)
  yu_lab <- place(useb, if (any(nzchar(use_capacity_label))) 0.088 else 0.052)
  leader <- function(x0, x1, y_bar, y_lab, col) {
    if (abs(y_lab - y_bar) < 0.004) return(invisible())
    lines(c(x0, x1), c(y_bar, y_lab), col = adjustcolor(col, 0.45), lwd = 0.7)
  }
  cur <- top
  for (i in seq_along(srcb)) {
    ribbon(xs[2], xt[1], srcb[[i]][["top"]], srcb[[i]][["bot"]], cur, cur - sources_n[[i]],
           pal[["navy"]], 0.30)
    chevron(0.398, srcb[[i]][["top"]], srcb[[i]][["bot"]], adjustcolor(pal[["navy"]], 0.85))
    cur <- cur - sources_n[[i]]
  }
  cur <- top
  for (i in seq_along(useb)) {
    ribbon(xt[2], xu[1], cur, cur - uses_n[[i]], useb[[i]][["top"]], useb[[i]][["bot"]],
           pal[["accent"]], 0.34)
    chevron(0.632, useb[[i]][["top"]], useb[[i]][["bot"]], adjustcolor(pal[["accent_ink"]], 0.85))
    cur <- cur - uses_n[[i]]
  }
  for (i in seq_along(srcb)) {
    rect(xs[1], srcb[[i]][["bot"]], xs[2], srcb[[i]][["top"]], col = pal[["navy"]], border = NA)
    leader(xs[1] - 0.004, xs[1] - 0.012, mean(srcb[[i]]), ys_lab[i], pal[["ink_soft"]])
    text(xs[1] - 0.016, ys_lab[i], lab_s[i], adj = c(1, 0.5), cex = 0.68, col = pal[["ink"]])
  }
  rect(xt[1], centre - 0.5, xt[2], top, col = pal[["accent"]], border = NA)
  for (i in seq_along(useb)) {
    rect(xu[1], useb[[i]][["bot"]], xu[2], useb[[i]][["top"]], col = pal[["pumpkin"]], border = NA)
    leader(xu[2] + 0.004, xu[2] + 0.012, mean(useb[[i]]), yu_lab[i], pal[["ink_soft"]])
    ym <- yu_lab[i]
    if (nzchar(use_capacity_label[i])) {
      text(xu[2] + 0.016, ym + 0.018, lab_u[i], adj = c(0, 0.5), cex = 0.68, col = pal[["ink"]])
      text(xu[2] + 0.016, ym - 0.020, use_capacity_label[i], adj = c(0, 0.5), cex = 0.56,
           col = pal[["ink_soft"]])
    } else {
      text(xu[2] + 0.016, ym, lab_u[i], adj = c(0, 0.5), cex = 0.68, col = pal[["ink"]])
    }
  }
  text(xs[2], head_y, "WHERE THE EXTRA MONEY CAME FROM", adj = c(1, 0.5), cex = 0.60,
       font = 2, col = pal[["navy"]])
  text(mean(xt), head_y, "$1.00 WINDFALL", adj = c(0.5, 0.5), cex = 0.62, font = 2,
       col = pal[["accent_ink"]])
  text(xu[1], head_y, "WHAT IT WAS USED FOR", adj = c(0, 0.5), cex = 0.60, font = 2,
       col = pal[["accent_ink"]])
  lines(c(0.250, 0.740), c(arrow_y, arrow_y), col = pal[["ink_soft"]], lwd = 1)
  polygon(c(0.750, 0.732, 0.732), c(arrow_y, arrow_y + 0.013, arrow_y - 0.013),
          col = pal[["ink_soft"]], border = NA)
  text(0.495, arrow_y + 0.022, "read left to right", adj = c(0.5, 0), cex = 0.58,
       font = 3, col = pal[["ink_soft"]])
  text(0.495, legend_y,
       paste0("Every bar is a positive amount of money. On the left it is revenue that arrived above ",
              "the organization's own\nbaseline. On the right it is spending above baseline, or a ",
              "balance-sheet stock that grew by more than it\nnormally does. The grey note under each ",
              "is the same amount as a share of what the organization already had."),
       adj = c(0.5, 1), cex = 0.57, col = pal[["ink_mid"]])
  if (length(dropped)) {
    text(0.495, foot_y + 0.055,
         paste0("Moved the other way during the windfall year, so not drawn: ",
                paste(dropped, collapse = "; "), ". The tables above keep the sign."),
         adj = c(0.5, 1), cex = 0.55, col = pal[["ink_soft"]], font = 3)
  }
  mtext(title, side = 3, line = 1.45, adj = 0, cex = 1.0, font = 2, col = pal[["ink"]])
  mtext(subtitle, side = 3, line = 0.45, adj = 0, cex = 0.64, col = pal[["ink_soft"]])
}

pick <- function(v) { v <- v[v > 0.005]; sort(v, decreasing = TRUE) }
src_v  <- setNames(src$cents_per_dollar, src$channel)
uses_v <- c(`Extra spending` = sp,
            setNames(bs$cents_per_dollar, bs$channel)[
              c("Cash", "Investments, other assets", "Receivables, prepaid",
                "Property and equipment")],
            `Debt paid down` = bs[channel == "Long-term debt paid down", cents_per_dollar] +
                               bs[channel == "Short-term liabilities paid down", cents_per_dollar] +
                               bs[channel == "Other liabilities paid down", cents_per_dollar])
cap_v <- c(`Extra spending` = pct_capacity(main_w, "spend"),
           Cash = pct_capacity(main_w, "bs_cash"),
           `Investments, other assets` = pct_capacity(main_w, "bs_othass"),
           `Receivables, prepaid` = pct_capacity(main_w, "bs_oca"),
           `Property and equipment` = pct_capacity(main_w, "bs_fixed"),
           `Debt paid down` = (sum(main_w$dol_bs_lt) + sum(main_w$dol_bs_st) +
                               sum(main_w$dol_bs_othli)) /
                              sum(main_w$boy_lt + main_w$boy_st + main_w$boy_other_liab))
cap_phrase <- c(`Extra spending` = "its normal budget", Cash = "the cash it held",
                `Investments, other assets` = "the investments it held",
                `Receivables, prepaid` = "the receivables it held",
                `Property and equipment` = "the property it held",
                `Debt paid down` = "the liabilities it owed")
sources_p <- pick(src_v); uses_p <- pick(uses_v)
dropped <- c(names(src_v)[src_v <= 0.005], names(uses_v)[uses_v <= 0.005])
capk <- cap_v[names(uses_p)]
flow_figure(
  sources_p, uses_p,
  use_capacity_label = ifelse(is.na(capk) | capk <= 0, "",
                              sprintf("%.0f%% of %s", 100 * capk, cap_phrase[names(uses_p)])),
  title = "Where a windfall comes from, and what it is used for",
  subtitle = paste0("cents per dollar of windfall  |  full-990 filers, December year-ends, revenue ",
                    pct(params$shock_threshold, 0), "+ above a ", BW,
                    "-year inflation-adjusted baseline, 2022-2023  |  n = ",
                    format(nrow(main_w), big.mark = ",")),
  dropped = tolower(dropped))
```

# 14. Findings

```{r findings, include = FALSE}
cash_c <- bs[channel == "Cash", cents_per_dollar]
cash_k <- bs[channel == "Cash", pct_of_capacity]
```

* **A windfall is mostly spent, and that is the sharpest contrast with H6.**
  `r fmt(sp)` of each windfall dollar is matched by spending above the
  organization's own baseline and `r fmt(kp)` is kept. On the loss side only
  `r fmt(two[channel == "Expenses", per_dollar_of_shortfall])` of a shortfall is
  matched by lower spending. Organizations adjust spending upward far more
  readily than downward.
* **What is kept goes to cash first.** `r fmt(cash_c)` per windfall dollar ends
  the year as cash, which is `r pct(cash_k)` of the cash the organization was
  holding when the year began.
* **A third of the year-over-year "windfalls" are recoveries.**
  `r pct(ov$t1_only / ov$t1_rule)` of the organization-years the *t-1* rule
  flags are rejected by the baseline rule; their prior year sat a median
  `r pct(rec[also_a_baseline_windfall == FALSE, median_prior_year_vs_baseline])`
  below their own baseline. The contamination is larger on the upside than the
  the 25% H6 found on the downside.
* **Windfalls are overwhelmingly concentrated, but concentration is not what
  drives the response.** The median windfall takes
  `r pct(median(main_w$top_share, na.rm = TRUE), 0)` of its good news from a
  single revenue line, and
  `r pct(mean(main_w$concentration == "single stream", na.rm = TRUE), 0)`
  of windfall organization-years are single-stream on the
  `r pct(params$conc_high, 0)` test. Yet the spending share is flat across the
  three concentration groups
  (`r fmt(conc[concentration == "single stream", spent])` /
  `r fmt(conc[concentration == "mostly one stream", spent])` /
  `r fmt(conc[concentration == "spread across lines", spent])`). Concentration
  shifts *which* block the kept portion lands in — more cash when it is one
  stream, more property when it is broad — but not how much is kept.
* **Which stream it arrives in is what matters.** Per dollar of extra program
  revenue, `r fmt(comp_alloc[component == "Program revenue", spent])` is spent
  and `r fmt(comp_alloc[component == "Program revenue", kept])` kept; per dollar
  of extra unrestricted contributions,
  `r fmt(comp_alloc[component == "Other contributions", kept])` is kept. That is
  an order-of-magnitude difference in retention driven entirely by the source,
  and it mirrors H6's loss-side finding that program revenue falls are absorbed
  by cutting while contributed falls land on the balance sheet.
* **Several "windfalls" are substitutions.** Between a fifth and a half of the
  rise in dues, events, other revenue and government grants is offset by other
  lines falling in the same year, so it never reaches the bottom line at all.

```{r save, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(gap_distribution = dist, gap_by_size = by_size, upper_tail = tail_tab,
               recovery = rec, overlap = ov, sources = src, allocation = alloc,
               spend_split = spl, blocks = bs, identity = ident,
               concentration = conc, components = comp_alloc, reserves = res,
               h5_continuity = h5_tab, two_tails = two, by_year = by_year,
               sensitivity = sens, window_sensitivity = alt,
               n = nrow(main_w), threshold = params$shock_threshold,
               max_gain = params$max_gain, base_window = BW, ref_year = params$ref_year),
          params$results_file)
}
```

```{r session}
sessionInfo()
```
