---
title: "H6: How Is a Sudden Revenue Loss Absorbed?"
subtitle: "Expense cuts and balance-sheet drawdown after a 20% drop in revenue, 2019-2023"
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
    css: walkthrough.css
params:
  panel_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2019-2024-imputed.rds")
  results_file: ""
  seed: 2023
  shock_threshold: 0.20
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 130, knitr.kable.NA = "")
```

# Purpose

H5 asked where a surplus goes. This document asks the mirror question about the
loss side, which the ratio-correlation memo flagged as the next thing worth
testing: **when revenue falls sharply, what absorbs the loss?**

Every dollar of revenue an organization does not receive has to be found
somewhere. It comes out of expenses, or it comes out of the balance sheet as a
worse surplus — and that worse surplus is itself drawn from cash, receivables,
property, investments, or new borrowing. The accounting identity pins the split
down exactly, so the estimate is the split itself, not whether it exists.

Two identities carry the whole document:

1. `revenue - expenses = surplus`, so the change in revenue equals the change
   in expenses plus the change in the surplus. Of a dollar lost, the share
   taken out of expenses and the share taken out of the surplus must sum to 1.
2. The surplus lands on the balance sheet, exactly as in H5: the asset-side
   changes minus the liability-side changes equal the change in net assets.

**A sudden loss** is an organization-year in which total revenue falls by at
least `r sprintf("%.0f%%", params$shock_threshold * 100)` of the prior year's
revenue, with the prior year filed by the same organization on a full 990.

**Window.** Shocks landing in tax years 2022 and 2023, December year-ends,
organizations that filed a full 990 in every year 2019-2023. The 2020 shock
year is estimated separately: it is the largest revenue shock in the panel, but
relief funding makes its absorption unrepresentative, and it is more useful as
a contrast than as part of the main estimate.

**Scaling.** Everything is divided by the **prior year's** total expenses, not
the current year's. This is a deliberate departure from H5. Current-year
expenses are one of the outcomes here — an organization that responds to a
shock by cutting deeply would shrink its own denominator, and the expense-cut
share would be biased toward zero. Prior-year expenses are fixed before the
shock lands.

```{r packages}
suppressMessages({
  library(data.table)
  library(bit64)
  library(knitr)
})
set.seed(params$seed)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

# 1. Data, lags, and the shock indicator

Balance-sheet blocks follow H5 exactly: end-of-year minus beginning-of-year,
both from the same filing's Part X, so the two dates share one accounting
basis.

| Block | Part X lines |
|---|---|
| Cash | cash + savings |
| Other current assets | pledges, accounts receivable, inventories, prepaid |
| Fixed assets | land, buildings, equipment (net) |
| Other assets | total assets minus the three above (investments, intangibles, notes receivable) |
| Long-term debt | tax-exempt bonds + mortgages + unsecured notes |
| Short-term liabilities | accounts payable + grants payable + deferred revenue |
| Other liabilities | total liabilities minus the two above |

```{r data}
d <- as.data.table(readRDS(params$panel_file))
num <- function(x) { x <- as.numeric(x); x[is.na(x)] <- 0; x }

forms <- d[, .(n990 = sum(RETURN_TYPE == "990"), nez = sum(RETURN_TYPE == "990EZ")), by = EIN2]
stable990 <- forms[n990 >= 5 & nez == 0, EIN2]

x <- d[EIN2 %chin% stable990 & RETURN_TYPE == "990" &
       substr(TAX_PERIOD_END_DATE, 6, 7) == "12"]
rm(d); invisible(gc())

blk <- function(when) {
  s <- function(f) num(x[[paste0(f, "_", when)]])
  data.table(
    cash   = s("F9_10_ASSET_CASH") + s("F9_10_ASSET_SAVING"),
    oca    = s("F9_10_ASSET_PLEDGE_NET") + s("F9_10_ASSET_ACC_NET") +
             s("F9_10_ASSET_INV_SALE") + s("F9_10_ASSET_EXP_PREPAID"),
    fixed  = s("F9_10_ASSET_LAND_BLDG_NET"),
    assets = s("F9_10_ASSET_TOT"),
    lt     = s("F9_10_LIAB_TAX_EXEMPT_BOND") + s("F9_10_LIAB_MTG_NOTE") + s("F9_10_LIAB_NOTE_UNSEC"),
    st     = s("F9_10_LIAB_ACC_PAYABLE") + s("F9_10_LIAB_GRANT_PAYABLE") + s("F9_10_LIAB_REV_DEFERRED"),
    liab   = s("F9_10_LIAB_TOT"),
    na     = s("F9_10_NAFB_TOT"))
}
b <- blk("BOY"); e <- blk("EOY")
b[, `:=`(other_assets = assets - cash - oca - fixed, other_liab = liab - lt - st)]
e[, `:=`(other_assets = assets - cash - oca - fixed, other_liab = liab - lt - st)]

dat <- data.table(
  EIN2 = x$EIN2, year = x$TAX_YEAR,
  E = num(x$F9_09_EXP_TOT_TOT), R = num(x$F9_08_REV_TOT_TOT),
  assets_boy = b$assets, cash_boy = b$cash,
  contributions = num(x$F9_08_REV_CONTR_TOT),
  program_revenue = num(x$F9_08_REV_PROG_TOT_TOT),
  investment_income = num(x$F9_08_REV_OTH_INVEST_INCOME_TOT) + num(x$F9_08_REV_OTH_INVEST_BOND_TOT),
  gains_on_sales = num(x$F9_08_REV_OTH_SALE_GAIN_NET_TOT),
  program_expenses = num(x$F9_09_EXP_TOT_PROG),
  admin_expenses = num(x$F9_09_EXP_TOT_MGMT),
  fundraising_expenses = num(x$F9_09_EXP_TOT_FUNDR))
dat[, surplus := R - E]
dat[, other_revenue := R - contributions - program_revenue - investment_income - gains_on_sales]
dat[, unallocated_expenses := E - program_expenses - admin_expenses - fundraising_expenses]
for (v in c("cash", "oca", "fixed", "other_assets", "lt", "st", "other_liab", "assets", "liab", "na")) {
  dat[[paste0("d_", v)]] <- e[[v]] - b[[v]]
}
setnames(dat, "d_oca", "d_other_current")
rm(b, e, x); invisible(gc())
```

The lag is taken within organization on consecutive tax years. An organization
missing a year contributes no pair across that gap.

**Balance-sheet blocks enter as second differences.** This is the one place
where copying H5 would be wrong. H5 regressed a block's movement over the year
on the *level* of the surplus, which is coherent: the surplus level is what
lands on the balance sheet. Here the treatment is a *change* in revenue, so the
matching outcome is the change in the block's movement — how much less cash the
organization added this year than last. Pairing a level change with a revenue
change instead puts the whole surplus on the left-hand side and the shock on the
right, and the identity fails: an early version of this document did exactly
that and reported a net block movement of -0.028 against a surplus hit of
+0.027, the same quantity with opposite signs. Second differences make the two
sides comparable and the identity holds.

```{r lags}
LAGGED <- c("R", "E", "surplus", "contributions", "program_revenue",
            "investment_income", "gains_on_sales", "other_revenue",
            "program_expenses", "admin_expenses", "fundraising_expenses",
            "unallocated_expenses",
            # the block movements are lagged too, to form the second differences noted above
            "d_cash", "d_other_current", "d_fixed", "d_other_assets",
            "d_lt", "d_st", "d_other_liab", "d_assets", "d_na")

stopifnot(anyDuplicated(dat, by = c("EIN2", "year")) == 0L)
setorder(dat, EIN2, year)
dat[, paste0("lag_", LAGGED) := shift(.SD), by = EIN2, .SDcols = LAGGED]
dat[, lag_year := shift(year), by = EIN2]

# Year-over-year changes, defined only on consecutive pairs.
pairs <- dat[!is.na(lag_year) & year - lag_year == 1L]
for (v in LAGGED) pairs[[paste0("ch_", v)]] <- pairs[[v]] - pairs[[paste0("lag_", v)]]

# The denominator is fixed before the shock: prior-year total expenses.
pairs <- pairs[lag_E > 0 & lag_R > 0 & assets_boy > 0]
pairs[, loss := -ch_R / lag_E]                  # positive when revenue falls
pairs[, drop_share := -ch_R / lag_R]            # the shock definition, in revenue units
pairs[, shock := drop_share >= params$shock_threshold]
pairs[, months_cash_boy := cash_boy / (lag_E / 12)]

kable(pairs[, .(org_years = .N, organizations = uniqueN(EIN2),
                shocks = sum(shock), shock_rate = round(mean(shock), 3),
                median_drop_if_shocked = round(median(drop_share[shock]), 3)), by = year][order(year)],
      format.args = list(big.mark = ","),
      caption = "Consecutive-year pairs and the incidence of a revenue shock")
```

# 2. Sample and trimming

Outliers are handled by **dropping organization-years, not by clipping values**
— the same rule H5 arrived at. Winsorizing each component separately breaks the
identity the whole design rests on: the clipped parts stop summing to the
clipped total and the coefficients stop adding up. Pairs outside the 1st/99th
percentile of the loss ratio or of the change in total assets are dropped
within year, and whatever survives keeps its raw values.

```{r sample}
SCALED <- c("ch_R", "ch_E", "ch_surplus", "ch_contributions", "ch_program_revenue",
            "ch_investment_income", "ch_gains_on_sales", "ch_other_revenue",
            "ch_program_expenses", "ch_admin_expenses", "ch_fundraising_expenses",
            "ch_unallocated_expenses",
            "ch_d_cash", "ch_d_other_current", "ch_d_fixed", "ch_d_other_assets",
            "ch_d_lt", "ch_d_st", "ch_d_other_liab", "ch_d_assets", "ch_d_na",
            "lag_program_expenses", "lag_admin_expenses", "lag_fundraising_expenses",
            "lag_unallocated_expenses")

scale_vars <- function(z, trim = TRUE) {
  out <- copy(z)
  for (v in SCALED) out[[v]] <- out[[v]] / out$lag_E
  if (!trim) return(out)
  # `:=` by group assigns back to the right rows; a grouped select would return
  # the flags in group order and misalign them when years are interleaved.
  out[, keep := {
    ql <- quantile(loss, c(0.01, 0.99), na.rm = TRUE)
    qa <- quantile(ch_d_assets, c(0.01, 0.99), na.rm = TRUE)
    loss >= ql[[1]] & loss <= ql[[2]] & ch_d_assets >= qa[[1]] & ch_d_assets <= qa[[2]]
  }, by = year]
  out <- out[keep == TRUE]
  out[, keep := NULL]
  out[]
}

shocked <- function(years) scale_vars(pairs[year %in% years & shock == TRUE])
main_s <- shocked(2022:2023)
cat("shocked organization-years, 2022-2023:", format(nrow(main_s), big.mark = ","),
    " organizations:", format(uniqueN(main_s$EIN2), big.mark = ","), "\n")
```

Trimming is applied *after* selecting shocked organization-years, so the
percentiles are those of the shocked distribution, not of all filers.

# 3. Two ways to read "cents per dollar lost"

Two estimands answer slightly different questions, and this document reports
both because they can disagree.

* **Share of the average (the headline).** The total dollars that flowed
  through a channel across the shocked sample, divided by the total dollars
  lost. Because each channel is a component of the loss row by row, these
  shares sum to exactly 1. This is the quantity the figure shows.
* **Marginal response (OLS).** The slope of the channel on the size of the
  loss, within the shocked sample. This answers "a deeper shock is absorbed how
  differently", which is not the same as "the average shock is absorbed how".
  Standard errors are HC1.

```{r estimands}
ols <- function(y, W) {
  XtXi <- solve(crossprod(W))
  bhat <- drop(XtXi %*% crossprod(W, y))
  res <- drop(y - W %*% bhat)
  V <- XtXi %*% crossprod(W * res) %*% XtXi * nrow(W) / (nrow(W) - ncol(W))
  list(b = bhat, se = sqrt(diag(V)))
}

# share of the average: exact by construction, since the channels sum to the loss
share_of_avg <- function(s, channel, sign = 1) sign * mean(s[[channel]]) / mean(s$loss)

marginal <- function(s, channel, sign = 1) {
  f <- ols(sign * s[[channel]], cbind(intercept = 1, loss = s$loss))
  c(b = f$b[["loss"]], se = f$se[["loss"]])
}

channel_table <- function(s, channels, signs = rep(1, length(channels))) {
  rbindlist(lapply(seq_along(channels), function(i) {
    m <- marginal(s, channels[[i]], signs[[i]])
    data.table(channel = names(channels)[[i]],
               cents_per_dollar_lost = share_of_avg(s, channels[[i]], signs[[i]]),
               marginal = m[["b"]], se = m[["se"]])
  }))
}
```

# 4. Where the loss comes from

Which revenue lines fell. The five lines sum to the change in total revenue by
construction, so the shares sum to 1.

```{r sources}
REV <- c(Contributions = "ch_contributions", `Program revenue` = "ch_program_revenue",
         `Investment income` = "ch_investment_income", `Gains on asset sales` = "ch_gains_on_sales",
         `Other revenue` = "ch_other_revenue")
src <- channel_table(main_s, REV, signs = rep(-1, length(REV)))   # -1: report losses as positive
kable(src, digits = 3, caption = "Which revenue lines fell, 2022-2023 shocks (cents per dollar lost)")
cat("sum of shares:", fmt(sum(src$cents_per_dollar_lost)), "\n")
```

# 5. First-line absorption: expenses or the balance sheet

The identity `change in revenue = change in expenses + change in surplus` means
these two shares must sum to exactly 1. That is a check on the code, not a
finding — if it fails, something upstream is wrong.

```{r absorption}
ABS <- c(`Expense cuts` = "ch_E", `Worse surplus` = "ch_surplus")
absorb <- channel_table(main_s, ABS, signs = c(-1, -1))
kable(absorb, digits = 3,
      caption = "Of each dollar of lost revenue, how much comes out of expenses vs. the surplus")
cat("identity check, shares sum to:", fmt(sum(absorb$cents_per_dollar_lost)),
    " | marginals sum to:", fmt(sum(absorb$marginal)), "\n")
```

## Which expenses are cut

The four functional lines sum to the change in total expenses.

```{r expense-split}
EXP <- c(`Program services` = "ch_program_expenses", `Management and general` = "ch_admin_expenses",
         Fundraising = "ch_fundraising_expenses", Unallocated = "ch_unallocated_expenses")
excut <- channel_table(main_s, EXP, signs = rep(-1, length(EXP)))
kable(excut, digits = 3, caption = "Where the expense cut falls (cents per dollar lost)")
cat("sums to the expense-cut share:", fmt(sum(excut$cents_per_dollar_lost)),
    " vs. ", fmt(absorb[channel == "Expense cuts", cents_per_dollar_lost]), "\n")
```

Program services carry most of the cut simply because they are most of the
budget. The more informative quantity is the cut relative to each line's own
prior-year size, which says whether administration and fundraising are
protected or cut harder than programs.

```{r expense-intensity}
intens <- rbindlist(lapply(names(EXP), function(nm) {
  lagv <- paste0("lag_", sub("^ch_", "", EXP[[nm]]))
  data.table(line = nm,
             share_of_prior_budget = mean(main_s[[lagv]]),
             cut_as_pct_of_own_line = -mean(main_s[[EXP[[nm]]]]) / mean(main_s[[lagv]]))
}))
kable(intens, digits = 3,
      caption = "Cut intensity: each line's share of the prior-year budget, and the cut as a share of that line")
```

# 6. Where the balance sheet absorbs it

The worse surplus has to be drawn from somewhere, and the same identity as H5
applies: asset-side changes minus liability-side changes equal the change in
net assets. Signs are set so that **a positive number is a source of funds**: an
asset that fell, or a liability that rose.

```{r blocks}
BLOCKS <- c(Cash = "ch_d_cash", `Receivables, prepaid` = "ch_d_other_current",
            `Property and equipment` = "ch_d_fixed", `Investments, other assets` = "ch_d_other_assets",
            `Long-term debt` = "ch_d_lt", `Short-term liabilities` = "ch_d_st",
            `Other liabilities` = "ch_d_other_liab")
SIDE <- c(Cash = "asset", `Receivables, prepaid` = "asset", `Property and equipment` = "asset",
          `Investments, other assets` = "asset", `Long-term debt` = "liability",
          `Short-term liabilities` = "liability", `Other liabilities` = "liability")
sgn <- ifelse(SIDE[names(BLOCKS)] == "asset", -1, 1)
bs <- channel_table(main_s, BLOCKS, signs = sgn)
bs[, side := SIDE[channel]]
kable(bs[, .(channel, side, cents_per_dollar_lost, marginal, se)], digits = 3,
      caption = "Which balance-sheet blocks fund the shock (positive = a source of funds)")

ident <- data.table(
  blocks_net = sum(bs$cents_per_dollar_lost),
  net_assets_direct = -mean(main_s$ch_d_na) / mean(main_s$loss),
  worse_surplus = absorb[channel == "Worse surplus", cents_per_dollar_lost],
  n = nrow(main_s))
kable(ident, digits = 3,
      caption = "Identity check: blocks net vs. the direct net-assets change vs. the surplus hit")
```

The first two columns must agree: they are the same quantity computed two ways.
The third need not — net assets also move with unrealized investment gains,
prior-period adjustments, and donated capital, none of which pass through the
surplus. The gap between them is that non-operating drift, and it is worth
reading rather than dismissing.

# 7. Do reserves buffer the cut?

This is the question the memo singled out. The reserves dimension of the index
is interpreted as a buffer, which is a claim about behaviour *within*
organizations: an organization with more months of cash on hand should be able
to absorb a shock without cutting programs as deeply. If high- and low-reserve
organizations cut the same way, the dimension is measuring size or sector, not
resilience.

Reserves are measured at the beginning of the shock year — months of cash on
hand against the prior year's expenses — so the grouping is fixed before the
shock lands.

```{r reserves}
# Rank-based terciles: quantile breaks collapse when many organizations report
# the same months of cash (zero, most often), which a cut() on values cannot survive.
main_s[, reserve_group := factor(c("low", "middle", "high")[
  pmin(3L, 1L + floor(3 * (frank(months_cash_boy, ties.method = "first") - 1) / .N))],
  levels = c("low", "middle", "high"))]
res <- rbindlist(lapply(levels(main_s$reserve_group), function(g) {
  s <- main_s[reserve_group == g]
  data.table(reserves = g, n = nrow(s),
             median_months_cash = median(s$months_cash_boy),
             expense_cut = share_of_avg(s, "ch_E", -1),
             program_cut = share_of_avg(s, "ch_program_expenses", -1),
             cash_drawdown = share_of_avg(s, "ch_d_cash", -1),
             borrowing = share_of_avg(s, "ch_d_lt", 1) + share_of_avg(s, "ch_d_st", 1))
}))
kable(res, digits = 3,
      caption = "Absorption by beginning-of-year reserves (cents per dollar lost)")
```

**Holding the comparison together.** The table above compares different
organizations, so a reserve difference could be a size or sector difference. A
shocked organization rarely appears twice, so organization fixed effects are not
available on the shocked sample alone; instead the shock indicator is
interacted with reserves on *all* 2022-2023 pairs, which keeps organizations
that had a shock in one year and not the other in the comparison.

```{r reserves-interaction}
all_pairs <- scale_vars(pairs[year %in% 2022:2023])
all_pairs[, high_reserve := as.numeric(months_cash_boy > median(months_cash_boy))]
all_pairs[, big_loss := as.numeric(shock)]
W <- cbind(intercept = 1, big_loss = all_pairs$big_loss, high = all_pairs$high_reserve,
           big_x_high = all_pairs$big_loss * all_pairs$high_reserve)
inter <- rbindlist(lapply(c("ch_E", "ch_program_expenses", "ch_d_cash"), function(v) {
  f <- ols(all_pairs[[v]], W)
  data.table(outcome = v, shock_effect = f$b[["big_loss"]], se = f$se[["big_loss"]],
             extra_if_high_reserve = f$b[["big_x_high"]], se_interaction = f$se[["big_x_high"]])
}))
kable(inter, digits = 4,
      caption = "Shock effect and its interaction with high reserves, all 2022-2023 pairs")
```

# 8. Contrasts and robustness

```{r contrasts}
windows <- list(`2020 (COVID shock)` = 2020, `2021 (relief year)` = 2021,
                `2022` = 2022, `2023` = 2023)
by_year <- rbindlist(lapply(names(windows), function(nm) {
  s <- shocked(windows[[nm]])
  if (nrow(s) < 100) return(NULL)
  data.table(window = nm, n = nrow(s),
             expense_cut = share_of_avg(s, "ch_E", -1),
             worse_surplus = share_of_avg(s, "ch_surplus", -1),
             cash = share_of_avg(s, "ch_d_cash", -1),
             investments = share_of_avg(s, "ch_d_other_assets", -1),
             property = share_of_avg(s, "ch_d_fixed", -1),
             borrowing = share_of_avg(s, "ch_d_lt", 1) + share_of_avg(s, "ch_d_st", 1))
}))
kable(by_year, digits = 3, caption = "Absorption by shock year")
```

**Threshold sensitivity.** The 20% cut point is a convention. If the split is a
real behavioural pattern it should not turn on the threshold.

```{r thresholds}
thr <- rbindlist(lapply(c(0.10, 0.20, 0.30, 0.50), function(t) {
  s <- scale_vars(pairs[year %in% 2022:2023 & drop_share >= t])
  data.table(threshold = t, n = nrow(s),
             expense_cut = share_of_avg(s, "ch_E", -1),
             worse_surplus = share_of_avg(s, "ch_surplus", -1),
             cash = share_of_avg(s, "ch_d_cash", -1),
             borrowing = share_of_avg(s, "ch_d_lt", 1) + share_of_avg(s, "ch_d_st", 1))
}))
kable(thr, digits = 3, caption = "Sensitivity to the shock threshold, 2022-2023")
```

**Trimming.** The untrimmed sample keeps every shocked organization-year.

```{r trimming}
untrimmed <- scale_vars(pairs[year %in% 2022:2023 & shock == TRUE], trim = FALSE)
trimcmp <- rbind(
  data.table(sample = "trimmed", n = nrow(main_s),
             expense_cut = share_of_avg(main_s, "ch_E", -1),
             worse_surplus = share_of_avg(main_s, "ch_surplus", -1),
             cash = share_of_avg(main_s, "ch_d_cash", -1)),
  data.table(sample = "untrimmed", n = nrow(untrimmed),
             expense_cut = share_of_avg(untrimmed, "ch_E", -1),
             worse_surplus = share_of_avg(untrimmed, "ch_surplus", -1),
             cash = share_of_avg(untrimmed, "ch_d_cash", -1)))
kable(trimcmp, digits = 3, caption = "Trimmed vs. untrimmed")
```

**The gain side, for comparison.** The same computation on organizations whose
revenue *rose* by at least the same threshold says whether the absorption of a
loss is the mirror image of the deployment of a gain. H5 found that deficits are
funded disproportionately out of cash; the prediction here is the same
asymmetry.

```{r gains}
gain_s <- scale_vars(pairs[year %in% 2022:2023 & drop_share <= -params$shock_threshold])
gain_share <- function(v, sign = 1) sign * mean(gain_s[[v]]) / mean(-gain_s$loss)
mirror <- data.table(
  channel = c("Expenses", "Surplus", "Cash", "Investments, other assets", "Borrowing"),
  per_dollar_lost = c(share_of_avg(main_s, "ch_E", -1), share_of_avg(main_s, "ch_surplus", -1),
                      share_of_avg(main_s, "ch_d_cash", -1), share_of_avg(main_s, "ch_d_other_assets", -1),
                      share_of_avg(main_s, "ch_d_lt", 1) + share_of_avg(main_s, "ch_d_st", 1)),
  per_dollar_gained = c(gain_share("ch_E"), gain_share("ch_surplus"), gain_share("ch_d_cash"),
                        gain_share("ch_d_other_assets"),
                        gain_share("ch_d_lt", -1) + gain_share("ch_d_st", -1)))
kable(mirror, digits = 3, caption = "Losing a dollar vs. gaining one, 2022-2023")
```

# 9. The flow figure

```{r flow-figure, fig.width = 8.8, fig.height = 5.3, fig.retina = 2}
pal <- c(ink = "#1f2429", ink_mid = "#4d565f", ink_soft = "#6d7681",
         accent = "#bd5416", accent_ink = "#8f3f10", navy = "#153243", pumpkin = "#dc965a")

pick <- function(v) { v <- pmax(v, 0); sort(v[v > 0.005], decreasing = TRUE) }
sources <- pick(setNames(src$cents_per_dollar_lost, src$channel))
uses <- pick(c(`Expense cuts` = absorb[channel == "Expense cuts", cents_per_dollar_lost],
               Cash = bs[channel == "Cash", cents_per_dollar_lost],
               `Investments, other assets` = bs[channel == "Investments, other assets", cents_per_dollar_lost],
               `Receivables, prepaid` = bs[channel == "Receivables, prepaid", cents_per_dollar_lost],
               `Property and equipment` = bs[channel == "Property and equipment", cents_per_dollar_lost],
               Borrowing = bs[channel == "Long-term debt", cents_per_dollar_lost] +
                           bs[channel == "Short-term liabilities", cents_per_dollar_lost] +
                           bs[channel == "Other liabilities", cents_per_dollar_lost]))
GREY <- c("Expense cuts", "Borrowing")

flow_figure <- function(sources, uses, note, subtitle, title) {
  op <- par(mar = c(0.4, 0.4, 2.6, 0.4), bg = "white", family = "sans")
  on.exit(par(op))
  plot.new(); plot.window(xlim = c(0, 1), ylim = c(-0.14, 1.30))
  gap <- 0.045; centre <- 0.56
  lab_s <- sprintf("%s  %.2f", names(sources), sources)
  lab_u <- sprintf("%s  %.2f", names(uses), uses)
  # each side is drawn to its own total so the ribbons meet the bar exactly
  sources <- sources / sum(sources); uses <- uses / sum(uses)
  stack <- function(v) {
    y <- centre + (sum(v) + gap * (length(v) - 1)) / 2
    out <- lapply(v, function(h) { r <- c(top = y, bot = y - h); y <<- y - h - gap; r })
    names(out) <- names(v); out
  }
  ribbon <- function(x0, x1, t0, b0, t1, b1, col, alpha) {
    t <- seq(0, 1, length.out = 80); s <- t * t * (3 - 2 * t); xs <- x0 + (x1 - x0) * t
    polygon(c(xs, rev(xs)), c(t0 + (t1 - t0) * s, rev(b0 + (b1 - b0) * s)),
            col = adjustcolor(col, alpha), border = NA)
  }
  xs <- c(0.275, 0.325); xt <- c(0.495, 0.545); xu <- c(0.715, 0.765)
  srcb <- stack(sources); useb <- stack(uses)
  top <- centre + 0.5
  cur <- top
  for (i in seq_along(srcb)) {
    ribbon(xs[2], xt[1], srcb[[i]][["top"]], srcb[[i]][["bot"]], cur, cur - sources[[i]],
           pal[["navy"]], 0.34); cur <- cur - sources[[i]]
  }
  cur <- top
  for (i in seq_along(useb)) {
    ribbon(xt[2], xu[1], cur, cur - uses[[i]], useb[[i]][["top"]], useb[[i]][["bot"]],
           if (names(uses)[i] %in% GREY) pal[["ink_mid"]] else pal[["accent"]], 0.40)
    cur <- cur - uses[[i]]
  }
  for (i in seq_along(srcb)) {
    rect(xs[1], srcb[[i]][["bot"]], xs[2], srcb[[i]][["top"]], col = pal[["navy"]], border = NA)
    text(xs[1] - 0.014, mean(srcb[[i]]), lab_s[i], adj = c(1, 0.5), cex = 0.68, col = pal[["ink"]])
  }
  rect(xt[1], centre - 0.5, xt[2], top, col = pal[["accent"]], border = NA)
  text(mean(xt), top + 0.045, "$1 of lost revenue", adj = c(0.5, 0), cex = 0.78, font = 2,
       col = pal[["accent_ink"]])
  for (i in seq_along(useb)) {
    rect(xu[1], useb[[i]][["bot"]], xu[2], useb[[i]][["top"]],
         col = if (names(uses)[i] %in% GREY) pal[["ink_mid"]] else pal[["pumpkin"]], border = NA)
    text(xu[2] + 0.014, mean(useb[[i]]), lab_u[i], adj = c(0, 0.5), cex = 0.68, col = pal[["ink"]])
  }
  text(xs[1] - 0.014, centre - 0.72, "revenue lines that fell", adj = c(1, 0.5),
       cex = 0.6, col = pal[["ink_soft"]], font = 3)
  text(mean(xt), centre - 0.72, note, adj = c(0.5, 0.5), cex = 0.6, col = pal[["ink_soft"]], font = 3)
  text(xu[2] + 0.014, centre - 0.72, "what absorbs it", adj = c(0, 0.5),
       cex = 0.6, col = pal[["ink_soft"]], font = 3)
  mtext(title, side = 3, line = 1.1, adj = 0, cex = 1.0, font = 2, col = pal[["ink"]])
  mtext(subtitle, side = 3, line = 0.1, adj = 0, cex = 0.66, col = pal[["ink_soft"]])
}

flow_figure(sources, uses,
            sprintf("%.0f cents of each dollar is cut from expenses",
                    100 * absorb[channel == "Expense cuts", cents_per_dollar_lost]),
            paste0("cents per dollar of lost revenue  |  full-990 filers, December year-ends, ",
                   sprintf("revenue down %.0f%%+", params$shock_threshold * 100),
                   ", 2022-2023  |  n = ", format(nrow(main_s), big.mark = ",")),
            "Where a sudden revenue loss comes from, and what absorbs it")
```

Labels carry the true cents-per-dollar values; the bars are normalised to each
side's own total so the ribbons meet the centre exactly. Any channel that moved
the wrong way — a block that grew during a shock — is dropped from the picture
and stays in the tables above, where its sign is readable.

# 10. Findings

```{r findings, include = FALSE}
ec <- absorb[channel == "Expense cuts", cents_per_dollar_lost]
ws <- absorb[channel == "Worse surplus", cents_per_dollar_lost]
```

* **The split.** Of each dollar of revenue lost, `r fmt(ec)` is absorbed by
  cutting expenses within the same year and `r fmt(ws)` runs through the
  surplus onto the balance sheet.
* **Cash.** `r fmt(bs[channel == "Cash", cents_per_dollar_lost])` per dollar
  lost comes out of cash — the H5 counterpart to cash being the shock absorber
  on the deficit side.
* **Borrowing.** `r fmt(bs[channel == "Long-term debt", cents_per_dollar_lost] + bs[channel == "Short-term liabilities", cents_per_dollar_lost])`
  per dollar lost is funded by new identified debt.
* **Reserves.** The buffer question is answered by the table in section 7: the
  expense-cut share for low-reserve organizations is
  `r fmt(res[reserves == "low", expense_cut])` against
  `r fmt(res[reserves == "high", expense_cut])` for high-reserve organizations.

```{r save, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(sources = src, absorption = absorb, expense_split = excut,
               expense_intensity = intens, blocks = bs, identity = ident,
               reserves = res, reserves_interaction = inter, by_year = by_year,
               thresholds = thr, trimming = trimcmp, mirror = mirror,
               n = nrow(main_s), threshold = params$shock_threshold,
               incidence = pairs[, .(.N, shocks = sum(shock)), by = year][order(year)]),
          params$results_file)
}
```

```{r session}
sessionInfo()
```
