---
title: "Reweighting the Fiscal Health Index"
subtitle: "How much do scores and rankings move when weights follow predictive value?"
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
    css: walkthrough.css
params:
  panel_file: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "PANEL", "panel-2019-2024-imputed.rds")
  results_file: ""
  seed: 2023
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 130, knitr.kable.NA = "")
```

# Purpose

`dimension-weighting.Rmd` found that reserves and performance predict different
outcomes, that leverage predicts none of them, and that weights proportional to
incremental AUC would be about 0.59 performance and 0.41 reserves rather than a
third each. This document rebuilds the index under those weights and asks the
practical question: **who moves, and does the reweighted score predict better?**

Three schemes, all over the same dimension scores:

| Scheme | Reserves | Performance | Leverage |
|---|---|---|---|
| Equal (current default) | 1/3 | 1/3 | 1/3 |
| Incremental AUC | 0.41 | 0.59 | 0 |
| Incremental AUC, excluding the cash outcome | 0.12 | 0.88 | 0 |

The third scheme drops the cash-drawdown outcome from the weighting exercise,
because an organization can only lose a month of cash if it had one, which
flatters reserves. It is shown as a sensitivity, not a recommendation.

Note on the package default: `fiscal_weights(include_standalone = TRUE)` also
gives equal weight to the standalone measures, so the shipped default would
spread weight across six measures, not three. The comparison here is against
the three-dimension version recommended in `dimension-discovery-2023.Rmd`.

```{r packages}
suppressMessages({
  library(data.table)
  library(bit64)
  library(knitr)
  library(splines)
})
source("dimension-scores.R")
set.seed(params$seed)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

# 1. Scores and index versions

```{r scores}
d <- as.data.table(readRDS(params$panel_file))
d <- d[RETURN_TYPE == "990" & substr(TAX_PERIOD_END_DATE, 6, 7) == "12" & TAX_YEAR %in% 2019:2024]
keep <- d[TAX_YEAR %in% 2019:2023, .N, by = EIN2][N == 5, EIN2]
in2024 <- unique(d[TAX_YEAR == 2024, EIN2])
d <- d[EIN2 %chin% keep]

WEIGHTS <- list(
  equal = c(reserves = 1/3, performance = 1/3, leverage = 1/3),
  incremental = c(reserves = 0.415, performance = 0.585, leverage = 0),
  performance_dominant = c(reserves = 0.121, performance = 0.879, leverage = 0))

s23 <- dimension_scores(d[TAX_YEAR == 2023], seed = params$seed)
for (nm in names(WEIGHTS)) s23[[paste0("index_", nm)]] <- composite_index(s23, WEIGHTS[[nm]])
kable(s23[, .(organizations = .N,
              mean_reserves = mean(reserves), mean_performance = mean(performance),
              mean_leverage = mean(leverage))], digits = 3, format.args = list(big.mark = ","))
```

# 2. How much do rankings move?

```{r rank-correlation}
idx <- paste0("index_", names(WEIGHTS))
kable(round(cor(s23[, ..idx], method = "spearman"), 4), caption = "Rank correlation between index versions, 2023")
```

```{r movement}
s23[, `:=`(decile_equal = cut(index_equal, breaks = seq(0, 100, 10), labels = 1:10, include.lowest = TRUE),
           decile_incr = cut(index_incremental, breaks = seq(0, 100, 10), labels = 1:10, include.lowest = TRUE))]
s23[, shift_pct := index_incremental - index_equal]
kable(s23[, .(`mean absolute move (percentile points)` = mean(abs(shift_pct)),
              `median absolute move` = median(abs(shift_pct)),
              `moved at least one decile` = mean(abs(as.integer(decile_incr) - as.integer(decile_equal)) >= 1),
              `moved at least two deciles` = mean(abs(as.integer(decile_incr) - as.integer(decile_equal)) >= 2),
              `moved at least three deciles` = mean(abs(as.integer(decile_incr) - as.integer(decile_equal)) >= 3))],
      digits = 3, caption = "Equal weights vs. incremental weights")
kable(dcast(s23[, .N, by = .(decile_equal, decile_incr)], decile_equal ~ decile_incr,
            value.var = "N", fill = 0),
      caption = "Decile under equal weights (rows) vs. incremental weights (columns)",
      format.args = list(big.mark = ","))
```

**Who moves?** The organizations that fall are the ones the old weighting
rewarded for a clean balance sheet despite weak operations, and the reverse.

```{r movers}
s23[, mover := fifelse(shift_pct >= quantile(shift_pct, 0.99), "biggest risers",
               fifelse(shift_pct <= quantile(shift_pct, 0.01), "biggest fallers", "middle"))]
kable(s23[, .(organizations = .N, mean_reserves = mean(reserves), mean_performance = mean(performance),
              mean_leverage = mean(leverage), median_expenses = median(E),
              median_surplus_margin = median(surplus_per_expense),
              median_cash_months = median(cash_months)), by = mover][order(mover)],
      digits = 3, format.args = list(big.mark = ","))
```

# 3. Does the reweighted index predict better?

The index is rebuilt on 2021 scores and used to predict 2022-2023, the same
outcomes as the weighting analysis.

```{r validation}
s21 <- dimension_scores(d[TAX_YEAR == 2021], seed = params$seed)
for (nm in names(WEIGHTS)) s21[[paste0("index_", nm)]] <- composite_index(s21, WEIGHTS[[nm]])

flows <- d[TAX_YEAR %in% 2021:2023, .(EIN2, year = TAX_YEAR,
                                      R = as.numeric(F9_08_REV_TOT_TOT),
                                      E = as.numeric(F9_09_EXP_TOT_TOT),
                                      na = as.numeric(F9_10_NAFB_TOT_EOY),
                                      cash = as.numeric(F9_10_ASSET_CASH_EOY) +
                                             as.numeric(F9_10_ASSET_SAVING_EOY))]
for (cl in c("R", "E", "na", "cash")) flows[[cl]][is.na(flows[[cl]])] <- 0
o <- dcast(flows, EIN2 ~ year, value.var = c("R", "E", "na", "cash"))
setnames(o, gsub("_(\\d{4})$", "\\1", names(o)))
o <- o[complete.cases(o)]
o[, `:=`(both_years_deficit = as.numeric(R2022 < E2022 & R2023 < E2023),
         any_deficit = as.numeric(R2022 < E2022 | R2023 < E2023),
         na_fell = as.numeric(na2023 < na2021),
         cash_fell_month = as.numeric(12 * (cash2023 - cash2021) / pmax(E2021, 1) < -1),
         no2024 = as.numeric(!(EIN2 %chin% in2024)))]
m <- merge(o, s21[, c("EIN2", idx, "E"), with = FALSE], by = "EIN2")

auc <- function(score, y) {
  r <- rank(score); n1 <- as.numeric(sum(y == 1)); n0 <- as.numeric(sum(y == 0))
  (sum(r[y == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
OUT <- c("both_years_deficit", "any_deficit", "na_fell", "cash_fell_month", "no2024")
val <- rbindlist(lapply(OUT, function(nm) {
  y <- m[[nm]]
  # a low index should mean trouble, so AUC is computed on the reversed score
  data.table(outcome = nm, rate = mean(y),
             equal = auc(-m$index_equal, y),
             incremental = auc(-m$index_incremental, y),
             performance_dominant = auc(-m$index_performance_dominant, y))
}))
val[, `:=`(gain_incremental = incremental - equal,
           gain_performance_dominant = performance_dominant - equal)]
kable(val, digits = 4, caption = "AUC of each index version (2021 index, 2022-2023 outcomes)")
```

**Distress concentration.** A practical read: how many of the organizations
that ran deficits in both years sit in the bottom decile of each index?

```{r bottom-decile}
bd <- rbindlist(lapply(names(WEIGHTS), function(nm) {
  col <- paste0("index_", nm)
  bottom <- m[[col]] <= 10
  data.table(scheme = nm,
             pct_of_both_year_deficits_in_bottom_decile = mean(bottom[m$both_years_deficit == 1]),
             deficit_rate_in_bottom_decile = mean(m$both_years_deficit[bottom]),
             deficit_rate_overall = mean(m$both_years_deficit))
}))
kable(bd, digits = 3)
```

# 4. Findings

```{r findings, include = FALSE}
rk <- cor(s23$index_equal, s23$index_incremental, method = "spearman")
mv <- mean(abs(as.integer(s23$decile_incr) - as.integer(s23$decile_equal)) >= 1)
```

* **Rankings move less than the weights do.** The equal-weighted and
  incremental-weighted indices correlate at `r fmt(rk, 3)` (Spearman), and
  `r fmt(100 * mv, 1)`% of organizations change at least one decile.
* **Prediction improves where performance matters** and is unchanged or worse
  for the cash outcome; the table in section 3 gives the sizes.
* **The bottom decile catches more distress** under the reweighted index, which
  is the number that matters for triage.

```{r save, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(weights = WEIGHTS,
               rank_cor = cor(s23[, ..idx], method = "spearman"),
               movement = s23[, .(mean_abs = mean(abs(shift_pct)), median_abs = median(abs(shift_pct)),
                                  one_decile = mean(abs(as.integer(decile_incr) - as.integer(decile_equal)) >= 1),
                                  two_deciles = mean(abs(as.integer(decile_incr) - as.integer(decile_equal)) >= 2),
                                  three_deciles = mean(abs(as.integer(decile_incr) - as.integer(decile_equal)) >= 3))],
               movers = s23[, .(n = .N, reserves = mean(reserves), performance = mean(performance),
                                leverage = mean(leverage), median_E = median(E),
                                median_margin = median(surplus_per_expense),
                                median_cash_months = median(cash_months)), by = mover],
               validation = val, bottom_decile = bd, n = nrow(m)), params$results_file)
}
```

```{r session}
sessionInfo()
```
