---
title: "Matching-Gap Indicator, Tax Year 2023"
subtitle: "Debt beyond what an organization's asset mix supports"
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
    css: walkthrough.css
params:
  data_dir: !r file.path(Sys.getenv("FISCAL_DATA_DIR", unset = file.path(path.expand("~"), "Documents", "fiscal-data")), "efile")
  efile_file: "efile-2023-bmf-dedup.rds"
  out_file: "matching-gap-2023.rds"
  results_file: ""
  seed: 2023
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 120, knitr.kable.NA = "")
```

# Purpose

`matching-hypotheses-2023.Rmd` confirmed two financing relationships that
survive both arithmetic and size structure:

* **H1:** long-term debt tracks fixed assets (elasticity 0.51);
* **H2:** short-term liabilities track non-cash current assets (elasticity
  0.20).

This document turns them into an indicator. For each organization, the H1 and
H2 models predict how much debt its asset mix and size support. The **matching
gap** is actual debt minus that benchmark, as a share of total assets. A
positive gap means more debt than similar organizations with the same assets
carry.

The construction and the validation expectations below were written into the
memo before the indicator was computed.

```{r packages}
library(knitr)
source("matching-gap.R")
set.seed(params$seed)
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

# 1. Plan (as written in advance)

**Construction.**

1. Refit the H1 and H2 two-part models on end-of-year lines: logistic for any
   debt, OLS on log amount with Duan's smearing factor. Same predictors and
   size controls as the hypothesis tests.
2. Predicted debt = P(any) x E(amount | any), for long-term debt and for
   short-term liabilities.
3. `gap_long` and `gap_short` = (actual - predicted) / total assets;
   `gap_total` = their sum.

**Validation, one year ahead within the filing.** Compute the gaps from
beginning-of-year lines using the end-of-year coefficients. Outcomes over the
year: an operating deficit, a decline in unrestricted net assets, and cash
falling by more than one month of expenses.

**Expected.** Each gap raises the probability of each outcome, net of
beginning-of-year `debt_assets` and size, with a positive coefficient whose
robust 99% interval excludes zero, and an AUC gain over `debt_assets` alone.
`gap_short` should predict the cash drawdown most strongly.

# 2. Data

```{r data}
efile <- readRDS(file.path(params$data_dir, params$efile_file))
eoy <- gap_lines(efile, "EOY")
boy <- gap_lines(efile, "BOY")
rm(efile); invisible(gc())
stopifnot(nrow(eoy) == nrow(boy))

fit_rows <- eoy$E > 0 & eoy$A > 0
val_rows <- eoy$E > 0 & eoy$A > 0 & boy$A > 0
data.frame(full_990_filers = nrow(eoy), benchmark_sample = sum(fit_rows),
           validation_sample = sum(val_rows))
```

# 3. Benchmarks

```{r benchmark}
mg <- matching_gap(eoy[fit_rows, ])
gaps <- mg$gaps
data.frame(model = c("long-term debt", "short-term liabilities"),
           filers = sapply(mg$fits, `[[`, "n"),
           with_positive_amount = sapply(mg$fits, `[[`, "n_positive"),
           logistic_converged = sapply(mg$fits, `[[`, "converged"),
           smearing_factor = sapply(mg$fits, `[[`, "smear"), row.names = NULL)
```

**Calibration.** Predicted dollars should match actual dollars on average,
overall and across the range of predictions.

```{r calibration}
m_fit <- eoy[fit_rows, ]
calib <- function(actual, pred) {
  dec <- cut(rank(pred, ties.method = "first"), 10, labels = FALSE)
  data.frame(decile = 1:10,
             mean_predicted = tapply(pred, dec, mean),
             mean_actual = tapply(actual, dec, mean),
             median_predicted = tapply(pred, dec, median),
             median_actual = tapply(actual, dec, median))
}
kable(data.frame(line = c("long-term debt", "short-term liabilities"),
                 total_actual = c(sum(m_fit$lt), sum(m_fit$st)),
                 total_predicted = c(sum(gaps$pred_long), sum(gaps$pred_short)),
                 ratio = c(sum(gaps$pred_long) / sum(m_fit$lt), sum(gaps$pred_short) / sum(m_fit$st))),
      digits = 3, format.args = list(big.mark = ","))
kable(calib(m_fit$lt, gaps$pred_long), digits = 0, format.args = list(big.mark = ","),
      caption = "Long-term debt by decile of prediction (dollars)")
kable(calib(m_fit$st, gaps$pred_short), digits = 0, format.args = list(big.mark = ","),
      caption = "Short-term liabilities by decile of prediction (dollars)")
```

# 4. The indicator

```{r distribution}
qs <- c(0.01, 0.05, 0.25, 0.5, 0.75, 0.95, 0.99)
kable(t(sapply(gaps[c("gap_long", "gap_short", "gap_total", "debt_assets")], quantile, qs)),
      digits = 3, caption = "Quantiles, end of year 2023")
kable(data.frame(
  indicator = c("gap_long", "gap_short", "gap_total"),
  share_positive = sapply(gaps[c("gap_long", "gap_short", "gap_total")], function(x) mean(x > 0)),
  spearman_with_debt_assets = sapply(gaps[c("gap_long", "gap_short", "gap_total")],
                                     function(x) cor(x, gaps$debt_assets, method = "spearman")),
  row.names = NULL), digits = 3)
```

Most organizations have no long-term debt, so `gap_long` is slightly negative
for them (they carry less than the average organization with the same assets)
and strongly positive for the minority who borrow beyond their property.

```{r save-indicator}
if (nzchar(params$out_file)) {
  out_path <- file.path(params$data_dir, params$out_file)
  saveRDS(gaps, out_path)
  out_path
}
```

# 5. Validation, one year ahead

```{r validation-data}
b <- boy[val_rows, ]
e <- eoy[val_rows, ]
gb <- apply_gap(b, mg$fits, mg$basis)
size_b <- gap_design(b, mg$basis)$X[, c(paste0("size_E", 1:4), paste0("size_other", 1:4), "no_other")]

wz <- function(x) {
  q <- quantile(x, c(0.01, 0.99), na.rm = TRUE)
  as.numeric(scale(pmin(pmax(x, q[[1]]), q[[2]])))
}
v <- data.frame(
  deficit     = as.numeric(e$R < e$E),
  una_decline = as.numeric(e$una < b$una),
  cash_drop   = as.numeric(e$cash - b$cash < -e$E / 12),
  debt_assets = wz(gb$debt_assets),
  gap_long    = wz(gb$gap_long),
  gap_short   = wz(gb$gap_short),
  gap_total   = wz(gb$gap_total))
kable(data.frame(outcome = c("deficit", "una_decline", "cash_drop"),
                 rate = colMeans(v[c("deficit", "una_decline", "cash_drop")]), row.names = NULL),
      digits = 3, caption = "Outcome base rates")
kable(round(cor(v[c("debt_assets", "gap_long", "gap_short", "gap_total")]), 2),
      caption = "Correlations among beginning-of-year predictors (winsorized, standardized)")
```

Each model is a linear probability model with size splines and HC1 standard
errors. Coefficients are per standard deviation of the predictor.

```{r validation-models}
lpm <- function(y, W) {
  XtXi <- solve(crossprod(W))
  bhat <- drop(XtXi %*% crossprod(W, y))
  fitted <- drop(W %*% bhat)
  res <- y - fitted
  V <- XtXi %*% crossprod(W * res) %*% XtXi * nrow(W) / (nrow(W) - ncol(W))
  list(coef = data.frame(term = colnames(W), estimate = bhat, se = sqrt(diag(V)), row.names = NULL),
       fitted = fitted)
}
auc <- function(score, y) {
  r <- rank(score)
  n1 <- as.numeric(sum(y == 1)); n0 <- as.numeric(sum(y == 0))   # doubles: n1 * n0 overflows integers
  (sum(r[y == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

specs <- list(
  `debt_assets only`             = c("debt_assets"),
  `debt_assets + gap_long + gap_short` = c("debt_assets", "gap_long", "gap_short"),
  `debt_assets + gap_total`      = c("debt_assets", "gap_total"))

validation <- do.call(rbind, lapply(c("deficit", "una_decline", "cash_drop"), function(o) {
  do.call(rbind, lapply(names(specs), function(s) {
    W <- cbind(size_b, as.matrix(v[specs[[s]]]))
    f <- lpm(v[[o]], W)
    cf <- f$coef[f$coef$term %in% c("debt_assets", "gap_long", "gap_short", "gap_total"), ]
    cbind(outcome = o, model = s, auc = auc(f$fitted, v[[o]]), cf,
          lower_99 = cf$estimate - qnorm(0.995) * cf$se, upper_99 = cf$estimate + qnorm(0.995) * cf$se)
  }))
}))
kable(validation, digits = 4, row.names = FALSE)

auc_gain <- do.call(rbind, lapply(split(validation, validation$outcome), function(x) {
  a <- tapply(x$auc, x$model, `[`, 1)
  data.frame(outcome = x$outcome[1], auc_debt_assets = a[["debt_assets only"]],
             auc_with_gaps = a[["debt_assets + gap_long + gap_short"]],
             gain_components = a[["debt_assets + gap_long + gap_short"]] - a[["debt_assets only"]],
             gain_total = a[["debt_assets + gap_total"]] - a[["debt_assets only"]])
}))
kable(auc_gain, digits = 4, row.names = FALSE, caption = "AUC with and without the gaps")
```

**Does the gap separate risk within a leverage level?** Outcome rates by
`gap_total` quintile, within terciles of `debt_assets`.

```{r quintiles}
tercile <- cut(rank(gb$debt_assets, ties.method = "first"), 3, labels = c("low", "middle", "high"))
quint <- ave(gb$gap_total, tercile, FUN = function(x) cut(rank(x, ties.method = "first"), 5, labels = FALSE))
rates <- aggregate(v[c("deficit", "una_decline", "cash_drop")],
                   by = list(debt_assets_tercile = tercile, gap_total_quintile = quint), FUN = mean)
kable(rates[order(rates$debt_assets_tercile, rates$gap_total_quintile), ], digits = 3, row.names = FALSE)
```

# 6. Verdicts against the plan

```{r verdicts}
with_gaps <- validation[validation$model == "debt_assets + gap_long + gap_short" &
                        validation$term %in% c("gap_long", "gap_short"), ]
verdicts <- data.frame(
  outcome = with_gaps$outcome, gap = with_gaps$term,
  estimate_per_sd = with_gaps$estimate, lower_99 = with_gaps$lower_99,
  meets_expectation = with_gaps$estimate > 0 & with_gaps$lower_99 > 0, row.names = NULL)
verdicts <- merge(verdicts, auc_gain[c("outcome", "gain_components")], by = "outcome")
kable(verdicts, digits = 4, row.names = FALSE)

short_cash <- with_gaps$estimate[with_gaps$outcome == "cash_drop" & with_gaps$term == "gap_short"]
short_strongest <- short_cash > max(with_gaps$estimate[with_gaps$term == "gap_short" & with_gaps$outcome != "cash_drop"])
```

`gap_short` `r if (short_strongest) "does" else "does not"` predict the cash
drawdown more strongly than it predicts the other two outcomes, as expected.

These are one-year associations within a filing. Organizations may carry debt
beyond their assets for reasons the benchmark cannot see: construction
projects in progress, leases recorded as liabilities, or deferred revenue from
multi-year contracts, which is a short-term liability but not a sign of strain.

# 7. Checks added after the first run

Sections 3-6 follow the plan. The first run exposed two problems the plan did
not anticipate, so this section adds checks. They are labeled as additions and
do not replace the planned results above.

1. **The two-part benchmark is badly calibrated.** Predicted totals are more
   than twice actual totals, and the top prediction decile overshoots. Log
   residuals are so dispersed that Duan's constant smearing factor is near 4,
   and it misallocates across the range. The standard alternative is **Poisson
   pseudo-maximum likelihood (PPML)** on dollars with zeros included. It
   estimates E[debt | assets] directly, with no retransformation.
2. **The cash-drawdown outcome has a floor.** Cash cannot fall by a month of
   expenses unless an organization starts the year with at least that much.
   Organizations with very negative gaps tend to hold few assets relative to
   expenses, and therefore little cash, so part of the gap-drawdown association
   could be mechanical. The added check controls for beginning-of-year months
   of cash and assets per dollar of expenses.

```{r ppml}
mg_p <- matching_gap(eoy[fit_rows, ], "ppml")
gaps_p <- mg_p$gaps
data.frame(model = c("long-term debt", "short-term liabilities"),
           converged = sapply(mg_p$fits, `[[`, "converged"), row.names = NULL)
kable(data.frame(line = c("long-term debt", "short-term liabilities"),
                 total_actual = c(sum(m_fit$lt), sum(m_fit$st)),
                 total_predicted = c(sum(gaps_p$pred_long), sum(gaps_p$pred_short)),
                 ratio = c(sum(gaps_p$pred_long) / sum(m_fit$lt), sum(gaps_p$pred_short) / sum(m_fit$st))),
      digits = 3, format.args = list(big.mark = ","), caption = "PPML calibration, totals")
kable(calib(m_fit$lt, gaps_p$pred_long), digits = 0, format.args = list(big.mark = ","),
      caption = "PPML: long-term debt by decile of prediction")
kable(calib(m_fit$st, gaps_p$pred_short), digits = 0, format.args = list(big.mark = ","),
      caption = "PPML: short-term liabilities by decile of prediction")
kable(t(sapply(gaps_p[c("gap_long", "gap_short", "gap_total")], quantile, qs)), digits = 3,
      caption = "PPML gaps, quantiles")
```

```{r validation-grid}
floor_controls <- cbind(size_b,
                        ns(log1p(12 * b$cash / b$E), df = 4),
                        ns(log(b$A / b$E), df = 4))
colnames(floor_controls) <- c(colnames(size_b), paste0("cash_months", 1:4), paste0("asset_intensity", 1:4))
gb_p <- apply_gap(b, mg_p$fits, mg_p$basis)

run_validation <- function(g, controls) {
  vv <- data.frame(v[c("deficit", "una_decline", "cash_drop")],
                   debt_assets = wz(g$debt_assets), gap_long = wz(g$gap_long),
                   gap_short = wz(g$gap_short), gap_total = wz(g$gap_total))
  do.call(rbind, lapply(c("deficit", "una_decline", "cash_drop"), function(o) {
    y <- vv[[o]]
    base <- lpm(y, cbind(controls, debt_assets = vv$debt_assets))
    comp <- lpm(y, cbind(controls, as.matrix(vv[c("debt_assets", "gap_long", "gap_short")])))
    tot  <- lpm(y, cbind(controls, as.matrix(vv[c("debt_assets", "gap_total")])))
    pick <- function(f, t) f$coef[f$coef$term == t, ]
    lo <- function(r) r$estimate - qnorm(0.995) * r$se
    data.frame(outcome = o,
               gap_long = pick(comp, "gap_long")$estimate, gap_long_lower99 = lo(pick(comp, "gap_long")),
               gap_short = pick(comp, "gap_short")$estimate, gap_short_lower99 = lo(pick(comp, "gap_short")),
               gap_total = pick(tot, "gap_total")$estimate, gap_total_lower99 = lo(pick(tot, "gap_total")),
               auc_debt_assets = auc(base$fitted, y),
               auc_gain_components = auc(comp$fitted, y) - auc(base$fitted, y),
               auc_gain_total = auc(tot$fitted, y) - auc(base$fitted, y))
  }))
}

grid <- rbind(
  cbind(benchmark = "two-part (planned)", controls = "planned", run_validation(gb, size_b)),
  cbind(benchmark = "two-part (planned)", controls = "+ cash months, asset intensity", run_validation(gb, floor_controls)),
  cbind(benchmark = "PPML", controls = "planned", run_validation(gb_p, size_b)),
  cbind(benchmark = "PPML", controls = "+ cash months, asset intensity", run_validation(gb_p, floor_controls)))
kable(grid, digits = 4, row.names = FALSE,
      caption = "Gap coefficients (per SD) and AUC gains, by benchmark and controls")
```

# 8. What the checks show

```{r findings-values, include = FALSE}
gr <- function(bm, ct, o, col) grid[grid$benchmark == bm & grid$controls == ct & grid$outcome == o, col]
P <- "planned"; F <- "+ cash months, asset intensity"; TP <- "two-part (planned)"; PP <- "PPML"
```

* **Cash drawdown.** With the planned controls, `gap_total` (two-part) raises
  the drawdown probability by `r fmt(gr(TP, P, "cash_drop", "gap_total"), 3)` per
  SD and adds `r fmt(gr(TP, P, "cash_drop", "auc_gain_total"), 3)` AUC. With
  beginning-of-year cash months and asset intensity controlled, the coefficient
  is `r fmt(gr(TP, F, "cash_drop", "gap_total"), 3)` and the AUC gain
  `r fmt(gr(TP, F, "cash_drop", "auc_gain_total"), 4)`. The PPML version gives
  `r fmt(gr(PP, F, "cash_drop", "gap_total"), 3)`. The planned association was
  the cash floor.
* **Deficits and net-asset declines.** `gap_long` stays positive under every
  benchmark and control set. With PPML and the added controls it is
  (`r fmt(gr(PP, F, "deficit", "gap_long"), 3)` for deficits,
  `r fmt(gr(PP, F, "una_decline", "gap_long"), 3)` for declines, per SD). The
  largest AUC gain in those rows is
  `r fmt(max(grid$auc_gain_components[grid$outcome != "cash_drop"]), 4)`.
* **Short-term gap.** `gap_short` is negative for deficits and net-asset
  declines in every row: short-term liabilities beyond what current assets
  support go with better operating results, not worse.
* **Calibration.** PPML matches total dollars by construction and tracks the
  top decile closely, but over-predicts the middle deciles, where most
  organizations carry no long-term debt.

```{r save-indicator-final}
if (nzchar(params$out_file)) {
  out <- cbind(gaps, setNames(gaps_p[c("pred_long", "pred_short", "gap_long", "gap_short", "gap_total")],
                              paste0(c("pred_long", "pred_short", "gap_long", "gap_short", "gap_total"), "_ppml")))
  saveRDS(out, file.path(params$data_dir, params$out_file))
}
```

```{r save-results, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(fits = mg$fits, calibration_long = calib(m_fit$lt, gaps$pred_long),
               calibration_short = calib(m_fit$st, gaps$pred_short),
               quantiles = t(sapply(gaps[c("gap_long", "gap_short", "gap_total", "debt_assets")], quantile, qs)),
               share_positive = sapply(gaps[c("gap_long", "gap_short", "gap_total")], function(x) mean(x > 0)),
               spearman = sapply(gaps[c("gap_long", "gap_short", "gap_total")],
                                 function(x) cor(x, gaps$debt_assets, method = "spearman")),
               totals = c(lt = sum(m_fit$lt), lt_pred = sum(gaps$pred_long),
                          st = sum(m_fit$st), st_pred = sum(gaps$pred_short)),
               base_rates = colMeans(v[c("deficit", "una_decline", "cash_drop")]),
               predictor_cor = cor(v[c("debt_assets", "gap_long", "gap_short", "gap_total")]),
               validation = validation, auc_gain = auc_gain, rates = rates, verdicts = verdicts,
               short_strongest = short_strongest, n_fit = sum(fit_rows), n_val = sum(val_rows),
               ppml = list(converged = sapply(mg_p$fits, `[[`, "converged"),
                           totals = c(lt_pred = sum(gaps_p$pred_long), st_pred = sum(gaps_p$pred_short)),
                           calibration_long = calib(m_fit$lt, gaps_p$pred_long),
                           calibration_short = calib(m_fit$st, gaps_p$pred_short),
                           quantiles = t(sapply(gaps_p[c("gap_long", "gap_short", "gap_total")], quantile, qs)),
                           spearman = sapply(gaps_p[c("gap_long", "gap_short", "gap_total")],
                                             function(x) cor(x, gaps_p$debt_assets, method = "spearman"))),
               grid = grid),
          params$results_file)
}
```

```{r session}
sessionInfo()
```
