---
title: "Asset-Liability Matching and Pecking Order, Tax Year 2023"
subtitle: "Pre-specified tests of H1-H3 from the ratio-correlation memo"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: kable
params:
  data_dir: !r file.path(tools::R_user_dir("fiscal", "cache"), "efile")
  efile_file: "efile-2023-bmf-dedup.rds"
  pkg_dir: "../.."
  seed: 2023
  perms: 200
  results_file: ""
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
options(width = 120)
```

# Purpose

`vignette("ratio-correlation-artifacts")` showed that the fiscal health
dimensions are largely predicted by how the ratios are built, and that a few
pairwise relationships survive a conditional construction null. The strongest
concern how organizations finance their assets. This document tests three of
them directly, as regressions on 990 dollar lines rather than as correlations
between ratios:

* **H1 Long-term matching.** Fixed assets are financed with long-term debt.
* **H1a Subsector.** Matching is stronger in capital-intensive subsectors
  (hospitals, universities, housing).
* **H2 Short-term matching.** Non-cash current assets are financed with
  short-term liabilities.
* **H3 Pecking order.** Organizations with more cash carry less long-term debt.

The specification, evidence rule, and support/rejection criteria were fixed in
the memo before any model below was estimated. They are repeated here and not
changed.

```{r packages}
library(knitr)
library(splines)
if (file.exists(file.path(params$pkg_dir, "DESCRIPTION"))) {
  pkgload::load_all(params$pkg_dir, quiet = TRUE)
} else {
  library(fiscal)
}
set.seed(params$seed)
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

# 1. Pre-specification

| | Outcome | Coefficient | Support if | Reject if |
|---|---|---|---|---|
| H1 | Long-term debt: any; log amount | log fixed assets | > 0 in both parts, and larger for long-term debt than for short-term liabilities (log amount) | <= 0 in either part, or not larger than in the short-term equation |
| H1a | Long-term debt: log amount | log fixed assets x capital-intensive subsector | > 0 | <= 0 |
| H2 | Short-term liabilities: any; log amount | log non-cash current assets | > 0 in both parts, and larger than in the long-term equation (log amount) | <= 0 in either part, or not larger than in the long-term equation |
| H3 | Long-term debt: any; log amount | log cash | < 0 in both parts | > 0 in both parts (supports the rival collateral account) |

* **Lines.** Long-term debt = tax-exempt bonds + secured mortgages and notes +
  unsecured notes (Part X lines 20, 23, 24). Short-term liabilities = accounts
  payable + grants payable + deferred revenue (17-19). Cash = cash + savings
  (1-2). Non-cash current assets = pledges + accounts receivable + inventories
  + prepaid expenses (3, 4, 8, 9). Fixed assets = land, buildings, and
  equipment, net (10c). Other assets = total assets minus those four, floored
  at 0.
* **Models.** Linear probability model for any positive amount; OLS on log
  amount among filers with a positive amount. Predictors enter as log(1 + x)
  plus an indicator for x > 0. Size controls: 4-df natural splines of log
  expenses and log(1 + other assets), plus an indicator for zero other assets.
* **Evidence rule.** A coefficient counts only if it is in the predicted
  direction and beyond the 0.5th or 99.5th percentile of `r params$perms`
  refits in which the predictor (its log and indicator together) is shuffled
  within 400 cells of expenses x assets per dollar of expenses. Cross-equation
  comparisons use the same shuffles. A result in the predicted direction that
  does not clear the bound is **inconclusive**.
* **Replication.** 2021 `dat10k`, with long-term debt = mortgages only and
  short-term liabilities = payables + grants payable (the other lines are not
  in that file). Success = same sign and beyond the 2.5th or 97.5th percentile.
* **Implementation details not in the memo.** Negative line values (rare
  reporting errors) are floored at 0 before logs. `dat10k` uses 100 cells
  (10 x 10) because it has about 6,000 filers.

# 2. Functions

```{r functions}
fix64 <- function(x) {
  if (inherits(x, "integer64")) return(as.numeric(x))
  if (is.double(x)) {
    nz <- x[!is.na(x) & x != 0]
    if (length(nz) && all(abs(nz) < 1e-300)) {
      class(x) <- "integer64"
      return(as.numeric(x))
    }
  }
  x
}

matching_lines <- function(d) {
  d <- as.data.frame(d)
  d <- d[d$RETURN_TYPE == "990", ]
  has <- function(f) f %in% names(d)
  fin <- function(f) {
    x <- if (has(f)) suppressWarnings(as.numeric(fix64(d[[f]]))) else rep(0, nrow(d))
    x[is.na(x)] <- 0
    x
  }
  ntee <- if (has("ntee_code_clean")) as.character(d$ntee_code_clean) else
          if (has("NTEE_NCCS")) as.character(d$NTEE_NCCS) else rep(NA_character_, nrow(d))
  sub  <- if (has("nteev2_subsector")) as.character(d$nteev2_subsector) else
          if (has("NTEEV2")) substr(as.character(d$NTEEV2), 1, 3) else rep(NA_character_, nrow(d))
  m <- data.frame(
    cash  = fin("F9_10_ASSET_CASH_EOY") + fin("F9_10_ASSET_SAVING_EOY"),
    ncca  = fin("F9_10_ASSET_PLEDGE_NET_EOY") + fin("F9_10_ASSET_ACC_NET_EOY") +
            fin("F9_10_ASSET_INV_SALE_EOY") + fin("F9_10_ASSET_EXP_PREPAID_EOY"),
    fixed = fin("F9_10_ASSET_LAND_BLDG_NET_EOY"),
    A     = fin("F9_10_ASSET_TOT_EOY"),
    E     = fin("F9_09_EXP_TOT_TOT"),
    lt    = fin("F9_10_LIAB_TAX_EXEMPT_BOND_EOY") + fin("F9_10_LIAB_MTG_NOTE_EOY") +
            fin("F9_10_LIAB_NOTE_UNSEC_EOY"),
    st    = fin("F9_10_LIAB_ACC_PAYABLE_EOY") + fin("F9_10_LIAB_GRANT_PAYABLE_EOY") +
            fin("F9_10_LIAB_REV_DEFERRED_EOY"),
    subsector = sub,
    capital = (sub %in% c("HOS", "UNI")) | (!is.na(ntee) & substr(ntee, 1, 1) == "L"))
  m <- m[m$E > 0 & m$A > 0, ]
  lines <- c("cash", "ncca", "fixed", "lt", "st")
  attr(m, "negative_share") <- sapply(m[lines], function(x) mean(x < 0))
  for (v in lines) m[[v]] <- pmax(m[[v]], 0)
  attr(m, "lines_present") <- c(
    bonds = has("F9_10_LIAB_TAX_EXEMPT_BOND_EOY"), unsecured_notes = has("F9_10_LIAB_NOTE_UNSEC_EOY"),
    deferred_revenue = has("F9_10_LIAB_REV_DEFERRED_EOY"))
  m
}

size_cells <- function(E, A, bins) {
  eb <- cut(rank(E, ties.method = "first"), bins, labels = FALSE)
  ab <- ave(A / E, eb, FUN = function(v) cut(rank(v, ties.method = "first"), bins, labels = FALSE))
  (eb - 1) * bins + ab
}

# p[i] = a random organization in the same cell as i.
permute_within <- function(cells) {
  n <- length(cells)
  p <- integer(n)
  p[order(cells, seq_len(n))] <- order(cells, runif(n))
  p
}

design <- function(m) {
  other <- pmax(m$A - m$cash - m$ncca - m$fixed, 0)
  sE <- ns(log(m$E), df = 4); colnames(sE) <- paste0("size_E", 1:4)
  sO <- ns(log1p(other), df = 4); colnames(sO) <- paste0("size_other", 1:4)
  list(
    blocks = list(
      fixed = cbind(log_fixed = log1p(m$fixed), any_fixed = as.numeric(m$fixed > 0)),
      cash  = cbind(log_cash  = log1p(m$cash),  any_cash  = as.numeric(m$cash > 0)),
      ncca  = cbind(log_ncca  = log1p(m$ncca),  any_ncca  = as.numeric(m$ncca > 0))),
    Z = cbind(intercept = 1, sE, sO, no_other = as.numeric(other == 0)))
}

outcome_models <- function(m) list(
  lt_any    = list(rows = rep(TRUE, nrow(m)), y = as.numeric(m$lt > 0)),
  lt_amount = list(rows = m$lt > 0,           y = log(m$lt[m$lt > 0])),
  st_any    = list(rows = rep(TRUE, nrow(m)), y = as.numeric(m$st > 0)),
  st_amount = list(rows = m$st > 0,           y = log(m$st[m$st > 0])))

# OLS with HC1 standard errors.
robust_fit <- function(y, W) {
  XtXi <- solve(crossprod(W))
  b <- drop(XtXi %*% crossprod(W, y))
  e <- drop(y - W %*% b)
  V <- XtXi %*% crossprod(W * e) %*% XtXi * nrow(W) / (nrow(W) - ncol(W))
  data.frame(term = colnames(W), estimate = b, se = sqrt(diag(V)), row.names = NULL)
}

# Coefficient on the first column of a shuffled block, holding everything else
# fixed (Frisch-Waugh-Lovell), for each outcome model, over B shuffles.
perm_block <- function(block, others, models, cells, B, build = function(Bp) Bp) {
  prep <- lapply(models, function(md) {
    q <- qr(others[md$rows, , drop = FALSE])
    list(q = q, ry = qr.resid(q, md$y))
  })
  coef_for <- function(Bfull) sapply(names(models), function(nm) {
    R <- qr.resid(prep[[nm]]$q, Bfull[models[[nm]]$rows, , drop = FALSE])
    solve(crossprod(R), crossprod(R, prep[[nm]]$ry))[1]
  })
  list(observed = coef_for(build(block)),
       perms = do.call(rbind, lapply(seq_len(B), function(i)
         coef_for(build(block[permute_within(cells), , drop = FALSE])))))
}

run_tests <- function(m, bins, B, alpha) {
  D <- design(m)
  models <- outcome_models(m)
  cells <- size_cells(m$E, m$A, bins)
  W <- cbind(do.call(cbind, D$blocks), D$Z)

  coefs <- do.call(rbind, lapply(names(models), function(nm) {
    f <- robust_fit(models[[nm]]$y, W[models[[nm]]$rows, , drop = FALSE])
    cbind(model = nm, n = sum(models[[nm]]$rows), f[f$term %in% c("log_fixed", "log_cash", "log_ncca"), ])
  }))

  perm <- lapply(names(D$blocks), function(b) {
    others <- cbind(do.call(cbind, D$blocks[setdiff(names(D$blocks), b)]), D$Z)
    perm_block(D$blocks[[b]], others, models, cells, B)
  })
  names(perm) <- names(D$blocks)

  # H1a: interaction with capital-intensive subsector, long-term amount only.
  cap <- as.numeric(m$capital)
  others_a <- cbind(D$blocks$cash, D$blocks$ncca, D$Z, capital = cap)
  h1a <- perm_block(D$blocks$fixed, others_a, models["lt_amount"], cells, B,
                    build = function(Bp) cbind(Bp[, "log_fixed"] * cap, Bp[, "any_fixed"] * cap, Bp))
  Wa <- cbind(fixed_x_capital = D$blocks$fixed[, "log_fixed"] * cap,
              any_fixed_x_capital = D$blocks$fixed[, "any_fixed"] * cap, W, capital = cap)
  h1a_fit <- robust_fit(models$lt_amount$y, Wa[models$lt_amount$rows, , drop = FALSE])

  lo <- alpha / 2; hi <- 1 - alpha / 2
  term_block <- c(log_fixed = "fixed", log_cash = "cash", log_ncca = "ncca")
  coefs$perm_lo <- mapply(function(t, md) quantile(perm[[term_block[[t]]]]$perms[, md], lo), coefs$term, coefs$model)
  coefs$perm_hi <- mapply(function(t, md) quantile(perm[[term_block[[t]]]]$perms[, md], hi), coefs$term, coefs$model)

  b <- function(block, md) perm[[block]]$observed[[md]]
  q <- function(x, p) unname(quantile(x, p))
  # The pre-specified rule: the estimate has the predicted sign AND lies beyond
  # the shuffle bound on that side. The shuffle distributions are not centered
  # at zero (size structure carries some association), so both conditions matter.
  beyond <- function(obs, dist, dir) sign(obs) == dir && (if (dir > 0) obs > q(dist, hi) else obs < q(dist, lo))
  diff_test <- function(block, a, c) {
    obs <- b(block, a) - b(block, c)
    dist <- perm[[block]]$perms[, a] - perm[[block]]$perms[, c]
    list(obs = obs, lo = q(dist, lo), mid = q(dist, 0.5), hi = q(dist, hi), pass = beyond(obs, dist, +1))
  }

  checks <- list(
    H1 = list(
      any    = list(obs = b("fixed", "lt_any"), dist = perm$fixed$perms[, "lt_any"], dir = +1),
      amount = list(obs = b("fixed", "lt_amount"), dist = perm$fixed$perms[, "lt_amount"], dir = +1),
      diff   = diff_test("fixed", "lt_amount", "st_amount")),
    H2 = list(
      any    = list(obs = b("ncca", "st_any"), dist = perm$ncca$perms[, "st_any"], dir = +1),
      amount = list(obs = b("ncca", "st_amount"), dist = perm$ncca$perms[, "st_amount"], dir = +1),
      diff   = diff_test("ncca", "st_amount", "lt_amount")),
    H3 = list(
      any    = list(obs = b("cash", "lt_any"), dist = perm$cash$perms[, "lt_any"], dir = -1),
      amount = list(obs = b("cash", "lt_amount"), dist = perm$cash$perms[, "lt_amount"], dir = -1)),
    H1a = list(
      interaction = list(obs = h1a$observed[[1]], dist = h1a$perms[, 1], dir = +1)))

  verdict <- function(h, ch) {
    parts <- ch[setdiff(names(ch), "diff")]
    pass  <- sapply(parts, function(x) beyond(x$obs, x$dist, x$dir))
    right <- sapply(parts, function(x) sign(x$obs) == x$dir)
    if (h == "H3") {
      wrong_beyond <- sapply(parts, function(x) beyond(x$obs, x$dist, -x$dir))
      if (all(pass)) return("supported")
      if (all(wrong_beyond)) return("rejected (collateral account supported)")
      return("inconclusive")
    }
    if (!is.null(ch$diff)) {
      if (all(right) && ch$diff$obs > 0 && all(pass) && ch$diff$pass) return("supported")
      if (!all(right) || ch$diff$obs <= 0) return("rejected")
      return("inconclusive")
    }
    if (all(pass)) "supported" else if (!all(right)) "rejected" else "inconclusive"
  }
  verdicts <- data.frame(hypothesis = names(checks),
                         verdict = mapply(verdict, names(checks), checks), row.names = NULL)

  detail <- do.call(rbind, lapply(names(checks), function(h) do.call(rbind, lapply(names(checks[[h]]), function(p) {
    x <- checks[[h]][[p]]
    if (p == "diff") return(data.frame(hypothesis = h, part = "long vs. short difference", estimate = x$obs,
                                       perm_lo = x$lo, perm_median = x$mid, perm_hi = x$hi, meets_rule = x$pass))
    data.frame(hypothesis = h, part = p, estimate = x$obs, perm_lo = q(x$dist, lo),
               perm_median = q(x$dist, 0.5), perm_hi = q(x$dist, hi),
               meets_rule = beyond(x$obs, x$dist, x$dir))
  }))))

  list(n = nrow(m), cells = length(unique(cells)), coefs = coefs, perm = perm, h1a = h1a,
       h1a_fit = h1a_fit[h1a_fit$term == "fixed_x_capital", ], verdicts = verdicts, detail = detail,
       elasticity_near_1 = b("fixed", "lt_amount") >= 0.5 && b("fixed", "lt_amount") <= 1.5)
}
```

# 3. Data

```{r data}
efile <- readRDS(file.path(params$data_dir, params$efile_file))
m23 <- matching_lines(efile)
rm(efile); invisible(gc())

describe <- function(m) data.frame(
  filers = nrow(m),
  share_long_term_debt = mean(m$lt > 0), share_short_term = mean(m$st > 0),
  share_fixed_assets = mean(m$fixed > 0), share_cash = mean(m$cash > 0),
  share_noncash_current = mean(m$ncca > 0), share_capital_subsector = mean(m$capital),
  median_expenses = median(m$E), median_long_term_debt_if_any = median(m$lt[m$lt > 0]))
kable(describe(m23), digits = 3, format.args = list(big.mark = ","))
kable(t(attr(m23, "negative_share")), digits = 4, caption = "Share of negative values floored at 0")
```

# 4. Results, 2023

```{r run-2023}
set.seed(params$seed)
res23 <- run_tests(m23, bins = 20, B = params$perms, alpha = 0.01)
kable(res23$verdicts, caption = "Verdicts under the pre-specified rule")
kable(res23$detail, digits = 3, row.names = FALSE,
      caption = "Carrying coefficients and the 0.5th/99.5th percentiles of within-cell shuffles")
```

The shuffle distributions are not centered at zero. Shuffling a line within a
cell keeps its link to the cell's total assets, and total assets are not a
control (they contain the predictors), so part of any size-driven association
survives the shuffle. `perm_median` shows that baseline. The rule therefore
asks for two things: the predicted sign, and a distance from that baseline.

```{r coefs-2023}
kable(res23$coefs, digits = 3, row.names = FALSE, format.args = list(big.mark = ","),
      caption = "All three predictors in all four models (HC1 standard errors; shuffle bounds)")
kable(res23$h1a_fit, digits = 3, row.names = FALSE, caption = "H1a interaction (HC1)")
```

The long-term-debt elasticity with respect to fixed assets is
`r fmt(res23$perm$fixed$observed[["lt_amount"]])`, which
`r if (res23$elasticity_near_1) "falls" else "does not fall"` in the 0.5-1.5
range read as "near 1".

# 5. Replication, 2021 `dat10k`

```{r replication}
data(dat10k, package = "fiscal")
m21 <- matching_lines(dat10k)
kable(t(attr(m21, "lines_present")), caption = "Lines available in dat10k")
kable(describe(m21), digits = 3, format.args = list(big.mark = ","))
set.seed(params$seed + 1)
res21 <- run_tests(m21, bins = 10, B = params$perms, alpha = 0.05)
kable(res21$verdicts, caption = "2021 verdicts (2.5th/97.5th percentile bounds)")
kable(res21$detail, digits = 3, row.names = FALSE)

replication <- merge(res23$detail[, c("hypothesis", "part", "estimate")],
                     res21$detail[, c("hypothesis", "part", "estimate", "meets_rule")],
                     by = c("hypothesis", "part"), suffixes = c("_2023", "_2021"))
replication$replicates <- sign(replication$estimate_2023) == sign(replication$estimate_2021) &
                          replication$meets_rule
kable(replication, digits = 3, row.names = FALSE,
      caption = "Replication: same sign as 2023, and the 2021 estimate meets the rule at the 2.5th/97.5th bound")
```

# 6. Summary

```{r summary-values, include = FALSE}
v23 <- setNames(res23$verdicts$verdict, res23$verdicts$hypothesis)
v21 <- setNames(res21$verdicts$verdict, res21$verdicts$hypothesis)
```

| Hypothesis | 2023 verdict | 2021 verdict |
|---|---|---|
| H1 Long-term matching | `r v23[["H1"]]` | `r v21[["H1"]]` |
| H1a Subsector | `r v23[["H1a"]]` | `r v21[["H1a"]]` |
| H2 Short-term matching | `r v23[["H2"]]` | `r v21[["H2"]]` |
| H3 Pecking order | `r v23[["H3"]]` | `r v21[["H3"]]` |

These are associations. Borrowing and acquiring property are decided together,
and organizations that own rather than lease both hold fixed assets and carry
mortgages, which inflates H1.

```{r save, include = FALSE}
if (nzchar(params$results_file)) {
  saveRDS(list(res23 = res23, res21 = res21, describe23 = describe(m23), describe21 = describe(m21),
               replication = replication), params$results_file)
}
```

```{r session}
sessionInfo()
```
